### 2023-09-25 02-data-analysis/xx-writeup-old.qmd
reorganize code files

---
title: "LLMs as probabilistic models for human experimental data"
author: "Michael Franke, Polina Tsvilodub and Fausto Carcassi"
format: 
  html: 
    code-fold: true
    code-summary: "Show the code"
    self-contained: true
execute:
  error: false
  warning: false
  message: false
  cache: false
callout-appearance: simple
editor:
  markdown:
    wrap: sentence
---

```{r preamble}
#| echo: false

# dependencies etc.
source("00-premable.r")

rerun = TRUE
source('00-stan-fit-helpers.R')

# which model to test
model = "probability"
```

# Motivation

Recent work comparing LLMs to human choice behavior in psychological experiments has focused on whether LLMs predict patterns of human answer behavior qualitatively.
The bulk of this work looks at the answers generated by the LLMs.
<!-- Especially when it comes to sophisticated LLMs, like the GPT-X family, this is problematic because it may conflate the core LLM behavior with potentially unkown sophisticated decoding strategies. -->
In general, much less attention has been payed to whether LLMs can also make adequate *quantitative predictions* that match human choice probabilities in suitable experiments.
(Notable exceptions at the interface of NLP and computational psycholinguistics (Linzen, Hu, Rabovsky ...).)

This work looks at whether probabilistic predictions of GTP-3.5 match human answer patterns also quantitatively.
Rather than looking at LLMs as behavior-generating black box, which we want to understand based on its input-output behvaior, we explore whether we can use LLMs as a predictive probabilistic model, or at least as a component of one, aiming to predict the full distribution of human experimental data.

# Experiment: Reference games

To keep matters simple, we use an established, well-understood and austere experimental paradigm to test human decision making in abstract communicative tasks, so-called **reference games**.
A reference game consists of two players, a speaker and an interpreter, who jointly observe a set of objects (the so-called context).
The speaker is assigned an object from the context set which they have to describe to the interpreter.
The interpreter observes the speaker's description, and chooses one of the objects from the context set.
The goal of the game is, for the speaker, to choose a description that enables the interpreter to choose the target object; and, for the listener, to guess correctly which object the speaker had in mind.

For example, Figure~<span style = "color:firebrick">XXX</span> shows an example, taken from <span style = "color:firebrick">Frank & Goodman (2012)</span>, which we refer to as the **classic reference game example**.
In this example, there are two features that differ across three objects (here shape and color).
One object shares both its color and shape with one other object, while the two other objects have one unique feature (e.g., being the only circle, or the only green object).
In a critical trial for the speaker, the target object to describe is one of the two objects with a unique feature.
... <span style = "color:firebrick">TODO: describe "classic example" </span> ...

The code for the experiment can be found [here](https://github.com/magpie-ea/magpie3-text-refgame), and a live version of the experiment can be tested [here](https://magpie-ea.github.io/magpie3-text-refgame/).

## Participants

Three hundred participants were recruited via Prolific for monetary compensation (0.45 pounds, corresponding to 15.40 pounds / hour).

## Materials & Design

While previous reference games with humans used pictorial representations of objects, and sometimes even pictorial representations of messages, we implemented a simple reference game in a completely text-based fashion.
See Figure <span style = "color:firebrick">XXX</span> for the precise text used.
Each context consisted of three objects.
Objects are defined by a triple of properties, namely a color, a shape and a texture.
The colors were ..., the shapes were ..., the textures were ....
...

To create the stimulus material for both the human experiments and the LLM simulations, we sampled 100 reference games, all of which are instances of the "classic reference game" show in Figure <span style = "color:firebrick">XXX</span>.
The sampled reference games differed in terms of the precise objects that instantiated the "classic example", and in terms of the order in which the objects and expression alternatives were presented in the text.

## Procedure

For each participant the experiment sampled four different instances from our list of 100 pre-created reference games.
Participants played two of these as a speaker (**production condition**), and the other two as an interpreter (**interpretation condition**).

## Results

The overall distribution of choices that correspond to the target, competitor, and distractor states is this:

```{r choice-counts-human-data}

tcd <- c("target", "competitor", "distractor")

d <- read_csv('02-data-prepped.csv') |> 
  mutate(
    response   = factor(response  , levels = tcd),
    prediction = factor(prediction, levels = tcd),
    condition  = factor(condition,  levels = c("production", "interpretation"))
    )

d_counts <- d |> 
  count(condition, response) |> 
  group_by(condition) |> 
  mutate(total = sum(n)) |> 
  ungroup() |> 
  mutate(proportion = n / total)

# d_counts |> 
#   ggplot(aes(x = response, y = n, fill = response)) +
#   geom_col(position = "dodge") +
#   facet_wrap(.~condition) +
#   ylab("") + xlab("") +
#   theme(legend.position="none")

refgame_counts <- d_counts |> 
  ggplot(aes(x = response, y = n, fill = response)) +
  geom_col(position = "dodge") +
  facet_grid(condition ~ .) +
  ylab("") + xlab("") +
  theme(legend.position="none") +
  ylim(c(0,604)) +
  ggtitle("human data") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
  theme(strip.text = element_text(size = 12)) + 
  theme(plot.title = element_text(size = 12))

refgame_counts

ggsave(plot = refgame_counts, 
       filename = "../04-paper/00-pics/refgame-counts.pdf", 
       height = 4, width = 3, scale = 1.1)

```

It is interesting that the distractor options were chosen rather often.
We also see that the number of target choices is higher in the production condition than in the interpretation condition.
This is in line with previous experimental results ... <span style = "color:firebrick">elaborate</span> ...

The next section looks at different ways of deriving quantitative predictions from an LLM to fit to the human data.

# LLM predictions for reference games

## Probabilistic predictions from LLMs

Since the reference game descriptions and choice options are all text-based, the quantitative predictions of LLMs can easily be derived form the conditional probabilities assigned the answer options after processing the context description.
Concretely, we prompted GPT-3.5 with each context description $C$ for all of the 100 pre-generated reference games, and then recorded the probabilities $P(o_i \mid C)$ assigned to all answer options.
If an answer option $o_i = w_{i1}, \dots, w_{in}$ consisted of more than one word, like in the interpretation condition, the conditional probabilities are taken to be the total word probabilities: $P(o_i \mid C) = \frac{1}{n} \prod_{j=1}^n P(w_{ij} \mid C, w_{i(j-1)})$.
Averaging over all contexts $C_1, \dots C_m$ which occurred in the experiment (including double occurrences, but separating production and interpretation conditions, so as to produce the aggregate predictions for exactly the set of conditions that the participant group saw), the model's (non-normalized) average prediction for choice option $o_i$ is:

$$P(o_i) = \frac{1}{m} \sum_{k=1}^{m} P(o_i \mid C_k)$$

The model's **vanilla prediction** for the probability of option $o_i$ is then the normalized probability over all available options:

$$\text{Prediction}(o_i) = \frac{P(o_i)}{\sum_{o_j} P(o_j)}$$

The vanilla prediction of the LLM is shown in the Figure below.

```{r stay-in-log-space}

softmax_row <- function(matrix) {
  exp_matrix <- exp(matrix)
  row_sums <- rowSums(exp_matrix)
  softmax_matrix <- exp_matrix / row_sums
  return(softmax_matrix)
}

# production
x <- d |> 
  # use only predictions for items in the relevant condition
  filter(condition == "production") |> 
  select(submission_id, trial_nr, item, condition, starts_with("scores_produ")) |> 
  mutate(scores_production_distractor = 
           log(exp(scores_production_distractor1) + exp(scores_production_distractor2))) |> 
  select(-scores_production_distractor1, -scores_production_distractor2) |> 
  select(starts_with("scores_produ")) |> as.matrix()

# old predictions
LLM_pred_old_prod <- apply(softmax_row(x), 2, mean)

# new predictions
y <- apply(x, 2, mean)
LLM_pred_new_prod <- exp(y)/ sum(exp(y))

alpha = 5
(z1 <- LLM_pred_new_prod ^ alpha / sum(LLM_pred_new_prod ^ alpha))
(z2 <- exp(alpha * y)/ sum(exp(alpha * y)))
z1 - z2
z1 == z2

# interpretation
x <- d |> 
  # use only predictions for items in the relevant condition
  filter(condition == "interpretation") |> 
  select(starts_with("scores_inter")) |> as.matrix()

# old predictions
LLM_pred_old_inter <- apply(softmax_row(x), 2, mean)

# new predictions
y <- apply(x, 2, mean)
LLM_pred_new_inter <- exp(y)/ sum(exp(y))

```

```{r get-and-plot-average-llm-predictions}

d_llm_prob_prod <- d |> 
  # use only predictions for items in the relevant condition
  filter(condition == "production") |> 
  select(item, starts_with("scores_produ")) |> 
  mutate(scores_production_distractor = 
           log(exp(scores_production_distractor1) + exp(scores_production_distractor2))) |> 
  select(-scores_production_distractor1, -scores_production_distractor2) |> 
  group_by(item) |> 
  mutate(n = n()) |>  
  pivot_longer(-c("item", "n")) |> 
  separate(name, into = c("Variable", "condition", "response"), sep = "_") |> 
  select(-Variable) |> 
  mutate(prob = exp(value)) |>
  # mutate(prob = value) |> 
  group_by(item, condition) |> 
  mutate(prob = prob / sum(prob) * n) |> 
  ungroup()

d_llm_prob_averages_prod <- d_llm_prob_prod |> 
  group_by(condition, response) |> 
  summarize(prob = mean(prob)) |>
  mutate(response = factor(response, levels = tcd)) |> 
  mutate(condition = 'production') |> 
  arrange(condition, response)

d_llm_prob_inter <- d |> 
  # use only predictions for items in the relevant condition
  filter(condition == "interpretation") |> 
  select(item, starts_with("scores_inter")) |> 
  group_by(item) |> 
  mutate(n = n()) |>  
  pivot_longer(-c("item", "n")) |> 
  separate(name, into = c("Variable", "condition", "response"), sep = "_") |> 
  select(-Variable) |> 
  mutate(prob = exp(value)) |>
  # mutate(prob = value) |> 
  group_by(item, condition) |> 
  mutate(prob = prob / sum(prob) * n) |> 
  ungroup()

d_llm_prob_averages_inter <- d_llm_prob_inter |> 
  group_by(condition, response) |> 
  summarize(prob = mean(prob)) |> 
  mutate(response = factor(response, levels = tcd)) |> 
  mutate(condition = 'interpretation') |> 
  arrange(condition, response)

d_llm_prob_averages <- 
  rbind(d_llm_prob_averages_prod,
        d_llm_prob_averages_inter) |> 
  mutate(condition = factor(condition, levels = c('production','interpretation')))

# which model to test?
if (model == "probability") {
  prob_prod  <- matrix(d_llm_prob_averages_prod  |> pull(prob), nrow = 1)
  prob_inter <- matrix(d_llm_prob_averages_inter |> pull(prob), nrow = 1)
} else {
  prob_prod  <- matrix(LLM_pred_new_prod,  nrow = 1)  
  prob_inter <- matrix(LLM_pred_new_inter, nrow = 1)
}

d_llm_prob_averages <- 
  tibble(condition = factor(c(rep("production",3), rep("interpretation",3)),
                            levels = c("production", "interpretation")),
         response = rep(d_llm_prob_averages_prod$response,2),
         prob = c(prob_prod, prob_inter))

model_predictions_vanilla <- d_llm_prob_averages |> 
  ggplot(aes(x = response, y = prob, fill = response)) +
  geom_col(position = "dodge") +
  facet_grid(condition ~ .) +
  ylab("") + xlab("") +
  theme(legend.position="none") +
  ylim(c(0,1)) +
  ggtitle("model predictions") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
  theme(strip.text = element_text(size = 12)) + 
  theme(plot.title = element_text(size = 12))

model_predictions_vanilla

ggsave(plot = model_predictions_vanilla, 
       filename = "../04-paper/00-pics/model-predictions-vanilla.pdf", 
       height = 4, width = 3, scale = 1.1)

```

## Adding a link function:  the $\alpha$-model

On the surface, the LLM predictions show a similar ordinal pattern as the human data.
The target option is preferred over the competitor, which is in turn preferred over the distractor option(s).
Moreover, the amount of target choices is higher in the production condition than in the interpretation condition.
However, the LLM predictions have lower entropy than the distribution of the human answers.
The LLM prediction could be said to be more "optimized", having far fewer distractor choices and a higher ratio of target choices than the human data.

But this does not mean that we should reject the idea that the LLM predictions could serve as a model of human choice data quite yet.
It depends on what we take the model's prediction to be.
If we take the model's prediction to be the precise unalterable probabilities shown in Figure <span style = "color:firebrick">XXX (vanilla predictions)</span>, the match to human answers is not great.
But we may be more lenient and modulate the model's predictions with an additional link function, thereby alterating -as it were- the entropy in the predicted probability vector.

One way of doing this is to use a power-law link function with variable parameter $\alpha$:

$$\text{Prediction}(o_i, \alpha) = \frac{P(o_i)^\alpha}{\sum_{o_j} P(o_j)^\alpha}$$

The previous vanilla predictions in Equation <span style = "color:firebrick">XXX</span> are a special case, assuming a fixed $\alpha =1$.
Notice that this is equivalent to a soft-max linking function on the underlying log-probabilities returned by the LLM.

We use Bayesian inference to obtain posterior estimates of credible values of $\alpha$, once for the production data and the model's predictions for the production condition, and once for the interpretation side.
In both cases, the model uses a single $\alpha$ with a log-normal prior, $\alpha \sim \text{log-Normal}(0.5,1)$ and the obvious multinomial likelihood function.
The usual Bayesian point and interval summary statistics for the two inferences are:

```{r stan-fit-average-predictions}
#| message: false
#| warning: false
#| error: false
#| output: false

# production
d_production <- d |> 
  filter(condition == "production") |> 
  mutate(response = case_when(response == "distractor1" ~ "distractor",
                              response == "distractor2" ~ "distractor",
                              TRUE ~ response)) |> 
  pull(response) |> 
  match(c("target", "competitor", "distractor"))

prob_production <- d_llm_prob_averages |> 
  filter(condition == "production") |> 
  pull(prob)

d_prod    <- matrix(table(d_production) |> as.integer(), nrow = 1)

fit_production   <- fit_data(d_prod, prob_prod) 
alpha_production <- fit_production$draws(format = "df") |> pull(alpha)

# interpretation
d_interpretation <- d |> 
  filter(condition == "interpretation") |> 
  pull(response) |> 
  match(c("target", "competitor", "distractor"))

prob_interpretation <- d_llm_prob_averages |> 
  filter(condition == "interpretation") |> 
  pull(prob)

d_inter    <- matrix(table(d_interpretation) |> as.integer(), nrow = 1)

fit_interpretation <- fit_data(d_inter, prob_inter) 
alpha_interpretation <- fit_interpretation$draws(format = "df") |> pull(alpha)

```

```{r show-summary-stats}
rbind(
  aida::summarize_sample_vector(alpha_production, name = "production"),
  aida::summarize_sample_vector(alpha_interpretation, name = "interpretation"),
  aida::summarize_sample_vector(alpha_production - alpha_interpretation, name = "diff. prod-inter")
)
```

We see that only values for $\alpha$ below 1 are credible, which is in line with the previous observation that the human data distribution is more entropic than the vanilla model predictions.
The table also shows posterior summary statistics for the difference vector of $\alpha$ values for the production case minus those for the interpretation case.
These differences are not credibly different from zero, suggesting that the the same $\alpha$ parameter might be used for both production and interpretation cases (at least for the data at hand).

Figure <span style = "color:firebrick">XXX</span> shows the posterior predictive distributions from the two model fits.
The colored bars show the observed counts.
The black dots are the means of the posterior predictives and the black bars given the 95% HDI of the posterior predictive distribution.
The data observations lie inside the posterior predictive 95% HDI for the interpretation condition, suggesting that the fitted model would predict data that looks similar to what the model was trained on.
For the production data, on the other hand, this is not the case.
Even with the ability to fit $\alpha$ to the data, the model is not able to reproduce the observed pattern in the data.

```{r get-posterior-predictives-alpha-model}

post_pred_prod <- get_posterior_predictives(fit_production, d_prod, "post_pred_prod") |> 
  mutate(condition = "production")
post_pred_inter <- get_posterior_predictives(fit_interpretation, d_inter, "post_pred_inter") |> 
  mutate(condition = "interpretation")

post_pred <- rbind(post_pred_inter, post_pred_prod) |> 
  mutate(condition = factor(condition, levels = c("production", "interpretation")))

post_pred |> 
  ggplot(aes(x = response, y = observed, fill = response)) +
  geom_col() +
  facet_grid(.~condition) +
  geom_pointrange(aes(x = response, y = mean, ymin = `|95%`, ymax = `95%|`), 
                  size = 0.7, linewidth = 1) +
  ylab("") + xlab("") +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
  theme(strip.text = element_text(size = 12))

ggsave(filename = "../04-paper/00-pics/PPC-alpha-model.pdf", width = 8, height = 3.5, scale = 0.9)

```

The interpreation of the visual PPC is corrobarete by a Bayesian posterior predictive check:

```{r Bayesian-p}
message("Bayesian p value for production: ", extract_bayesian_p(fit_production))
message("Bayesian p value for interpretation: ", extract_bayesian_p(fit_interpretation))
```

Intuitively, there are two reasons for the model's failure on production data.
First, since the model predicts a very low rate of distractor choices, we would need a rather low $\alpha$ to increas this probability.
But, second, for low-ish values of $\alpha$, the model cannot predict the large probability ratio in favor of target choice over competitor choices.
In sum, the model is unable to capture the quantitive patter (i.e., the set of probability ratios among choice options) adequately for the production case.

<span style = "color:firebrick">could include:</span>

- doing soft-max on probabilities (not log-probs) is a far worse model; maybe mention?

## Adding noise: the $\alpha \epsilon$-model

The LLM's failure to predict the human production is due, at least in part, to under-predicting the rate at which human participants selected the distractor.
But a human selection of a distractor is, most naturally, explained as an error anyway.
In experimental modeling work, it is therefore common to equip the predictive models also with a low error-probability, e.g., an $\epsilon$ chance of selecting a random option.

We therefore fit another model with two parameters: $\alpha$ as before and $\epsilon$ with prior $\epsilon \sim \text{Exponential}(1)$ and define the multinomial choice probability for option $o_i$ as:

$$\text{Prediction}(o_i, \alpha, \epsilon) = \frac{P(o_i)^\alpha + \epsilon}{\sum_{o_j} \left( P(o_j)^\alpha + \epsilon \right)}$$

Summary statistics for the Bayesian posterior are shown in the table below.
We see that $\alpha$ now has higher estimates for the production model, because $\epsilon$-errors can help explain the distractor choices.
The $\alpha$ estimated for the production data is now credibly higher than for the interpretation data, which makes intuitive sense, given that the prduction condition is inuitively easier.
There is no credible difference in estimated $\epsilon$ between production and interpretation model.

```{r}
#| output: false

fit_production_eps <- fit_data(d_prod, prob_prod, 
                           model_name = '../01-code/llm-average-matrix-epsilon-arrayed.stan')
fit_interpretation_eps <- fit_data(d_inter, prob_inter,
                               model_name = '../01-code/llm-average-matrix-epsilon.stan')
```

```{r}

produce_summary_prodInt_epsilonAlpha(fit_production_eps, fit_interpretation_eps)

```

Figure <span style = "color:firebrick">XXX</span> shows a visual posterior predictive check for the $\alpha\epsilon$-model.
When accommodating for an $\epsilon$ chance of random error, the LLM-based models do not fail the visual check: once trained on the data, they are no longer surprised by it.
This is a low bar to pass, but it is nonetheless reassuring that probabilistic predictions obtained from an LLM can serve as input to a probabilistic model predicting quantitative pattersn in human choice data.

```{r}
post_pred_prod_eps <- get_posterior_predictives(fit_production_eps, d_prod, 
                                                "post_pred_prod_eps") |> 
  mutate(condition = "production")
post_pred_inter_eps <- get_posterior_predictives(fit_interpretation_eps, d_inter,
                                                 "post_pred_inter_eps") |> 
  mutate(condition = "interpretation")

post_pred_eps <- rbind(post_pred_inter_eps, post_pred_prod_eps) |> 
  mutate(condition = factor(condition, levels = c("production", "interpretation")))

post_pred_eps |> 
  ggplot(aes(x = response, y = observed, fill = response)) +
  geom_col() +
  facet_grid(.~condition) +
  geom_pointrange(aes(x = response, y = mean, ymin = `|95%`, ymax = `95%|`), size = 0.7, linewidth = 1) +
  ylab("") + xlab("") +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
  theme(strip.text = element_text(size = 12))

ggsave(filename = "../04-paper/00-pics/PPC-alpha-eps-model.pdf", width = 8, height = 3.5, scale = 0.9)

```

We can corroborate this intuition with a Bayesian posterior $p$-value for both model fits:

```{r Bayesian-p-alpha-epsilon}
message("Bayesian p value for production:", extract_bayesian_p(fit_production_eps))
message("Bayesian p value for interpretation:", extract_bayesian_p(fit_interpretation_eps))
```

### Sanity check: Range of prediction of the $\alpha\epsilon$-model

see file `05-prediction-range.R`

# Individual-item analysis

## Look at arg-max model predictions

Human choice frequencies for when arg-max model would predict target or non-target; indeed slightly different outcomes also for humans.

```{r}
d |> 
  mutate(response_binary = ifelse(response == "target", "target", "other")) |> 
  count(condition, prediction, response) |>  
  group_by(condition, prediction) |> 
  mutate(total = sum(n),
         proportion = n / total) |> 
  # pivot_wider(id_cols = 1:2, names_from = response, values_from = proportion) |> 
  ggplot(aes(x = response, y = proportion, fill = response)) + 
  facet_grid(condition ~ prediction) +
  geom_col()
```

## Fit an $\alpha \epsilon$ model to the by-item aggregate data

Check if an $\alpha \epsilon$ model predicts correctly when it has to predict the data from each item individually.

First get the predictions for each each item:

```{r extract-byItem-LLM-predictions}
d_item_analysis <- d_llm_prob_prod |>
  rbind(d_llm_prob_inter) |>
  mutate(condition = factor(condition, levels = c('production', 'interpretation'))) |>
  full_join(d |>
              select(item, feature_set, position_production, position_interpretation) |>
              unique(),
            by = 'item') |>
  mutate(position = case_when(condition == "production" ~ position_production,
                              TRUE ~ position_interpretation)) |>
  # filter(response != "distractor") |>
  unique() |>
  pivot_wider(id_cols = c(item, condition, feature_set, position_production, position_interpretation, position),
              names_from = response, values_from = prob) |> 
  arrange(condition, item)

```

Prepare by-item data to fit, starting with production.

```{r prepare-data-for-byItem-fitting-production}

# production

d_counts_items <- d |> 
  count(condition, item, response) |> 
  group_by(condition, item) |> 
  mutate(total = sum(n)) |> 
  ungroup() |> 
  mutate(proportion = n / total) |> 
  arrange(condition, item, response)

d_counts_items_matrix_prod <- d_counts_items |> 
  filter(condition == "production") |> 
  pivot_wider(id_cols = item, names_from = response, values_from = n, values_fill = 0) |> 
  select(-item) |> 
  as.matrix()

prob_items_prod <- d_item_analysis |>
  filter(condition == "production") |> 
  group_by(item) |> 
  summarize(
    target = mean(target),
    competitor = mean(competitor),
    distractor = mean(distractor)
    ) |> 
  ungroup() |> 
  select(-item) |> 
  as.matrix()

```

And for interpretation.

```{r prepare-data-for-byItem-fitting-interpretation}

# interpretation

d_counts_items_matrix_inter <- d_counts_items |> 
  filter(condition == "interpretation") |> 
  pivot_wider(id_cols = item, names_from = response, values_from = n, values_fill = 0) |> 
  select(-item) |> 
  as.matrix()

prob_items_inter <- d_item_analysis |>
  filter(condition == "interpretation") |> 
  group_by(item) |> 
  summarize(
    target = mean(target),
    competitor = mean(competitor),
    distractor = mean(distractor)
    ) |> 
  ungroup() |> 
  select(-item) |> 
  as.matrix()

```

```{r model-fit-byItem}
#| message: false
#| warning: false
#| error: false
#| output: false

fit_items_prod <- fit_data(
  d_counts_items_matrix_prod, 
  prob_items_prod, 
  model_name = '../01-code/llm-average-matrix-epsilon.stan')

fit_items_inter <- fit_data(
  d_counts_items_matrix_inter, 
  prob_items_inter, 
  model_name = '../01-code/llm-average-matrix-epsilon.stan')
```

The summary stats for the feature-wise model are:
( <span style = "color:firebrick"> TODO: interpret this </span>)

```{r stats-for-fits-feature-sets}
produce_summary_prodInt_epsilonAlpha(fit_items_prod, fit_items_inter)
```

The visual posterior predictive fit for the feature-wise $\alpha\epsilon$-model is:

```{r PPC-items}

# rerun = FALSE

# production
post_pred_items_prod <- 
  get_posterior_predictives(
    fit_items_prod,
    d_counts_items_matrix_prod,
    "post-pred-prod-item"
    ) |> 
  group_by(row) |> 
  mutate(total = sum(observed)) |> 
  filter(response == "target") |> 
  mutate(row = factor(row))

post_pred_items_prod |> 
  ggplot(aes(x = fct_reorder(row, mean/total), y = observed/total)) +
  geom_pointrange(aes(y = mean/total, ymin = `|95%`/total, ymax = `95%|`/total), 
                  size = 0.6, linewidth = 1, color = "lightgray") +
  geom_point(aes(y = mean/total), color = "darkgray") +
  geom_point(color = project_colors[2]) +
  coord_flip() +
  xlab("item") +
  ylab("proportion of target") +
  theme(axis.text.y = element_blank(),
           axis.ticks.y = element_blank()) +
  ggtitle("production")

ggsave(filename = "~/Desktop/post-pred-production.png", scale = 0.6)

post_pred_items_prod |> 
  ggplot(aes(x = mean/total, observed / total)) +
  geom_segment((aes(x = 0, y = 0, xend = 1, yend=1)), color = "gray") +
  geom_point(alpha = 0.8) +
  xlim(c(0,1)) +
  ylim(c(0,1)) + 
  ylab("observed") +
  xlab("predicted") +
  ggtitle("production")

ggsave(filename = "~/Desktop/obs-pred-production.png", scale = 0.6)

# interpretation
post_pred_items_inter <- 
  get_posterior_predictives(
    fit_items_inter,
    d_counts_items_matrix_inter,
    "post-pred-inter-item"
    ) |> 
  group_by(row) |> 
  mutate(total = sum(observed)) |> 
  filter(response == "target") |> 
  mutate(row = factor(row))

post_pred_items_inter |> 
  ggplot(aes(x = fct_reorder(row, mean/total), y = observed/total)) +
  geom_pointrange(aes(y = mean/total, ymin = `|95%`/total, ymax = `95%|`/total), 
                  size = 0.6, linewidth = 1, color = "lightgray") +
  geom_point(aes(y = mean/total), color = "darkgray") +
  geom_point(color = project_colors[2]) +
  coord_flip() +
  xlab("item") +
  ylab("proportion of target") +
  theme(axis.text.y = element_blank(),
           axis.ticks.y = element_blank()) +
  ggtitle("interpretation")

ggsave(filename = "~/Desktop/post-pred-interpretation.png", scale = 0.6)

post_pred_items_inter |> 
  ggplot(aes(x = mean/total, y =  observed/ total)) +
  # geom_segment(aes(x = `|95%`/total, xend = `95%|`/total, yend = observed/total), color = "gray", alpha = 0.7) +
  geom_segment((aes(x = 0, y = 0, xend = 1, yend=1)), color = "gray") +
  geom_point(alpha = 0.8) +
  xlim(c(0,1)) +
  ylim(c(0,1)) + 
  ylab("observed") +
  xlab("predicted") +
  ggtitle("interpretation")

ggsave(filename = "~/Desktop/obs-pred-interpretation.png", scale = 0.6)
```

Bayesian $p$-values for these by-item level fits:

```{r Bayesian-p-alpha-epsilon-byItem}
message("Bayesian p value for production:", extract_bayesian_p(fit_items_prod))
message("Bayesian p value for interpretation:", extract_bayesian_p(fit_items_inter))
```

# Beyond grand aggregation

The LLM-based $\alpha \epsilon$-model makes different predictions for different prompts.
The following plot shows the vanilla predictions for the target choice option for different `feature sets`.
A feature set is a pair consisting of the trigger feature and the nuisance feature. 
The `trigger feature` is the feature (color, shape or texture) that should be selected in the production condition and that the trigger word in the interpretation condition is an instance of. <span style = "color:firebrick"> TODO: give example</span>.
The `nuisance feature` is the feature whose level all objects share, so that it can, strictly speaking, be completely ignored.
The figure below shows that, when `color` is the trigger feature and `texture` is the nuisance feature, the models' predictions are consistently close to one in the production condition.
But for the feature set `shape-texture`, the model may assign an almost zero probability to the target choice.

```{r}

d_item_analysis |>
  rename(`feature set` = feature_set) |> 
  ggplot(aes(x = `feature set`, y = target, color = `feature set`, shape = `feature set`)) +
  geom_jitter(width = 0.3, size = 1.8) + 
  facet_grid(. ~ condition) +
  xlab("") +
  ylab("predicted target probability") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(strip.text = element_text(size = 12)) + 
  theme(plot.title = element_text(size = 12)) +
  theme(legend.position="none")

ggsave(filename = "../04-paper/00-pics/feature-set-variation.pdf",
       width = 8, height = 4, scale = 0.95)

```

Likewise, the LLM's predictions seem to depend on the order of the choice options.
This effect is less pronounced for the production condition, as shown in the following figure (where a sequence like `[t,d,c,d]` represents a context in which the target option was presented first, then the first distractor, than the competitor and finally the second distractor).

```{r}
d_item_analysis |>
  filter(condition == "production") |> 
  # mutate(position = droplevels(position)) |> 
  ggplot(aes(x = position_production, y = target, color = position)) +
  geom_jitter(width = 0.3) + 
  ylab("predicted target probability") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(strip.text = element_text(size = 12)) + 
  theme(plot.title = element_text(size = 12)) +
  theme(legend.position="none") +
  xlab("order of choice options") +
  ggtitle("production")

ggsave(filename = "../04-paper/00-pics/order-variation-production.pdf",
       width = 16*0.575, height = 4, scale = 0.7)
```

The effect is more pronounced for the interpretation condition.
For example instances of the ordering `[c,t,d]` are assigned very low target probabilities.

```{r}
d_item_analysis |>
  filter(condition == "interpretation") |> 
  ggplot(aes(x = position, y = target, color = position, shape = position)) +
  geom_jitter(width = 0.3) + 
  ylab("predicted target probability") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(strip.text = element_text(size = 12)) + 
  theme(plot.title = element_text(size = 12)) +
  theme(legend.position="none") +
  xlab("order of choice options") +
  ggtitle("interpretation")

ggsave(filename = "../04-paper/00-pics/order-variation-interpretation.pdf",
       width = 16*0.375, height = 4, scale = 0.7)
```

This demonstrates that the model's predictions are quite sensitive to differences in the context, such as the kinds of features that matter for the language task, or the order of presentation of the choice options.
The questions to be asked is therefore: do we see similar patterns also in the human data?; or should we conceive of LLMs as predictors only after aggregating over "spurious contextual elements" like order of presentation of alternatives?

## Effects of relevant features

We first investigate whether the features relevant for the communicative task have an influence on human choice behavior, and whether the LLM predictions specific to each level of `feature set` yield adequate predictions for the human data.

### Human data

First, here are the choice counts for the human data broken up by `feature set`, starting with the production condition:

```{r human-data-counts-per-featureSet}
d_counts_features <- d |> 
  count(condition, feature_set, response) |> 
  group_by(condition, feature_set) |> 
  mutate(total = sum(n)) |> 
  ungroup() |> 
  mutate(proportion = n / total) |> 
  arrange(condition, feature_set, response)

d_counts_features |> 
  filter(condition == "production") |> 
  ggplot(aes(x = response, y = n/total, fill = response)) +
  geom_col(position = "dodge") +
  # facet_grid(condition ~ feature_set) +
  ylab("proportion") + xlab("") +
  facet_wrap(. ~ feature_set, ncol = 3) +
  theme(legend.position="none") +
  ggtitle("production") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
```

There are some slight numerical differences, but maybe not too noteworthy.
To test whether differences for different features matter to human answer behavior in the production task, we ran a Bayesian logistic regression and checked whether the estimated proportions for conditions with, respectively, the highest (`shape-color`) and lowest observed target choice proportion (`shape-texture`) are credibly different.
We find no credible difference.
This suggests that human data has no noteworthy systematic variation based on which features are relevant for the production task.

```{r BDA-features-production}
#| output: false

# shape-color mas max. target proportion
# shape-texture is min target proportion

fit_features_prod <- 
  brms::brm(response ~ feature_set,
    data = d_counts_features |> 
      filter(condition == "interpretation") |> 
      uncount(n) |> 
      mutate(response = case_when(response == "distractor" ~ "competitor",
                                  TRUE ~ response)),
    family = bernoulli()
  )

```

```{r BDA-compare-prod}
faintr::compare_groups(
  fit_features_prod,
  higher = feature_set == "shape-color",
  lower  = feature_set == "shape-texture"
)
```

The picture for the interpretation task looks different.
There are conditions where the target is clearly chosen more frequently (`shape-color`), but others where the modal choice is the distractor (`color-texture`).

```{r human-data-counts-per-featureSet-interpretation}

d_counts_features |> 
  filter(condition == "interpretation") |> 
  ggplot(aes(x = response, y = n/total, fill = response)) +
  geom_col(position = "dodge") +
  # facet_grid(condition ~ feature_set) +
  ylab("proportion") + xlab("") +
  facet_wrap(. ~ feature_set, ncol = 3) +
  theme(legend.position="none") +
  ggtitle("interpretation") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))

```

To test whether these differences are noteworthy, we ran another a Bayesian logistic regression comparing the estimated proportions for conditions with, respectively, the highest (`shape-color`) and lowest observed target choice proportion (`color-texture`).
Indeed, it turns out that they are credibly different.

```{r BDA-features-interpretation}
#| output: false

# max target prop: shape-color
# min target prop: color-texture

fit_features_inter <- 
  brms::brm(response ~ feature_set,
    data = d_counts_features |> 
      filter(condition == "interpretation") |> 
      uncount(n) |> 
      mutate(response = case_when(response == "distractor" ~ "competitor",
                                  TRUE ~ response)),
    family = bernoulli()
  )

```

```{r BDA-features-compare-groups-interpretations}
faintr::compare_groups(
  fit_features_inter,
  higher = feature_set == "shape-color",
  lower  = feature_set == "color-texture"
)
```

Using leave-one-out model comparison to pin an intercept-only model against the the previous model with a single predictor `feature_set`, we find that the intercept-only model is numerically better, but not noteworthily so.

```{r BDA-features-LOO-inter}
#| message: false
#| warning: false
#| error: false
#| output: false

fit_features_inter_Intercept <- 
  brms::brm(response ~ 1,
    data = d_counts_features |> 
      filter(condition == "interpretation") |> 
      uncount(n) |> 
      mutate(response = case_when(response == "distractor" ~ "competitor",
                                  TRUE ~ response)),
    family = bernoulli()
  )

```

```{r BDA-features-LOO-inter-results}
loo_compare(fit_features_inter |> loo(), fit_features_inter_Intercept |> loo())
```

In sum, while there is not indication that human production data depends on the kind of task-relevant features, the human interpration data may be susceptible to variation in task-relevant features.

### Model predictions

The model's predictions do seem to vary based on task-relevant features, at least for the interpretation condition.
We therefore fit another $\alpha\epsilon$-model which differs from the previous in that it does not aim to explain the grand-average of choice proportions, but the choice proportions for each constellation of task-relevant features.
(For clarity: we use a single parameter pair $\alpha$ and $\epsilon$, but use it to predict count data for six different conditions based on six potentially different predictions made by the LLM.)

```{r}
d_counts_features_matrix_prod <- d_counts_features |> 
  filter(condition == "production") |> 
  pivot_wider(id_cols = feature_set, names_from = response, values_from = n) |> 
  select(-feature_set) |> 
  as.matrix()
dimnames(d_counts_features_matrix_prod)[[1]] <- 
  d_counts_features |> 
  pull(feature_set) |> unique()

d_counts_features_matrix_inter <- d_counts_features |> 
  filter(condition == "interpretation") |> 
  pivot_wider(id_cols = feature_set, names_from = response, values_from = n) |> 
  select(-feature_set) |> 
  as.matrix()
dimnames(d_counts_features_matrix_inter)[[1]] <- 
  d_counts_features |> 
  pull(feature_set) |> unique()

prob_features_prod <- d_item_analysis |>
  filter(condition == "production") |> 
  group_by(feature_set) |> 
  summarize(
    target = mean(target),
    competitor = mean(competitor),
    distractor = mean(distractor)
    ) |> 
  ungroup() |> 
  select(-feature_set) |> 
  as.matrix()

prob_features_inter <- d_item_analysis |>
  filter(condition == "interpretation") |> 
  group_by(feature_set) |> 
  summarize(
    target = mean(target),
    competitor = mean(competitor),
    distractor = mean(distractor)
    ) |> 
  ungroup() |> 
  select(-feature_set) |> 
  as.matrix()

```

```{r fit-alpha-epsilon-model-feature-sets}
#| message: false
#| warning: false
#| error: false
#| output: false

fit_features_prod <- fit_data(
  d_counts_features_matrix_prod, 
  prob_features_prod, 
  model_name = '../01-code/llm-average-matrix-epsilon.stan')

fit_features_inter <- fit_data(
  d_counts_features_matrix_inter, 
  prob_features_inter, 
  model_name = '../01-code/llm-average-matrix-epsilon.stan')
```

The summary stats for the feature-wise model are:
( <span style = "color:firebrick"> TODO: interpret this </span>)

```{r stats-for-fits-feature-sets}
produce_summary_prodInt_epsilonAlpha(fit_features_prod, fit_features_inter)
```

The visual posterior predictive fit for the feature-wise $\alpha\epsilon$-model is:

```{r PPC-features}
# production
post_pred_features_prod <- 
  get_posterior_predictives(
    fit_features_prod,
    d_counts_features_matrix_prod
    )

post_pred_features_prod |> 
  ggplot(aes(x = response, y = observed, fill = response)) +
  geom_col() +
  facet_wrap(.~ rownames, ncol=3) +
  geom_pointrange(aes(x = response, y = mean, ymin = `|95%`, ymax = `95%|`), 
                  size = 0.7, linewidth = 1) +
  ylab("") +
  theme(legend.position="none") +
  ggtitle("production") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))

# interpretation
post_pred_features_inter <- 
  get_posterior_predictives(
    fit_features_inter,
    d_counts_features_matrix_inter
    )

post_pred_features_inter |> 
  ggplot(aes(x = response, y = observed, fill = response)) +
  geom_col() +
  facet_wrap(.~ rownames, ncol=3) +
  geom_pointrange(aes(x = response, y = mean, ymin = `|95%`, ymax = `95%|`), 
                  size = 0.7, linewidth = 1) +
  ylab("") +
  theme(legend.position="none") +
  ggtitle("interpretation") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))

```

For the production data, where there was little ground for suspecting variation in the human data, the model's feature-wise prediction are reasonable, with the sole expection of perhaps the `shape-texture` condition.
But for the interpretation data, there is mild reason for diagnosing a potential systematic difference: in conditions where the human data suggests that target and competitor choices are roughly equal (`color-texture` and `texture-shape`), the model still predicts that target choices should be preferred; reversely, where the model predicts that target and distractor choices are rather similar, the `texture-color` condition, the human data is biased towards the target.
( <span style = "color:firebrick">TODO: rethink conclusion</span>)

## Effects of order of choice options

### Human data

Human choice counts for each order of answer alternatives are shown below.

- for production: there is some variation but the target is consistently the preferred option (by a margin)
- for interpretation: there are cases where the competitor is chose more frequently than the target

```{r human-data-counts-per-choiceOrder}

d_counts_choice <- d |> 
  count(condition, position, response) |> 
  group_by(condition, position) |> 
  mutate(total = sum(n)) |> 
  ungroup() |> 
  mutate(proportion = n / total) |> 
  arrange(condition, position, response)

d_counts_choice |> 
  filter(condition == "production") |> 
  ggplot(aes(x = response, y = n, fill = response)) +
  geom_col(position = "dodge") +
  facet_wrap(. ~ position, ncol = 4) +
  ylab("") + xlab("") +
  theme(legend.position="none") +
  ggtitle("production") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
  coord_flip()
  # ggtitle("production")

d_counts_choice |> 
  filter(condition == "interpretation") |> 
  ggplot(aes(x = response, y = n, fill = response)) +
  geom_col(position = "dodge") +
  facet_wrap(. ~ position, ncol = 3) +
  ylab("") + xlab("") +
  theme(legend.position="none") +
  ggtitle("interpretation") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
  # coord_flip() + 
  # ggtitle("interpretation")
```

BDA logistic comparing numerically highest and lowest options ... not credibly different for production.

```{r BDA-choices-production}
#| output: false

d_counts_choice |>
  filter(condition == "production", response == "target") |>
  arrange(proportion)

# max target proportion: [d,t,c,d]
# min target proportion: [d,t,d,c]

fit_choice_prod <- 
  brms::brm(response ~ position,
    data = d_counts_choice |> 
      filter(condition == "production") |> 
      uncount(n) |> 
      mutate(response = case_when(response == "distractor" ~ "competitor",
                                  TRUE ~ response)),
    family = bernoulli()
  )

```

```{r BDA-choices-production-results}
faintr::compare_groups(
  fit_choice_prod,
  higher = position == "[d,t,c,d]",
  lower  = position == "[d,t,d,c]"
)

```

... but for interpretation ...

```{r BDA-choices-interpretation}
#| output: false

# d_counts_choice |>
#   filter(condition == "interpretation", response == "target") |>
#   arrange(proportion)

# max target proportion: [d,t,c]
# min target proportion: [d,c,t]

fit_choice_inter <- 
  brms::brm(response ~ position,
    data = d_counts_choice |> 
      filter(condition == "interpretation") |> 
      uncount(n) |> 
      mutate(response = case_when(response == "distractor" ~ "competitor",
                                  TRUE ~ response)),
    family = bernoulli()
  )

```

```{r BDA-choices-interpretation-results}
faintr::compare_groups(
  fit_choice_inter,
  higher = position == "[d,t,c]",
  lower  = position == "[d,c,t]"
)
```

Loo-based model comparison for interpretation against intercept-only model shows that a model which includes order of choice options as an explanatory factor is significantly better than an intercept-only model.

```{r BDA-choice-LOO-inter}
#| message: false
#| warning: false
#| error: false
#| output: false

fit_choice_inter_Intercept <- 
  brms::brm(response ~ 1,
    data = d_counts_choice |> 
      filter(condition == "interpretation") |> 
      uncount(n) |> 
      mutate(response = case_when(response == "distractor" ~ "competitor",
                                  TRUE ~ response)),
    family = bernoulli()
  )

```

```{r BDA-choice-LOO-inter-results}
loo_compare(fit_choice_inter |> loo(), fit_choice_inter_Intercept |> loo())
```

We may conclude from this that there is reaonable support for the supposition that the human interpretation data is influenced by the order of presentation of choice alternatives.

### Model predictions

It remains to inspect whether the model, which also seems susceptible to choice-order, is compatible with the human data on a by choice-order level.

```{r extract-by-choiceOrder-counts}
d_counts_choice_matrix_prod <- d_counts_choice |> 
  filter(condition == "production") |> 
  pivot_wider(id_cols = position, names_from = response, values_from = n) |> 
  select(-position) |> 
  as.matrix()
d_counts_choice_matrix_prod[is.na(d_counts_choice_matrix_prod)] <- 0
dimnames(d_counts_choice_matrix_prod)[[1]] <- 
  d_counts_choice |> 
  filter(condition == "production") |> 
  pull(position) |> unique()

d_counts_choice_matrix_inter <- d_counts_choice |> 
  filter(condition == "interpretation") |> 
  pivot_wider(id_cols = position, names_from = response, values_from = n) |> 
  select(-position) |> 
  as.matrix()
dimnames(d_counts_choice_matrix_inter)[[1]] <- 
  d_counts_choice |> 
  filter(condition == "interpretation") |> 
  pull(position) |> unique()

prob_choice_prod <- d_item_analysis |>
  filter(condition == "production") |> 
  group_by(position) |> 
  summarize(
    target = mean(target),
    competitor = mean(competitor),
    distractor = mean(distractor)
    ) |> 
  ungroup() |> 
  select(-position) |> 
  as.matrix()

prob_choice_inter <- d_item_analysis |>
  filter(condition == "interpretation") |> 
  group_by(position) |> 
  summarize(
    target = mean(target),
    competitor = mean(competitor),
    distractor = mean(distractor)
    ) |> 
  ungroup() |> 
  select(-position) |> 
  as.matrix()

```

```{r fit-alpha-epsilon-model-choice}
#| message: false
#| warning: false
#| error: false
#| output: false

fit_choice_prod <- fit_data(
  d_counts_choice_matrix_prod, 
  prob_choice_prod, 
  model_name = '../01-code/llm-average-matrix-epsilon.stan')

fit_choice_inter <- fit_data(
  d_counts_choice_matrix_inter, 
  prob_choice_inter, 
  model_name = '../01-code/llm-average-matrix-epsilon.stan')
```

The summary stats for the feature-wise model are:
( <span style = "color:firebrick"> TODO: interpret this </span>)

```{r stats-for-fits-choice}
produce_summary_prodInt_epsilonAlpha(fit_choice_prod, fit_choice_inter)
```

The visual posterior predictive fits for the feature-wise $\alpha\epsilon$-model for the production model is:

```{r PPC-choice}
# production
post_pred_choice_prod <- 
  get_posterior_predictives(
    fit_choice_prod,
    d_counts_choice_matrix_prod
    )

post_pred_choice_prod |> 
  ggplot(aes(x = response, y = observed, fill = response)) +
  geom_col() +
  facet_wrap(.~ rownames, ncol=3) +
  geom_pointrange(aes(x = response, y = mean, ymin = `|95%`, ymax = `95%|`), 
                  size = 0.7, linewidth = 1) +
  ylab("") +
  theme(legend.position="none") +
  ggtitle("production") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))
```

The posterior predictives fo the LLM-based model are decent, except for one case: in order condition `[t,d,d,c]` the model predicts a higher competitor choice probability, but this pattern is not supported by the human data.

The PPC for the interpretation data is:

```{r PPC-choice-interpretation}
# interpretation
post_pred_choice_inter <- 
  get_posterior_predictives(
    fit_choice_inter,
    d_counts_choice_matrix_inter
    )

post_pred_choice_inter |> 
  ggplot(aes(x = response, y = observed, fill = response)) +
  geom_col() +
  facet_wrap(.~ rownames, ncol=3) +
  geom_pointrange(aes(x = response, y = mean, ymin = `|95%`, ymax = `95%|`), 
                  size = 0.7, linewidth = 1) +
  ylab("") +
  theme(legend.position="none") +
  ggtitle("interpretation") +
  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1))

```

Here, the model's predictions are strictly speaking off in a number of cases (e.g., severely overpredicting the distractor choice rate), but they are qualitatively quire remarkable.
In cases where the human data favors the competitor (top row), the model also predicts relatively higher choice rates for the competitor, and vice versa.

# Conclusion

It is not entirely ludicrous to use numerical predictions from LLMs are part of predictive probabilistic models.
But we have to be careful.
Ideally, we average out over low-level variation, such as from order of presentation or similar "nuisance," at least as long as we do not understand what causes this variation in the predictions of models and further research that investigates when exactly this variation accords with empirically observed patterns.

This also implies that we should likely *not* (yet) aspire to use LLMs are models or individual- or item-level predictors.

( <span style = "color:firebrick">TODO: rethink conclusion</span>)

<span style = "color:firebrick">to be continued</span>

<!-- - there is not enough data to investigate `target ~ feature_set * position` -->

<!-- ```{r} -->
<!-- x <- d_item_analysis |> -->
<!--   filter( condition == "production" ) |> -->
<!--   group_by(feature_set, position_production) |> -->
<!--   tidyboot_mean(target) -->

<!-- x |> -->
<!--   ggplot(aes(x = feature_set, y = mean, fill = feature_set)) + -->
<!--   geom_col() + -->
<!--   geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper)) + -->
<!--   ggplot2::coord_flip() + -->
<!--   facet_wrap(~position_production) -->

<!-- ``` -->

### 2023-09-25 02-data-analysis/04-prediction-range.R

@@ -1,0 +1,82 @@
+source('00-premable.r')
+
+plot_predictive_range <- function(p, title = "production", gridsize = 100) {
+  
+  vertices <- data.frame(
+    x = c(0, 1/3, 0.5),
+    y = c(1, 1/3, 0.5)
+  )
+  
+  vertices_simplex <- data.frame(
+    x = c(0, 0  , 1),
+    y = c(0, 1  , 0)
+  )
+  
+  # get predictions for a grid of alpha and epsilon
+  grid_ae <- expand.grid(
+    alpha = seq(0,7, length.out = gridsize),
+    epsilon = seq(0,1, length.out = gridsize)
+  ) |> as_tibble() 
+  pred_matrix = matrix(0, nrow = nrow(grid_ae), ncol = 3)
+  for (i in 1:nrow(pred_matrix)) {
+    q <- p^grid_ae$alpha[i] / sum(p^grid_ae$alpha[i])
+    pred_matrix[i,] <- (1-grid_ae$epsilon[i]) * q + grid_ae$epsilon[i] * c(1/3,1/3,1/3)
+  }
+  dimnames(pred_matrix)[[2]] <- tcd
+  
+  cbind(grid_ae, as_tibble(pred_matrix)) |> 
+    ggplot(aes(y = target, x = competitor)) + 
+    # simplex
+    geom_polygon(data = vertices_simplex, aes(x = x, y = y), fill = "lightgray", alpha = 0.25) +
+    # simplex subset that satisfies constraints 
+    geom_polygon(data = vertices, aes(x = x, y = y), fill = "gray", alpha = 0.5) +
+    # geom_segment(aes(x = 1/3, y = 1/3, xend = p[2], yend = p[1]), color = project_colors[3], size = 1.5) +
+    xlim(c(0,1)) +   ylim(c(0,1)) +
+    # simplex boundaries
+    geom_segment(aes(x= 0, y=1, xend=1, yend=0), color = "firebrick") +
+    geom_segment(aes(x= 0, y=0, xend=1, yend=0), color = "firebrick") +
+    geom_segment(aes(x= 0, y=0, xend=0, yend=1), color = "firebrick") +
+    # boundaries of region satisfying the constraint
+    # constraint: competitor is bigger than distractor
+    geom_segment(
+      aes(x = 0, xend=1/3, y=1, yend=1/3),
+      color = "gray"
+    ) +
+    # constraint: target is bigger than competitor
[…]

### 2023-09-25 01-code/openai_logprobs.py
remove key

@@ -11,7 +11,7 @@ from pprint import pprint
 from make_material import sample_vignette
 
 # openai.api_key = os.getenv("OPENAI_API_KEY")
-OAI = "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"
+# OAI = "bla"
 openai.api_key = OAI
 
 def getLogProbContinuation(initialSequence, continuation, preface = ''):

### 2023-09-25 02-data-analysis/00-stan-fit-helpers.R
rename folder, new modeling section

@@ -8,7 +8,7 @@
 #' @return Fit object.
 #'
 #' @export
-fit_data <- function(d, prob, model_name = '../01-code/llm-average-matrix-epsilon-arrayed.stan') {
+fit_data <- function(d, prob, model_name = '00-stan-files/llm-average-matrix-epsilon-arrayed.stan') {
   n_chunk <- dim(prob)[1]
   n_item  <- dim(prob)[2]
   data_list <- list(d=d, n_chunk=n_chunk, n_item=n_item, prob=prob, uniform = c(1/3, 1/3, 1/3))

### 2023-09-25 02-data-analysis/03-data-analysis.R

@@ -138,7 +138,7 @@ model_predictions_vanilla
 
 ggsave(plot = model_predictions_vanilla, 
        filename = "../04-paper/00-pics/model-predictions-vanilla.pdf", 
-       height = 4, width = 3, scale = 1.1)
+       height = 4, width = 6, scale = 1.1)
 
 #######################################################
 ## prepare data
@@ -213,13 +213,32 @@ produce_summary_prodInt_epsilonAlpha(fit_prod_wide, fit_inter_wide)
 ## get posterior predictives
 #######################################################
 
-pp_prod_narrow <- get_posterior_predictives(fit_prod_narrow, d_prod_global, filename = "post_pred_prod_narrow")
-pp_prod_intermediate <- get_posterior_predictives(fit_prod_intermediate, d_prod_global, filename = "post_pred_prod_intermediate")
-pp_prod_wide <- get_posterior_predictives(fit_prod_wide, d_prod_global, filename = "post_pred_prod_wide")
+pp_prod_narrow <- get_posterior_predictives(fit_prod_narrow, d_prod_global, filename = "post_pred_prod_narrow") |> 
+  mutate(condition = "production", model = "narrow")
+pp_prod_intermediate <- get_posterior_predictives(fit_prod_intermediate, d_prod_global, filename = "post_pred_prod_intermediate") |> 
+  mutate(condition = "production", model = "intermediate")
+pp_prod_wide <- get_posterior_predictives(fit_prod_wide, d_prod_global, filename = "post_pred_prod_wide") |> 
+  mutate(condition = "production", model = "wide")
 
-pp_inter_narrow <- get_posterior_predictives(fit_inter_narrow, d_inter_global, filename = "post_pred_inter_narrow")
-pp_inter_intermediate <- get_posterior_predictives(fit_inter_intermediate, d_inter_global, filename = "post_pred_inter_intermediate")
-pp_inter_wide <- get_posterior_predictives(fit_inter_wide, d_inter_global, filename = "post_pred_inter_wide")
+pp_inter_narrow <- get_posterior_predictives(fit_inter_narrow, d_inter_global, filename = "post_pred_inter_narrow") |> 
+  mutate(condition = "interpretation", model = "narrow")
+pp_inter_intermediate <- get_posterior_predictives(fit_inter_intermediate, d_inter_global, filename = "post_pred_inter_intermediate") |> 
[…]

### 2023-09-25 02-data-analysis/03-data-analysis.R
updated model section (new plots)

@@ -296,7 +296,7 @@ PPC_data |>
   ) +
   theme(legend.position = "right")
 
-ggsave(filename = "../03-paper/00-pics/PPC-alpha-eps-model.pdf", width = 8, height = 3.5, scale = 0.9)
+ggsave(filename = "../03-paper/00-pics/PPC-alpha-eps-model.pdf", width = 8, height = 3.5, scale = 1.0)
 
 #######################################################
 ## Bayesian p-values

### 2023-09-27 02-data-analysis/03-data-analysis.R
added RSA model predictions

@@ -42,7 +42,7 @@ refgame_counts <- d_counts |>
 refgame_counts
 
 ggsave(plot = refgame_counts, 
-       filename = "../04-paper/00-pics/refgame-counts.pdf", 
+       filename = "../03-paper/00-pics/refgame-counts.pdf", 
        height = 4, width = 3, scale = 1.1)
 
 #######################################################
@@ -190,24 +190,51 @@ d_inter_item <- d |>
 d_inter_item[is.na(d_inter_item)] = 0
 
 #######################################################
+## RSA model predictions (Vanilla, alpha = 0)
+#######################################################
+
+RSA_pred_prod  <- array(c(0.89, 0.11, 0), dim=c(1,1,3))
+RSA_pred_inter <- array(c(0.82, 0.18, 0), dim=c(1,1,3))
+
+#######################################################
 ## fit data
 #######################################################
 
 fit_prod_narrow <- fit_data(d_prod_global, array(LLM_pred_prod_narrow, dim=c(1,1,3)))
 fit_prod_intermediate <- fit_data(d_prod_global, array(LLM_pred_prod_intermediate, dim=c(1,1,3)))
 fit_prod_wide <- fit_data(d_prod_global, array(LLM_pred_prod_wide, dim=c(1,nrow(LLM_pred_prod_wide),3)))
+fit_prod_RSA  <- fit_data(d_prod_global, RSA_pred_prod)
 
 fit_inter_narrow <- fit_data(d_inter_global, array(LLM_pred_inter_narrow, dim=c(1,1,3)))
 fit_inter_intermediate <- fit_data(d_inter_global, array(LLM_pred_inter_intermediate, dim=c(1,1,3)))
 fit_inter_wide <- fit_data(d_inter_global, array(LLM_pred_inter_wide, dim=c(1,nrow(LLM_pred_inter_wide),3)))
+fit_inter_RSA  <- fit_data(d_inter_global, RSA_pred_inter)
 
 #######################################################
 ## Bayesian stats
 #######################################################
 
-produce_summary_prodInt_epsilonAlpha(fit_prod_narrow, fit_inter_narrow)
-produce_summary_prodInt_epsilonAlpha(fit_prod_intermediate, fit_inter_intermediate)
-produce_summary_prodInt_epsilonAlpha(fit_prod_wide, fit_inter_wide)
+posterior_stats <- rbind(
+  produce_summary_prodInt_epsilonAlpha(fit_prod_narrow, fit_inter_narrow) |> 
[…]

### 2023-09-27 02-data-analysis/03-data-analysis.R
draft section §4 (item-level preds)

@@ -1,6 +1,6 @@
 # dependencies etc.
 source("00-premable.r")
-rerun = TRUE
+rerun = FALSE
 source('00-stan-fit-helpers.R')
 
 #######################################################
@@ -236,6 +236,8 @@ posterior_stats |>
   facet_grid(condition ~ Parameter, scales = "free") +
   coord_flip()
 
+ggsave(filename = "../03-paper/00-pics/posterior-stats.pdf", width = 8, height = 4, scale = 1.0)
+
 #######################################################
 ## get posterior predictives
 #######################################################
@@ -353,18 +355,239 @@ tibble(
   pivot_wider(id_cols = model, names_from = condition, values_from = Bppp_value) |> 
   xtable::xtable()
 
+#######################################################
+## item-level analysis (LLM)
+#######################################################
+
+# prepare data (item-level)
+
+d_llm_prob_prod <- d |> 
+  filter(condition == "production") |> 
+  select(item, starts_with("scores_produ")) |> 
+  mutate(scores_production_distractor = 
+           log(exp(scores_production_distractor1) + exp(scores_production_distractor2))) |> 
+  select(-scores_production_distractor1, -scores_production_distractor2) |> 
+  group_by(item) |> 
+  mutate(n = n()) |>  
+  pivot_longer(-c("item", "n")) |> 
+  separate(name, into = c("Variable", "condition", "response"), sep = "_") |> 
+  select(-Variable) |> 
+  mutate(prob = exp(value)) |>
+  group_by(item, condition) |> 
+  mutate(prob = prob / sum(prob) * n) |> 
+  ungroup()
+
+d_llm_prob_inter <- d |> 
+  filter(condition == "interpretation") |> 
+  select(item, starts_with("scores_inter")) |> 
+  group_by(item) |> 
+  mutate(n = n()) |>  
+  pivot_longer(-c("item", "n")) |> 
+  separate(name, into = c("Variable", "condition", "response"), sep = "_") |> 
+  select(-Variable) |> 
+  mutate(prob = exp(value)) |>
+  group_by(item, condition) |> 
+  mutate(prob = prob / sum(prob) * n) |> 
+  ungroup() 
+
+d_item_analysis <- d_llm_prob_prod |>
+  rbind(d_llm_prob_inter) |>
[…]

### 2023-09-29 02-data-analysis/03-data-analysis.R
updated with correct vanilla RSA predictions

@@ -1,6 +1,6 @@
 # dependencies etc.
 source("00-premable.r")
-rerun = FALSE
+rerun = TRUE
 source('00-stan-fit-helpers.R')
 
 #######################################################
@@ -137,7 +137,7 @@ model_predictions_vanilla <- d_llm_prob_averages |>
 model_predictions_vanilla
 
 ggsave(plot = model_predictions_vanilla, 
-       filename = "../04-paper/00-pics/model-predictions-vanilla.pdf", 
+       filename = "../03-paper/00-pics/model-predictions-vanilla.pdf", 
        height = 4, width = 6, scale = 1.1)
 
 #######################################################
@@ -193,8 +193,8 @@ d_inter_item[is.na(d_inter_item)] = 0
 ## RSA model predictions (Vanilla, alpha = 0)
 #######################################################
 
-RSA_pred_prod  <- array(c(0.89, 0.11, 0), dim=c(1,1,3))
-RSA_pred_inter <- array(c(0.82, 0.18, 0), dim=c(1,1,3))
+RSA_pred_prod  <- array(c(2/3, 1/3, 0), dim=c(1,1,3))
+RSA_pred_inter <- array(c(0.6, 0.4, 0), dim=c(1,1,3))
 
 #######################################################
 ## fit data
@@ -228,6 +228,7 @@ posterior_stats <- rbind(
   model     = factor(model, levels = rev(c("narrow", "wide", "intermediate", "RSA")))
 )
 
+# plot
 posterior_stats |> 
   filter(condition != "diff. prod-inter") |> 
   ggplot() +
@@ -237,6 +238,14 @@ posterior_stats |>
   coord_flip()
 
 ggsave(filename = "../03-paper/00-pics/posterior-stats.pdf", width = 8, height = 4, scale = 1.0)
+
+# summary stats as table
+
+posterior_stats |> 
+  filter(!is.na(condition)) |> 
+  select(c(6,5,1,2,3,4)) |> 
+  xtable::xtable()
+  
 
 #######################################################
 ## get posterior predictives
@@ -475,7 +484,7 @@ produce_summary_prodInt_epsilonAlpha(fit_items_prod, fit_items_inter)
 post_pred_items_prod <- 
   get_posterior_predictives(
     fit_items_prod,
-    d_counts_items_matrix_prod,
+    data_items_prod,
     "post-pred-prod-item"
[…]

### 2023-10-03 03-paper/paper-LLM-probmod.tex
draft intro

\documentclass[fleqn]{article}

\usepackage{arxiv}

% \usepackage[style=authoryear-comp, % Citation marks as [Jef65]
%             natbib=true,      % Natbib-style cite macros \citeauthor &c.
%             hyperref=true,    % Cites in pdf are links to bib
%                               % (hyperref conf.)
%             maxnames=3,       % truncate name lists if more than 2
%                               % names appear
%             doi=false,        % no doi's
%             url=false,        % no url's
%             sortcites=false,  % do NOT sort cites in the style of the
%                               % bibliography
%             %backref=true     % insert backrefs in reference section
%             ]{biblatex}
\usepackage{mypackages}
\bibliography{~/Desktop/icloud/MyRefGlobal}
\usepackage{mycommands}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{cleveref}       % smart cross-referencing
\usepackage{lipsum}         % Can be removed after putting your text content
\usepackage{graphicx}
\usepackage{doi}

\usepackage{float}

% Statistical models based on LLM predictions
% LLMs fuel statistical models for full distributional predictions of human experimental data
% Large language models for full distributional predictions of human experimental data
% Statistical modeling with predictors from large language models
\title{Beyond the best guess: Distributional predictions from large language models for human forced-choice data}

% Here you can change the date presented in the paper title
%\date{September 9, 1985}
% Or remove it
\date{}

\newif\ifuniqueAffiliation
% Uncomment to use multiple affiliations variant of author block
% \uniqueAffiliationtrue

\ifuniqueAffiliation % Standard variant of author block
  \author{ Michael Franke\thanks{Use footnote for providing further
		information about author (webpage, alternative
		address)---\emph{not} for acknowledging funding agencies.} \\
	Department of Linguistics\\
	University of Tübingen\\
	\texttt{michael.franke@uni-tuebingen.de} \\
	%% examples of more authors
	\And
	Polina Tsvilodub\thanks{Use footnote for providing further
		information about author (webpage, alternative
		address)---\emph{not} for acknowledging funding agencies.} \\
	Department of Linguistics\\
	University of Tübingen\\
	\texttt{polina.tsvilodub@gmail.com} \\
	\And
	Fausto Carcassi\thanks{Use footnote for providing further
		information about author (webpage, alternative
		address)---\emph{not} for acknowledging funding agencies.} \\
	Department of Linguistics\\
	University of Tübingen\\
	\texttt{fausto.carcassi@gmail.com} \\
	%% \AND
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
	%% \And
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
	%% \And
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
}
\else
% Multiple affiliations variant of author block
\usepackage{authblk}
\renewcommand\Authfont{\bfseries}
\setlength{\affilsep}{0em}
% box is needed for correct spacing with authblk
% \newbox{\orcid}\sbox{\orcid}{\includegraphics[scale=0.06]{orcid.pdf}}
\author{Michael Franke, Polina Tsvilodub, Fausto Carcassi}
\affil{Department of Linguistics\\University of Tübingen\\
\texttt{[michael.franke|polina.tsvilodub|fausto.carcassi]@uni-tuebingen.de}}
\fi

% running right header
\renewcommand{\headeright}{}
% small title under big title
\renewcommand{\undertitle}{}
\renewcommand{\shorttitle}{Statistical modeling with LLM predictors}

%%% Add PDF metadata to help others organize their library
%%% Once the PDF is generated, you can check the metadata with
%%% $ pdfinfo template.pdf
\hypersetup{
pdftitle={Statistical modeling with LLM predictors},
pdfsubject={},
pdfauthor={Michael Franke, Polina Tsvilodub, Fausto Carcassi},
pdfkeywords={statistical models, language use, large language models, Bayesian data analysis, reference games},
}

\begin{document}
\maketitle

\begin{abstract}
	{\textcolor{gray}{fill me}}
\end{abstract}

% keywords can be removed
% \keywords{First keyword \and Second keyword \and More}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} enabled a new generation of powerful large language models (LLMs) \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and} which excel on standard benchmarks and promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}, especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
Recently an increasing number of diverse applications has gone beyond using LLMs out-of-the-box, i.e., based on their single-run input-output behavior, and instead utilize LLMs as a part of a larger computational process.
Examples aimed at improving output quality include sophisticated prompting \citep{LiuLiu2022:Generated-Knowl}, structured reasoning or ``neuro-symnolic AI'' \citep{CreswellShanahan2022:Selection-Infer}, or applications of LLMs as scoring functions in open-ended applications \citep{ZhangLehman2023:OMNI:-Open-ende}.
Other work, uses LLMs as part of bigger programs to build towards something more akin to explanatory cognitive models \citep{WongGrand2023:From-Word-Model}.
\mf{enlarge briefly, include more references}

For all of these applications, it is crucial to understand what LLMs can or cannot reliable do.
The prevalent approach towards understanding the performance of LLMs relies on benchmark testing which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``true answer'' exists, averaged over many instances of this task.
Benchmark-driven assessments are very useful to systematically study the effects of scale, e.g., in an engineering-oriented context \citep[e.g.,][]{srivastava2023-BIGbench}.
Other recent work is more psychologically oriented and asks to what extent LLM performace is ``human-like.''
LLM performance is therefore compared to human choice behavior in psychological experiments to investigate whether LLMs predict patterns of human answer behavior qualitatively \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
Notably, some work goes even further and asks whether LLMs can also make adequate \emph{quantitative predictions}.
For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as measures of (self-paced) reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.

This work seeks to extend the investigation in the human-likeness of LLM predictions even further, by exploring whether LLM-derived measures can feed into probabilistic models predicting the full distribution of human answers in a forced-choice tasks.

\begin{itemize}
  \item distributional predictions are more informative, so beneficial for ``understanding'' and for ``applications''
  \item for ``understanding LLMs'' more informative, richer predictions can be more diagnostic about LLM capabilities (risks and potentials)
  \item reliable full-distributional predictions from LLMs can be useful in applications, e.g., relative scoring based on elusive intuitive concepts like ``relevance'' or ``interestingness''
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Motivation}
\label{motivation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experiment: Reference games}
\label{experiment-reference-games}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}
  \centering

  \includegraphics[width = 0.8\textwidth]{00-pics/reference-game.png}

  \caption{Structure of a reference game with human participants. Each trial consists of a set of objects, the so-called context. In production trials, participants choose a single word to describe a target object from the context. In interpretation trials, an object is selected as the likely object a trigger word referring to.}
  \label{fig:ref-game}
\end{figure}

To keep matters simple, we use an established, well-understood and austere experimental paradigm to test human decision making in abstract communicative tasks, so-called \textbf{reference games} \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,SikosVenhuizen2021:Reevaluating-pr}.
A reference game consists of two players, a speaker and an interpreter, who jointly observe a set of objects, usually referred to as context (see Figure~\ref{fig:ref-game}).
In the \textbf{production condition}, the speaker is assigned a \emph{target object} from the context set which they have to describe to the interpreter.
In the \textbf{interpretation condition}, the interpreter observes a description, here called \emph{trigger word}, and chooses one of the objects from the context set.
The goal of the game is, for the speaker, to choose a description that enables the interpreter to choose the target object; and, for the interpreter, to guess correctly which object the speaker had in mind.

The example in Figure~\ref{fig:ref-game} is a standard case, which we will use throughout, where choices are informative about the pragmatic reasoning that decision makers engage in.
In this example, there are two features that differ across three objects (here shape and color).
One object shares both its color and shape with one other object, while the two other objects have one unique feature (e.g., being the only circle, or the only green object).
%
In a critical production trial, the target object to describe is one of the two objects with a unique feature.
The speaker has four words to choose from.
The \textbf{target utterance} is the word which uniquely describes the target object.
The \textbf{distractor utterance} is the word that is true of the target object, but also true of another object.
The other utterances, both of which are false of the target are \textbf{competitor utterance}.
%
In a critical interpretation trial, the trigger word is one that is true of two of the three objects.
If participants engage in pragmatic thought, they might reason that \emph{if} the speaker had wanted to refer to one of the two objects of which the trigger word is true (blue square and blue circle in Figure~\ref{fig:ref-game}), the speaker could have used a more informative word for exactly one of those two objects (``circle''), so they are more likely to refer to the \textbf{target object} (the blue square in Figure~\ref{fig:ref-game}).
The \textbf{competitor object} is the other object of which the trigger word is true.
The \textbf{distractor object} is the object of which the trigger word is false.

We replicated a simple reference game with human participants in which each trial instantiated the structure of the example shown in Figure~\ref{fig:ref-game}.
While previous reference games with human participants used pictorial representations of objects, and sometimes even pictorial representations of messages, we implemented a text-only version in order to be able to compare the predictions of LLMs for the exact same stimuli.
The experiment was realized as an online task using \texttt{magpie} \citep{FrankeJi:magpie:-Minimal}.\footnote{
  The code for the experiment can be found at \href{https://github.com/magpie-ea/magpie3-text-refgame}{https://github.com/magpie-ea/magpie3-text-refgame}, and a live version of the experiment can be tested at \href{https://magpie-ea.github.io/magpie3-text-refgame/}{https://magpie-ea.github.io/magpie3-text-refgame/}.
}

\subsection{Participants}
\label{participants}

A total of 302 participants were recruited via Prolific for monetary
compensation (\textsterling0.45, corresponding to roughly \textsterling 15.40 per hour).
All participants self-identified as native speakers of English.

\subsection{Materials \& design}
\label{materials-design}

We created 100 different items as stimulus material via a stochastic process.
Each item is a different textual description of a reference game with the same logical structure as the example from Figure~\ref{fig:ref-game}.
For each item, the context consisted of three objects.
Objects are defined by a triple of properties, namely a color, a shape and a texture.
For each property, there were four possible values, e.g., blue, green, red, and orange for color.
The sampled items differed in terms of the properties of the objects in the context set, and in terms of the order in which the objects and expression alternatives were presented in the text.
Figures~\ref{fig:refgame-screenshot-production} and \ref{fig:refgame-screenshot-interpretation} from Appendix~\ref{sec:scre-from-online} show example screenshots from the experiment.

\subsection{Procedure}
\label{procedure}

For each participant the experiment sampled four different items.
Participants first played two of these in the production condition, then the other two in the interpretation condition.

\subsection{Results}\label{results}

The overall distribution of choices that correspond to the target, competitor, and distractor states is shown in Figure~\ref{fig:refgame-counts} on the left.\footnote{
  The production condition actually has two distractor choices.
  Here and in the following, these are lumped together as a single category, also when modeling random errors in later models.
  This is a simplification but does not change anything of substance.}
It is interesting that the distractor options were chosen rather often.
We also see that the number of target choices is higher in the production condition than in the interpretation condition.
This is in line with previous experimental results on human reference games.
% \citep{QingFranke2013:Variations-on-a}.
For example, in previous forced-choice reference games with human participants with pictorial presentations of objects, \citet{QingFranke2013:Variations-on-a} observed the following proportions: $\tuple{0.882, 0.118, 0}$ in the production and $\tuple{0.492, 0.506, 0.003}$ in the interpretation condition (for 288 observations in each condition).

\begin{figure}
  \centering

    \includegraphics[width=0.9\linewidth]{00-pics/PPC-alpha-eps-model.pdf}

    \caption{Counts of choices from reference games with human participants (colored bars), with summary statistics from the posterior predictive distribution of three models (shapes and error bars).
      Shapes show the mean of the posterior predictive distributions of three models, differing in the scope of the item-level aggregation with respect to soft-maximizing and scaling.
      Error-bars show corresponding 95\% credible intervals of the posterior predictive.
    }
  \label{fig:refgame-counts}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Model predictions from probabilistic pragmatics}
\label{sec:model-pred-from}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Data from reference games with human participants have been variously analyzed with probabilistic models using inspiration from behavioral game theory \citep[e.g.,][]{DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a}, probabilistic Bayesian modeling \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,FrankeDegen2015:Reasoning-in-Re} or other forms of probabilistic modeling \citep[e.g.,][]{GattGompel2013:Are-we-Bayesian}.
Common to these approaches is that they derive or define, based on some explicit conceptual motivation, a parameterized stochastic policy, $P_{S}(u \mid s; \theta_{S})$, modulated by parameters $\theta_{S}$, for a speaker's choice of expression or utterance $u$ given a referent or state $s$, which the speaker wants to communicate;
and a stochastic listener policy, $P_{L}(s \mid u; \theta_{L})$, capturing the probability of choosing a referent $s$ for utterance $u$.

As a concrete example, we introduce the Rational Speech Act (RSA) model first described in this form by \citet{FrankGoodman2012:Predicting-Prag} \citep[for overview see][]{FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu,StevensBenz2018:Game-Theoretic-,ScontrasTessler2021:A-practical-int,Degen2023:The-Rational-Sp}.
The RSA model defines pragmatic reasoning as a sequence of iterated (soft-)optmization of policies, grounding out in literal interpretation.
If $\mathfrak{L}(s,u) \mapsto \set{0,1}$ is a semantic meaning function mapping each pair of state $s$ and utterance $u$ to a (binary) truth-value, and if $P_{s}$ is a prior over states, a literal listener policy is defined as:
%
\begin{align*}
 P_{L_{0}}(s \mid u) \propto \mathfrak{L}(s,u) P(s)\,.
\end{align*}
%
The pragmatic speaker soft-optimizes the choice of utterance to minimize the literal listener's surprisal for the to-be-communicated state:
%
\begin{align*}
  P_{S}(u_{i} \mid s, \alpha) & = \text{SoftMax}(\myvec{u}_{s}, \alpha)_{i}\,, \text{ where } \\
  \myvec{u}_{s} & = \tuple{\log P_{L_{0}}(s \mid u_{1}), \dots , \log P_{L_{0}}(s \mid u_{n})}\,.
\end{align*}
%
The softmax function \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun} takes a vector of non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{n}}$ and outputs a probability distribution $\myvec{p} = \tuple{p_{1}, \dots, p_{n}}$ in which odds are proportional to score differences, modulated by optimality parameter $\alpha$:
%
\begin{align*}
 \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )
\end{align*}
%
Finally, the pragmatic listener is defined as the policy resulting from applying Bayes rule, solving the inverse-problem for the previously defined speaker policy:
%
\begin{align*}
  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P(s)
\end{align*}

\begin{figure}
  \centering
  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
  \caption{
    Example of predictions from the RSA model (with $\alpha=1$).
    The semantic meaning is shown as a matrix of binary truth-values.
    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents), assume $\alpha=1$ and are shown as row-stochastic matrices.
  }
  \label{fig:RSA-example}
\end{figure}

Figure~\ref{fig:RSA-example} gives example calculations (assuming flat priors and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
For $\alpha=1$, the model predicts that the probability of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{0.6, 0.4, 0}$ for the interpretation condition.
Increasing $\alpha$ will increase the odds of target over competitor choices.
Yet, the model also predicts probability zero for distractor choices, so that the human data shown in Figur~\ref{fig:refgame-counts}e, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
It is therefore common to include a random error probability for each choice \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}, such as via Laplace smoothing: if $P_{r}(R_{l}, C, \alpha_{c})$ is the RSA model's probabilistic prediction for response category $R_{l}$ (target, competitor, or distractor) for condition $C$ (production or interpretation) and condition-specific optimality $\alpha_{c}$, the random-error smoothed prediction is:
\begin{align*}
  P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c}) \propto P_{r}(R_{l}, C; \alpha_{c}) +  \frac{\epsilon_{c}}{3}\,,
\end{align*}
where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
The result is a four-parameter model, one pair of parameters per condition, which provides a likelihood function for the categorical choice data and so can be fitted to the data and compared against other probabilistic models providing a likelihood function for the same data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{LLM predictions for reference games}
\label{llm-predictions-for-reference-games}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section explores different ways of deriving a probabilistic likelihood function for the human data from a Large Language Model.
For a set-up in which the task descriptions and choice options are all text-based, predictions of LLMs at the item-level can be derived as a function of the (log) probabilities assigned to the continuations corresponding to each choice option after processing the task description.
The previous literature has shown that scoring longer sequences of text for an autoregressive LLM in terms of \emph{average next-token log probabilities}, or \emph{average log-probs} for short, gives improved predictions in relevant benchmark tests \citep[e.g.,][]{BrownMann2020:Language-Models}.
Using average log-probs for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
In the following, we first describe the different options of deriving LLM-based predictor terms.
We then compare all probabilistic models based on their adequacy of explaining the human choice data.

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{00-pics/measures-overview.png}
  \caption{
    Schematic overview over three different approaches of obtaining condition-level predictions by aggregating over item-level scores.
    The basic item-level score is average next-word log probability.
    This can be taken as-is into averaging (narrow-scope aggregation), or first transformed from log- into probability-space, either with optimization before (wide scope) or after averaging (intermediate scope).
    The notation for the condition-level options uses pseudo-code to more clearly reflect the relevant operator scope differences (where triangles represent the pipe operator).
  }
  \label{fig:measures-overview}
\end{figure}

\subsection{LLM-based probabilistic predictions for forced-choice data}
\label{sec:notat--term}

Let $C$ be the experimental condition of interest (here: the production or interpretation condition).
Let \(\set{I_1, \dots I_m}\) be the multi-set of items that occurred in the human experiment for condition $C$.\footnote{
  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.
}
Each item $I_{k}$ consists of a task description $T_{k}$ and $l$ choice options $O_{k1}, \dots, O_{kl}$, which are treated as textual continuations of the task description (see Appendix~\ref{sec:examples-items-llm} for an example).
For each item $I_{k}$, let $F(R_{l}, I_{k})$ be the choice option $O_{ki}$ that corresponds to \emph{response type} $R_{l}$ (here: target, competitor, distractor).\footnote{In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
We will define an item-level (non-normalized) score $S(O_{ki})$ for each choice option of a given item $k$, and use this to define a condition-level probabilistic predictions for response types by averaging over all items in a given condition.

\paragraph{Item-level score.}
If a choice option $O_{ki} = w$ is a single word, like in the production condition, an LLM's item-level prediction can simply be defined as the next-word log probability $\log P_{\text{LLM}} (w \mid T_{k})$.
If choice options are strings of more than one word, as in the interpretation condition, we follow the previous literature \citep[e.g.,][]{BrownMann2020:Language-Models} and define the \emph{item-level score} $S(O_{{ki}}, I_{k})$ for option $O_{ki} = w_{ki1}, \dots w_{kin}$ of item $k$, in terms of the average log-probs:
%
\begin{align*}
S(O_{ki}, I_{k}) =  \frac{1}{n} \sum_{j=1}^n \log P_{\text{LLM}} \left(w_{kij} \mid T_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%
The item-level score of response type $R_{l}$ is $S(R_{l}, I_{k}) = S(F(R_{l}, I_{k}))$.

\paragraph{Condition-level predictions.}
Condition-level predictions are a probability distribution over response types, obtained from item-level scores $S(R_{l}, I_{k})$ by averaging over all items belonging to the relevant condition.
To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$, defined as:
%
\begin{align*}
 \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )
\end{align*}
%
The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation:
%
\begin{align*}
  \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
  \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
\end{align*}
%
This means that there are three conceivable scope-sites for aggregating item-level information as shown in Figure~\ref{fig:measures-overview}.
Narrow-scope aggregation first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities:
%
\begin{align*}
  % & \textbf{narrow-scope aggregation} \\
  & P_{n}(R_{l}, C ; \alpha) \propto \expo \left [  \alpha \ \frac{1}{m} \ \sum_{i = k}^{m} S(R_{l}, I_{k})  \right ]
    \tag*{\textcolor{gray}{[narrow-scope aggregation]}}
\end{align*}
%
Wide-scope aggregation, first transposes scores into probabilities, scales them and only aggregates over items last.
\begin{align*}
  % & \textbf{wide-scope aggregation} \\
  & P_{w}(R_{l}, C ; \alpha) \propto \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( \alpha \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha \ S(R_{l'}, I_{k}) \right )}
    \tag*{\textcolor{gray}{[wide-scope aggregation]}}
\end{align*}
%
Intermediate-scope aggregation performs item-level averaging after mapping scores onto probabilities (using softmax with $\alpha=1$), but before scaling (with a power-law transformation with variable $\alpha$):
\begin{align*}
  % & \textbf{intermediate-scope aggregation} \\
  & P_{i}(R_{l}, C ; \alpha) \propto  \left [ \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( S(R_{l'}, I_{k}) \right )} \right ]^{\alpha}
    \tag*{\textcolor{gray}{[intermediate-scope aggregation]}}
\end{align*}

All three condition-level predictors coincide when there is only one item, of course.
For more items, wide-scope and intermediate-scope aggregation coincide when $\alpha = 1$.
In all other cases, the predictors are not guaranteed to be identical.
Conceptually, the main difference is what each measure considers to be the basic item-level unit to aggregate over.
Narrow-scope aggregation considers raw scores, not \emph{not} probabilistic predictions at the item level.
Wide-scope aggregation considers $\alpha$-optimized probabilities at the item level, which is compatible with a procedure of sampling, via softmax decoding, item-level answer options.
Similarly, intermediate-scope aggregation is also compatible with a sampling based picture, but would use pure decoding (without $\alpha$-optimization) at the item level.

\subsection{Model fitting, criticims \& comparison}
\label{sec:model-fitting}

Parameterized predictions $P_{m}(R_{l}, C ; \alpha_{c})$ for models $m \in \set{n,w,i,r}$, where $r$ is the RSA model introduced in Section~\ref{sec:model-pred-from}, can be assessed in the light of the empirical data from human participants with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
For each condition $C$ (production and interpretation), each model has a free optimality parameter $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ with a log-Normal prior.
Following the rationale from Section~\ref{llm-predictions-for-reference-games}, we also assume that each model has one epsilon parameter per condition, $\epsilon_{c}$, which accounts for random errors in the human data via Laplace smoothing, with a prior $\epsilon_{c} \sim \text{Beta}(1,5)$ favoring small values.
The resulting likelihood function for model $m$ and data from condition $C$ is a multinomial distribution with predictor for response $R_{l}$ given by:
%
\begin{align*}
  P_{m}(R_{l}, C; \alpha_{c}, \epsilon_{c}) \propto P_{m}(R_{l}, C; \alpha_{c}) +  \frac{\epsilon_{c}}{3}
\end{align*}

Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of each $\alpha_{c}$ and $\epsilon_{c}$ for each model.
The usual Bayesian summary statistics for posteriors over model parameters are shown in Figure~\ref{fig:posterior-stats}.

\begin{table}[ht]
  \centering

  % \includegraphics[width=0.9\linewidth]{00-pics/posterior-stats.pdf}

  \begin{tabular}{llllcr}
    \toprule \addlinespace[1ex]
    model        & condition      & Parameter  & $|$95\% & mean & 95\%$|$ \\
    \midrule  \addlinespace[1ex]
    narrow       & production     & $\alpha$   & 0.19    & 0.22 & 0.26 \\
                 &                & $\epsilon$ & 0.00    & 0.04 & 0.11 \\
                 & interpretation & $\alpha$   & 0.15    & 0.18 & 0.20 \\
                 &                & $\epsilon$ & 0.00    & 0.02 & 0.06 \\ \addlinespace[0.75ex]
    intermediate & production     & $\alpha$   & 0.86    & 1.04 & 1.22 \\
                 &                & $\epsilon$ & 0.08    & 0.13 & 0.18 \\
                 & interpretation & $\alpha$   & 0.36    & 0.48 & 0.60 \\
                 &                & $\epsilon$ & 0.00    & 0.05 & 0.12 \\ \addlinespace[0.75ex]
    wide         & production     & $\alpha$   & 0.22    & 2.69 & 8.79 \\
                 &                & $\epsilon$ & 0.08    & 0.13 & 0.19 \\
                 & interpretation & $\alpha$   & 0.19    & 1.08 & 6.03 \\
                 &                & $\epsilon$ & 0.00    & 0.06 & 0.18 \\ \addlinespace[0.75ex]
    RSA          & production     & $\alpha$   & 2.59    & 3.16 & 3.71 \\
                 &                & $\epsilon$ & 0.08    & 0.13 & 0.18 \\
                 & interpretation & $\alpha$   & 0.24    & 0.66 & 1.07 \\
                 &                & $\epsilon$ & 0.10    & 0.15 & 0.20 \\ \addlinespace[0.25ex]
    \bottomrule \\
  \end{tabular}

  \caption{
    Summary statistics of posterior samples.
    Number indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
    \mf{maybe not necessary, but can we explain the high values for $\alpha$ in the wide model?}
  }
  \label{fig:posterior-stats}
\end{table}

To assess goodness-of-fit, Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution, i.e., the model's predictions about data of the same size and structure as the training data.
As a minimal bar, we would require a model trained on observed data $D_{\text{obs}}$ to not be surprised by $D_{\text{obs}}$.
Figure~\ref{fig:refgame-counts} shows that only the theoretical model (RSA) and the intermediate-scope model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate.
To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
Consequently, the results from Table~\ref{tab:Bppp-values} suggest that the narrow-scope model fails on the interpretation data, and is at most borderline compatible with the production data; that the wide-scope model is able to reproduce the production data, but fails on the interpretation data; and that only the intermediate-scope model is able to fully recover the data from both conditions.

\begin{table}[ht]
\centering

% \begin{tabular}{lrr}
%   \toprule
%   model        & production & interpretation \\ \midrule
%   narrow       & 0.05       & 0.00 \\
%   wide         & 0.51       & 0.00 \\
%   intermediate & 0.49       & 0.26 \\
%   RSA          & 0.49       & 0.52 \\
%   \bottomrule \\
% \end{tabular}

\begin{tabular}{lrrrr}
  \toprule
  & narrow & wide & intermediate & RSA \\ \midrule
  production & 0.05 & 0.51 & 0.49 & 0.49 \\
  interpretation & 0.00 & 0.00 & 0.26 & 0.52 \\ \bottomrule \\
\end{tabular}

\caption{Sample-based estimates of Bayesian posterior predictive $p$-values for each model and each condition, based on likelihood of the observed data as test statistic.}
\label{tab:Bppp-values}
\end{table}

These results tell us that not all ways of deriving condition-level predictions by averaging over item-level variation are equally good.
Some approaches clearly fail basic checks for statistical goodness-of-fit.
On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the intermediate-scope model, which is able to recover the patterns in the data.
In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.\footnote{
  A potential worry is that the intermediate-scope model is trivial in the sense that it could have predicted \emph{any} data observation.
  Appendix~\ref{sec:range-prior-pred} shows that this is not the case.
}
This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Item-level predictions}
\label{sec:item-level-pred}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The ability to predict a condition-level averages is useful for applications, and may also be theoretically insightful.
Nevertheless, for human data, averages over item- or individual-level variation can also be misleading \citep[e.g.,][]{StanovichWest2000:Individual-diff,EstesMaddox2005:Risks-of-Drawin,HaafRouder2019:Some-do-and-som}, as aggregate patterns may be artifacts of systematic variation at the subject- or item-level.
LLMs do not make individual-level predictions, at least not without further intervention, such as special prompting.
But the question remains whether the variation in item-level predictions made by LLMs is reflected in the human data.

To address this question, we fit a model to the human data at the item-level.
The data to be explained is the set of all response type counts $\tuple{t_{k}, c_{k}, d_{k}}$ for each item $I_{k}$.
The model's predictions for the data from item $I_{k}$ from condition $C$ (production or interpretation) are:
%
\begin{align*}
  P_{\text{item}}(R_{l}, C, I_{k}, \alpha_{c}, \epsilon_{c}) \propto \frac{\expo \left( \alpha_{c} \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha_{c} \ S(R_{l'}, I_{k}) \right )} + \epsilon_{c}
\end{align*}
%
In words, we use a single pair or parameters $\alpha_{c}$ and $\epsilon_{c}$, one for each condition, take the LLM's item-level scores for each item, and derive the response probabilities in the exact same way as before.

Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM predictor predicts variance which is not borne out by the human data.
Concretely, the plot shows, for each item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data.
There is ample variation in the model predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
So, the model itself predicts systematic variability at the item level.
The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
In fact, sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (0.0001 for production, 0 for interpretation), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
In contrast, using the global, item-independent RSA model predictions to fit the item-level data, we obtain estimates of posterior predictive $p$-values that do not discredit the model ($~0.338$ for each condition).
These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.

\begin{figure}
  \centering

  \includegraphics[width = 0.9\linewidth]{00-pics/item-combined-obs-pred.pdf}

  \caption{
    Item-level prediction-observation plot.
    Each dot represents an item.
    The $x$-coordinate represents the mean of the posterior prediction for the target choice probability for the given item.
    The $y$-coordinate represents the observed proportion of target choices for that item.
    The gray line (identity function) is the ideal prediction.
  }
  \label{fig:item-level-obs-pred}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

It is not entirely ludicrous to use numerical predictions from LLMs are
part of predictive probabilistic models. But we have to be careful.
Ideally, we average out over low-level variation, such as from order of
presentation or similar ``nuisance,'' at least as long as we do not
understand what causes this variation in the predictions of models and
further research that investigates when exactly this variation accords
with empirically observed patterns.

This also implies that we should likely \emph{not} (yet) aspire to use
LLMs are models or individual- or item-level predictors.

( {TODO: rethink conclusion})

{to be continued}

\newpage
\appendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Screenshots from the online experiment with human participants}
\label{sec:scre-from-online}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Figure~\ref{fig:refgame-screenshot-production} shows a trial from the production condition, Figure~\ref{fig:refgame-screenshot-interpretation} one for the interpretation condition of the online experiment.

\begin{figure}[H]
  \centering
  \includegraphics[width = 0.8\textwidth]{00-pics/refgame-production.png}

  \caption{Screen shot from a production trial of the online experiment.}
  \label{fig:refgame-screenshot-production}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width = 0.8\textwidth]{00-pics/refgame-interpretation.png}

  \caption{Screen shot from an interpretation trial of the online experiment.}
  \label{fig:refgame-screenshot-interpretation}
\end{figure}

\section{Example item for the LLM experiment}
\label{sec:examples-items-llm}

The text-based input for the LLM predictions mirrors the text in the human experiment, except that the LLM input also lists the set of all available choice options (which for the human experiment is unnecessary since this information is given by the buttons for the forced-choice selection).
For example, the task description $T_{k}$ for the item that corresponds to the production trial shown in Figure~\ref{fig:refgame-screenshot-production} is shown below (the actual input has no line breaks in the first paragraph):

\begin{verbatim}
Your task is to play a conversation game. There are three objects that
you and your friend can see. You have to choose a single word to identify
one of the three objects for your friend.

The three objects are:

a blue triangle with spades
a blue circle with stripes
a blue triangle with stripes

Your task is to make your friend pick out the following target object:

the blue circle with stripes

Which of the following words would you choose:

spades
stripes
circle
triangle

Your answer:

I would choose the word
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Range of prior predictions of intermediate-scope model}
\label{sec:range-prior-pred}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

It may seem that, given the freedom to adjust optimization $\alpha$ and random guessing rate $\epsilon$, some models might be able to explain \emph{any} data observation.
The intermediate-scope model, which was not discredited by model criticism in terms of recovery-based posterior predictive checks, is not trivial in this sense.
Figure~\ref{fig:prediction-range} shows the range of predictions that the intermediate-scope model is, in principle, capable of making (for the whole possible range of \(\alpha\) and \(\epsilon\) values, disregarding Bayesian priors).
There are distributions over response types which the model does not explain predict \emph{ex ante}.
For example, the model does not predict cases where the number of competitor choices is almost as high as the number of target choice, as would be the case if participants would consistently \emph{not} engage in pragmatic reasoning.

\begin{figure}[t]
  \centering

  \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-prod.png}
  %
  \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-inter.png}

  \caption{
    Range of predictions the intermediate-scope model makes for any pair of parameter values $\alpha_{c}$ and $\epsilon_{c}$.
    The light gray triangle with the red boundary is the probability simplex, i.e., the space of all possible 3-place probability vectors.
    The darker gray area in gray boundary lines is the subspace that is compatible with the ordering that the probability of choosing the target is bigger than that of the competitor, which in turn is bigger than that of the distractor.
    The blue area is the subspace of probabilistic predictions the intermediate-scope model makes under any value of its parameters.
    The orange dot in the middle is the ``Laplace point'' of equal probability for all three options, which the model predicts for $\alpha=0$ or $\epsilon=1$.
    The yellow dot is the ``vanilla prediction'' for $\alpha_{c}=1$ and $\epsilon_{c}=0$.
  }
  \label{fig:prediction-range}
\end{figure}

\printbibliography[heading=bibintoc]

\end{document}

### 2023-10-03 03-paper/paper-LLM-probmod.tex
draft new section 3

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Motivation}
+\section{Motivation: Why care for distributional predictions?}
 \label{motivation}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+The prevalent approach to assessing LLM performance based on benchmark-testing quantifies accuracy for a task based on a ``winner-takes-all'' (WTS) strategy.
+Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances all belonging to the same task, or items belonging to the same (logical) condition in a behavioral experiment.
+% For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
+Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice alternatives $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
+The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:
+%
+\begin{align*}
+  S\left( y_{ki}, x_{k} \right) = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right) =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
+\end{align*}
+%
+The WTS approach considers item $I_{k}$ to be answered correctly if the designated ``goal answer'' $y_{ki}$ is among the options that maximize the item-level score.
+The accuracy of the LLM on the given task is the proportion of items in which the correct answer is selected.
+
+While generally a pragmatic and useful approach, there are several reasons why going beyond the best guess can be beneficial.
+Consider first a high-level conceptual argument.
+If task performance is categorical (in the most extreme case: binary), somewhere along the path from a numerical item-level score to accuracy a discontinuity has to be introduced.
+Every such discontinuity bottleneck entails loss of information.
+As this information might be useful, discontinuity should ideally happen as late as possible; or even better: not at all, unless it is irrelevant.
+Consequently, the question becomes: when is the additional information inherent in the scores (not) relevant for assessing LLM performance?
+
+Imagine that there are two options, and that the goal answer is a small $\epsilon$ better in 80\% of the $m$ items, and otherwise worse.
+The WTS strategy predicts an accuracy of 0.8.
+This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTS strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
+For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
+As only the best option will be chosen, that information is irrelevant.
+But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
+If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy with virtual certainty, so the accuracy for choosing the goal answer in such an application scenario would around 0.4.\footnote{The probability of the second option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 1 in the remaining 20\%.}
+This is a categorical shift from predominantly goal answer to predominantly non-goal answer.
+The argument holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two accuracy scores would in principle be more favorable for selecting the goal answers.
+Two things matter:
+(i)
+whether the LLM would be more likely used in cases where greedy-like evaluation is required; and
+(ii)
+what the nature of the item-level variation is.
+
+\bigskip
+
+\mf{to be continued}
 

### 2023-10-04 03-paper/paper-LLM-probmod.tex
minor revision §1-2

@@
 
-The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} enabled a new generation of powerful large language models (LLMs) \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and} which excel on standard benchmarks and promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}, especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
+The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} enabled a new generation of powerful large language models (LLMs) \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and} which excel on standard benchmarks and promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
+% especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
 Recently an increasing number of diverse applications has gone beyond using LLMs out-of-the-box, i.e., based on their single-run input-output behavior, and instead utilize LLMs as a part of a larger computational process.
@@
 Benchmark-driven assessments are very useful to systematically study the effects of scale, e.g., in an engineering-oriented context \citep[e.g.,][]{srivastava2023-BIGbench}.
-Other recent work is more psychologically oriented and asks to what extent LLM performace is ``human-like.''
+Other recent work is more psychologically oriented and asks to what extent LLM performance is ``human-like.''
 LLM performance is therefore compared to human choice behavior in psychological experiments to investigate whether LLMs predict patterns of human answer behavior qualitatively \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
-Notably, some work goes even further and asks whether LLMs can also make adequate \emph{quantitative predictions}.
+There is also some work investigating whether LLMs can make adequate \emph{quantitative predictions}.
 For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as measures of (self-paced) reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.
 
-This work seeks to extend the investigation in the human-likeness of LLM predictions even further, by exploring whether LLM-derived measures can feed into probabilistic models predicting the full distribution of human answers in a forced-choice tasks.
+This work seeks to extend the investigation of the human-likeness of LLM predictions even further, by exploring whether LLM-derived measures can feed into probabilistic models predicting the full distribution of human answers in a forced-choice tasks.
 
@@
 
-The prevalent approach to assessing LLM performance based on benchmark-testing quantifies accuracy for a task based on a ``winner-takes-all'' (WTS) strategy.
-Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances all belonging to the same task, or items belonging to the same (logical) condition in a behavioral experiment.
+The prevalent approach to assessing LLM performance based on benchmark-testing quantifies accuracy for a multiple-choice task with a ``winner-takes-all'' (WTS) strategy.
+Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
 % For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
 Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice alternatives $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
-The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:
+The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
+  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
+  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
+  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
+}
 %
@@
 
-Imagine that there are two options, and that the goal answer is a small $\epsilon$ better in 80\% of the $m$ items, and otherwise worse.
+Imagine that there are two options, and that the goal answer is a small $\epsilon$ better in 80\% of the task's items, and otherwise worse.
 The WTS strategy predicts an accuracy of 0.8.
@@
 This is a categorical shift from predominantly goal answer to predominantly non-goal answer.
-The argument holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two accuracy scores would in principle be more favorable for selecting the goal answers.
-Two things matter:
+The argument holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the goal answers.
+Consequently, this example shows that two things matter to the question of whether a WTS strategy provides a good signal for LLM performance in a task:
 (i)
-whether the LLM would be more likely used in cases where greedy-like evaluation is required; and
+whether the designated applications are foremost cases where greedy-like evaluation is required; and
 (ii)

### 2023-10-05 03-paper/paper-LLM-probmod.tex
revised §2

@@
 
-The prevalent approach to assessing LLM performance based on benchmark-testing quantifies accuracy for a multiple-choice task with a ``winner-takes-all'' (WTS) strategy.
+The prevalent approach to assessing LLM performance based on benchmark-testing quantifies accuracy for a multiple-choice task with a ``winner-takes-all'' (WTA) strategy.
 Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
@@
 \begin{align*}
-  S\left( y_{ki}, x_{k} \right) = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right) =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
+  S_{\text{item}}\left( y_{ki}, x_{k} \right) = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right) =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
 \end{align*}
 %
-The WTS approach considers item $I_{k}$ to be answered correctly if the designated ``goal answer'' $y_{ki}$ is among the options that maximize the item-level score.
-The accuracy of the LLM on the given task is the proportion of items in which the correct answer is selected.
+The WTA approach considers item $I_{k}$ to be answered correctly if the designated ``goal answer'' $y_{ki^{*}}$ is among the options that maximize the item-level score.
+The accuracy of the LLM on the given task is the proportion of items in which the correct answer is selected. \mf{Rephrase or add FN on tie-breaking.}
 
@@
 
-Imagine that there are two options, and that the goal answer is a small $\epsilon$ better in 80\% of the task's items, and otherwise worse.
-The WTS strategy predicts an accuracy of 0.8.
-This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTS strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
+The WTA approach implicitly assumes that item-level choices are resolved by a greedy-like choice of the best alternative.
+It is therefore a good measure of accuracy when the LLM's main use cases are of a greedy-like nature as well.
+When applications rely on a less extreme sampling based strategy, the WTA approach can be generalized to softmax sampling.
+The probability of choosing option $y_{ki}$ for item $I_{k}$ is:
+%
+\begin{align*}
+P_{\text{choose}}(y_{ki}) \propto \expo \left [ \alpha \ S_{\text{item}}\left( y_{ki}, x_{k} \right) \right ]\,,
+\end{align*}
+%
+and the accuracy is the average probability of choosing the designated target option $y_{ki^{*}}$.
+For $\alpha \rightarrow \infty$, this softmax strategy converges to the WTA strategy.
+But in general, these two approaches can yield even categorically different predictions.
+
+Imagine that there are two options, and that the target option is a small $\epsilon$ better in 80\% of the task's items, and otherwise worse.
+The WTA-based accuracy is 0.8.
+This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
 For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
@@
 But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
-If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy with virtual certainty, so the accuracy for choosing the goal answer in such an application scenario would around 0.4.\footnote{The probability of the second option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 1 in the remaining 20\%.}
+If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the second option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 1 in the remaining 20\%.}
 This is a categorical shift from predominantly goal answer to predominantly non-goal answer.
 The argument holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the goal answers.
-Consequently, this example shows that two things matter to the question of whether a WTS strategy provides a good signal for LLM performance in a task:
-(i)
-whether the designated applications are foremost cases where greedy-like evaluation is required; and
-(ii)
-what the nature of the item-level variation is.
 
-\bigskip
+Differences in accuracy scores between a WTA-based or a more general softmax-based strategy depend on the properties of the variation in item-level scores, in particular the relation between score-ordering and score-differences.
+(The example was set up deliberately to have one score-ordering for very small score-differences, and another for huge score-differences.)
+This raises an important general question, namely how to deal with item-level variation in a system's predictions in the first place.
+Data from forced-choice task with human participants will likely show item-level variation, which is why hierarchical modeling accommodating item-level differences is recommended in statistical analyses \citep[e.g.,][]{Jaeger2008:Categorical-dat,barr2013}.
+But whether item-level variation in LLM scores aligns with differences in human choice rates for each item is an empirical question that has to be addressed by explicit modeling and comparison; an exercise that this paper starts to take on.
+On \emph{a priori} grounds, the WTA-based strategy may be deemed implausible as a predictor of human-level choice behavior for a simple item: it is unlikely that, for each item, there is a single choice option that (almost) \emph{all} participants will choose.
 
-\mf{to be continued}
+To the extent that machine predictions show extreme, unsystematic, erratic item-level variation which is quite unlike that of human participants in comparable experiments, an alternative strategy suggests itself, namely to \emph{not} judge a system by its average item-level predictions, but based on its task- or condition-level predictions obtained from different ways of aggregating over item-level variation.
+Section~\ref{llm-predictions-for-reference-games} therefore introduces and compares different ways of deriving condition-level predictions from (variable) item-level scores.
+Properly constructed condition-level predictions might be more aligned with average human choice rates, despite the fact that each complex system (machine and human) shows quite different underlying item-level variation.
+In other words, we should ask whether modern LLMs can predict aggregate human data by \emph{some} process of aggregating its item-level scores, because it might turn out that predicting aggregates is the best the LLMs can do (in particular cases).
+
+\mf{To address whether LLM's condition-level predictions are adequate as item-level predictors, I should re-run the item-level analysis with the mean posterior predictive of the intermediate-scope model (instead of, or in addition to RSA)!}
 
@@
 \begin{align*}
-  P_{S}(u_{i} \mid s, \alpha) & = \text{SoftMax}(\myvec{u}_{s}, \alpha)_{i}\,, \text{ where } \\
-  \myvec{u}_{s} & = \tuple{\log P_{L_{0}}(s \mid u_{1}), \dots , \log P_{L_{0}}(s \mid u_{n})}\,.
-\end{align*}
-%
-The softmax function \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun} takes a vector of non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{n}}$ and outputs a probability distribution $\myvec{p} = \tuple{p_{1}, \dots, p_{n}}$ in which odds are proportional to score differences, modulated by optimality parameter $\alpha$:
-%
-\begin{align*}
- \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )
+  P_{S}(u_{i} \mid s, \alpha) & \propto \expo \left [ \log P_{L_{0}}(s \mid u_{i}) \right ] \,.
 \end{align*}
@@
 \begin{align*}
-  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P(s)
+  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P(s) \,.
 \end{align*}
@@
 Condition-level predictions are a probability distribution over response types, obtained from item-level scores $S(R_{l}, I_{k})$ by averaging over all items belonging to the relevant condition.
-To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$, defined as:
+To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
 %
 \begin{align*}
- \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )
+ \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
 \end{align*}
 %
-The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation:
+The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
 %

### 2023-10-05 03-paper/paper-LLM-probmod.tex
streamlined code and section on Bpppv item-level

@@
 For a set-up in which the task descriptions and choice options are all text-based, predictions of LLMs at the item-level can be derived as a function of the (log) probabilities assigned to the continuations corresponding to each choice option after processing the task description.
-The previous literature has shown that scoring longer sequences of text for an autoregressive LLM in terms of \emph{average next-token log probabilities}, or \emph{average log-probs} for short, gives improved predictions in relevant benchmark tests \citep[e.g.,][]{BrownMann2020:Language-Models}.
-Using average log-probs for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
+% The previous literature has shown that scoring longer sequences of text for an autoregressive LLM in terms of \emph{average next-token log probabilities}, or \emph{average log-probs} for short, gives improved predictions in relevant benchmark tests \citep[e.g.,][]{BrownMann2020:Language-Models}.
+Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
 In the following, we first describe the different options of deriving LLM-based predictor terms.
@@
 Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
-We will define an item-level (non-normalized) score $S(O_{ki})$ for each choice option of a given item $k$, and use this to define a condition-level probabilistic predictions for response types by averaging over all items in a given condition.
+We will define an item-level (non-normalized) score $S(R_{l}, I_{k})$ for each response type $R_{l}$ and item $I_{k}$, and use this to define a condition-level probabilistic predictions for response types by averaging over all items in a given condition.
 
 \paragraph{Item-level score.}
-If a choice option $O_{ki} = w$ is a single word, like in the production condition, an LLM's item-level prediction can simply be defined as the next-word log probability $\log P_{\text{LLM}} (w \mid T_{k})$.
-If choice options are strings of more than one word, as in the interpretation condition, we follow the previous literature \citep[e.g.,][]{BrownMann2020:Language-Models} and define the \emph{item-level score} $S(O_{{ki}}, I_{k})$ for option $O_{ki} = w_{ki1}, \dots w_{kin}$ of item $k$, in terms of the average log-probs:
+The \emph{item-level score} $S(O_{{ki}}, I_{k})$ for option $O_{ki} = w_{ki1}, \dots w_{kin}$ of item $k$ is defined as the log-probability of continuation $O_{ki}$ after prompt $I_{k}$:\footnote{Numerically, we worked with \emph{average} log-probabilities.
+As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition.
+Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words.
+In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
 %
 \begin{align*}
-S(O_{ki}, I_{k}) =  \frac{1}{n} \sum_{j=1}^n \log P_{\text{LLM}} \left(w_{kij} \mid T_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
+S(O_{ki}, I_{k}) =  \sum_{j=1}^n \log P_{\text{LLM}} \left(w_{kij} \mid T_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
 \end{align*}
@@
 Conceptually, the main difference is what each measure considers to be the basic item-level unit to aggregate over.
-Narrow-scope aggregation considers raw scores, not \emph{not} probabilistic predictions at the item level.
+Narrow-scope aggregation considers raw scores, which are \emph{not} probabilistic predictions at the item level.
 Wide-scope aggregation considers $\alpha$-optimized probabilities at the item level, which is compatible with a procedure of sampling, via softmax decoding, item-level answer options.
-Similarly, intermediate-scope aggregation is also compatible with a sampling based picture, but would use pure decoding (without $\alpha$-optimization) at the item level.
+Wide-scope aggregation contains the ``winner-takes-all'' procedure for accuracy scoring in common benchmark test for $\alpha \rightarrow \infty$.
+Finally, like wide-scope aggregation, intermediate-scope aggregation is also compatible with a sampling based picture, but would use pure decoding (without $\alpha$-optimization) at the item level and reserves the scaling of probability distributions to a stage \emph{after} aggregation.
 
@@
 
-The ability to predict a condition-level averages is useful for applications, and may also be theoretically insightful.
+The ability to predict a condition-level average is useful for applications, and may also be theoretically insightful.
 Nevertheless, for human data, averages over item- or individual-level variation can also be misleading \citep[e.g.,][]{StanovichWest2000:Individual-diff,EstesMaddox2005:Risks-of-Drawin,HaafRouder2019:Some-do-and-som}, as aggregate patterns may be artifacts of systematic variation at the subject- or item-level.
@@
 \begin{align*}
-  P_{\text{item}}(R_{l}, C, I_{k}, \alpha_{c}, \epsilon_{c}) \propto \frac{\expo \left( \alpha_{c} \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha_{c} \ S(R_{l'}, I_{k}) \right )} + \epsilon_{c}
+  P_{\text{item}}(R_{l}, C, I_{k}, \alpha_{c}, \epsilon_{c}) \propto \frac{\expo \left( \alpha_{c} \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha_{c} \ S(R_{l'}, I_{k}) \right )} + \epsilon_{c}\,.
 \end{align*}
 %
-In words, we use a single pair or parameters $\alpha_{c}$ and $\epsilon_{c}$, one for each condition, take the LLM's item-level scores for each item, and derive the response probabilities in the exact same way as before.
+In words, we use a single pair of parameters $\alpha_{c}$ and $\epsilon_{c}$, one for each condition, take the LLM's item-level scores for each item, and derive the response probabilities in the exact same way as before.
+Notice that item-level scores are the same for all three types of models introduced and compared in the previous section, so that this by-item analysis applies to the raw LLM prediction irrespective of down-stream aggregation.
 
@@
 In fact, sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (0.0001 for production, 0 for interpretation), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
-In contrast, using the global, item-independent RSA model predictions to fit the item-level data, we obtain estimates of posterior predictive $p$-values that do not discredit the model ($~0.338$ for each condition).
+In contrast, using the global, item-independent RSA model predictions to fit the item-level data, we obtain estimates of posterior predictive $p$-values that do not discredit the model ($~0.338$ for production, $~0.252$ for interpretation).
 These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.

### 2023-10-05 02-data-analysis/03-data-analysis.R

@@ -1,6 +1,6 @@
 # dependencies etc.
 source("00-premable.r")
-rerun = TRUE
+rerun = FALSE
 source('00-stan-fit-helpers.R')
 
 #######################################################
@@ -583,8 +583,27 @@ message("Bayesian p value for interpretation (LLM, by-item):", extract_bayesian_
 
 
 #######################################################
-## item-level analysis (RSA)
+## item-level analysis (w/ condition-level predictors)
 #######################################################
+
+get_Bpppv_itemLevel_glbPredictor <- function(pred_prod, pred_inter) {
+  
+  fit_items_prod_glblPredictor <- fit_data(
+    data_items_prod, 
+    array(rep(pred_prod, each = nrow(data_items_prod)), dim = c(nrow(data_items_prod),1, 3)), 
+    model_name = '00-stan-files/llm-average-matrix-epsilon-arrayed.stan')
+  
+  fit_items_inter_glblPredictor <- fit_data(
+    data_items_inter, 
+    array(rep(pred_inter, each = nrow(data_items_inter)), dim = c(nrow(data_items_inter),1, 3)),
+    model_name = '00-stan-files/llm-average-matrix-epsilon-arrayed.stan')
+  
+  return(list(
+    production     = extract_bayesian_p(fit_items_prod_glblPredictor), 
+    interpretation = extract_bayesian_p(fit_items_inter_glblPredictor)))
+}
+
+get_Bpppv_itemLevel_glbPredictor
 
 fit_items_prod_RSA <- fit_data(
   data_items_prod, 

### 2023-10-06 03-paper/paper-LLM-probmod.tex
revised §1

@@
 % especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
-Recently an increasing number of diverse applications has gone beyond using LLMs out-of-the-box, i.e., based on their single-run input-output behavior, and instead utilize LLMs as a part of a larger computational process.
-Examples aimed at improving output quality include sophisticated prompting \citep{LiuLiu2022:Generated-Knowl}, structured reasoning or ``neuro-symnolic AI'' \citep{CreswellShanahan2022:Selection-Infer}, or applications of LLMs as scoring functions in open-ended applications \citep{ZhangLehman2023:OMNI:-Open-ende}.
-Other work, uses LLMs as part of bigger programs to build towards something more akin to explanatory cognitive models \citep{WongGrand2023:From-Word-Model}.
-\mf{enlarge briefly, include more references}
+Recent works increasingly go beyond using LLMs based on single-run input-output behavior, and instead utilize LLMs as a part of a larger computational process.
+Examples include sophisticated prompting strategies \citep[e.g.,][]{LiuLiu2022:Generated-Knowl}, or structured reasoning models \citep[e.g.,][]{CreswellShanahan2022:Selection-Infer,GaoMadaan2023:PAL:-Program-ai,ParanjapeLundberg2023:ART:-Automatic-}.
+Information from LLMs is also used  to rank or numerically score options in open-ended applications, e.g., to mimic human judgements of relevance or interestingness \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
+Other work, uses LLMs as part of bigger programs to build towards something more akin to explanatory cognitive models \citep[e.g.,][]{WongGrand2023:From-Word-Model}.
 
 For all of these applications, it is crucial to understand what LLMs can or cannot reliable do.
-The prevalent approach towards understanding the performance of LLMs relies on benchmark testing which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``true answer'' exists, averaged over many instances of this task.
+The prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' exists, averaged over many instances of this task.
 Benchmark-driven assessments are very useful to systematically study the effects of scale, e.g., in an engineering-oriented context \citep[e.g.,][]{srivastava2023-BIGbench}.
@@
 LLM performance is therefore compared to human choice behavior in psychological experiments to investigate whether LLMs predict patterns of human answer behavior qualitatively \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
-There is also some work investigating whether LLMs can make adequate \emph{quantitative predictions}.
+Going further, there is also work investigating whether LLMs can make adequate \emph{quantitative predictions}.
 For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as measures of (self-paced) reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.
 
-This work seeks to extend the investigation of the human-likeness of LLM predictions even further, by exploring whether LLM-derived measures can feed into probabilistic models predicting the full distribution of human answers in a forced-choice tasks.
+This work seeks to extend the investigation of the human-likeness of quantitative information derived from LLMs even further, by exploring whether LLM-derived measures can feed into probabilistic models predicting the full distribution of human answers in a forced-choice tasks.
+The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
+Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
 
-\begin{itemize}
-  \item distributional predictions are more informative, so beneficial for ``understanding'' and for ``applications''
-  \item for ``understanding LLMs'' more informative, richer predictions can be more diagnostic about LLM capabilities (risks and potentials)
-  \item reliable full-distributional predictions from LLMs can be useful in applications, e.g., relative scoring based on elusive intuitive concepts like ``relevance'' or ``interestingness''
-\end{itemize}
+A major contribution of this work is methodological.
+By embedding numerical scores from LLMs into probabilistic models for human multiple-choice data, we can use Bayesian data analysis to train, test and compare different LLM-derived predictor scores.
+Concretely, after motivating the desire to assess quantitative information from LLMs in Section~\ref{motivation}, Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a salient probabilistic cognitive model \citep{FrankGoodman2012:Predicting-Prag}.
+Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions are adequate to capture the human data at the item-level and finds that they are not, even though a condition-level predictor from the RSA model is.
+Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the task- or condition-level and compares them against the human data and each other.
+We find that not every way of constructing condition-level predictions from item-level scores is equally good.
+\mf{grant conclusion?}
 

### 2023-10-06 03-paper/paper-LLM-probmod.tex
revised §2

@@
 
-The prevalent approach to assessing LLM performance based on benchmark-testing quantifies accuracy for a multiple-choice task with a ``winner-takes-all'' (WTA) strategy.
+The prevalent approach to assessing LLM performance is based on benchmark-testing, using large data sets for tasks in which a designated goal answer exists.
+Tasks often contain multiple-choice options, and accuracy for a task is determined with a ``winner-takes-all'' (WTA) strategy \citep[e.g.,][]{srivastava2023-BIGbench}.
+
 Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
 % For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
-Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice alternatives $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
+Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
+Among the choice options is a designated \emph{target option} $y_{ki^{*}}$ that is assumed to be the ``true'' or to-be-selected goal answer.
+
 The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
@@
 \begin{align*}
-  S_{\text{item}}\left( y_{ki}, x_{k} \right) = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right) =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
+  S\left( y_{ki}, x_{k} \right) = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right) =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
 \end{align*}
 %
-The WTA approach considers item $I_{k}$ to be answered correctly if the designated ``goal answer'' $y_{ki^{*}}$ is among the options that maximize the item-level score.
-The accuracy of the LLM on the given task is the proportion of items in which the correct answer is selected. \mf{Rephrase or add FN on tie-breaking.}
+The WTA approach considers item $I_{k}$ to be answered correctly if the target option $y_{ki^{*}}$ is the option that maximizes the item-level score.
+The WTA-based accuracy of the LLM on the given task is the proportion of items which are answered correctly.
 
+\mf{the following paragraphs should be shortened and sharpened}
 While generally a pragmatic and useful approach, there are several reasons why going beyond the best guess can be beneficial.
@@
 \begin{align*}
-P_{\text{choose}}(y_{ki}) \propto \expo \left [ \alpha \ S_{\text{item}}\left( y_{ki}, x_{k} \right) \right ]\,,
+P_{\text{choose}}(y_{ki}) \propto \expo \left [ \alpha \ S \left( y_{ki}, x_{k} \right) \right ]\,,
 \end{align*}
@@
 
-Imagine that there are two options, and that the target option is a small $\epsilon$ better in 80\% of the task's items, and otherwise worse.
+Imagine that there are two options, and that the target option's score is a small $\epsilon$ better in 80\% of the task's items, and otherwise worse.
 The WTA-based accuracy is 0.8.
@@
 But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
-If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the second option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 1 in the remaining 20\%.}
+If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases.
+This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
 This is a categorical shift from predominantly goal answer to predominantly non-goal answer.
-The argument holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the goal answers.
+The argument holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
 
-Differences in accuracy scores between a WTA-based or a more general softmax-based strategy depend on the properties of the variation in item-level scores, in particular the relation between score-ordering and score-differences.
-(The example was set up deliberately to have one score-ordering for very small score-differences, and another for huge score-differences.)
+Differences in accuracy scores between a WTA-based or a more general softmax-based strategy depend on the variation in item-level scores, in particular the relation between score-ordering and score-differences.
+(The example in the previous paragraph was set up deliberately to have one score-ordering for very small score-differences, and another for huge score-differences.)
 This raises an important general question, namely how to deal with item-level variation in a system's predictions in the first place.
-Data from forced-choice task with human participants will likely show item-level variation, which is why hierarchical modeling accommodating item-level differences is recommended in statistical analyses \citep[e.g.,][]{Jaeger2008:Categorical-dat,barr2013}.
+Data from forced-choice tasks with human participants will likely show item-level variation as well, which is why hierarchical modeling accommodating item-level differences is recommended in statistical analyses \citep[e.g.,][]{Jaeger2008:Categorical-dat,barr2013}.
 But whether item-level variation in LLM scores aligns with differences in human choice rates for each item is an empirical question that has to be addressed by explicit modeling and comparison; an exercise that this paper starts to take on.
@@
 In other words, we should ask whether modern LLMs can predict aggregate human data by \emph{some} process of aggregating its item-level scores, because it might turn out that predicting aggregates is the best the LLMs can do (in particular cases).
-
-\mf{To address whether LLM's condition-level predictions are adequate as item-level predictors, I should re-run the item-level analysis with the mean posterior predictive of the intermediate-scope model (instead of, or in addition to RSA)!}
 

### 2023-10-06 02-data-analysis/03-data-analysis.R
function for item-level testing

@@ -603,19 +603,10 @@ get_Bpppv_itemLevel_glbPredictor <- function(pred_prod, pred_inter) {
     interpretation = extract_bayesian_p(fit_items_inter_glblPredictor)))
 }
 
-get_Bpppv_itemLevel_glbPredictor
+# Bpppc_item_wide <- get_Bpppv_itemLevel_glbPredictor(c(2/3,1/3,0), c(0.6,0.4,0))
+Bpppc_item_RSA  <- get_Bpppv_itemLevel_glbPredictor(c(2/3,1/3,0), c(0.6,0.4,0))
 
-fit_items_prod_RSA <- fit_data(
-  data_items_prod, 
-  array(rep(c(2/3,1/3,0), each = nrow(data_items_prod)), dim = c(nrow(data_items_prod),1, 3)), 
-  model_name = '00-stan-files/llm-average-matrix-epsilon-arrayed.stan')
+# # Bayesian posterior predictive p-values
 
-fit_items_inter_RSA <- fit_data(
-  data_items_inter, 
-  array(rep(c(0.6,0.4,0), each = nrow(data_items_inter)), dim = c(nrow(data_items_inter),1, 3)),
-  model_name = '00-stan-files/llm-average-matrix-epsilon-arrayed.stan')
-
-# Bayesian posterior predictive p-values
-
-message("Bayesian p value for production (RSA, by-item):", extract_bayesian_p(fit_items_prod_RSA))
-message("Bayesian p value for interpretation (RSA, by-item):", extract_bayesian_p(fit_items_prod_RSA))
+message("Bayesian p value for production (RSA, by-item):", Bpppc_item_RSA$production)
+message("Bayesian p value for interpretation (RSA, by-item):", Bpppc_item_RSA$interpretation)

### 2023-10-06 03-paper/paper-LLM-probmod.tex
rearranged and revised §5 & 6

@@
 
-To keep matters simple, we use an established, well-understood and austere experimental paradigm to test human decision making in abstract communicative tasks, so-called \textbf{reference games} \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,SikosVenhuizen2021:Reevaluating-pr}.
+Reference games are an established, well-understood and austere experimental paradigm to test human decision making in abstract communicative tasks \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,SikosVenhuizen2021:Reevaluating-pr}.
 A reference game consists of two players, a speaker and an interpreter, who jointly observe a set of objects, usually referred to as context (see Figure~\ref{fig:ref-game}).
@@
 
-The overall distribution of choices that correspond to the target, competitor, and distractor states is shown in Figure~\ref{fig:refgame-counts} on the left.\footnote{
+The overall distribution of choices that correspond to the target, competitor, and distractor states is shown in Figure~\ref{fig:refgame-counts}.\footnote{
   The production condition actually has two distractor choices.
@@
 
-\begin{figure}
+\begin{figure}[t]
   \centering
@@
 
-    \caption{Counts of choices from reference games with human participants (colored bars), with summary statistics from the posterior predictive distribution of three models (shapes and error bars).
-      Shapes show the mean of the posterior predictive distributions of three models, differing in the scope of the item-level aggregation with respect to soft-maximizing and scaling.
+    \caption{Counts of choices from reference games with human participants (colored bars), with summary statistics from the posterior predictive distribution of four models (shapes and error bars).
+      Shapes show the mean of the posterior predictive distributions of the RSA model and three further models derived from item-level LLM-supplied scores.
       Error-bars show corresponding 95\% credible intervals of the posterior predictive.
@@
 
-Data from reference games with human participants have been variously analyzed with probabilistic models using inspiration from behavioral game theory \citep[e.g.,][]{DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a}, probabilistic Bayesian modeling \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,FrankeDegen2015:Reasoning-in-Re} or other forms of probabilistic modeling \citep[e.g.,][]{GattGompel2013:Are-we-Bayesian}.
-Common to these approaches is that they derive or define, based on some explicit conceptual motivation, a parameterized stochastic policy, $P_{S}(u \mid s; \theta_{S})$, modulated by parameters $\theta_{S}$, for a speaker's choice of expression or utterance $u$ given a referent or state $s$, which the speaker wants to communicate;
+Data from reference games with human participants have been variously analyzed with probabilistic models using inspiration from behavioral game theory \citep[e.g.,][]{DegenFranke2013:Cost-Based-Prag}, probabilistic Bayesian modeling \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag} or other forms of probabilistic modeling \citep[e.g.,][]{GattGompel2013:Are-we-Bayesian}.
+Common to these approaches is that they derive or define, based on some explicit conceptual motivation, a parameterized stochastic speaker policy, $P_{S}(u \mid s; \theta_{S})$, modulated by parameters $\theta_{S}$, for a speaker's choice of expression or utterance $u$ given a referent or state $s$, which the speaker wants to communicate;
 and a stochastic listener policy, $P_{L}(s \mid u; \theta_{L})$, capturing the probability of choosing a referent $s$ for utterance $u$.
@@
 Figure~\ref{fig:RSA-example} gives example calculations (assuming flat priors and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
-For $\alpha=1$, the model predicts that the probability of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{0.6, 0.4, 0}$ for the interpretation condition.
+For $\alpha=1$, the model predicts that the probabilities of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{0.6, 0.4, 0}$ for the interpretation condition.
 Increasing $\alpha$ will increase the odds of target over competitor choices.
-Yet, the model also predicts probability zero for distractor choices, so that the human data shown in Figur~\ref{fig:refgame-counts}e, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
+Yet, the model also predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
 It is therefore common to include a random error probability for each choice \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}, such as via Laplace smoothing: if $P_{r}(R_{l}, C, \alpha_{c})$ is the RSA model's probabilistic prediction for response category $R_{l}$ (target, competitor, or distractor) for condition $C$ (production or interpretation) and condition-specific optimality $\alpha_{c}$, the random-error smoothed prediction is:
@@
 
+Parameterized predictions like $P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c})$ can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
+Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
+Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{fig:posterior-stats}).
+
+To assess goodness-of-fit, we can use the posterior predictive distribution, i.e., the model's predictions about data of the same size and structure as the training data.
+As a minimal bar, we would require a model trained on observed data $D_{\text{obs}}$ to not be surprised by $D_{\text{obs}}$.
+Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models to be introduced later).
+We see that the RSA model passes this ``visual posterior predictive check'' \citep{Kruschke2021:Bayesian-Analys} for both conditions.
+To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
+These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
+Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
+The results from Table~\ref{tab:Bppp-values} show that the RSA model passes this numerical test of model adequacy.
+
+The following sections apply these test of Bayesian model criticism also to models built around predictor values from LLMs.
+Section~\ref{sec:item-level-pred} first looks at the data from each item individually, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{LLM predictions for reference games}
+\section{Item-level predictions from LLMs}
+\label{sec:item-level-pred}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Figure~\ref{fig:refgame-counts} shows the condition-level counts for choices for all of the 100 different items combined.
+An LLM, on the other hand, first and foremost makes predictions about each item.
+The items of the reference game experiment differ in the task-relevant features (color, shape, texture), as well as the order of presentation of states and words.
+We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelvant material \citep{DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options.
+\mf{references from the NLP literature on effects of prompt variation?}
+The main question to be addressed in this section is whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.
+
+For a set-up in which the task descriptions and choice options are all text-based, predictions of LLMs at the item-level can be derived as a function of the (log) probabilities assigned to the continuations corresponding to each choice option after processing the task description.
+Let $C$ be the experimental condition of interest (here: the production or interpretation condition).
+Let \(\set{I_1, \dots I_m}\) be the multi-set of items that occurred in the human experiment for condition $C$.\footnote{
+  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.
+}
+Each item $I_{k}$ consists of a task description $T_{k}$ and $l$ choice options $O_{k1}, \dots, O_{kl}$, which are treated as textual continuations of the task description (see Appendix~\ref{sec:examples-items-llm} for an example).
+For each item $I_{k}$, let $F(R_{l}, I_{k})$ be the choice option $O_{ki}$ that corresponds to \emph{response type} $R_{l}$ (here: target, competitor, distractor).\footnote{In the current set-up the response type ``distractor'' has two instantiations in the production condition.
+Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
+The \emph{item-level score} $S(O_{{ki}}, I_{k})$ for option $O_{ki} = w_{ki1}, \dots w_{kin}$ of item $I_{k}$ is defined as the log-probability of continuation $O_{ki}$ after prompt $T_{k}$:\footnote{Numerically, we worked with \emph{average} log-probabilities.
+As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition.
+Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words.
+In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
+%
+\begin{align*}
+S(O_{ki}, I_{k}) =  \sum_{j=1}^n \log P_{\text{LLM}} \left(w_{kij} \mid T_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
+\end{align*}
+%
+The item-level score of response type $R_{l}$ is $S(R_{l}, I_{k}) = S(F(R_{l}, I_{k}))$.
+
+Using item-level scores from GPT-3.5, we fit a model to the human data at the item-level.
+The data to be explained is the set of all triples of response type counts $\tuple{t_{k}, c_{k}, d_{k}}$ (target, competitor, distractor), separately for each item $I_{k}$.
+The model's predictions for the data from item $I_{k}$ from condition $C$ (production or interpretation) are:
+%
+\begin{align*}
+  P_{\text{item}}(R_{l}, C, I_{k}, \alpha_{c}, \epsilon_{c}) \propto \frac{\expo \left( \alpha_{c} \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha_{c} \ S(R_{l'}, I_{k}) \right )} + \epsilon_{c}\,.
+\end{align*}
+%
+In words, we use a single pair of parameters $\alpha_{c}$ and $\epsilon_{c}$, one for each condition, take the LLM's item-level scores for each item, and derive the response probabilities in the exact same way as before.
+% Notice that item-level scores are the same for all three types of models introduced and compared in the previous section, so that this by-item analysis applies to the raw LLM prediction irrespective of down-stream aggregation.
+Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.
+
+\begin{figure}[t]
+  \centering
+
+  \includegraphics[width = 0.9\linewidth]{00-pics/item-combined-obs-pred.pdf}
+
+  \caption{
+    Item-level prediction-observation plot.
+    Each dot represents an item.
+    The $x$-coordinate represents the mean of the posterior prediction for the target choice probability for the given item.
+    The $y$-coordinate represents the observed proportion of target choices for that item.
+    The gray line (identity function) is the ideal prediction.
+  }
+  \label{fig:item-level-obs-pred}
+\end{figure}
+
+Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
+Concretely, the plot shows, for each item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
+There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
+So, the model itself predicts systematic variability at the item level.
+The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
+
+Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (0.0001 for production, 0 for interpretation), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
+In contrast, using the global, item-independent RSA model predictions to fit the item-level data, we obtain estimates of posterior predictive $p$-values that do not discredit the model ($\sim 0.338$ for production, $\sim 0.252$ for interpretation).
+These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
+Put more strongly, a model that seeks to predict what human participants choose on a by-item level, based on the most obvious item-level score, is ruled out by experimental data.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Condition-level predictions}
 \label{llm-predictions-for-reference-games}
@@
 
-This section explores different ways of deriving a probabilistic likelihood function for the human data from a Large Language Model.
-For a set-up in which the task descriptions and choice options are all text-based, predictions of LLMs at the item-level can be derived as a function of the (log) probabilities assigned to the continuations corresponding to each choice option after processing the task description.
-% The previous literature has shown that scoring longer sequences of text for an autoregressive LLM in terms of \emph{average next-token log probabilities}, or \emph{average log-probs} for short, gives improved predictions in relevant benchmark tests \citep[e.g.,][]{BrownMann2020:Language-Models}.
+This section explores different ways of deriving a probabilistic likelihood function for the human data from a Large Language Model at the aggregate, condition-level.
 Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
@@
 
-\subsection{LLM-based probabilistic predictions for forced-choice data}
-\label{sec:notat--term}
-
-Let $C$ be the experimental condition of interest (here: the production or interpretation condition).
-Let \(\set{I_1, \dots I_m}\) be the multi-set of items that occurred in the human experiment for condition $C$.\footnote{
-  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.
-}
-Each item $I_{k}$ consists of a task description $T_{k}$ and $l$ choice options $O_{k1}, \dots, O_{kl}$, which are treated as textual continuations of the task description (see Appendix~\ref{sec:examples-items-llm} for an example).
-For each item $I_{k}$, let $F(R_{l}, I_{k})$ be the choice option $O_{ki}$ that corresponds to \emph{response type} $R_{l}$ (here: target, competitor, distractor).\footnote{In the current set-up the response type ``distractor'' has two instantiations in the production condition.
-Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
-We will define an item-level (non-normalized) score $S(R_{l}, I_{k})$ for each response type $R_{l}$ and item $I_{k}$, and use this to define a condition-level probabilistic predictions for response types by averaging over all items in a given condition.
-
-\paragraph{Item-level score.}
-The \emph{item-level score} $S(O_{{ki}}, I_{k})$ for option $O_{ki} = w_{ki1}, \dots w_{kin}$ of item $k$ is defined as the log-probability of continuation $O_{ki}$ after prompt $I_{k}$:\footnote{Numerically, we worked with \emph{average} log-probabilities.
-As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition.
-Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words.
-In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
-%
-\begin{align*}
-S(O_{ki}, I_{k}) =  \sum_{j=1}^n \log P_{\text{LLM}} \left(w_{kij} \mid T_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
-\end{align*}
-%
-The item-level score of response type $R_{l}$ is $S(R_{l}, I_{k}) = S(F(R_{l}, I_{k}))$.
-
-\paragraph{Condition-level predictions.}
 Condition-level predictions are a probability distribution over response types, obtained from item-level scores $S(R_{l}, I_{k})$ by averaging over all items belonging to the relevant condition.
@@
 Wide-scope aggregation considers $\alpha$-optimized probabilities at the item level, which is compatible with a procedure of sampling, via softmax decoding, item-level answer options.
-Wide-scope aggregation contains the ``winner-takes-all'' procedure for accuracy scoring in common benchmark test for $\alpha \rightarrow \infty$.
+Wide-scope aggregation contains the ``winner-takes-all'' procedure for accuracy scoring in common benchmark tests for $\alpha \rightarrow \infty$.
 Finally, like wide-scope aggregation, intermediate-scope aggregation is also compatible with a sampling based picture, but would use pure decoding (without $\alpha$-optimization) at the item level and reserves the scaling of probability distributions to a stage \emph{after} aggregation.
 
-\subsection{Model fitting, criticims \& comparison}
-\label{sec:model-fitting}
-
-Parameterized predictions $P_{m}(R_{l}, C ; \alpha_{c})$ for models $m \in \set{n,w,i,r}$, where $r$ is the RSA model introduced in Section~\ref{sec:model-pred-from}, can be assessed in the light of the empirical data from human participants with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
-For each condition $C$ (production and interpretation), each model has a free optimality parameter $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ with a log-Normal prior.
-Following the rationale from Section~\ref{llm-predictions-for-reference-games}, we also assume that each model has one epsilon parameter per condition, $\epsilon_{c}$, which accounts for random errors in the human data via Laplace smoothing, with a prior $\epsilon_{c} \sim \text{Beta}(1,5)$ favoring small values.
-The resulting likelihood function for model $m$ and data from condition $C$ is a multinomial distribution with predictor for response $R_{l}$ given by:
-%
-\begin{align*}
-  P_{m}(R_{l}, C; \alpha_{c}, \epsilon_{c}) \propto P_{m}(R_{l}, C; \alpha_{c}) +  \frac{\epsilon_{c}}{3}
-\end{align*}
-
-Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of each $\alpha_{c}$ and $\epsilon_{c}$ for each model.
-The usual Bayesian summary statistics for posteriors over model parameters are shown in Figure~\ref{fig:posterior-stats}.
-
-\begin{table}[ht]
+\begin{table}[t]
   \centering
@@
     Number indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
-    \mf{maybe not necessary, but can we explain the high values for $\alpha$ in the wide model?}
   }
@@
 
-To assess goodness-of-fit, Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution, i.e., the model's predictions about data of the same size and structure as the training data.
-As a minimal bar, we would require a model trained on observed data $D_{\text{obs}}$ to not be surprised by $D_{\text{obs}}$.
+The three condition-level predictions yield three different probabilistic models, each with two pairs of condition-dependent parameters $\alpha_{c}$, and $\epsilon_{c}$.
+Using priors and Bayesian posterior estimation as previously for the RSA (see Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and posterior predictive distributions.
+The usual Bayesian summary statistics for posteriors over model parameters are shown in Table~\ref{fig:posterior-stats}.
+Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
+What is noteworthy about these estimates is that the $\alpha_{c}$ parameters for the wide-scope model have a very large credible range in both conditions.
+This is because, already for values of $\alpha_{c} = 1$, the item-level predictions are virtually ``winner-takes-all'' choices, selecting the option with the maximum score with probability almost 1 for almost all items.
+This implies that, by studying the posterior predictive distribution of the wide-scope model, we are virtually testing condition-level predictions from the standard WTA-strategy used for benchmark testing (see Section~\ref{motivation}).
+
 Figure~\ref{fig:refgame-counts} shows that only the theoretical model (RSA) and the intermediate-scope model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate.
 To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
-These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
-Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
 Consequently, the results from Table~\ref{tab:Bppp-values} suggest that the narrow-scope model fails on the interpretation data, and is at most borderline compatible with the production data; that the wide-scope model is able to reproduce the production data, but fails on the interpretation data; and that only the intermediate-scope model is able to fully recover the data from both conditions.
@@
 This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Item-level predictions}
-\label{sec:item-level-pred}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The ability to predict a condition-level average is useful for applications, and may also be theoretically insightful.
-Nevertheless, for human data, averages over item- or individual-level variation can also be misleading \citep[e.g.,][]{StanovichWest2000:Individual-diff,EstesMaddox2005:Risks-of-Drawin,HaafRouder2019:Some-do-and-som}, as aggregate patterns may be artifacts of systematic variation at the subject- or item-level.
-LLMs do not make individual-level predictions, at least not without further intervention, such as special prompting.
-But the question remains whether the variation in item-level predictions made by LLMs is reflected in the human data.
-
-To address this question, we fit a model to the human data at the item-level.
-The data to be explained is the set of all response type counts $\tuple{t_{k}, c_{k}, d_{k}}$ for each item $I_{k}$.
-The model's predictions for the data from item $I_{k}$ from condition $C$ (production or interpretation) are:
-%
-\begin{align*}
-  P_{\text{item}}(R_{l}, C, I_{k}, \alpha_{c}, \epsilon_{c}) \propto \frac{\expo \left( \alpha_{c} \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha_{c} \ S(R_{l'}, I_{k}) \right )} + \epsilon_{c}\,.
-\end{align*}
-%
-In words, we use a single pair of parameters $\alpha_{c}$ and $\epsilon_{c}$, one for each condition, take the LLM's item-level scores for each item, and derive the response probabilities in the exact same way as before.
-Notice that item-level scores are the same for all three types of models introduced and compared in the previous section, so that this by-item analysis applies to the raw LLM prediction irrespective of down-stream aggregation.
-
-Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM predictor predicts variance which is not borne out by the human data.
-Concretely, the plot shows, for each item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data.
-There is ample variation in the model predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
-So, the model itself predicts systematic variability at the item level.
-The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
-In fact, sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (0.0001 for production, 0 for interpretation), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
-In contrast, using the global, item-independent RSA model predictions to fit the item-level data, we obtain estimates of posterior predictive $p$-values that do not discredit the model ($~0.338$ for production, $~0.252$ for interpretation).
-These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
-
-\begin{figure}
-  \centering
-
-  \includegraphics[width = 0.9\linewidth]{00-pics/item-combined-obs-pred.pdf}
-
-  \caption{
-    Item-level prediction-observation plot.
-    Each dot represents an item.
-    The $x$-coordinate represents the mean of the posterior prediction for the target choice probability for the given item.
-    The $y$-coordinate represents the observed proportion of target choices for that item.
-    The gray line (identity function) is the ideal prediction.
-  }
-  \label{fig:item-level-obs-pred}
-\end{figure}
 

### 2023-10-06 03-paper/paper-LLM-probmod.tex
minor rev throughout

@@
 
+\mf{would be good to add references here}
+
 The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} enabled a new generation of powerful large language models (LLMs) \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and} which excel on standard benchmarks and promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
@@
 Going further, there is also work investigating whether LLMs can make adequate \emph{quantitative predictions}.
-For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as measures of (self-paced) reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.
+For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.
 
@@
 The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
-Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
+Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
 
@@
 By embedding numerical scores from LLMs into probabilistic models for human multiple-choice data, we can use Bayesian data analysis to train, test and compare different LLM-derived predictor scores.
-Concretely, after motivating the desire to assess quantitative information from LLMs in Section~\ref{motivation}, Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a salient probabilistic cognitive model \citep{FrankGoodman2012:Predicting-Prag}.
+\mf{describe high-level results here}.
+
+The paper is structured as follows.
+Section~\ref{motivation} further motivates the desire to assess quantitative information from LLMs, and it introduces the distinction between item-level and condition-level predictions.
+Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a salient probabilistic cognitive model \citep{FrankGoodman2012:Predicting-Prag}.
 Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions are adequate to capture the human data at the item-level and finds that they are not, even though a condition-level predictor from the RSA model is.
-Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the task- or condition-level and compares them against the human data and each other.
+Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
 We find that not every way of constructing condition-level predictions from item-level scores is equally good.
@@
 Tasks often contain multiple-choice options, and accuracy for a task is determined with a ``winner-takes-all'' (WTA) strategy \citep[e.g.,][]{srivastava2023-BIGbench}.
-
 Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
@@
 Among the choice options is a designated \emph{target option} $y_{ki^{*}}$ that is assumed to be the ``true'' or to-be-selected goal answer.
-
 The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
@@
 Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{fig:posterior-stats}).
+\mf{include information on MCMC, samples, warm-up etc.; follow Kruschke recipe}
 
-To assess goodness-of-fit, we can use the posterior predictive distribution, i.e., the model's predictions about data of the same size and structure as the training data.
+To assess goodness-of-fit, we can use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
+\mf{maybe insert formal definition of post-pred distribution?}
 As a minimal bar, we would require a model trained on observed data $D_{\text{obs}}$ to not be surprised by $D_{\text{obs}}$.
 Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models to be introduced later).
-We see that the RSA model passes this ``visual posterior predictive check'' \citep{Kruschke2021:Bayesian-Analys} for both conditions.
+We see that the RSA model passes this ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-} for both conditions.
 To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
+\mf{maybe insert formal definition of BPPPV?}
 These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
@@
 The items of the reference game experiment differ in the task-relevant features (color, shape, texture), as well as the order of presentation of states and words.
-We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelvant material \citep{DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options.
+We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelvant material \citep{DegenHawkins2020:When-redundancy} \mf{more refs on (color) over-specification}; and machines may be susceptible to the presentation of the order of choice options.
 \mf{references from the NLP literature on effects of prompt variation?}
@@
 Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (0.0001 for production, 0 for interpretation), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
-In contrast, using the global, item-independent RSA model predictions to fit the item-level data, we obtain estimates of posterior predictive $p$-values that do not discredit the model ($\sim 0.338$ for production, $\sim 0.252$ for interpretation).
+In contrast, using the global, item-independent RSA model predictions to fit the item-level data, we obtain estimates of posterior predictive $p$-values that do not discredit the model ($0.338$ for production, $0.252$ for interpretation).
 These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
@@
 
-This section explores different ways of deriving a probabilistic likelihood function for the human data from a Large Language Model at the aggregate, condition-level.
+This section explores different ways of deriving a probabilistic likelihood function for the human data from an LLM at the aggregate, condition-level.
 Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
@@
 The three condition-level predictions yield three different probabilistic models, each with two pairs of condition-dependent parameters $\alpha_{c}$, and $\epsilon_{c}$.
-Using priors and Bayesian posterior estimation as previously for the RSA (see Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and posterior predictive distributions.
+Using priors and Bayesian posterior estimation as previously described for the RSA (see Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
 The usual Bayesian summary statistics for posteriors over model parameters are shown in Table~\ref{fig:posterior-stats}.
-Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
-What is noteworthy about these estimates is that the $\alpha_{c}$ parameters for the wide-scope model have a very large credible range in both conditions.
+What is noteworthy about these estimates is that the $\alpha_{c}$ parameters for the wide-scope model have a very large credible range.
 This is because, already for values of $\alpha_{c} = 1$, the item-level predictions are virtually ``winner-takes-all'' choices, selecting the option with the maximum score with probability almost 1 for almost all items.
@@
 
-Figure~\ref{fig:refgame-counts} shows that only the theoretical model (RSA) and the intermediate-scope model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate.
+Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
+We find that only the theoretical model (RSA) and the intermediate-scope model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
 To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.

### 2023-11-01 03-paper/paper-LLM-probmod.tex
revised §1

@@
 
-% \usepackage[style=authoryear-comp, % Citation marks as [Jef65]
-%             natbib=true,      % Natbib-style cite macros \citeauthor &c.
-%             hyperref=true,    % Cites in pdf are links to bib
-%                               % (hyperref conf.)
-%             maxnames=3,       % truncate name lists if more than 2
-%                               % names appear
-%             doi=false,        % no doi's
-%             url=false,        % no url's
-%             sortcites=false,  % do NOT sort cites in the style of the
-%                               % bibliography
-%             %backref=true     % insert backrefs in reference section
-%             ]{biblatex}
 \usepackage{mypackages}
@@
 
-% Statistical models based on LLM predictions
-% LLMs fuel statistical models for full distributional predictions of human experimental data
-% Large language models for full distributional predictions of human experimental data
-% Statistical modeling with predictors from large language models
-\title{Beyond the best guess: Distributional predictions from large language models for human forced-choice data}
+\title{Language models as probabilistic cognitive models}
 
-% Here you can change the date presented in the paper title
-%\date{September 9, 1985}
-% Or remove it
 \date{}
@@
 % Uncomment to use multiple affiliations variant of author block
-% \uniqueAffiliationtrue
+\uniqueAffiliationtrue
 
 \ifuniqueAffiliation % Standard variant of author block
-  \author{ Michael Franke\thanks{Use footnote for providing further
-		information about author (webpage, alternative
-		address)---\emph{not} for acknowledging funding agencies.} \\
+  \author{ Michael Franke\thanks{Corresponding author.} \\
 	Department of Linguistics\\
 	University of Tübingen\\
-	\texttt{michael.franke@uni-tuebingen.de} \\
+	\texttt{mchfranke@uni-tuebingen.de} \\
 	%% examples of more authors
 	\And
-	Polina Tsvilodub\thanks{Use footnote for providing further
-		information about author (webpage, alternative
-		address)---\emph{not} for acknowledging funding agencies.} \\
+	Polina Tsvilodub \\
 	Department of Linguistics\\
@@
 	\And
-	Fausto Carcassi\thanks{Use footnote for providing further
-		information about author (webpage, alternative
-		address)---\emph{not} for acknowledging funding agencies.} \\
-	Department of Linguistics\\
-	University of Tübingen\\
+	Fausto Carcassi \\
+	ILLC\\
+	University of Amsterdam\\
 	\texttt{fausto.carcassi@gmail.com} \\
-	%% \AND
-	%% Coauthor \\
-	%% Affiliation \\
-	%% Address \\
-	%% \texttt{email} \\
-	%% \And
-	%% Coauthor \\
-	%% Affiliation \\
-	%% Address \\
-	%% \texttt{email} \\
-	%% \And
-	%% Coauthor \\
-	%% Affiliation \\
-	%% Address \\
-	%% \texttt{email} \\
 }
@@
 \begin{abstract}
-	{\textcolor{gray}{fill me}}
+  State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks.
+  But targeted assessment with methods inspired by experimental psychology highlight dissimilarities between human behavior and LLM predictions.
+  Going even a step further, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of probabilistic cognitive modeling.
+  We identify core differences, in terms of conceptual interpretation and usual practical evaluation, between LLMs and probabilistic cognitive models (PCMs).
+  Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level).
+  Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs fail to capture the variance in the human data at the item-level.
+  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, similar to those that a probabilistic cognitive model might make, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to our data.
+  These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at an aggregate level, for which they are, however, not designed to make predictions in the first place.
 \end{abstract}
@@
 
-\mf{would be good to add references here}
-
-The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} enabled a new generation of powerful large language models (LLMs) \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and} which excel on standard benchmarks and promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
+The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} has enabled a new generation of powerful large language models (LLMs) \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and}.
+State-of-the-art LLMs excel on many benchmark tests \mf{refs for standard and non-standard tests here} and promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
 % especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
@@
 
-For all of these applications, it is crucial to understand what LLMs can or cannot reliable do.
+For all of these applications, it is crucial to understand what LLMs can or cannot reliably do.
 The prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' exists, averaged over many instances of this task.
@@
 For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.
+This work seeks to extend the investigation of the human-likeness of LLMs even further, by comparing quantitative predictions derivable from LLMs for empirical data from human multiple-choice behavior to the predictions of probabilistic cognitive models, which are usually used to analyze, or explain, these data.
+In slogan form, while a lot of previous work has used targeted assessment of LLM capabilities from the point of view of the experimental psychologist, we here adopt the perspective of the cognitive modeller.
 
-This work seeks to extend the investigation of the human-likeness of quantitative information derived from LLMs even further, by exploring whether LLM-derived measures can feed into probabilistic models predicting the full distribution of human answers in a forced-choice tasks.
-The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
-Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
+For the purposes of this investigation, a probabilistic cognitive model (PCM) is characterized by providing a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each observable data $D_{C}$ for any experimental condition $C$ and numerical parameters $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
+Usually, conditions $C$ provide a categorization of different tasks which is meaningful against the theoretical background of experimental investigation, and parameters $\theta$ modulate the predicted likelihood of the data in human interpretable ways (see Section~\ref{motivation} for more detail).
+There are non-trivial, insightful differences between LLMs and PCMs, both at the conceptual level and in terms of common practices in evaluation, which we will explore in this paper to help better delineate the potential and limitations of LLMs.
+For one, the ``primitive predictions'' that LLMs provide are not at the level of a condition $C$, but at a lower item-based level.
+In this sense, LLMs could be said to be less explanatory than PCMs by their very nature.
+For another, LLMs are often just assessed based on their ability to identify a target answer, not to predict the full distribution observed in the data.
+Nevertheless, distributional predictions for condition-level data can in principle be derived from LLMs, but there are many ways of doing so, yet with little consensus of how to do so.
+In this paper, we therefore introduce and systematically compare different ways of carving numerical information obtained from an LLM into a parameterized PCM for condition-level data.
 
 A major contribution of this work is methodological.
-By embedding numerical scores from LLMs into probabilistic models for human multiple-choice data, we can use Bayesian data analysis to train, test and compare different LLM-derived predictor scores.
-\mf{describe high-level results here}.
+We subjected human participants and an LLM (OpenAI's GPT-3.5 model \mf{add ref}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
+We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare different LLM-derived PCMs, which we also compare to an established, theoretically informed PCM based on standard likelihood-based measures of statistical goodness-of-fit.
+This is quite different, and arguably more stringent and insightful, than the standard practice of evaluating LLMs based on accuracy scores, determined by a winner-takes-all strategy to identify a single target response.
+We find that variability predicted by the LLM at the item-level is not borne out by the human data, and that some but not all ways of constructing condition-level predictions are equally good.
+These results suggest that it is not impossible to build predictive models for condition-level data, similar to PCMs, but that seemingly subtle details in the derivation of likelihoods matter.
+The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not.
 
 The paper is structured as follows.
-Section~\ref{motivation} further motivates the desire to assess quantitative information from LLMs, and it introduces the distinction between item-level and condition-level predictions.
+Section~\ref{motivation} further elaborates on the conceptual differences between LLMs and PCMs and the practical differences between prevalent evaluation methods for the two.
 Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a salient probabilistic cognitive model \citep{FrankGoodman2012:Predicting-Prag}.
@@
 Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
-We find that not every way of constructing condition-level predictions from item-level scores is equally good.
-\mf{grant conclusion?}
 
@@
 
+\newpage
+
+\section*{Snippets}
+
+\begin{itemize}
+  \item for conclusion?
+  \begin{itemize}
+    \item The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
+    Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
+  \end{itemize}
+\end{itemize}
+
 \end{document}

### 2023-11-01 03-paper/paper-LLM-probmod.tex
revisions §2

@@
 For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.
-This work seeks to extend the investigation of the human-likeness of LLMs even further, by comparing quantitative predictions derivable from LLMs for empirical data from human multiple-choice behavior to the predictions of probabilistic cognitive models, which are usually used to analyze, or explain, these data.
+The work presented in this paper seeks to extend the investigation of the human-likeness of LLMs even further, by comparing quantitative predictions derivable from LLMs for empirical data from human multiple-choice behavior to the predictions of probabilistic cognitive models, which are usually used to analyze, or explain, these data.
 In slogan form, while a lot of previous work has used targeted assessment of LLM capabilities from the point of view of the experimental psychologist, we here adopt the perspective of the cognitive modeller.
@@
 For the purposes of this investigation, a probabilistic cognitive model (PCM) is characterized by providing a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each observable data $D_{C}$ for any experimental condition $C$ and numerical parameters $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
-Usually, conditions $C$ provide a categorization of different tasks which is meaningful against the theoretical background of experimental investigation, and parameters $\theta$ modulate the predicted likelihood of the data in human interpretable ways (see Section~\ref{motivation} for more detail).
+% Usually, conditions $C$ provide a categorization of different tasks which is meaningful against the theoretical background of experimental investigation, and parameters $\theta$ modulate the predicted likelihood of the data in human interpretable ways (see Section~\ref{motivation} for more detail).
 There are non-trivial, insightful differences between LLMs and PCMs, both at the conceptual level and in terms of common practices in evaluation, which we will explore in this paper to help better delineate the potential and limitations of LLMs.
@@
 A major contribution of this work is methodological.
-We subjected human participants and an LLM (OpenAI's GPT-3.5 model \mf{add ref}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
-We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare different LLM-derived PCMs, which we also compare to an established, theoretically informed PCM based on standard likelihood-based measures of statistical goodness-of-fit.
+We introduce different ways of defining condition-level, distributional predictions based on LLM-provided numerical measures, and compare these LLM-fuelled probabilistic models against each other, but also against an established, theoretically informed PCM, the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
+To obtain data for this investigation, we subjected human participants and an LLM (OpenAI's GPT-3.5 model \mf{add ref}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
+We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare PCMs based on standard likelihood-based measures of statistical goodness-of-fit to experimental data.
 This is quite different, and arguably more stringent and insightful, than the standard practice of evaluating LLMs based on accuracy scores, determined by a winner-takes-all strategy to identify a single target response.
@@
 The paper is structured as follows.
-Section~\ref{motivation} further elaborates on the conceptual differences between LLMs and PCMs and the practical differences between prevalent evaluation methods for the two.
+Section~\ref{motivation} motivates the investigation in this paper, by enlarging on the importance of deriving distributional predictions for conditional-level forced-choice data, and further elaborating on the conceptual differences between LLMs and PCMs and the practical differences between prevalent evaluation methods for the two.
 Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a salient probabilistic cognitive model \citep{FrankGoodman2012:Predicting-Prag}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Motivation: Why care for distributional predictions?}
+\section{Motivation: Distributional predictions for condition-level forced-choice data}
 \label{motivation}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Psychological research into the workings of the human mind aims to find generalizable, systematic patterns in information processing in different domains of cognition.
+Experimental work therefore often compares human performance in different \emph{experimental conditions}, which reflect the general factors that are hypothesized to influence behavior, but might instantiate each experimental condition with different \emph{experimental items}, which are not under scrutiny for any systematic, predictable effect on the observed measurements.
+For example, classical research on human memory \citep{AtkinsonShiffrin1968:Human-memory:-A} investigated the effect of rehearsal on memory consolidation.
+Relevant experiments compare recall with and without rehearsal (experimental conditions), while using different items (words, numbers, etc., to be memorized) in each instantiation of a memory task.
+Likewise, when studying how hearing a color word can facilitate a same-or-different judgement of color swatches \citep{Rosch1975:The-Nature-of-M}, the main experimental manipulation, reflected in different experimental conditions, concerns the typicality of shown color swatches, while the variability between different color words like ``blue'' or ``green'', is less important to this research question and so treated as \emph{item-level variation}.
+Consequently, a typical psychological experiment is mainly interested in assessing behavior at the level of the experimental condition, because that is where the distinctions relevant to the research question reside.
+Nevertheless, each experimental condition $C$ can be, and often is, instantiated with different instantiations of $C$, variation among which is deemed less relevant to the research question at hand.
+
+Data from human participants for experiments of this kind usually shows some variability between items, and also variability between participants.
+This variability is commonly incorporated in statistical analyses as random stochastic variation, e.g., by using hierarchical (regression) models \citep{Jaeger2008:Categorical-dat,barr2013,SorensenHohensteinb2016:Bayesian-linear}.
+Still, the focus of interest usually remains at the condition-level effects, because it is this more abstract level of behavioral aggregation that is relevant for generalizable theory building.
+Similarly, when analyzing or explaining data from psychological experiments with a probabilistic cognitive model (PCM), the PCM will naturally be set-up to predict data at the level of the relevant experimental manipulation, i.e., at the condition-level.
+Even though item-level (and individual-level) variation can, and sometimes should, be incorporated into probabilistic cognitive modeling \citep[e.g.,][]{NilsonRieskamp2011:Hierarchical-Ba,Lee2011:How-Cognitive-M,ScheibehenneRieskamp2013:Testing-the-Ada,FrankeDablander2016:What-does-the-c}, the focus of cognitive modeling is likewise to predict or explain systematic variation between experimental conditions.
+Therefore, a basic PCM is essentially characterized by a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each logically possible data observation $D_{C}$ for any experimental condition $C$ and numerical parameters $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
+The difference between a theory-driven PCM and a standard, theory-neutral statistical model (like a regression model), which may also define a parameterized likelihood function of condition-level data, is that the PCM's likelihood function is defined in terms that are considered to be conceptually meaningful for the research domain at hand (see Section~\ref{sec:model-pred-from} for a concrete example).
+
+PCMs, thus construed, differ strongly from LLMs.
+LLMs make predictions about the likelihood of words given a preceding or surrounding text.
+They do not, without further derivation, provide predictions at a conceptually meaningful level of abstraction, similar to experimental conditions.
+In this way, LLMs are not explanatory models in the same sense as PCMs might be: they simply do not operate at the level of generalizable abstraction necessary for a common understanding of ``explanation.'' \mf{refs from PhilOScience?}
+Nevertheless, we can derive predictions from LLM's word-level predictions, first at the item-level and also at the condition-level.
+Indeed, this is what standard benchmark-testing does, albeit with a focus on a winner-take-all strategy, and not with an eye towards deriving a full likelihood function for empirical data.
 
@@
 
-\mf{the following paragraphs should be shortened and sharpened}
-While generally a pragmatic and useful approach, there are several reasons why going beyond the best guess can be beneficial.
-Consider first a high-level conceptual argument.
-If task performance is categorical (in the most extreme case: binary), somewhere along the path from a numerical item-level score to accuracy a discontinuity has to be introduced.
+While generally a pragmatic and useful approach, assessing LLM performance with a WTA-based accuracy measure can misleading, because it ignores potentially relevant distributional information.
+At an abstract level, the problem is that if task performance is categorical (in the most extreme case: binary), somewhere along the path from a numerical item-level score to accuracy a discontinuity has to be introduced.
 Every such discontinuity bottleneck entails loss of information.
-As this information might be useful, discontinuity should ideally happen as late as possible; or even better: not at all, unless it is irrelevant.
-Consequently, the question becomes: when is the additional information inherent in the scores (not) relevant for assessing LLM performance?
+As this information might be useful, discontinuity should ideally happen as late as possible; or even better: not at all, unless we can be sure that it is always irrelevant.
+But since we hardly can be sure that it is irrelevant for all circumstances, we should better assess performance based on a models' full distributional predictions.
 
 The WTA approach implicitly assumes that item-level choices are resolved by a greedy-like choice of the best alternative.
-It is therefore a good measure of accuracy when the LLM's main use cases are of a greedy-like nature as well.
-When applications rely on a less extreme sampling based strategy, the WTA approach can be generalized to softmax sampling.
-The probability of choosing option $y_{ki}$ for item $I_{k}$ is:
+% It is therefore a good measure of accuracy when the LLM's main use cases are of a greedy-like nature as well.
+To generalize to less extreme sampling scenarios, the WTA approach can be generalized to softmax sampling, where the probability of choosing option $y_{ki}$ for item $I_{k}$ is:
 %
@@
 For $\alpha \rightarrow \infty$, this softmax strategy converges to the WTA strategy.
-But in general, these two approaches can yield even categorically different predictions.
 
+WTA-based accuracy measures can differ form accuracy measures obtained from the more general softmax-based strategy.
+Whether there is a difference, depends on the variation in item-level scores, in particular the relation between score-ordering and score-differences.
 Imagine that there are two options, and that the target option's score is a small $\epsilon$ better in 80\% of the task's items, and otherwise worse.
@@
 
-Differences in accuracy scores between a WTA-based or a more general softmax-based strategy depend on the variation in item-level scores, in particular the relation between score-ordering and score-differences.
-(The example in the previous paragraph was set up deliberately to have one score-ordering for very small score-differences, and another for huge score-differences.)
-This raises an important general question, namely how to deal with item-level variation in a system's predictions in the first place.
-Data from forced-choice tasks with human participants will likely show item-level variation as well, which is why hierarchical modeling accommodating item-level differences is recommended in statistical analyses \citep[e.g.,][]{Jaeger2008:Categorical-dat,barr2013}.
-But whether item-level variation in LLM scores aligns with differences in human choice rates for each item is an empirical question that has to be addressed by explicit modeling and comparison; an exercise that this paper starts to take on.
-On \emph{a priori} grounds, the WTA-based strategy may be deemed implausible as a predictor of human-level choice behavior for a simple item: it is unlikely that, for each item, there is a single choice option that (almost) \emph{all} participants will choose.
+The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.
+Instead, the alternative softmax-based strategy conservatively generalizes WTA-based performance assessment, but also yields full distributional predictions at the item-level, i.e., a non-degenerate likelihood prediction for \emph{all} choice options, whenever $\alpha$ is not infinite.
+When aspiring to predict experimental data with LLMs, the WTA-based strategy may almost sure be ruled out as a predictor of human-level choice behavior for a simple item: for many experimental set-ups, it is highly unlikely that, for each item, there is a single choice option that (almost) \emph{all} participants will choose.
+But whether the softmax-based item-level predictions, with variable $\alpha$, align with item-level variation in human choice data is an empirical question that has to be addressed by explicit modeling and comparison; an exercise that this paper starts to take on (see Section~\ref{sec:item-level-pred}).
 
-To the extent that machine predictions show extreme, unsystematic, erratic item-level variation which is quite unlike that of human participants in comparable experiments, an alternative strategy suggests itself, namely to \emph{not} judge a system by its average item-level predictions, but based on its task- or condition-level predictions obtained from different ways of aggregating over item-level variation.
-Section~\ref{llm-predictions-for-reference-games} therefore introduces and compares different ways of deriving condition-level predictions from (variable) item-level scores.
-Properly constructed condition-level predictions might be more aligned with average human choice rates, despite the fact that each complex system (machine and human) shows quite different underlying item-level variation.
-In other words, we should ask whether modern LLMs can predict aggregate human data by \emph{some} process of aggregating its item-level scores, because it might turn out that predicting aggregates is the best the LLMs can do (in particular cases).
+Whether item-level predictions by an LLM are empirically vindicated or not, they may serve as the backbone for deriving condition-level predictions, just as the usual accuracy scores are obtained by averaging over all items belonging to a task or experimental condition.
+However, if we generalize to softmax-based distributional predictions, we find that there are different ways of deriving condition-level predictions (see Section~\ref{llm-predictions-for-reference-games}).
+We should therefore address the also the empirical question of which, if any, of these constructions of condition-level predictions provides a good fit to human data.
 

### 2023-11-01 03-paper/paper-LLM-probmod.tex
integrated feedback §3-6

@@
 We introduce different ways of defining condition-level, distributional predictions based on LLM-provided numerical measures, and compare these LLM-fuelled probabilistic models against each other, but also against an established, theoretically informed PCM, the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
-To obtain data for this investigation, we subjected human participants and an LLM (OpenAI's GPT-3.5 model \mf{add ref}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
+To obtain data for this investigation, we subjected human participants and an LLM (OpenAI's GPT-3.5 model \mf{which model exactly, when assessed, cite paper}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
 We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare PCMs based on standard likelihood-based measures of statistical goodness-of-fit to experimental data.
@@
 Section~\ref{motivation} motivates the investigation in this paper, by enlarging on the importance of deriving distributional predictions for conditional-level forced-choice data, and further elaborating on the conceptual differences between LLMs and PCMs and the practical differences between prevalent evaluation methods for the two.
-Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a salient probabilistic cognitive model \citep{FrankGoodman2012:Predicting-Prag}.
-Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions are adequate to capture the human data at the item-level and finds that they are not, even though a condition-level predictor from the RSA model is.
+Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a commonly used, theoretically-motivate PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
+Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions are adequate to capture the human data at the item-level and finds that they are not, even though the theoretically-motivated RSA model is.
 Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
@@
 Whether there is a difference, depends on the variation in item-level scores, in particular the relation between score-ordering and score-differences.
-Imagine that there are two options, and that the target option's score is a small $\epsilon$ better in 80\% of the task's items, and otherwise worse.
+Imagine that there are two options, and that the target option's score is a small $\epsilon$ higher in 80\% of the task's items, and otherwise lower.
 The WTA-based accuracy is 0.8.
@@
 Each trial consists of a set of objects, the so-called context.
-In production trials, participants choose a single word to describe a target object from the context.
+In production trials, participants choose a single word to describe a trigger object from the context.
 In interpretation trials, an object is selected as the likely object a trigger word referring to.}
@@
 A reference game consists of two players, a speaker and an interpreter, who jointly observe a set of objects, usually referred to as context (see Figure~\ref{fig:ref-game}).
-In the \textbf{production condition}, the speaker is assigned a \emph{target object} from the context set which they have to describe to the interpreter.
+In the \textbf{production condition}, the speaker is assigned a \emph{trigger object} from the context set which they have to describe to the interpreter.
 In the \textbf{interpretation condition}, the interpreter observes a description, here called \emph{trigger word}, and chooses one of the objects from the context set.
-The goal of the game is, for the speaker, to choose a description that enables the interpreter to choose the target object; and, for the interpreter, to guess correctly which object the speaker had in mind.
+The goal of the game is, for the speaker, to choose a description that enables the interpreter to choose the trigger object; and, for the interpreter, to guess correctly which object the speaker had in mind when using the trigger word.
 
-The example in Figure~\ref{fig:ref-game} is a standard case, which we will use throughout, where choices are informative about the pragmatic reasoning that decision makers engage in.
+The example in Figure~\ref{fig:ref-game} is a standard case, which we will use throughout, where human choices are informative about the pragmatic reasoning that human decision makers engage in.
 In this example, there are two features that differ across three objects (here shape and color).
@@
 %
-In a critical production trial, the target object to describe is one of the two objects with a unique feature.
+In a critical production trial, the trigger object to describe is one of the two objects with a unique feature.
 The speaker has four words to choose from.
-The \textbf{target utterance} is the word which uniquely describes the target object.
-The \textbf{distractor utterance} is the word that is true of the target object, but also true of another object.
-The other utterances, both of which are false of the target are \textbf{competitor utterance}.
+The \textbf{target utterance} is the word which uniquely describes the trigger object.
+The \textbf{distractor utterance} is the word that is true of the trigger object, but also true of another object.
+The other utterances, both of which are false of the trigger object are \textbf{competitor utterance}.
 %
@@
 We replicated a simple reference game with human participants in which each trial instantiated the structure of the example shown in Figure~\ref{fig:ref-game}.
-While previous reference games with human participants used pictorial representations of objects, and sometimes even pictorial representations of messages, we implemented a text-only version in order to be able to compare the predictions of LLMs for the exact same stimuli.
+While previous reference games with human participants used pictorial representations of objects, and sometimes even pictorial representations of messages, we implemented a text-only version in order to be able to compare the predictions of LLMs for human data, when both LLMs and humans processed the same textual representation of the stimuli.
 The experiment was realized as an online task using \texttt{magpie} \citep{FrankeJi:magpie:-Minimal}.\footnote{
@@
 For each item, the context consisted of three objects.
-Objects are defined by a triple of properties, namely a color, a shape and a texture.
+As in the original paper by \citet{FrankGoodman2012:Predicting-Prag}, objects are defined by a triple of properties, namely a color, a shape and a texture.
 For each property, there were four possible values, e.g., blue, green, red, and orange for color.
@@
 % \citep{QingFranke2013:Variations-on-a}.
-For example, in previous forced-choice reference games with human participants with pictorial presentations of objects, \citet{QingFranke2013:Variations-on-a} observed the following proportions: $\tuple{0.882, 0.118, 0}$ in the production and $\tuple{0.492, 0.506, 0.003}$ in the interpretation condition (for 288 observations in each condition).
+For example, in previous forced-choice reference games with human participants with pictorial presentations of objects, \citet{QingFranke2013:Variations-on-a} observed the following proportions of target, competitor and distractor options: $\tuple{0.882, 0.118, 0}$ in the production and $\tuple{0.492, 0.506, 0.003}$ in the interpretation condition (for 288 observations in each condition).
 
@@
 %
-The pragmatic speaker soft-optimizes the choice of utterance to minimize the literal listener's surprisal for the to-be-communicated state:
+The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated (e.g., the trigger object/nicefra):
 %
@@
     The semantic meaning is shown as a matrix of binary truth-values.
-    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents), assume $\alpha=1$ and are shown as row-stochastic matrices.
+    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents) for $\alpha=1$ are shown as row-stochastic matrices.
   }
@@
 
-Figure~\ref{fig:RSA-example} gives example calculations (assuming flat priors and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
-For $\alpha=1$, the model predicts that the probabilities of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{0.6, 0.4, 0}$ for the interpretation condition.
+Figure~\ref{fig:RSA-example} gives example calculations (assuming a flat prior $P(s)$ and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
+For $\alpha=1$, the model predicts that the probabilities of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{\nicefrac{3}{5}, \nicefrac{2}{5}, 0}$ for the interpretation condition.
 Increasing $\alpha$ will increase the odds of target over competitor choices.
 Yet, the model also predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
-It is therefore common to include a random error probability for each choice \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}, such as via Laplace smoothing: if $P_{r}(R_{l}, C, \alpha_{c})$ is the RSA model's probabilistic prediction for response category $R_{l}$ (target, competitor, or distractor) for condition $C$ (production or interpretation) and condition-specific optimality $\alpha_{c}$, the random-error smoothed prediction is:
+It is therefore common to include a small error probability $\epsilon$, with which a choice would be made at random \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}: if $P_{r}(R_{l}, C, \alpha_{c})$ is the RSA model's probabilistic prediction for response category $R_{l}$ (target, competitor, or distractor) for condition $C$ (production or interpretation) and condition-specific optimality $\alpha_{c}$, the random-error smoothed prediction is:
 \begin{align*}
-  P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c}) \propto P_{r}(R_{l}, C; \alpha_{c}) +  \frac{\epsilon_{c}}{3}\,,
+  P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
 \end{align*}
-where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
+where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.\mf{FC objected to ``condition-specificity'' which is fair, but I leave it as it was; my reasoning: we use model checking to rule out models; in this setting the conservative strategy is to err on the side of making the models too powerful (even at the cost of conceptual plausibility)}
 The result is a four-parameter model, one pair of parameters per condition, which provides a likelihood function for the categorical choice data and so can be fitted to the data and compared against other probabilistic models providing a likelihood function for the same data.
@@
 To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
-\mf{maybe insert formal definition of BPPPV?}
 These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
-Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
-The results from Table~\ref{tab:Bppp-values} show that the RSA model passes this numerical test of model adequacy.
+% Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
+% The results from Table~\ref{tab:Bppp-values} show that the RSA model passes this numerical test of model adequacy.
 
-The following sections apply these test of Bayesian model criticism also to models built around predictor values from LLMs.
+The following sections apply these tests of Bayesian model criticism also to models built around predictor values from LLMs.
 Section~\ref{sec:item-level-pred} first looks at the data from each item individually, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.
@@
 
-Figure~\ref{fig:refgame-counts} shows the condition-level counts for choices for all of the 100 different items combined.
+Figure~\ref{fig:refgame-counts} shows the condition-level counts for human choices for all of the 100 different items combined.
 An LLM, on the other hand, first and foremost makes predictions about each item.
-The items of the reference game experiment differ in the task-relevant features (color, shape, texture), as well as the order of presentation of states and words.
-We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelvant material \citep{DegenHawkins2020:When-redundancy} \mf{more refs on (color) over-specification}; and machines may be susceptible to the presentation of the order of choice options.
-\mf{references from the NLP literature on effects of prompt variation?}
+As mentioned previously, the items of the reference game experiment differ in the task-relevant features (color, shape, texture), as well as the order of presentation of states and words.
+We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelevant material \citep{DegenHawkins2020:When-redundancy} \mf{more refs on (color) over-specification}; and machines may be susceptible to the presentation of the order of choice options.
+\mf{@PT references from the NLP literature on effects of prompt variation?}
 The main question to be addressed in this section is whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.
@@
 %
+\mf{check if formula is redundant; already introduced?}
 The item-level score of response type $R_{l}$ is $S(R_{l}, I_{k}) = S(F(R_{l}, I_{k}))$.
@@
 Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
-In the following, we first describe the different options of deriving LLM-based predictor terms.
+In the following, we first describe the different options of deriving LLM-based predictors.
 We then compare all probabilistic models based on their adequacy of explaining the human choice data.
@@
     Number indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
+    \mf{make wide format}
   }
@@
 Figure~\ref{fig:prediction-range} shows the range of predictions that the intermediate-scope model is, in principle, capable of making (for the whole possible range of \(\alpha\) and \(\epsilon\) values, disregarding Bayesian priors).
-There are distributions over response types which the model does not explain predict \emph{ex ante}.
+There are distributions over response types which the model does not predict \emph{ex ante}.
 For example, the model does not predict cases where the number of competitor choices is almost as high as the number of target choice, as would be the case if participants would consistently \emph{not} engage in pragmatic reasoning.
@@
   \end{itemize}
+  \item PT: the high-level take-away that item-level LLM predictions don't fit human predictions well might, as always, be met with the comment that the task is rather artificial.
+maybe this should go into the discussion
+  \item FC: The fact that the aggregate predictions track aggregate human behaviour means that the discrepancies are washing out.
+But there's also the interesting question of finding the systematic differences between LLMs and people.
+The task then would be to find sets of stimuli where the differences do \emph{not} wash out and ask: What if anything do these stimuli share?
+This paper could also be a jumping board for this line of research.
+  \item PT: maybe a conclusion in the spirit of "cogmod in the age of LLMs"/ "computational psycholinguistics \& careful assessment of reasoning / linguistic abilities of LLMs against human performance" might be not entirely ludicrous too
 \end{itemize}

### 2023-11-02 03-paper/paper-LLM-probmod.tex
revision §1

@@
   State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks.
-  But targeted assessment with methods inspired by experimental psychology highlight dissimilarities between human behavior and LLM predictions.
+  But targeted assessment with methods inspired by experimental psychology highlights dissimilarities between human behavior and LLM predictions.
   Going even a step further, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of probabilistic cognitive modeling.
@@
   Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs fail to capture the variance in the human data at the item-level.
-  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, similar to those that a probabilistic cognitive model might make, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to our data.
+  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, thus building PCM-like predictions from LLM-derived information, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to our data.
   These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at an aggregate level, for which they are, however, not designed to make predictions in the first place.
@@
 
-The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} has enabled a new generation of powerful large language models (LLMs) \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and}.
-State-of-the-art LLMs excel on many benchmark tests \mf{refs for standard and non-standard tests here} and promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
+The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} has enabled a new generation of powerful large language models  \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and}.
+State-of-the-art LLMs excel on many benchmark data sets \citep{srivastava2023-BIGbench,PerezRinger2023:Discovering-Lan}, and so promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
 % especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
@@
 The prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' exists, averaged over many instances of this task.
-Benchmark-driven assessments are very useful to systematically study the effects of scale, e.g., in an engineering-oriented context \citep[e.g.,][]{srivastava2023-BIGbench}.
+Benchmark-driven assessments are very useful for a systematic study of the effects of scale, e.g., in an engineering-oriented context \citep[e.g.,][]{srivastava2023-BIGbench}.
 Other recent work is more psychologically oriented and asks to what extent LLM performance is ``human-like.''
-LLM performance is therefore compared to human choice behavior in psychological experiments to investigate whether LLMs predict patterns of human answer behavior qualitatively \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
+LLM performance is therefore compared to human choice behavior in psychological experiments and investigates whether LLMs predict patterns of human answer behavior \emph{qualitatively} \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
 Going further, there is also work investigating whether LLMs can make adequate \emph{quantitative predictions}.
-For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.
-The work presented in this paper seeks to extend the investigation of the human-likeness of LLMs even further, by comparing quantitative predictions derivable from LLMs for empirical data from human multiple-choice behavior to the predictions of probabilistic cognitive models, which are usually used to analyze, or explain, these data.
+For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of the N400 component of event-related-potentials in EEG measurements \citep{LindborgRabovsky2021:Meaning-in-brai}.
+The work presented in this paper seeks to extend the investigation of the human-likeness of LLMs by looking at empirical data from human multiple-choice behavior and comparing an established PCM, which is commonly use to predict or explain this kind of data, to structurally similar quantitative predictions derived from LLMs.
 In slogan form, while a lot of previous work has used targeted assessment of LLM capabilities from the point of view of the experimental psychologist, we here adopt the perspective of the cognitive modeller.
 
-For the purposes of this investigation, a probabilistic cognitive model (PCM) is characterized by providing a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each observable data $D_{C}$ for any experimental condition $C$ and numerical parameters $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
+For the purposes of this investigation, a probabilistic cognitive model (PCM) is characterized by providing a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each observable data $D_{C}$ for any experimental condition $C$ and parameter values $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
 % Usually, conditions $C$ provide a categorization of different tasks which is meaningful against the theoretical background of experimental investigation, and parameters $\theta$ modulate the predicted likelihood of the data in human interpretable ways (see Section~\ref{motivation} for more detail).
-There are non-trivial, insightful differences between LLMs and PCMs, both at the conceptual level and in terms of common practices in evaluation, which we will explore in this paper to help better delineate the potential and limitations of LLMs.
+There are non-trivial, insightful differences between LLMs and PCMs, both at the conceptual level and in terms of common practices in evaluation, which we will explore in this paper in order to contribute to a better delineation of the potential and limitations of LLMs.
 For one, the ``primitive predictions'' that LLMs provide are not at the level of a condition $C$, but at a lower item-based level.
-In this sense, LLMs could be said to be less explanatory than PCMs by their very nature.
+% In this sense, LLMs could be said to be less explanatory than PCMs by their very nature (see Section~\ref{motivation}).
 For another, LLMs are often just assessed based on their ability to identify a target answer, not to predict the full distribution observed in the data.
@@
 We introduce different ways of defining condition-level, distributional predictions based on LLM-provided numerical measures, and compare these LLM-fuelled probabilistic models against each other, but also against an established, theoretically informed PCM, the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
-To obtain data for this investigation, we subjected human participants and an LLM (OpenAI's GPT-3.5 model \mf{which model exactly, when assessed, cite paper}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
+To obtain data for this investigation, we subjected human participants and an LLM (a version of OpenAI's GPT-3.5; see Section~\ref{sec:item-level-pred}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
 We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare PCMs based on standard likelihood-based measures of statistical goodness-of-fit to experimental data.
@@
 We find that variability predicted by the LLM at the item-level is not borne out by the human data, and that some but not all ways of constructing condition-level predictions are equally good.
-These results suggest that it is not impossible to build predictive models for condition-level data, similar to PCMs, but that seemingly subtle details in the derivation of likelihoods matter.
+These results suggest that it is not impossible to build predictive models for condition-level data, similar to PCMs, on top of LLM-supplied information, but that seemingly subtle details in the derivation of likelihoods matter.
 The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not.
@@
 
-Using item-level scores from GPT-3.5, we fit a model to the human data at the item-level.
+Using item-level scores from GPT-3.5, we fit a model to the human data at the item-level \mf{which model exactly, when assessed, cite paper}.
 The data to be explained is the set of all triples of response type counts $\tuple{t_{k}, c_{k}, d_{k}}$ (target, competitor, distractor), separately for each item $I_{k}$.

### 2023-11-05 03-paper/paper-LLM-probmod.tex
updated §1

@@
 The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} has enabled a new generation of powerful large language models  \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and}.
-State-of-the-art LLMs excel on many benchmark data sets \citep{srivastava2023-BIGbench,PerezRinger2023:Discovering-Lan}, and so promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
+State-of-the-art LLMs excel on many benchmark data sets \citep[e.g.,][]{srivastava2023-BIGbench,PerezRinger2023:Discovering-Lan}, and so promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
 % especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
@@
 The prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' exists, averaged over many instances of this task.
-Benchmark-driven assessments are very useful for a systematic study of the effects of scale, e.g., in an engineering-oriented context \citep[e.g.,][]{srivastava2023-BIGbench}.
-Other recent work is more psychologically oriented and asks to what extent LLM performance is ``human-like.''
-LLM performance is therefore compared to human choice behavior in psychological experiments and investigates whether LLMs predict patterns of human answer behavior \emph{qualitatively} \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
-Going further, there is also work investigating whether LLMs can make adequate \emph{quantitative predictions}.
+Benchmark-driven assessments are very useful for engineering purposes, e.g., to systematically study of the effects of scale \citep[e.g.,][]{srivastava2023-BIGbench}, but is also used to ask to what extent LLM performance is ``human-like.''
+In more psychologically oriented work, LLM performance is therefore compared to human choice behavior in psychological experiments and investigates whether LLMs predict patterns of human answer behavior  \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
+The main focus is often to compare qualitative patterns in LLM predictions and human data, but there is also work investigating whether LLMs can make adequate \emph{quantitative predictions}.
 For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of the N400 component of event-related-potentials in EEG measurements \citep{LindborgRabovsky2021:Meaning-in-brai}.
-The work presented in this paper seeks to extend the investigation of the human-likeness of LLMs by looking at empirical data from human multiple-choice behavior and comparing an established PCM, which is commonly use to predict or explain this kind of data, to structurally similar quantitative predictions derived from LLMs.
-In slogan form, while a lot of previous work has used targeted assessment of LLM capabilities from the point of view of the experimental psychologist, we here adopt the perspective of the cognitive modeller.
+The work presented in this paper seeks to extend the investigation of the human-likeness of LLMs by looking at empirical data from human multiple-choice behavior and comparing an established, theoretically motivated probabilistic cognitive model, which is commonly use to predict or explain this kind of data, to structurally similar quantitative predictions derived from LLMs.
+In short, while a lot of previous work has used targeted assessment of LLM capabilities from the point of view of the experimental psychologist, we here adopt the more specific perspective of the cognitive modeller.
 
-For the purposes of this investigation, a probabilistic cognitive model (PCM) is characterized by providing a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each observable data $D_{C}$ for any experimental condition $C$ and parameter values $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
+For the purposes of this investigation, a probabilistic cognitive model (PCM) is characterized by providing a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each possible data observation $D_{C}$ for any experimental condition $C$ and parameter values $\theta$ \citep[e.g.,][]{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
 % Usually, conditions $C$ provide a categorization of different tasks which is meaningful against the theoretical background of experimental investigation, and parameters $\theta$ modulate the predicted likelihood of the data in human interpretable ways (see Section~\ref{motivation} for more detail).
@@
 For another, LLMs are often just assessed based on their ability to identify a target answer, not to predict the full distribution observed in the data.
-Nevertheless, distributional predictions for condition-level data can in principle be derived from LLMs, but there are many ways of doing so, yet with little consensus of how to do so.
-In this paper, we therefore introduce and systematically compare different ways of carving numerical information obtained from an LLM into a parameterized PCM for condition-level data.
+% Nevertheless, distributional predictions for condition-level data can in principle be derived from LLMs, but there are many ways of doing so, yet with little consensus of how to do so.
 
-A major contribution of this work is methodological.
-We introduce different ways of defining condition-level, distributional predictions based on LLM-provided numerical measures, and compare these LLM-fuelled probabilistic models against each other, but also against an established, theoretically informed PCM, the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
+% A major contribution of this work is methodological.
+This paper introduces and systematically compare different ways of carving numerical information obtained from an LLM into a structure similar to a parameterized PCM and address the empirical question of whether such an ``LLM-based PCM'' adequately captures the distribution of human forced-choice data at different levels of aggregation.
 To obtain data for this investigation, we subjected human participants and an LLM (a version of OpenAI's GPT-3.5; see Section~\ref{sec:item-level-pred}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
-We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare PCMs based on standard likelihood-based measures of statistical goodness-of-fit to experimental data.
+We introduce different ways of defining condition-level, distributional predictions based on LLM-provided numerical measures.
+We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare these ``LLM-based PCMs'' against each other, and against an established, theoretically informed PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
 This is quite different, and arguably more stringent and insightful, than the standard practice of evaluating LLMs based on accuracy scores, determined by a winner-takes-all strategy to identify a single target response.
@@
 These results suggest that it is not impossible to build predictive models for condition-level data, similar to PCMs, on top of LLM-supplied information, but that seemingly subtle details in the derivation of likelihoods matter.
-The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not.
+The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not. \mf{double-check formulation of results}
 
 The paper is structured as follows.
-Section~\ref{motivation} motivates the investigation in this paper, by enlarging on the importance of deriving distributional predictions for conditional-level forced-choice data, and further elaborating on the conceptual differences between LLMs and PCMs and the practical differences between prevalent evaluation methods for the two.
+Section~\ref{motivation} motivates the investigation in this paper, by enlarging on the importance of deriving distributional predictions for conditional-level forced-choice data, and further elaborating on the relevant differences between LLMs and PCMs.
 Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a commonly used, theoretically-motivate PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.

### 2023-11-05 03-paper/paper-LLM-probmod.tex
revised §2

@@
 
-Psychological research into the workings of the human mind aims to find generalizable, systematic patterns in information processing in different domains of cognition.
-Experimental work therefore often compares human performance in different \emph{experimental conditions}, which reflect the general factors that are hypothesized to influence behavior, but might instantiate each experimental condition with different \emph{experimental items}, which are not under scrutiny for any systematic, predictable effect on the observed measurements.
+Psychological research into the workings of the human mind aims to find generalizable patterns in the way information is processed within or across different domains of cognition.
+Experimental work therefore often compares human performance in different \emph{experimental conditions}, which reflect the general factors that are hypothesized to influence behavior.
+Yet a single experimental condition is often instantiated with different \emph{experimental items}, which are not under scrutiny for any systematic, predictable effect on the observed measurements.
 For example, classical research on human memory \citep{AtkinsonShiffrin1968:Human-memory:-A} investigated the effect of rehearsal on memory consolidation.
-Relevant experiments compare recall with and without rehearsal (experimental conditions), while using different items (words, numbers, etc., to be memorized) in each instantiation of a memory task.
-Likewise, when studying how hearing a color word can facilitate a same-or-different judgement of color swatches \citep{Rosch1975:The-Nature-of-M}, the main experimental manipulation, reflected in different experimental conditions, concerns the typicality of shown color swatches, while the variability between different color words like ``blue'' or ``green'', is less important to this research question and so treated as \emph{item-level variation}.
+Relevant experiments compare recall with and without rehearsal (experimental conditions), while using different items (words, numbers, etc., to be memorized) in each instantiation of the same memory task.
+Likewise, when studying how hearing a color word can facilitate a same-or-different judgement of color swatches \citep{Rosch1975:The-Nature-of-M}, the main experimental manipulation concerns the typicality of shown color swatches, while the variability between different color words like ``blue'' or ``green'', is less important to this research question and so treated as \emph{item-level variation}.
 Consequently, a typical psychological experiment is mainly interested in assessing behavior at the level of the experimental condition, because that is where the distinctions relevant to the research question reside.
-Nevertheless, each experimental condition $C$ can be, and often is, instantiated with different instantiations of $C$, variation among which is deemed less relevant to the research question at hand.
+Nevertheless, each experimental condition can be, and often is, instantiated with different items, variation among which is deemed less relevant to the research question at hand.
 
@@
 Similarly, when analyzing or explaining data from psychological experiments with a probabilistic cognitive model (PCM), the PCM will naturally be set-up to predict data at the level of the relevant experimental manipulation, i.e., at the condition-level.
-Even though item-level (and individual-level) variation can, and sometimes should, be incorporated into probabilistic cognitive modeling \citep[e.g.,][]{NilsonRieskamp2011:Hierarchical-Ba,Lee2011:How-Cognitive-M,ScheibehenneRieskamp2013:Testing-the-Ada,FrankeDablander2016:What-does-the-c}, the focus of cognitive modeling is likewise to predict or explain systematic variation between experimental conditions.
+Even though item-level (and individual-level) variation can, and sometimes should, be incorporated into probabilistic cognitive modeling \citep[e.g.,][]{NilsonRieskamp2011:Hierarchical-Ba,Lee2011:How-Cognitive-M,ScheibehenneRieskamp2013:Testing-the-Ada}, the focus of cognitive modeling is to explain systematic variation between experimental conditions.
 Therefore, a basic PCM is essentially characterized by a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each logically possible data observation $D_{C}$ for any experimental condition $C$ and numerical parameters $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
@@
 
-PCMs, thus construed, differ strongly from LLMs.
+PCMs, thus construed, differ markedly from LLMs.
 LLMs make predictions about the likelihood of words given a preceding or surrounding text.
 They do not, without further derivation, provide predictions at a conceptually meaningful level of abstraction, similar to experimental conditions.
-In this way, LLMs are not explanatory models in the same sense as PCMs might be: they simply do not operate at the level of generalizable abstraction necessary for a common understanding of ``explanation.'' \mf{refs from PhilOScience?}
+In this way, LLMs are not explanatory models in the same sense as PCMs might be: they simply do not operate at a level of generalizable abstraction which is arguably necessary for increasing human understanding of the phenomenon in question \citep{Dellsen2020:Beyond-Explanat,Grimm2021:Understanding}.
 Nevertheless, we can derive predictions from LLM's word-level predictions, first at the item-level and also at the condition-level.
-Indeed, this is what standard benchmark-testing does, albeit with a focus on a winner-take-all strategy, and not with an eye towards deriving a full likelihood function for empirical data.
+Whether or not these predictions are deemed explanatory, the empirical question arises of whether any of these predictions are empirically adequate.
+This empirical question can be addressed by subjecting ``LLM-based PCMs'' to the same statistical model checking as the more usual theoretically motivated PCMs.
 
-The prevalent approach to assessing LLM performance is based on benchmark-testing, using large data sets for tasks in which a designated goal answer exists.
-Tasks often contain multiple-choice options, and accuracy for a task is determined with a ``winner-takes-all'' (WTA) strategy \citep[e.g.,][]{srivastava2023-BIGbench}.
+Standard benchmark-testing on multiple-choice tasks usually uses a winner-take-all (WTA) strategy \citep[e.g.,][]{srivastava2023-BIGbench} to define accuracy scores.
+The WTA-strategy can also be used to derive probabilistic item-level and condition-level predictions from LLMs.
 Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
 % For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
-Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
-Among the choice options is a designated \emph{target option} $y_{ki^{*}}$ that is assumed to be the ``true'' or to-be-selected goal answer.
+Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$, and each of which .
+For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same answer category $R_{i}$, e.g., that $y_{k1}$ is always the designated \emph{target option} that is assumed to be the ``true'' or to-be-selected goal answer, and that the other choice options are equally assignable into meaningful categories of answer types.
+
 The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
+  \mf{How to extend to masked language models?}
   More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
@@
 \begin{align*}
-  S\left( y_{ki}, x_{k} \right) = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right) =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
+  % S\left( y_{ki}, x_{k} \right) =
+  S_{ki}
+  % = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right)
+  =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
 \end{align*}
 %
-The WTA approach considers item $I_{k}$ to be answered correctly if the target option $y_{ki^{*}}$ is the option that maximizes the item-level score.
-The WTA-based accuracy of the LLM on the given task is the proportion of items which are answered correctly.
-
-While generally a pragmatic and useful approach, assessing LLM performance with a WTA-based accuracy measure can misleading, because it ignores potentially relevant distributional information.
-At an abstract level, the problem is that if task performance is categorical (in the most extreme case: binary), somewhere along the path from a numerical item-level score to accuracy a discontinuity has to be introduced.
-Every such discontinuity bottleneck entails loss of information.
-As this information might be useful, discontinuity should ideally happen as late as possible; or even better: not at all, unless we can be sure that it is always irrelevant.
-But since we hardly can be sure that it is irrelevant for all circumstances, we should better assess performance based on a models' full distributional predictions.
-
-The WTA approach implicitly assumes that item-level choices are resolved by a greedy-like choice of the best alternative.
-% It is therefore a good measure of accuracy when the LLM's main use cases are of a greedy-like nature as well.
-To generalize to less extreme sampling scenarios, the WTA approach can be generalized to softmax sampling, where the probability of choosing option $y_{ki}$ for item $I_{k}$ is:
+The WTA-based approach selects answers that maximize the item-level score (with random tie-breaking).
+Concretely, if $B_{k} = \arg \max_{i} S_{ki}$ is the set of all options that maximize the item-level score for item $k$, then the \emph{item-level prediction} for item $I_{k}$ of the WTA is:
 %
 \begin{align*}
-P_{\text{choose}}(y_{ki}) \propto \expo \left [ \alpha \ S \left( y_{ki}, x_{k} \right) \right ]\,,
+  P_{\text{item}}^{\text{WTA}} =
+  \begin{cases}
+    \frac{1}{\card{B_{k}}} & \text{if } y_{ki} \in B_{k} \text{, and} \\
+    0                     & \text{otherwise.}
+  \end{cases}
 \end{align*}
 %
-and the accuracy is the average probability of choosing the designated target option $y_{ki^{*}}$.
-For $\alpha \rightarrow \infty$, this softmax strategy converges to the WTA strategy.
+Notice that the item-level WTA-prediction is usually a degenerate probability distribution placing all probability mass on a single choice option.
 
-WTA-based accuracy measures can differ form accuracy measures obtained from the more general softmax-based strategy.
-Whether there is a difference, depends on the variation in item-level scores, in particular the relation between score-ordering and score-differences.
+The \emph{condition-level prediction} of the WTA approach is obtained by averaging over the item-level predictions for each answer category:
+%
+\begin{align*}
+  P_{\text{cond}}^{\text{WTA}}\left(R_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{WTA}} \left(y_{ki} \right)\,.
+\end{align*}
+%
+The \emph{WTA-based accuracy} is the probability $P_{\text{cond}}^{\text{WTA}}\left(R_{1} \right)$ of selecting the target option over all instances of the task.
+This is the most prevalent measure of the quality of LLM predictions.
+
+% While generally a pragmatic and useful approach, assessing LLM performance with a WTA-based accuracy measure can misleading, because it ignores potentially relevant distributional information.
+% At an abstract level, the problem is that if task performance is categorical (in the most extreme case: binary), somewhere along the path from a numerical item-level score to accuracy a discontinuity has to be introduced.
+% Every such discontinuity bottleneck entails loss of information.
+% As this information might be useful, discontinuity should ideally happen as late as possible; or even better: not at all, unless we can be sure that it is always irrelevant.
+% But since we hardly can be sure that it is irrelevant for all circumstances, we should better assess performance based on a models' full distributional predictions.
+
+The WTA approach implicitly assumes that item-level choices are resolved by a greedy-like choice of the best alternatives.
+This procedure ignores information about relative differences between scores among choice options; information which may be conceptually and practically relevant.
+To generalize to less extreme sampling scenarios, the WTA approach can be generalized to softmax sampling, where the \emph{item-level prediction} of choosing option $y_{ki}$ is:
+%
+\begin{align*}
+P_{\text{item}}^{\text{SM}}\left ( y_{ki} \right ) \propto \expo \left [ \alpha \ S \left( y_{ki}, x_{k} \right) \right ]\,.
+\end{align*}
+%
+The \emph{condition-level prediction} can be defined analogously to the above by averaging over items:
+\begin{align*}
+  P_{\text{cond}}^{\text{SM}}\left(R_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{ki} \right)\,.
+\end{align*}
+As before, the \emph{softmax-based accuracy}, $P_{\text{cond}}^{\text{SM}}\left(R_{i} \right)$, is the average probability of choosing the designated target option $y_{k1}$.
+For $\alpha \rightarrow \infty$, the softmax strategy converges to the WTA strategy, but for finite $\alpha$ the approaches differ.
+This has interesting entailments for accuracy assessment in more engineering-oriented contexts, as well as for questions of exploring the ``human-likeness'' of LLM predictions.
+
+Looking at accuracy assessment, WTA-based accuracy measures can even differ qualitatively from accuracy measures obtained from the more general softmax-based strategy.
+Differences in accuracy measures depend on the variation in item-level scores, in particular the relation between score-ordering and score-differences.
+The example below shows that categorically different results are, in principle, possible when using WTA- or softmax based accuracy.
+The example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
+%
+\begin{quote}
+  % \ex \label{bsp:exmpl-diff-WTA-SM}
+  \textbf{Example:}
   Imagine that there are two options, and that the target option's score is a small $\epsilon$ higher in 80\% of the task's items, and otherwise lower.
@@
 This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
-This is a categorical shift from predominantly goal answer to predominantly non-goal answer.
-The argument holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
+\end{quote}
+%
+The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.
 
-The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.
-Instead, the alternative softmax-based strategy conservatively generalizes WTA-based performance assessment, but also yields full distributional predictions at the item-level, i.e., a non-degenerate likelihood prediction for \emph{all} choice options, whenever $\alpha$ is not infinite.
-When aspiring to predict experimental data with LLMs, the WTA-based strategy may almost sure be ruled out as a predictor of human-level choice behavior for a simple item: for many experimental set-ups, it is highly unlikely that, for each item, there is a single choice option that (almost) \emph{all} participants will choose.
-But whether the softmax-based item-level predictions, with variable $\alpha$, align with item-level variation in human choice data is an empirical question that has to be addressed by explicit modeling and comparison; an exercise that this paper starts to take on (see Section~\ref{sec:item-level-pred}).
-
-Whether item-level predictions by an LLM are empirically vindicated or not, they may serve as the backbone for deriving condition-level predictions, just as the usual accuracy scores are obtained by averaging over all items belonging to a task or experimental condition.
-However, if we generalize to softmax-based distributional predictions, we find that there are different ways of deriving condition-level predictions (see Section~\ref{llm-predictions-for-reference-games}).
-We should therefore address the also the empirical question of which, if any, of these constructions of condition-level predictions provides a good fit to human data.
+When aspiring to predict experimental data from human participants with LLMs, the WTA-based strategy makes very strong predictions at the item-level, where it usually predicts a degenerate probability distribution according to which every data observation for a particular item should fall into the exact same response category.
+This may seem like an \emph{a priori} unlikely prediction, but needs to be tested empirically, just as the item-level predictions by the softmax-based approach should (see Section~\ref{sec:item-level-pred}).
+Whether item-level predictions by an LLM are empirically vindicated or not, is logically independent of whether these item-level predictions serve well as the backbone for deriving condition-level predictions.
+We therefore also need to test, against empirical data, whether different ways of deriving condition-level predictions are adequate (see Section~\ref{llm-predictions-for-reference-games}).
 

### 2023-11-05 03-paper/paper-LLM-probmod.tex
revised §4

@@
 %
-The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated (e.g., the trigger object/nicefra):
+The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated, i.e., the trigger object:
 %
@@
 \end{align*}
-where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.\mf{FC objected to ``condition-specificity'' which is fair, but I leave it as it was; my reasoning: we use model checking to rule out models; in this setting the conservative strategy is to err on the side of making the models too powerful (even at the cost of conceptual plausibility)}
+where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
+\mf{FC objected to ``condition-specificity'' which is fair, but I leave it as it was; my reasoning: we use model checking to rule out models; in this setting the conservative strategy is to err on the side of making the models too powerful (even at the cost of conceptual plausibility)}
 The result is a four-parameter model, one pair of parameters per condition, which provides a likelihood function for the categorical choice data and so can be fitted to the data and compared against other probabilistic models providing a likelihood function for the same data.
@@
 Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
-Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{fig:posterior-stats}).
-\mf{include information on MCMC, samples, warm-up etc.; follow Kruschke recipe}
+Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{fig:posterior-stats}).\footnote{
+  Posterior samples where generated for four chains with 2000 samples each, after a warm-up of 1000 samples.
+The ``adapt-delta'' value was set to 0.99.
+Converge was checked with $\hat{R}$-statistics \citep{GelmanRubin1992:Inference-from-}.
+}
 
 To assess goodness-of-fit, we can use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
-\mf{maybe insert formal definition of post-pred distribution?}
-As a minimal bar, we would require a model trained on observed data $D_{\text{obs}}$ to not be surprised by $D_{\text{obs}}$.
-Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models to be introduced later).
-We see that the RSA model passes this ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-} for both conditions.
+Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models, which are to be introduced later).
+We see that the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-} for both conditions, in the sense that, when trained on the data, the predictions the model would generate include the observed choice rates for each answer option.
 To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
 These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
+A very small value indicates that the model might be inadequate for reproducing the data it was trained on.
+This is clearly not the case for the RSA model.
 % Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
@@
     Summary statistics of posterior samples.
-    Number indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
+    Numbers indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
     \mf{make wide format}

### 2023-11-05 03-paper/paper-LLM-probmod.tex
revisions §4 (add item-level preds)

@@
 \begin{align*}
-  P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
+  P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
 \end{align*}
@@
 \mf{FC objected to ``condition-specificity'' which is fair, but I leave it as it was; my reasoning: we use model checking to rule out models; in this setting the conservative strategy is to err on the side of making the models too powerful (even at the cost of conceptual plausibility)}
-The result is a four-parameter model, one pair of parameters per condition, which provides a likelihood function for the categorical choice data and so can be fitted to the data and compared against other probabilistic models providing a likelihood function for the same data.
+The data $D_{C}$ from condition $C$, see Figure~\ref{fig:refgame-counts}, consists of counts for each response category.
+The parameterized likelihood function entailed by the RSA model for $D_{C}$ is:
+%
+\begin{align*}
+ P_{r}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \text{Multinomial}(D_{C}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
+\end{align*}
+%
+The result is a four-parameter model, one pair of parameters per condition.
 
-Parameterized predictions like $P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c})$ can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
+Parameterized predictions like $P_{r}(D_{C} \mid C, \alpha_{C}, \epsilon_{C})$ can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
 Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
@@
 
-To assess goodness-of-fit, we can use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
+To assess goodness-of-fit, we use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
 Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models, which are to be introduced later).
-We see that the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-} for both conditions, in the sense that, when trained on the data, the predictions the model would generate include the observed choice rates for each answer option.
+We see that for both conditions the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-}, namely that the distribution of posterior predictions includes the observed choice rates for each answer option.
 To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
@@
 A very small value indicates that the model might be inadequate for reproducing the data it was trained on.
-This is clearly not the case for the RSA model.
-% Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
-% The results from Table~\ref{tab:Bppp-values} show that the RSA model passes this numerical test of model adequacy.
+This is clearly not the case for the RSA model with posterior predictive $p$-values close to 0.5.
+In sum, the condition-level predictions by the RSA model, a theoretically motivated PCM, are not discredited by the condition-level data.
+
+We can also use the condition-level predictions for predicting the item-level data.
+If the whole data set for condition $C$ is chunked into item-level data, $D_{C} = \tuple{D^{1}_{C}, \dots, D^{m}_{C} }$, the likelihood function for analysis at the item-level uses the same (condition-level) predictor to separately predict data from each item:
+%
+\begin{align*}
+ P_{r}^{\text{item}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \prod_{k = 1}^{m} \text{Multinomial}(D_{C}^{k}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
+\end{align*}
+Fitting for the partitioned data set, and estimates of posterior predictive $p$-values that do not discredit the model ($0.338$ for production, $0.252$ for interpretation).
+This suggests that the item-level variation in the human data is reasonably well captured by a single condition-level predictor from a theoretically motivated PCM.
 

### 2023-11-05 03-paper/paper-LLM-probmod.tex
1/2 revisions §5

@@
   % S\left( y_{ki}, x_{k} \right) =
-  S_{ki}
+  \text{S}_{ki}
   % = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right)
@@
 The WTA-based approach selects answers that maximize the item-level score (with random tie-breaking).
-Concretely, if $B_{k} = \arg \max_{i} S_{ki}$ is the set of all options that maximize the item-level score for item $k$, then the \emph{item-level prediction} for item $I_{k}$ of the WTA is:
+Concretely, if $B_{k} = \arg \max_{i} \text{S}_{ki}$ is the set of all options that maximize the item-level score for item $k$, then the \emph{item-level prediction} for item $I_{k}$ of the WTA is:
 %
@@
 \begin{align*}
-P_{\text{item}}^{\text{SM}}\left ( y_{ki} \right ) \propto \expo \left [ \alpha \ S \left( y_{ki}, x_{k} \right) \right ]\,.
+P_{\text{item}}^{\text{SM}}\left ( y_{ki} \right ) \propto \expo \left (\alpha \ \text{S}_{ki} \right)\,.
 \end{align*}
@@
 \begin{align*}
- P_{r}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \text{Multinomial}(D_{C}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
+ P^{\text{RSA}}_{\text{cond}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \text{Multinomial}(D_{C}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
 \end{align*}
@@
 \begin{align*}
- P_{r}^{\text{item}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \prod_{k = 1}^{m} \text{Multinomial}(D_{C}^{k}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
+ P^{\text{RSA}}_{\text{item}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \prod_{k = 1}^{m} \text{Multinomial}(D_{C}^{k}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
 \end{align*}
+%
 Fitting for the partitioned data set, and estimates of posterior predictive $p$-values that do not discredit the model ($0.338$ for production, $0.252$ for interpretation).
-This suggests that the item-level variation in the human data is reasonably well captured by a single condition-level predictor from a theoretically motivated PCM.
+This suggests that the item-level variation in the human data is not so pronounced as to provide strong evidence against the condition-level predictor from a theoretically motivated PCM.
 
@@
 
-Figure~\ref{fig:refgame-counts} shows the condition-level counts for human choices for all of the 100 different items combined.
+The RSA model's predictions are defined at the condition-level.
 An LLM, on the other hand, first and foremost makes predictions about each item.
 As mentioned previously, the items of the reference game experiment differ in the task-relevant features (color, shape, texture), as well as the order of presentation of states and words.
-We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelevant material \citep{DegenHawkins2020:When-redundancy} \mf{more refs on (color) over-specification}; and machines may be susceptible to the presentation of the order of choice options.
+We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelevant material \citep[e.g.,][]{DegenHawkins2020:When-redundancy} \mf{more refs on (color) over-specification}; and machines may be susceptible to the presentation of the order of choice options.
 \mf{@PT references from the NLP literature on effects of prompt variation?}
-The main question to be addressed in this section is whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.
+It therefore becomes an empirical question of whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.
 
-For a set-up in which the task descriptions and choice options are all text-based, predictions of LLMs at the item-level can be derived as a function of the (log) probabilities assigned to the continuations corresponding to each choice option after processing the task description.
-Let $C$ be the experimental condition of interest (here: the production or interpretation condition).
-Let \(\set{I_1, \dots I_m}\) be the multi-set of items that occurred in the human experiment for condition $C$.\footnote{
-  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.
-}
-Each item $I_{k}$ consists of a task description $T_{k}$ and $l$ choice options $O_{k1}, \dots, O_{kl}$, which are treated as textual continuations of the task description (see Appendix~\ref{sec:examples-items-llm} for an example).
-For each item $I_{k}$, let $F(R_{l}, I_{k})$ be the choice option $O_{ki}$ that corresponds to \emph{response type} $R_{l}$ (here: target, competitor, distractor).\footnote{In the current set-up the response type ``distractor'' has two instantiations in the production condition.
-Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
-The \emph{item-level score} $S(O_{{ki}}, I_{k})$ for option $O_{ki} = w_{ki1}, \dots w_{kin}$ of item $I_{k}$ is defined as the log-probability of continuation $O_{ki}$ after prompt $T_{k}$:\footnote{Numerically, we worked with \emph{average} log-probabilities.
+Since the WTA-based item-level predictions are a limiting case of the more general softmax-approach, we consider the latter, repeated from above with slightly updated notation:
+%
+\begin{align*}
+P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C, \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
+\end{align*}
+%
+Item-level scores $\text{S}_{ki}$, as defined in Section~\ref{motivation}, are obtained from GPT-3.5 \mf{which model exactly, when assessed, cite paper}.
+% \footnote{
+  % Numerically, we worked with \emph{average} log-probabilities.
 As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition.
@@
 In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
+The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
+The choice options are categorized as in the human experiment.\footnote{
+  In the current set-up the response type ``distractor'' has two instantiations in the production condition.
+Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
+Allowing for random errors with probability $\epsilon_{C}$, as before,:
 %
 \begin{align*}
-S(O_{ki}, I_{k}) =  \sum_{j=1}^n \log P_{\text{LLM}} \left(w_{kij} \mid T_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
+  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C, \alpha_{c}, \epsilon_{c} \right )
+  = \epsilon_{c} \  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C, \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \
+  (1 = \epsilon_{c}) \  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C, \alpha_{c}, \epsilon_{c} \right )
+  \,,
 \end{align*}
 %
-\mf{check if formula is redundant; already introduced?}
-The item-level score of response type $R_{l}$ is $S(R_{l}, I_{k}) = S(F(R_{l}, I_{k}))$.
-
-Using item-level scores from GPT-3.5, we fit a model to the human data at the item-level \mf{which model exactly, when assessed, cite paper}.
-The data to be explained is the set of all triples of response type counts $\tuple{t_{k}, c_{k}, d_{k}}$ (target, competitor, distractor), separately for each item $I_{k}$.
-The model's predictions for the data from item $I_{k}$ from condition $C$ (production or interpretation) are:
+the resulting likelihood function for item-level data with item-level predictions is:
 %
 \begin{align*}
-  P_{\text{item}}(R_{l}, C, I_{k}, \alpha_{c}, \epsilon_{c}) \propto \frac{\expo \left( \alpha_{c} \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha_{c} \ S(R_{l'}, I_{k}) \right )} + \epsilon_{c}\,.
+ P^{\text{SM}}_{\text{item}}(D_{c} \mid C, \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial}(D_{c}^{k}, \tuple{P^{\text{SM}}_{\text{item}}(y_{ki} \mid C,  \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
 \end{align*}
 %
-In words, we use a single pair of parameters $\alpha_{c}$ and $\epsilon_{c}$, one for each condition, take the LLM's item-level scores for each item, and derive the response probabilities in the exact same way as before.
-% Notice that item-level scores are the same for all three types of models introduced and compared in the previous section, so that this by-item analysis applies to the raw LLM prediction irrespective of down-stream aggregation.
 Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.
@@
 
+Posterior credible values for parameters are.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rlrrrrrr}
+  \toprule
+ && $\alpha$ &&& \(\epsilon\) & \\
+ & condition & $|$95 & mean & 95\%$|$ & $|$95\% & mean\_epsilon & 95\%$|$\_epsilon \\ \midrule
+1 & production & 0.15 & 0.21 & 0.29 & 0.00 & 0.10 & 0.21 \\
+  2 & interpretation & 0.08 & 0.13 & 0.18 & 0.00 & 0.17 & 0.33 \\
+   \bottomrule
+\end{tabular}
+\end{table}
+
+\mf{Posteriors over $\alpha$ and $\epsilon$}
+
 Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
-Concretely, the plot shows, for each item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
+Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
 There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{itemize}
+  \item mention multi-set averages
+  \begin{itemize}
+    \item Let \(\set{I_1, \dots I_m}\) be the multi-set of items that occurred in the human experiment for condition $C$.\footnote{
+  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.
+}
+
+  \end{itemize}
+\end{itemize}
 

### 2023-11-06 03-paper/paper-LLM-probmod.tex
master table, rev §4

@@
 The RSA model defines pragmatic reasoning as a sequence of iterated (soft-)optmization of policies, grounding out in literal interpretation.
-If $\mathfrak{L}(s,u) \mapsto \set{0,1}$ is a semantic meaning function mapping each pair of state $s$ and utterance $u$ to a (binary) truth-value, and if $P_{s}$ is a prior over states, a literal listener policy is defined as:
+If $\mathfrak{L}(s,u) \mapsto \set{0,1}$ is a semantic meaning function mapping each pair of state $s$ and utterance $u$ to a (binary) truth-value, and if $P_{\text{prior}}(s)$ is a prior over states, a literal listener policy is defined as:
 %
 \begin{align*}
- P_{L_{0}}(s \mid u) \propto \mathfrak{L}(s,u) P(s)\,.
+ P_{L_{0}}(s \mid u) \propto \mathfrak{L}(s,u)\  P_{\text{prior}}(s)\,.
 \end{align*}
 %
-The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated, i.e., the trigger object:
+The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated, i.e., to maximize the log-probability of the trigger object:
 %
-\begin{align*}
+\begin{align}
+  \label{eq:RSA-speaker}
   P_{S}(u_{i} \mid s, \alpha) & \propto \expo \left [ \log P_{L_{0}}(s \mid u_{i}) \right ] \,.
-\end{align*}
+\end{align}
 %
@@
 %
-\begin{align*}
-  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P(s) \,.
-\end{align*}
+\begin{align}
+  \label{eq:RSA-interpreter}
+  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P_{\text{prior}}(s) \,.
+\end{align}
 
@@
 
-Figure~\ref{fig:RSA-example} gives example calculations (assuming a flat prior $P(s)$ and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
+Figure~\ref{fig:RSA-example} gives example calculations (assuming a flat prior and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
 For $\alpha=1$, the model predicts that the probabilities of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{\nicefrac{3}{5}, \nicefrac{2}{5}, 0}$ for the interpretation condition.
 Increasing $\alpha$ will increase the odds of target over competitor choices.
-Yet, the model also predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
-It is therefore common to include a small error probability $\epsilon$, with which a choice would be made at random \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}: if $P_{r}(R_{l}, C, \alpha_{c})$ is the RSA model's probabilistic prediction for response category $R_{l}$ (target, competitor, or distractor) for condition $C$ (production or interpretation) and condition-specific optimality $\alpha_{c}$, the random-error smoothed prediction is:
+In sum, the condition-level predictions of the RSA model are a parameterized function $P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c})$, assigning a probability to each response category $R_{l}$ in each condition $C$ (production or interpretation).
+
+The model constructed so far predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
+It is therefore common to include a small error probability $\epsilon$, with which a choice would be made at random \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}, so that we get:
+%
 \begin{align*}
@@
 \end{align*}
+%
 where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
 \mf{FC objected to ``condition-specificity'' which is fair, but I leave it as it was; my reasoning: we use model checking to rule out models; in this setting the conservative strategy is to err on the side of making the models too powerful (even at the cost of conceptual plausibility)}
+
 The data $D_{C}$ from condition $C$, see Figure~\ref{fig:refgame-counts}, consists of counts for each response category.
-The parameterized likelihood function entailed by the RSA model for $D_{C}$ is:
+The parameterized likelihood function entailed by the RSA model for condition-level data $D_{C}$ is:
 %
-\begin{align*}
+\begin{align}
+  \label{eq:RSA-likelihood}
  P^{\text{RSA}}_{\text{cond}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \text{Multinomial}(D_{C}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
-\end{align*}
+\end{align}
 %
@@
 
-Parameterized predictions like $P_{r}(D_{C} \mid C, \alpha_{C}, \epsilon_{C})$ can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
+Parameterized predictions, like in Equation~(\ref{eq:RSA-likelihood}), can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
 Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
-Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{fig:posterior-stats}).\footnote{
+Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{tab:sumStats}).\footnote{
   Posterior samples where generated for four chains with 2000 samples each, after a warm-up of 1000 samples.
@@
 We see that for both conditions the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-}, namely that the distribution of posterior predictions includes the observed choice rates for each answer option.
-To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
+To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
 These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
 A very small value indicates that the model might be inadequate for reproducing the data it was trained on.
-This is clearly not the case for the RSA model with posterior predictive $p$-values close to 0.5.
+This is clearly not the case for the RSA model with posterior predictive $p$-values close to 0.5, as shown in Table~\ref{tab:sumStats}.
 In sum, the condition-level predictions by the RSA model, a theoretically motivated PCM, are not discredited by the condition-level data.
@@
 %
-Fitting for the partitioned data set, and estimates of posterior predictive $p$-values that do not discredit the model ($0.338$ for production, $0.252$ for interpretation).
+Fitting for the partitioned data set, and estimates of posterior predictive $p$-values that do not discredit the model (see Table~\ref{tab:sumStats}).
 This suggests that the item-level variation in the human data is not so pronounced as to provide strong evidence against the condition-level predictor from a theoretically motivated PCM.
@@
 Section~\ref{sec:item-level-pred} first looks at the data from each item individually, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.
+
+\begin{table}[t]
+\centering
+\begin{tabular}{llllcrlcrcl}
+  \toprule
+  &&&& $\alpha$ &&& $\epsilon$ & \\ \cmidrule(r){4-6} \cmidrule(l){7-9}
+  data & condition & model & $|$95\%\ & mean\ & 95\%$|$\ & $|$95\%\ & mean\ & 95\%$|$\ & Bpppv & \\
+  \midrule
+
+  cond.-level & avg. scores        & production     & 0.19 & 0.00 & 0.22 & 0.04 & 0.26 & 0.11 & 0.05 & * \\
+  cond.-level & avg. scores        & interpretation & 0.15 & 0.00 & 0.18 & 0.02 & 0.20 & 0.06 & 0.00 & * \\
+  cond.-level & avg. probabilities & production     & 0.86 & 0.08 & 1.04 & 0.13 & 1.22 & 0.18 & 0.51 &  \\
+  cond.-level & avg. probabilities & interpretation & 0.36 & 0.00 & 0.48 & 0.05 & 0.60 & 0.12 & 0.00 & * \\
+  cond.-level & avg. WTA           & production     & 0.87 & 0.08 & 1.04 & 0.13 & 1.23 & 0.17 & 0.49 &  \\
+  cond.-level & avg. WTA           & interpretation & 0.17 & 0.10 & 0.39 & 0.15 & 0.61 & 0.20 & 0.26 &  \\
+  cond.-level & RSA                & production     & 2.59 & 0.08 & 3.16 & 0.13 & 3.71 & 0.18 & 0.49 &  \\
+  cond.-level & RSA                & interpretation & 0.24 & 0.10 & 0.66 & 0.15 & 1.07 & 0.20 & 0.52 &  \\
+  item-level  & LLM                & production     & 0.15 & 0.00 & 0.21 & 0.10 & 0.29 & 0.21 & 0.00 & * \\
+  item-level  & LLM                & interpretation & 0.08 & 0.00 & 0.13 & 0.17 & 0.18 & 0.33 & 0.00 & * \\
+  item-level  & RSA                & production     & 2.62 & 0.08 & 3.15 & 0.13 & 3.69 & 0.17 & 0.34 &  \\
+  item-level  & RSA                & interpretation & 0.24 & 0.10 & 0.66 & 0.15 & 1.05 & 0.20 & 0.25 & \\
+
+   \bottomrule \\
+\end{tabular}
+\caption{
+  Summary statistics for model fits.
+  Columns show the means and 95\% credible intervals for posterior samples for each parameter, as well as the Bayesian posterior predictive $p$-value for each model, condition and data level.
+  Asterisks mark $p$-values not above $0.05$, flagging relative inability to capture the structure of the data.
+}
+\label{tab:sumStats}
+\end{table}
 
@@
 Posterior credible values for parameters are.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{rlrrrrrr}
-  \toprule
- && $\alpha$ &&& \(\epsilon\) & \\
- & condition & $|$95 & mean & 95\%$|$ & $|$95\% & mean\_epsilon & 95\%$|$\_epsilon \\ \midrule
-1 & production & 0.15 & 0.21 & 0.29 & 0.00 & 0.10 & 0.21 \\
-  2 & interpretation & 0.08 & 0.13 & 0.18 & 0.00 & 0.17 & 0.33 \\
-   \bottomrule
-\end{tabular}
-\end{table}
-
-\mf{Posteriors over $\alpha$ and $\epsilon$}
 
@@
 
-\begin{table}[t]
-  \centering
+% \begin{table}[t]
+%   \centering
 
-  % \includegraphics[width=0.9\linewidth]{00-pics/posterior-stats.pdf}
+%   % \includegraphics[width=0.9\linewidth]{00-pics/posterior-stats.pdf}
 
-  \begin{tabular}{llllcr}
-    \toprule \addlinespace[1ex]
-    model        & condition      & Parameter  & $|$95\% & mean & 95\%$|$ \\
-    \midrule  \addlinespace[1ex]
-    narrow       & production     & $\alpha$   & 0.19    & 0.22 & 0.26 \\
-                 &                & $\epsilon$ & 0.00    & 0.04 & 0.11 \\
-                 & interpretation & $\alpha$   & 0.15    & 0.18 & 0.20 \\
-                 &                & $\epsilon$ & 0.00    & 0.02 & 0.06 \\ \addlinespace[0.75ex]
-    intermediate & production     & $\alpha$   & 0.86    & 1.04 & 1.22 \\
-                 &                & $\epsilon$ & 0.08    & 0.13 & 0.18 \\
-                 & interpretation & $\alpha$   & 0.36    & 0.48 & 0.60 \\
-                 &                & $\epsilon$ & 0.00    & 0.05 & 0.12 \\ \addlinespace[0.75ex]
-    wide         & production     & $\alpha$   & 0.22    & 2.69 & 8.79 \\
-                 &                & $\epsilon$ & 0.08    & 0.13 & 0.19 \\
-                 & interpretation & $\alpha$   & 0.19    & 1.08 & 6.03 \\
-                 &                & $\epsilon$ & 0.00    & 0.06 & 0.18 \\ \addlinespace[0.75ex]
-    RSA          & production     & $\alpha$   & 2.59    & 3.16 & 3.71 \\
-                 &                & $\epsilon$ & 0.08    & 0.13 & 0.18 \\
-                 & interpretation & $\alpha$   & 0.24    & 0.66 & 1.07 \\
-                 &                & $\epsilon$ & 0.10    & 0.15 & 0.20 \\ \addlinespace[0.25ex]
-    \bottomrule \\
-  \end{tabular}
+%   \begin{tabular}{llllcr}
+%     \toprule \addlinespace[1ex]
+%     model        & condition      & Parameter  & $|$95\% & mean & 95\%$|$ \\
+%     \midrule  \addlinespace[1ex]
+%     narrow       & production     & $\alpha$   & 0.19    & 0.22 & 0.26 \\
+%                  &                & $\epsilon$ & 0.00    & 0.04 & 0.11 \\
+%                  & interpretation & $\alpha$   & 0.15    & 0.18 & 0.20 \\
+%                  &                & $\epsilon$ & 0.00    & 0.02 & 0.06 \\ \addlinespace[0.75ex]
+%     intermediate & production     & $\alpha$   & 0.86    & 1.04 & 1.22 \\
+%                  &                & $\epsilon$ & 0.08    & 0.13 & 0.18 \\
+%                  & interpretation & $\alpha$   & 0.36    & 0.48 & 0.60 \\
+%                  &                & $\epsilon$ & 0.00    & 0.05 & 0.12 \\ \addlinespace[0.75ex]
+%     wide         & production     & $\alpha$   & 0.22    & 2.69 & 8.79 \\
+%                  &                & $\epsilon$ & 0.08    & 0.13 & 0.19 \\
+%                  & interpretation & $\alpha$   & 0.19    & 1.08 & 6.03 \\
+%                  &                & $\epsilon$ & 0.00    & 0.06 & 0.18 \\ \addlinespace[0.75ex]
+%     RSA          & production     & $\alpha$   & 2.59    & 3.16 & 3.71 \\
+%                  &                & $\epsilon$ & 0.08    & 0.13 & 0.18 \\
+%                  & interpretation & $\alpha$   & 0.24    & 0.66 & 1.07 \\
+%                  &                & $\epsilon$ & 0.10    & 0.15 & 0.20 \\ \addlinespace[0.25ex]
+%     \bottomrule \\
+%   \end{tabular}
 
-  \caption{
-    Summary statistics of posterior samples.
-    Numbers indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
-    \mf{make wide format}
-  }
-  \label{fig:posterior-stats}
-\end{table}
+%   \caption{
+%     Summary statistics of posterior samples.
+%     Numbers indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
+%     \mf{make wide format}
+%   }
+%   \label{fig:posterior-stats}
+% \end{table}
 
@@
 Using priors and Bayesian posterior estimation as previously described for the RSA (see Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
-The usual Bayesian summary statistics for posteriors over model parameters are shown in Table~\ref{fig:posterior-stats}.
+The usual Bayesian summary statistics for posteriors over model parameters are shown in Table~\ref{tab:sumStats}.
 What is noteworthy about these estimates is that the $\alpha_{c}$ parameters for the wide-scope model have a very large credible range.
@@
 We find that only the theoretical model (RSA) and the intermediate-scope model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
-To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
-Consequently, the results from Table~\ref{tab:Bppp-values} suggest that the narrow-scope model fails on the interpretation data, and is at most borderline compatible with the production data; that the wide-scope model is able to reproduce the production data, but fails on the interpretation data; and that only the intermediate-scope model is able to fully recover the data from both conditions.
+To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
+Consequently, the results from Table~\ref{tab:sumStats} suggest that the narrow-scope model fails on the interpretation data, and is at most borderline compatible with the production data; that the wide-scope model is able to reproduce the production data, but fails on the interpretation data; and that only the intermediate-scope model is able to fully recover the data from both conditions.
 
-\begin{table}[ht]
-\centering
+% \begin{table}[ht]
+% \centering
 
-% \begin{tabular}{lrr}
+% % \begin{tabular}{lrr}
+% %   \toprule
+% %   model        & production & interpretation \\ \midrule
+% %   narrow       & 0.05       & 0.00 \\
+% %   wide         & 0.51       & 0.00 \\
+% %   intermediate & 0.49       & 0.26 \\
+% %   RSA          & 0.49       & 0.52 \\
+% %   \bottomrule \\
+% % \end{tabular}
+
+% \begin{tabular}{lrrrr}
 %   \toprule
-%   model        & production & interpretation \\ \midrule
-%   narrow       & 0.05       & 0.00 \\
-%   wide         & 0.51       & 0.00 \\
-%   intermediate & 0.49       & 0.26 \\
-%   RSA          & 0.49       & 0.52 \\
-%   \bottomrule \\
+%   & narrow & wide & intermediate & RSA \\ \midrule
+%   production & 0.05 & 0.51 & 0.49 & 0.49 \\
+%   interpretation & 0.00 & 0.00 & 0.26 & 0.52 \\ \bottomrule \\
 % \end{tabular}
 
-\begin{tabular}{lrrrr}
-  \toprule
-  & narrow & wide & intermediate & RSA \\ \midrule
-  production & 0.05 & 0.51 & 0.49 & 0.49 \\
-  interpretation & 0.00 & 0.00 & 0.26 & 0.52 \\ \bottomrule \\
-\end{tabular}
-
-\caption{Sample-based estimates of Bayesian posterior predictive $p$-values for each model and each condition, based on likelihood of the observed data as test statistic.}
-\label{tab:Bppp-values}
-\end{table}
+% \caption{Sample-based estimates of Bayesian posterior predictive $p$-values for each model and each condition, based on likelihood of the observed data as test statistic.}
+% \label{tab:Bppp-values}
+% \end{table}
 

### 2023-11-07 03-paper/paper-LLM-probmod.tex
minor revs §4

@@
   \label{eq:RSA-speaker}
-  P_{S}(u_{i} \mid s, \alpha) & \propto \expo \left [ \log P_{L_{0}}(s \mid u_{i}) \right ] \,.
+  P_{S}(u \mid s, \alpha) & \propto \expo \left [ \log P_{L_{0}}(s \mid u) \right ] \,.
 \end{align}
@@
 Increasing $\alpha$ will increase the odds of target over competitor choices.
-In sum, the condition-level predictions of the RSA model are a parameterized function $P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c})$, assigning a probability to each response category $R_{l}$ in each condition $C$ (production or interpretation).
+In sum, the condition-level predictions of the RSA model are a parameterized function $P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c})$, assigning a probability to each response category $R_{l}$ (target, competitor, or distractor) in each condition $C$ (production or interpretation).
 
@@
 \begin{align*}
-  P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
+  P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon_{c}) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
 \end{align*}
@@
 To assess goodness-of-fit, we use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
-Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models, which are to be introduced later).
-We see that for both conditions the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-}, namely that the distribution of posterior predictions includes the observed choice rates for each answer option.
-To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
-These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
-A very small value indicates that the model might be inadequate for reproducing the data it was trained on.
-This is clearly not the case for the RSA model with posterior predictive $p$-values close to 0.5, as shown in Table~\ref{tab:sumStats}.
+Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models for condition-level data, which are to be introduced later).
+We see that for both conditions the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-}, which require that the distribution of posterior predictions includes the observed choice rates for each answer option.
+To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values (Bppp values), using likelihood of the observed data as a test statistics.
+These Bppp values approximate the probability that a model conditioned on observed data $D_{\text{obs}}$ predicts future data $D_{rep}$, of the same size and format as $D_{obs}$, that is at least as likely as the data $D_{\text{obs}}$ itself is (given the posterior predictive distribution).
+Very small Bppp values indicate that the model might be inadequate for reproducing the data it was trained on, so to speak.
+This is clearly not the case for the RSA model with Bppp values close to 0.5, as shown in Table~\ref{tab:sumStats}.
 In sum, the condition-level predictions by the RSA model, a theoretically motivated PCM, are not discredited by the condition-level data.
+Reversely, the RSA model seems to adequately capture the condition-level data.
 
 We can also use the condition-level predictions for predicting the item-level data.
-If the whole data set for condition $C$ is chunked into item-level data, $D_{C} = \tuple{D^{1}_{C}, \dots, D^{m}_{C} }$, the likelihood function for analysis at the item-level uses the same (condition-level) predictor to separately predict data from each item:
+If the whole data set for condition $C$ is chunked into item-level data, $D_{C} = \tuple{D^{1}_{C}, \dots, D^{m}_{C} }$, the likelihood function for analysis at the item-level uses the same (condition-level) predictor to separately predict data from each item, so that the likelihood function becomes:
 %
@@
 %
-Fitting for the partitioned data set, and estimates of posterior predictive $p$-values that do not discredit the model (see Table~\ref{tab:sumStats}).
+Fitting the RSA model to the partitioned data set, we find estimates of Bppp values that do not discredit the model (see Table~\ref{tab:sumStats}).
 This suggests that the item-level variation in the human data is not so pronounced as to provide strong evidence against the condition-level predictor from a theoretically motivated PCM.
+In other words, the condition-level RSA predictions seem to adequately capture also the item-level data.
 
 The following sections apply these tests of Bayesian model criticism also to models built around predictor values from LLMs.
-Section~\ref{sec:item-level-pred} first looks at the data from each item individually, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.
+Section~\ref{sec:item-level-pred} first looks at the item-level data, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.
 
@@
   Summary statistics for model fits.
-  Columns show the means and 95\% credible intervals for posterior samples for each parameter, as well as the Bayesian posterior predictive $p$-value for each model, condition and data level.
-  Asterisks mark $p$-values not above $0.05$, flagging relative inability to capture the structure of the data.
+  Columns show the means and 95\% credible intervals for posterior samples for each parameter, as well as the Bayesian posterior predictive $p$-values for each model, condition and data level.
+  Asterisks mark Bppp-values no higher than $0.05$, flagging relative inability to capture the structure of the data.
 }

### 2023-11-11 03-paper/paper-LLM-probmod.tex
added analysis of alpha-values for item-level fits

@@
 \begin{align*}
- P^{\text{RSA}}_{\text{item}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \prod_{k = 1}^{m} \text{Multinomial}(D_{C}^{k}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
+ P^{\text{RSA}}_{\text{item}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \prod_{k = 1}^{m} \text{Multinomial} \left (D_{C}^{k}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3} \right )\,.
 \end{align*}
@@
   &&&& $\alpha$ &&& $\epsilon$ & \\ \cmidrule(r){4-6} \cmidrule(l){7-9}
-  data & condition & model & $|$95\%\ & mean\ & 95\%$|$\ & $|$95\%\ & mean\ & 95\%$|$\ & Bpppv & \\
+  data & model & condition & $|$95\%\ & mean\ & 95\%$|$\ & $|$95\%\ & mean\ & 95\%$|$\ & Bpppv & \\
   \midrule
 
-  cond.-level & avg. scores        & production     & 0.19 & 0.00 & 0.22 & 0.04 & 0.26 & 0.11 & 0.05 & * \\
-  cond.-level & avg. scores        & interpretation & 0.15 & 0.00 & 0.18 & 0.02 & 0.20 & 0.06 & 0.00 & * \\
-  cond.-level & avg. probabilities & production     & 0.86 & 0.08 & 1.04 & 0.13 & 1.22 & 0.18 & 0.51 &  \\
-  cond.-level & avg. probabilities & interpretation & 0.36 & 0.00 & 0.48 & 0.05 & 0.60 & 0.12 & 0.00 & * \\
-  cond.-level & avg. WTA           & production     & 0.87 & 0.08 & 1.04 & 0.13 & 1.23 & 0.17 & 0.49 &  \\
-  cond.-level & avg. WTA           & interpretation & 0.17 & 0.10 & 0.39 & 0.15 & 0.61 & 0.20 & 0.26 &  \\
-  cond.-level & RSA                & production     & 2.59 & 0.08 & 3.16 & 0.13 & 3.71 & 0.18 & 0.49 &  \\
-  cond.-level & RSA                & interpretation & 0.24 & 0.10 & 0.66 & 0.15 & 1.07 & 0.20 & 0.52 &  \\
-  item-level  & LLM                & production     & 0.15 & 0.00 & 0.21 & 0.10 & 0.29 & 0.21 & 0.00 & * \\
-  item-level  & LLM                & interpretation & 0.08 & 0.00 & 0.13 & 0.17 & 0.18 & 0.33 & 0.00 & * \\
-  item-level  & RSA                & production     & 2.62 & 0.08 & 3.15 & 0.13 & 3.69 & 0.17 & 0.34 &  \\
-  item-level  & RSA                & interpretation & 0.24 & 0.10 & 0.66 & 0.15 & 1.05 & 0.20 & 0.25 & \\
+  item-level  & RSA                & production     & 2.62 & 3.15 & 3.69 & 0.08 & 0.13 & 0.17 & 0.34 &  \\
+  item-level  & RSA                & interpretation & 0.24 & 0.66 & 1.05 & 0.10 & 0.15 & 0.20 & 0.25 &  \\
+  item-level  & LLM                & production     & 0.15 & 0.21 & 0.29 & 0.00 & 0.10 & 0.21 & 0.00 & * \\
+  item-level  & LLM                & interpretation & 0.08 & 0.13 & 0.18 & 0.00 & 0.17 & 0.33 & 0.00 & * \\
+  cond.-level & RSA                & production     & 2.59 & 3.16 & 3.71 & 0.08 & 0.13 & 0.18 & 0.49 &  \\
+  cond.-level & RSA                & interpretation & 0.24 & 0.66 & 1.07 & 0.10 & 0.15 & 0.20 & 0.52 &  \\
+  cond.-level & avg. scores        & production     & 0.19 & 0.22 & 0.26 & 0.00 & 0.04 & 0.11 & 0.05 & * \\
+  cond.-level & avg. scores        & interpretation & 0.15 & 0.18 & 0.20 & 0.00 & 0.02 & 0.06 & 0.00 & * \\
+  cond.-level & avg. probabilities & production     & 0.86 & 1.04 & 1.22 & 0.08 & 0.13 & 0.18 & 0.51 &  \\
+  cond.-level & avg. probabilities & interpretation & 0.36 & 0.48 & 0.60 & 0.00 & 0.05 & 0.12 & 0.00 & * \\
+  cond.-level & avg. WTA           & production     & 0.87 & 1.04 & 1.23 & 0.08 & 0.13 & 0.17 & 0.49 &  \\
+  cond.-level & avg. WTA           & interpretation & 0.17 & 0.39 & 0.61 & 0.10 & 0.15 & 0.20 & 0.26 &  \\
 
@@
 An LLM, on the other hand, first and foremost makes predictions about each item.
-As mentioned previously, the items of the reference game experiment differ in the task-relevant features (color, shape, texture), as well as the order of presentation of states and words.
-We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelevant material \citep[e.g.,][]{DegenHawkins2020:When-redundancy} \mf{more refs on (color) over-specification}; and machines may be susceptible to the presentation of the order of choice options.
+As mentioned previously, the items of the reference game experiment differ in which levels of features (color, shape, texture) instantiate the structure of the task shown in Figure \ref{fig:ref-game}, as well as the order of presentation of objects and words.
+We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options.
 \mf{@PT references from the NLP literature on effects of prompt variation?}
@@
 
-Since the WTA-based item-level predictions are a limiting case of the more general softmax-approach, we consider the latter, repeated from above with slightly updated notation:
+As described in Section~\ref{motivation}, the standard procedure for assessing LLM predictions in multi-choice tasks is to use a winner-takes-all (WTA) strategy that just picks the best option for each item.
+But since this strategy has conceptual and empirical weaknesses, we define the item-level LLM predictions using the more general softmax-approach, which contains the WTA-based approach as a limiting case, repeated here from above with slightly updated notation:
 %
 \begin{align*}
-P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C, \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
+P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
 \end{align*}
@@
 \begin{align*}
-  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C, \alpha_{c}, \epsilon_{c} \right )
-  = \epsilon_{c} \  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C, \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \
-  (1 = \epsilon_{c}) \  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C, \alpha_{c}, \epsilon_{c} \right )
-  \,,
+  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
+  = \epsilon_{c} \  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
 \end{align*}
 %
-the resulting likelihood function for item-level data with item-level predictions is:
+The resulting likelihood function for item-level data with item-level predictions is:
 %
 \begin{align*}
- P^{\text{SM}}_{\text{item}}(D_{c} \mid C, \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial}(D_{c}^{k}, \tuple{P^{\text{SM}}_{\text{item}}(y_{ki} \mid C,  \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
+ P^{\text{SM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{SM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
 \end{align*}
@@
 
-Posterior credible values for parameters are.
+Summary statistics for samples from the posterior distribution over parameters are shown in Table~\ref{tab:sumStats}.
+Credible values of $\alpha$ parameters are comparatively low for the LLM-based model, suggesting that the predicted scores have rather large differences, which have to be compensated for by the softmax link-function to adequately fit the data.
+Figure~\ref{fig:closeness-target-item-level} shows the mean predicted target probabilities for different parameter values of $\alpha$, highlighting that for posterior credible values, the predictions are clearly distinct from the predictions of a WTA strategy.
+This suggests that the WTA strategy, which is the special limiting case for $\alpha \rightarrow \infty$, provides a substantially worse fit for item-level choices than those provided by less extreme values of $\alpha$.
 
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\textwidth]{00-pics/closeness-target-by-alpha-item-level.pdf}
+  \caption{
+    Mean predicted target probability for different values of softmax parameter $\alpha$.
+    The solid black lines show the probability of the target options, predicted for a fixed value of $\alpha$ ($x$-axis), averaged over all items for each condition.
+    The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$.
+    For reference, the target probability under a random choice strategy and under the WTA strategy are shown with dashed lines.
+    For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
+  }
+  \label{fig:closeness-target-item-level}
+\end{figure}
+
+Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
 Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
@@
 The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
-
-Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (0.0001 for production, 0 for interpretation), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
-In contrast, using the global, item-independent RSA model predictions to fit the item-level data, we obtain estimates of posterior predictive $p$-values that do not discredit the model ($0.338$ for production, $0.252$ for interpretation).
 These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2023-11-11 02-data-analysis/03-data-analysis.R

@@ -49,11 +49,21 @@ ggsave(plot = refgame_counts,
 ## prepare model predictions (-> move to prep script?)
 #######################################################
 
-softmax_row <- function(matrix, alpha = 1) {
-  exp_matrix <- exp(alpha * matrix)
+softmax_row <- function(matrix_of_scores, alpha = 1) {
+  exp_matrix <- exp(alpha * matrix_of_scores)
   row_sums <- rowSums(exp_matrix)
   softmax_matrix <- exp_matrix / row_sums
   return(softmax_matrix)
+}
+
+wta_row <- function(matrix_of_scores) {
+  result_matrix <- matrix(0, nrow = nrow(matrix_of_scores), ncol = ncol(matrix_of_scores))
+  for (i in 1:nrow(matrix_of_scores)) {
+    max_val <- max(matrix_of_scores[i, ])
+    max_indices <- which(matrix_of_scores[i, ] == max_val)
+    result_matrix[i, max_indices] <- 1/length(max_indices)
+  }
+  return(result_matrix)
 }
 
 #################
@@ -61,7 +71,7 @@ softmax_row <- function(matrix, alpha = 1) {
 #################
 
 # wrangle data
-x <- d |> 
+x_prod <- d |> 
   filter(condition == "production") |> 
   select(submission_id, trial_nr, item, condition, starts_with("scores_produ")) |> 
   mutate(scores_production_distractor = 
@@ -69,41 +79,43 @@ x <- d |>
   select(-scores_production_distractor1, -scores_production_distractor2)
 
 # narrow- and intermediate-scope predictions for production
-matrix_global <- x |> 
+matrix_global_prod <- x_prod |> 
   select(starts_with("scores_produ")) |> as.matrix()
-LLM_pred_prod_narrow       <- softmax_row(matrix(apply(matrix_global, 2, mean), nrow = 1))
-LLM_pred_prod_intermediate <- matrix(apply(softmax_row(matrix_global), 2, mean), nrow = 1)
+LLM_pred_prod_narrow       <- softmax_row(matrix(apply(matrix_global_prod, 2, mean), nrow = 1))
+LLM_pred_prod_intermediate <- matrix(apply(softmax_row(matrix_global_prod), 2, mean), nrow = 1)
+LLM_pred_prod_WTA          <- matrix(apply(wta_row(matrix_global_prod), 2, mean), nrow = 1)
 
 # wide-scope predictions
[…]

### 2023-11-11 02-data-analysis/03-data-analysis.R
updated plotting script after new terminology

@@ -283,9 +283,9 @@ pp_prod_RSA <- get_posterior_predictives(fit_prod_RSA, d_prod_global, filename =
 pp_inter_narrow <- get_posterior_predictives(fit_inter_narrow, d_inter_global, filename = "post_pred_inter_narrow") |> 
   mutate(condition = "interpretation", model = "avg. scores")
 pp_inter_intermediate <- get_posterior_predictives(fit_inter_intermediate, d_inter_global, filename = "post_pred_inter_intermediate") |> 
-  mutate(condition = "interpretation", model = "avg. probabilities")
+  mutate(condition = "interpretation", model = "intermediate")
 pp_inter_wide <- get_posterior_predictives(fit_inter_wide, d_inter_global, filename = "post_pred_inter_wide") |> 
-  mutate(condition = "interpretation", model = "wide")
+  mutate(condition = "interpretation", model = "avg. probabilities")
 pp_inter_WTA <- get_posterior_predictives(fit_inter_WTA, d_inter_global, filename = "post_pred_inter_WTA") |> 
   mutate(condition = "interpretation", model = "avg. WTA")
 pp_inter_RSA <- get_posterior_predictives(fit_inter_RSA, d_inter_global, filename = "post_pred_inter_RSA") |> 
@@ -344,10 +344,13 @@ plot_PPC(pp_prod_RSA, pp_inter_RSA)
 
 # all models in one
 
+d_counts
+
 PPC_data |> 
   ggplot() +
-  geom_col(data = PPC_data |> filter(model == "narrow"),
-           aes(x = response, y = observed, fill = response)) +
+  # geom_col(data = PPC_data |> filter(model == "narrow"),
+  #          aes(x = response, y = observed, fill = response)) +
+  geom_col(data = d_counts, aes(x= response, y = n, fill = response)) +
   facet_grid(.~condition) +
   geom_pointrange(aes(x = response, y = mean, ymin = `|95%`, ymax = `95%|`, shape = model, group = model), 
                   position = position_dodge(width = 0.75), size = 0.6, linewidth = 0.8, color = project_colors[6]) +

### 2023-11-18 03-paper/paper-LLM-probmod.tex
revision formal notation §5 & 6

@@
 
-To assess goodness-of-fit, we use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
-Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models for condition-level data, which are to be introduced later).
-We see that for both conditions the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-}, which require that the distribution of posterior predictions includes the observed choice rates for each answer option.
-To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values (Bppp values), using likelihood of the observed data as a test statistics.
-These Bppp values approximate the probability that a model conditioned on observed data $D_{\text{obs}}$ predicts future data $D_{rep}$, of the same size and format as $D_{obs}$, that is at least as likely as the data $D_{\text{obs}}$ itself is (given the posterior predictive distribution).
-Very small Bppp values indicate that the model might be inadequate for reproducing the data it was trained on, so to speak.
-This is clearly not the case for the RSA model with Bppp values close to 0.5, as shown in Table~\ref{tab:sumStats}.
-In sum, the condition-level predictions by the RSA model, a theoretically motivated PCM, are not discredited by the condition-level data.
-Reversely, the RSA model seems to adequately capture the condition-level data.
-
-We can also use the condition-level predictions for predicting the item-level data.
-If the whole data set for condition $C$ is chunked into item-level data, $D_{C} = \tuple{D^{1}_{C}, \dots, D^{m}_{C} }$, the likelihood function for analysis at the item-level uses the same (condition-level) predictor to separately predict data from each item, so that the likelihood function becomes:
-%
-\begin{align*}
- P^{\text{RSA}}_{\text{item}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \prod_{k = 1}^{m} \text{Multinomial} \left (D_{C}^{k}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3} \right )\,.
-\end{align*}
-%
-Fitting the RSA model to the partitioned data set, we find estimates of Bppp values that do not discredit the model (see Table~\ref{tab:sumStats}).
-This suggests that the item-level variation in the human data is not so pronounced as to provide strong evidence against the condition-level predictor from a theoretically motivated PCM.
-In other words, the condition-level RSA predictions seem to adequately capture also the item-level data.
-
-The following sections apply these tests of Bayesian model criticism also to models built around predictor values from LLMs.
-Section~\ref{sec:item-level-pred} first looks at the item-level data, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.
-
 \begin{table}[t]
@@
 
+To assess goodness-of-fit, we use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
+Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models for condition-level data, which are to be introduced later).
+We see that for both conditions the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-}, which require that the distribution of posterior predictions includes the observed choice rates for each answer option.
+To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values (Bppp values), using likelihood of the observed data as a test statistics.
+These Bppp values approximate the probability that a model conditioned on observed data $D_{\text{obs}}$ predicts future data $D_{rep}$, of the same size and format as $D_{obs}$, that is at least as likely as the data $D_{\text{obs}}$ itself is (given the posterior predictive distribution).
+Very small Bppp values indicate that the model might be inadequate for reproducing the data it was trained on, so to speak.
+This is clearly not the case for the RSA model with Bppp values close to 0.5, as shown in Table~\ref{tab:sumStats}.
+In sum, the condition-level predictions by the RSA model, a theoretically motivated PCM, are not discredited by the condition-level data.
+Reversely, the RSA model seems to adequately capture the condition-level data.
+
+We can also use the condition-level predictions for predicting the item-level data.
+If the whole data set for condition $C$ is chunked into item-level data, $D_{C} = \tuple{D^{1}_{C}, \dots, D^{m}_{C} }$, the likelihood function for analysis at the item-level uses the same (condition-level) predictor to separately predict data from each item, so that the likelihood function becomes:
+%
+\begin{align*}
+ P^{\text{RSA}}_{\text{item}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \prod_{k = 1}^{m} \text{Multinomial} \left (D_{C}^{k}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3} \right )\,.
+\end{align*}
+%
+Fitting the RSA model to the partitioned data set, we find estimates of Bppp values that do not discredit the model (see Table~\ref{tab:sumStats}).
+This suggests that the item-level variation in the human data is not so pronounced as to provide strong evidence against the condition-level predictor from a theoretically motivated PCM.
+In other words, the condition-level RSA predictions seem to adequately capture also the item-level data.
+
+The following sections apply these tests of Bayesian model criticism also to models built around predictor values from LLMs.
+Section~\ref{sec:item-level-pred} first looks at the item-level data, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 As mentioned previously, the items of the reference game experiment differ in which levels of features (color, shape, texture) instantiate the structure of the task shown in Figure \ref{fig:ref-game}, as well as the order of presentation of objects and words.
-We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options.
+We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, such as over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options.
 \mf{@PT references from the NLP literature on effects of prompt variation?}
 It therefore becomes an empirical question of whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\textwidth]{00-pics/closeness-target-by-alpha-item-level.pdf}
+  \caption{
+    Mean predicted target probability for different values of softmax parameter $\alpha$.
+    The solid black lines show the probability of the target options, predicted for a fixed value of $\alpha$ ($x$-axis), averaged over all items for each condition.
+    The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$.
+    For reference, the target probability under a random choice strategy and under the WTA strategy are shown with dashed lines.
+    For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
+  }
+  \label{fig:closeness-target-item-level}
+\end{figure}
 
@@
 \begin{align*}
-P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
+P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
 \end{align*}
@@
 \begin{align*}
-  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
-  = \epsilon_{c} \  P_{\text{item}}^{\text{SM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
+  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
+  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
 \end{align*}
@@
 \begin{align*}
- P^{\text{SM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{SM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
+ P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
 \end{align*}
@@
 Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.
+
+Summary statistics for samples from the posterior distribution over parameters are shown in Table~\ref{tab:sumStats}.
+Credible values of $\alpha$ parameters are comparatively low for the LLM-based model, suggesting that the predicted scores have rather large differences, which have to be compensated for by the softmax link-function to adequately fit the data.
+Figure~\ref{fig:closeness-target-item-level} shows the mean predicted target probabilities for different parameter values of $\alpha$, highlighting that for posterior credible values, the predictions are clearly distinct from the predictions of a WTA strategy.
+This suggests that the WTA strategy, which is the special limiting case for $\alpha \rightarrow \infty$, provides a substantially worse fit for item-level choices than those provided by less extreme values of $\alpha$.
 
@@
 
-Summary statistics for samples from the posterior distribution over parameters are shown in Table~\ref{tab:sumStats}.
-Credible values of $\alpha$ parameters are comparatively low for the LLM-based model, suggesting that the predicted scores have rather large differences, which have to be compensated for by the softmax link-function to adequately fit the data.
-Figure~\ref{fig:closeness-target-item-level} shows the mean predicted target probabilities for different parameter values of $\alpha$, highlighting that for posterior credible values, the predictions are clearly distinct from the predictions of a WTA strategy.
-This suggests that the WTA strategy, which is the special limiting case for $\alpha \rightarrow \infty$, provides a substantially worse fit for item-level choices than those provided by less extreme values of $\alpha$.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\textwidth]{00-pics/closeness-target-by-alpha-item-level.pdf}
-  \caption{
-    Mean predicted target probability for different values of softmax parameter $\alpha$.
-    The solid black lines show the probability of the target options, predicted for a fixed value of $\alpha$ ($x$-axis), averaged over all items for each condition.
-    The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$.
-    For reference, the target probability under a random choice strategy and under the WTA strategy are shown with dashed lines.
-    For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
-  }
-  \label{fig:closeness-target-item-level}
-\end{figure}
-
 Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
@@
 These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
-Put more strongly, a model that seeks to predict what human participants choose on a by-item level, based on the most obvious item-level score, is ruled out by experimental data.
+Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by experimental data.
 
@@
   \caption{
-    Schematic overview over three different approaches of obtaining condition-level predictions by aggregating over item-level scores.
-    The basic item-level score is average next-word log probability.
-    This can be taken as-is into averaging (narrow-scope aggregation), or first transformed from log- into probability-space, either with optimization before (wide scope) or after averaging (intermediate scope).
-    The notation for the condition-level options uses pseudo-code to more clearly reflect the relevant operator scope differences (where triangles represent the pipe operator).
+    Schematic overview over three different approaches of obtaining condition-level predictions by aggregating item-level information.
+    Approaches differ in the kind of item-level information they take into account and in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
   }
@@
 
-Condition-level predictions are a probability distribution over response types, obtained from item-level scores $S(R_{l}, I_{k})$ by averaging over all items belonging to the relevant condition.
+Condition-level predictions are a probability distribution over response types, obtained from item-level scores $\text{S}_{ki}$ by averaging over all items belonging to the relevant condition.
 To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
@@
 \begin{align*}
-  % & \textbf{narrow-scope aggregation} \\
-  & P_{n}(R_{l}, C ; \alpha) \propto \expo \left [  \alpha \ \frac{1}{m} \ \sum_{i = k}^{m} S(R_{l}, I_{k})  \right ]
-    \tag*{\textcolor{gray}{[narrow-scope aggregation]}}
+  & P_{\text{cond}}^{\text{SCR}}(r_{i} \mid C, \alpha_{c})
+    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{ki}  \right ]
+    \tag*{\textcolor{gray}{[average scores (narrow-scope aggregation)]}}
 \end{align*}
@@
   % & \textbf{wide-scope aggregation} \\
-  & P_{w}(R_{l}, C ; \alpha) \propto \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( \alpha \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha \ S(R_{l'}, I_{k}) \right )}
-    \tag*{\textcolor{gray}{[wide-scope aggregation]}}
+  & P_{\text{cond}}^{\text{PRB}}(r_{i} \mid C, \alpha_{c})
+    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha_{c})
+    \tag*{\textcolor{gray}{[average probabilities (wide-scope aggregation)]}}
 \end{align*}
@@
   % & \textbf{intermediate-scope aggregation} \\
-  & P_{i}(R_{l}, C ; \alpha) \propto  \left [ \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( S(R_{l'}, I_{k}) \right )} \right ]^{\alpha}
-    \tag*{\textcolor{gray}{[intermediate-scope aggregation]}}
+  & P_{\text{cond}}^{\text{WTA}}(r_{i} \mid C, \alpha_{c})
+    \propto  \left [ \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) \right ]^{\alpha_{c}}\,,
+    \tag*{\textcolor{gray}{[average WTA (intermediate-scope aggregation)]}}
 \end{align*}
+where $P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) = \lim_{\alpha} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha)$.
+\bigskip
 

### 2023-11-18 03-paper/paper-LLM-probmod.tex
revisions §6

@@
 
-\begin{itemize}
-  \item mention multi-set averages
-  \begin{itemize}
-    \item Let \(\set{I_1, \dots I_m}\) be the multi-set of items that occurred in the human experiment for condition $C$.\footnote{
-  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.
-}
-
-  \end{itemize}
-\end{itemize}
-
-This section explores different ways of deriving a probabilistic likelihood function for the human data from an LLM at the aggregate, condition-level.
-Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
-In the following, we first describe the different options of deriving LLM-based predictors.
-We then compare all probabilistic models based on their adequacy of explaining the human choice data.
+% This section explores different ways of deriving a probabilistic likelihood function for the human data from an LLM at the aggregate, condition-level.
+% Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
+% In the following, we first describe the different options of deriving LLM-based predictors.
+% We then compare all probabilistic models based on their adequacy of explaining the human choice data.
 
@@
 
-Condition-level predictions are a probability distribution over response types, obtained from item-level scores $\text{S}_{ki}$ by averaging over all items belonging to the relevant condition.
-To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
-%
-\begin{align*}
- \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
-\end{align*}
-%
-The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
-%
-\begin{align*}
-  \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
-  \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
-\end{align*}
-%
-This means that there are three conceivable scope-sites for aggregating item-level information as shown in Figure~\ref{fig:measures-overview}.
-Narrow-scope aggregation first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities:
+While LLMs do not make conditional-level predictions as such, they can be derived from item-level scores $\text{S}_{ki}$ by averaging over all items belonging to the relevant condition.
+There are many ways of averaging item-level information.
+Figure~\ref{fig:measures-overview} shows three salient approaches, which differ in what the underlying item-level measure is: the raw scores $\text{S}_{ki}$, the item-level probability derived from it (as introduced in Section~\ref{motivation} and used in Section~\ref{sec:item-level-pred}), or the predictions from the winner-takes-all strategy commonly used in benchmark testing (see Section~\ref{motivation}).
+
+% In general, to map a vector non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
+% %
+% \begin{align*}
+%  \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
+% \end{align*}
+% %
+% The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
+% %
+% \begin{align*}
+%   \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
+%   \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
+% \end{align*}
+
+The \emph{average-scores predictor} first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities using the usual parameterize softmax function:\footnote{
+  Technically, we average over the multi-set \(\set{I_1, \dots I_m}\) of items that occurred in the human experiment for condition $C$.
+  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.}
 %
@@
 %
-Wide-scope aggregation, first transposes scores into probabilities, scales them and only aggregates over items last.
+The \emph{average-probabilities predictor} first transposes scores into probabilities with a parameterized softmax function, and only aggregates over items last.
 \begin{align*}
@@
 %
-Intermediate-scope aggregation performs item-level averaging after mapping scores onto probabilities (using softmax with $\alpha=1$), but before scaling (with a power-law transformation with variable $\alpha$):
+Finally, the \emph{average-WTA predictor} considers the prediction of the WTA strategy as the basic item-level unit to aggregate over.
+To add parameterized scaling to this method a power-law transformation with variable $\alpha$ is a natural choice:
 \begin{align*}
@@
 
-All three condition-level predictors coincide when there is only one item, of course.
-For more items, wide-scope and intermediate-scope aggregation coincide when $\alpha = 1$.
-In all other cases, the predictors are not guaranteed to be identical.
-Conceptually, the main difference is what each measure considers to be the basic item-level unit to aggregate over.
-Narrow-scope aggregation considers raw scores, which are \emph{not} probabilistic predictions at the item level.
-Wide-scope aggregation considers $\alpha$-optimized probabilities at the item level, which is compatible with a procedure of sampling, via softmax decoding, item-level answer options.
-Wide-scope aggregation contains the ``winner-takes-all'' procedure for accuracy scoring in common benchmark tests for $\alpha \rightarrow \infty$.
-Finally, like wide-scope aggregation, intermediate-scope aggregation is also compatible with a sampling based picture, but would use pure decoding (without $\alpha$-optimization) at the item level and reserves the scaling of probability distributions to a stage \emph{after} aggregation.
+The \emph{average-scores} and \emph{average-probabilities} predictions are equivalent if there is only one item, of course.
+The prediction of the \emph{average-WTA} method is then the special case of $\alpha \rightarrow \infty$.
+For cases with more than one item, the predictions of these different predictors are not guaranteed to be the same.
+Conceptually, the \emph{average-probabilities} and the \emph{average-WTA} predictors, but not the \emph{averages-scores} predictor, are compatible with a picture in which condition-level predictions result from the actual predictions at the item level.
+The difference between the \emph{average-probabilities} and the \emph{average-WTA} predictors is that the latter is an aggregate of an item-level predictor that is demonstrably bad, as shown in the previous Section~\ref{sec:item-level-pred}.
 
@@
 
-The three condition-level predictions yield three different probabilistic models, each with two pairs of condition-dependent parameters $\alpha_{c}$, and $\epsilon_{c}$.
-Using priors and Bayesian posterior estimation as previously described for the RSA (see Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
+Using the same approach as for the RSA model in Section~\ref{sec:model-pred-from}, the three condition-level predictions yield three different likelihood functions for the condition-level data.
+Using priors and Bayesian posterior estimation as previously described in Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
 The usual Bayesian summary statistics for posteriors over model parameters are shown in Table~\ref{tab:sumStats}.
-What is noteworthy about these estimates is that the $\alpha_{c}$ parameters for the wide-scope model have a very large credible range.
-This is because, already for values of $\alpha_{c} = 1$, the item-level predictions are virtually ``winner-takes-all'' choices, selecting the option with the maximum score with probability almost 1 for almost all items.
-This implies that, by studying the posterior predictive distribution of the wide-scope model, we are virtually testing condition-level predictions from the standard WTA-strategy used for benchmark testing (see Section~\ref{motivation}).
+What is noteworthy is that for the \emph{average-WTA} model, the estimates of $\alpha_{c}$ in the production condition do not rule out, in fact lie close to, the special value $\alpha_{c}=1$, for which the power-law transformation is the identity function.
+This means that just averaging WTA-responses at the item-level yields a reasonable predictor for the production data at the condition-level.
+However, for the interpretation condition, the value $\alpha_{c}=1$ is clearly outside the range of credible parameter values, so that a simple recipe like ``always average WTA-repsonses without transformation'' is not a viable strategy for good condition-level predictions in general.
+It is also worth noting that for the \emph{average-probability} model values of $\alpha_{c}$ close to one are virtually indistinguishable from the item-level predictions of the WTA strategy (see Figure~\ref{fig:closeness-target-item-level}).
+This suggests that, for the production data, the aggregating item-level predictions from the WTA-strategy gives good condition-level predictions.
+Since the production data is also very biased towards almost exclusively target choices by human participants, this is not too surprising.
 
 Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
-We find that only the theoretical model (RSA) and the intermediate-scope model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
+We find that only the theoretical model (RSA) and the \emph{average-WTA} model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
 To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
-Consequently, the results from Table~\ref{tab:sumStats} suggest that the narrow-scope model fails on the interpretation data, and is at most borderline compatible with the production data; that the wide-scope model is able to reproduce the production data, but fails on the interpretation data; and that only the intermediate-scope model is able to fully recover the data from both conditions.
+Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} model fails on the interpretation data, and is at most borderline compatible with the production data; that the \emph{average-probabilities} model is able to reproduce the production data, but fails on the interpretation data; and that only the \emph{average-WTA} model is able to fully recover the data from both conditions.
 
@@
 Some approaches clearly fail basic checks for statistical goodness-of-fit.
-On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the intermediate-scope model, which is able to recover the patterns in the data.
+On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the \emph{average-WTA} model, which is able to recover the patterns in the data.
 In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.\footnote{
-  A potential worry is that the intermediate-scope model is trivial in the sense that it could have predicted \emph{any} data observation.
+  A potential worry is that the \emph{average-WTA} model is trivial in the sense that it could have predicted \emph{any} data observation.
   Appendix~\ref{sec:range-prior-pred} shows that this is not the case.
+  \mf{todo: redo these calculations with actual WTA!}
 }
 This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.
+On the other hand, there is a stark contrast between the results of item-level and condition-level data.
+The model that was able to properly fit the condition-level data relies on a model of item-level choice that is clearly incompatible with the human item-level data.
+The model seems to be ``right for the wrong reasons.''
 
@@
 
-It is not entirely ludicrous to use numerical predictions from LLMs are
-part of predictive probabilistic models. But we have to be careful.
-Ideally, we average out over low-level variation, such as from order of
-presentation or similar ``nuisance,'' at least as long as we do not
-understand what causes this variation in the predictions of models and
-further research that investigates when exactly this variation accords
-with empirically observed patterns.
+It is not entirely ludicrous to use numerical predictions from LLMs are part of predictive probabilistic models.
+But we have to be careful.
+Ideally, we average out over low-level variation, such as from order of presentation or similar ``nuisance,'' at least as long as we do not understand what causes this variation in the predictions of models and further research that investigates when exactly this variation accords with empirically observed patterns.
 

### 2023-11-21 03-paper/paper-LLM-probmod.tex
minor revisions throughout whole doc

@@
 \begin{abstract}
-  State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks.
-  But targeted assessment with methods inspired by experimental psychology highlights dissimilarities between human behavior and LLM predictions.
-  Going even a step further, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of probabilistic cognitive modeling.
+  State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks and are increasingly used as components in larger applications, where LLM-based predictions serve as proxies for human judgements.
+  This raises questions about the human-likeness of LLM-derived information and whether LLMs, if they deliver human-like behavior, could possibly be considered explanatory models of (aspects of) human cognition or language use.
+  To shed more light on these issues, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of probabilistic cognitive modeling.
   We identify core differences, in terms of conceptual interpretation and usual practical evaluation, between LLMs and probabilistic cognitive models (PCMs).
-  Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level).
+  Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level), and that this is a sense in which LLMs may be said to be not explanatory.
   Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs fail to capture the variance in the human data at the item-level.
@@
 
-% A major contribution of this work is methodological.
-This paper introduces and systematically compare different ways of carving numerical information obtained from an LLM into a structure similar to a parameterized PCM and address the empirical question of whether such an ``LLM-based PCM'' adequately captures the distribution of human forced-choice data at different levels of aggregation.
+This paper introduces and systematically compares different ways of carving numerical information obtained from an LLM into a structure similar to a parameterized PCM and addresses the empirical question of whether such an ``LLM-based PCM'' adequately captures the distribution of human forced-choice data at different levels of aggregation.
 To obtain data for this investigation, we subjected human participants and an LLM (a version of OpenAI's GPT-3.5; see Section~\ref{sec:item-level-pred}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
@@
 We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare these ``LLM-based PCMs'' against each other, and against an established, theoretically informed PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
-This is quite different, and arguably more stringent and insightful, than the standard practice of evaluating LLMs based on accuracy scores, determined by a winner-takes-all strategy to identify a single target response.
+This is quite different, and arguably more stringent and insightful, than the standard practice of evaluating LLMs based on accuracy scores.
 We find that variability predicted by the LLM at the item-level is not borne out by the human data, and that some but not all ways of constructing condition-level predictions are equally good.
 These results suggest that it is not impossible to build predictive models for condition-level data, similar to PCMs, on top of LLM-supplied information, but that seemingly subtle details in the derivation of likelihoods matter.
-The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not. \mf{double-check formulation of results}
+The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not.
 
 The paper is structured as follows.
-Section~\ref{motivation} motivates the investigation in this paper, by enlarging on the importance of deriving distributional predictions for conditional-level forced-choice data, and further elaborating on the relevant differences between LLMs and PCMs.
+Section~\ref{motivation} motivates the investigation in this paper by highlighting relevant differences in the way  LLMs and PCMs make predictions about data from forced-choice tasks.
 Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a commonly used, theoretically-motivate PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
-Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions are adequate to capture the human data at the item-level and finds that they are not, even though the theoretically-motivated RSA model is.
+Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions adequately capture the human data at the item-level.
 Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Motivation: Distributional predictions for condition-level forced-choice data}
+\section{Predictions for multiple-choice tasks from PCMs and LLMs}
 \label{motivation}
@@
 
+This section argues that a major conceptual difference between LLMs and PCMs is that they predict data at different levels of abstraction: the item-level and the condition-level, as introduced below.
+Another important difference is that PCMs are usually evaluated on a non-degenerate likelihood function, whereas common practice in LLM-evaluation is to use a winner-takes-all strategy.
+Problems with this latter strategy are highlighted and alternative ways of defining item- and condition-level predictions from LLMs are introduced for subsequent empirical testing.
+
+\paragraph{Item- and condition-level data.}
 Psychological research into the workings of the human mind aims to find generalizable patterns in the way information is processed within or across different domains of cognition.
@@
 
+\paragraph{Probabilistic cognitive models.}
 Data from human participants for experiments of this kind usually shows some variability between items, and also variability between participants.
@@
 
+\paragraph{PCMs vs.~LLMs.}
 PCMs, thus construed, differ markedly from LLMs.
@@
 Nevertheless, we can derive predictions from LLM's word-level predictions, first at the item-level and also at the condition-level.
-Whether or not these predictions are deemed explanatory, the empirical question arises of whether any of these predictions are empirically adequate.
-This empirical question can be addressed by subjecting ``LLM-based PCMs'' to the same statistical model checking as the more usual theoretically motivated PCMs.
+Whether any of these predictions are adequate is ultimately an empirical question, which can be addressed by subjecting ``LLM-based PCMs'' to the same statistical model checking as the more usual theoretically motivated PCMs.
 
+\paragraph{Predictions from LLMs.}
 Standard benchmark-testing on multiple-choice tasks usually uses a winner-take-all (WTA) strategy \citep[e.g.,][]{srivastava2023-BIGbench} to define accuracy scores.
@@
 Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$, and each of which .
-For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same answer category $R_{i}$, e.g., that $y_{k1}$ is always the designated \emph{target option} that is assumed to be the ``true'' or to-be-selected goal answer, and that the other choice options are equally assignable into meaningful categories of answer types.
+For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same response category $r_{i}$ and that $y_{k1}$ always corresponds to the designated \emph{target option}, $r_{1}$, that is assumed to be the ``true'' or to-be-selected goal answer.
 
 The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
-  \mf{How to extend to masked language models?}
+  \mf{TODO: Add quick word on how this can in principle be extended to masked language models.}
   More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
@@
 
-The \emph{condition-level prediction} of the WTA approach is obtained by averaging over the item-level predictions for each answer category:
+The \emph{condition-level prediction} of the WTA approach is obtained by averaging over the item-level predictions for each response category:
 %
 \begin{align*}
-  P_{\text{cond}}^{\text{WTA}}\left(R_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{WTA}} \left(y_{ki} \right)\,.
+  P_{\text{cond}}^{\text{WTA}}\left(r_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{WTA}} \left(y_{ki} \right)\,.
 \end{align*}
 %
-The \emph{WTA-based accuracy} is the probability $P_{\text{cond}}^{\text{WTA}}\left(R_{1} \right)$ of selecting the target option over all instances of the task.
+The \emph{WTA-based accuracy} is the probability $P_{\text{cond}}^{\text{WTA}}\left(r_{1} \right)$ of selecting the target option $r_{1}$ over all instances of the task.
 This is the most prevalent measure of the quality of LLM predictions.
@@
 \begin{align*}
-  P_{\text{cond}}^{\text{SM}}\left(R_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{ki} \right)\,.
+  P_{\text{cond}}^{\text{SM}}\left(r_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{ki} \right)\,.
 \end{align*}
-As before, the \emph{softmax-based accuracy}, $P_{\text{cond}}^{\text{SM}}\left(R_{i} \right)$, is the average probability of choosing the designated target option $y_{k1}$.
+As before, the \emph{softmax-based accuracy}, $P_{\text{cond}}^{\text{SM}}\left(r_{1} \right)$, is the average probability of choosing the designated target response option $r_{1}$.
+
 For $\alpha \rightarrow \infty$, the softmax strategy converges to the WTA strategy, but for finite $\alpha$ the approaches differ.
-This has interesting entailments for accuracy assessment in more engineering-oriented contexts, as well as for questions of exploring the ``human-likeness'' of LLM predictions.
-
-Looking at accuracy assessment, WTA-based accuracy measures can even differ qualitatively from accuracy measures obtained from the more general softmax-based strategy.
-Differences in accuracy measures depend on the variation in item-level scores, in particular the relation between score-ordering and score-differences.
-The example below shows that categorically different results are, in principle, possible when using WTA- or softmax based accuracy.
-The example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
+Indeed, WTA-based accuracy can even differ qualitatively from softmax-based accuracy, as shown by the following example.
+The example shows that differences between accuracy measures depend on the variation in item-level scores, in particular on the relation between score-ordering and score-differences.
+Note that the example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
 %
@@
 %
-Item-level scores $\text{S}_{ki}$, as defined in Section~\ref{motivation}, are obtained from GPT-3.5 \mf{which model exactly, when assessed, cite paper}.
+Item-level scores $\text{S}_{ki}$, as defined in Section~\ref{motivation}, were obtained from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} .
 % \footnote{
@@
     Schematic overview over three different approaches of obtaining condition-level predictions by aggregating item-level information.
-    Approaches differ in the kind of item-level information they take into account and in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
+    Approaches differ in the kind of item-level information they take into account, which entails differences in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
   }
@@
 The \emph{average-scores predictor} first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities using the usual parameterize softmax function:\footnote{
-  Technically, we average over the multi-set \(\set{I_1, \dots I_m}\) of items that occurred in the human experiment for condition $C$.
-  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.}
+  The reported results average over the multi-set \(\set{I_1, \dots I_m}\) of items that occurred in the human experiment for condition $C$.
+  By using a multi-set, which may contain a single item multiple times, we produce aggregate predictions for exactly the set of items that the participant group saw, which provides the most fitting counterpart to the human data.}
 %
@@
   & P_{\text{cond}}^{\text{SCR}}(r_{i} \mid C, \alpha_{c})
-    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{ki}  \right ]
+    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{ki}  \right ] \,.
     \tag*{\textcolor{gray}{[average scores (narrow-scope aggregation)]}}
@@
 %
-The \emph{average-probabilities predictor} first transposes scores into probabilities with a parameterized softmax function, and only aggregates over items last.
+The \emph{average-probabilities predictor} first transposes scores into probabilities with a parameterized softmax function, and only aggregates over items last:
 \begin{align*}
@@
   & P_{\text{cond}}^{\text{PRB}}(r_{i} \mid C, \alpha_{c})
-    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha_{c})
+    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha_{c}) \,.
     \tag*{\textcolor{gray}{[average probabilities (wide-scope aggregation)]}}
@@
 Finally, the \emph{average-WTA predictor} considers the prediction of the WTA strategy as the basic item-level unit to aggregate over.
-To add parameterized scaling to this method a power-law transformation with variable $\alpha$ is a natural choice:
+To add parameterized scaling to this method, a power-law transformation with variable $\alpha$ is a natural choice:
 \begin{align*}
@@
 
-The \emph{average-scores} and \emph{average-probabilities} predictions are equivalent if there is only one item, of course.
-The prediction of the \emph{average-WTA} method is then the special case of $\alpha \rightarrow \infty$.
+The \emph{average-scores} and \emph{average-probabilities} predictions are equivalent if there is only one item, in which case the prediction of the \emph{average-WTA} method is the special case of $\alpha \rightarrow \infty$.
 For cases with more than one item, the predictions of these different predictors are not guaranteed to be the same.
@@
 
-% \begin{table}[t]
-%   \centering
-
-%   % \includegraphics[width=0.9\linewidth]{00-pics/posterior-stats.pdf}
-
-%   \begin{tabular}{llllcr}
-%     \toprule \addlinespace[1ex]
-%     model        & condition      & Parameter  & $|$95\% & mean & 95\%$|$ \\
-%     \midrule  \addlinespace[1ex]
-%     narrow       & production     & $\alpha$   & 0.19    & 0.22 & 0.26 \\
-%                  &                & $\epsilon$ & 0.00    & 0.04 & 0.11 \\
-%                  & interpretation & $\alpha$   & 0.15    & 0.18 & 0.20 \\
-%                  &                & $\epsilon$ & 0.00    & 0.02 & 0.06 \\ \addlinespace[0.75ex]
-%     intermediate & production     & $\alpha$   & 0.86    & 1.04 & 1.22 \\
-%                  &                & $\epsilon$ & 0.08    & 0.13 & 0.18 \\
-%                  & interpretation & $\alpha$   & 0.36    & 0.48 & 0.60 \\
-%                  &                & $\epsilon$ & 0.00    & 0.05 & 0.12 \\ \addlinespace[0.75ex]
-%     wide         & production     & $\alpha$   & 0.22    & 2.69 & 8.79 \\
-%                  &                & $\epsilon$ & 0.08    & 0.13 & 0.19 \\
-%                  & interpretation & $\alpha$   & 0.19    & 1.08 & 6.03 \\
-%                  &                & $\epsilon$ & 0.00    & 0.06 & 0.18 \\ \addlinespace[0.75ex]
-%     RSA          & production     & $\alpha$   & 2.59    & 3.16 & 3.71 \\
-%                  &                & $\epsilon$ & 0.08    & 0.13 & 0.18 \\
-%                  & interpretation & $\alpha$   & 0.24    & 0.66 & 1.07 \\
-%                  &                & $\epsilon$ & 0.10    & 0.15 & 0.20 \\ \addlinespace[0.25ex]
-%     \bottomrule \\
-%   \end{tabular}
-
-%   \caption{
-%     Summary statistics of posterior samples.
-%     Numbers indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
-%     \mf{make wide format}
-%   }
-%   \label{fig:posterior-stats}
-% \end{table}
-
-Using the same approach as for the RSA model in Section~\ref{sec:model-pred-from}, the three condition-level predictions yield three different likelihood functions for the condition-level data.
-Using priors and Bayesian posterior estimation as previously described in Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
-The usual Bayesian summary statistics for posteriors over model parameters are shown in Table~\ref{tab:sumStats}.
+Using the same approach as for the RSA model in Section~\ref{sec:model-pred-from}, we can build likelihood functions for condition-level data around the three predictors introduced above.
+With the same priors and methods used before, we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
+Summary statistics for posteriors over model parameters are shown in Table~\ref{tab:sumStats}.
 What is noteworthy is that for the \emph{average-WTA} model, the estimates of $\alpha_{c}$ in the production condition do not rule out, in fact lie close to, the special value $\alpha_{c}=1$, for which the power-law transformation is the identity function.
@@
 It is also worth noting that for the \emph{average-probability} model values of $\alpha_{c}$ close to one are virtually indistinguishable from the item-level predictions of the WTA strategy (see Figure~\ref{fig:closeness-target-item-level}).
-This suggests that, for the production data, the aggregating item-level predictions from the WTA-strategy gives good condition-level predictions.
+This suggests that, for the production data, aggregation of item-level predictions from the WTA-strategy gives good condition-level predictions.
 Since the production data is also very biased towards almost exclusively target choices by human participants, this is not too surprising.
@@
 Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
-We find that only the theoretical model (RSA) and the \emph{average-WTA} model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
+We find that only the theoretical model (RSA) and the \emph{average-WTA} model pass this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
 To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
-Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} model fails on the interpretation data, and is at most borderline compatible with the production data; that the \emph{average-probabilities} model is able to reproduce the production data, but fails on the interpretation data; and that only the \emph{average-WTA} model is able to fully recover the data from both conditions.
-
-% \begin{table}[ht]
-% \centering
-
-% % \begin{tabular}{lrr}
-% %   \toprule
-% %   model        & production & interpretation \\ \midrule
-% %   narrow       & 0.05       & 0.00 \\
-% %   wide         & 0.51       & 0.00 \\
-% %   intermediate & 0.49       & 0.26 \\
-% %   RSA          & 0.49       & 0.52 \\
-% %   \bottomrule \\
-% % \end{tabular}
-
-% \begin{tabular}{lrrrr}
-%   \toprule
-%   & narrow & wide & intermediate & RSA \\ \midrule
-%   production & 0.05 & 0.51 & 0.49 & 0.49 \\
-%   interpretation & 0.00 & 0.00 & 0.26 & 0.52 \\ \bottomrule \\
-% \end{tabular}
-
-% \caption{Sample-based estimates of Bayesian posterior predictive $p$-values for each model and each condition, based on likelihood of the observed data as test statistic.}
-% \label{tab:Bppp-values}
-% \end{table}
+Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} model fails on the interpretation data, and is at most borderline compatible with the production data; that the \emph{average-probabilities} model is able to reproduce the production data, but fails on the interpretation data; and that only the \emph{average-WTA} model does not fail to capture the data from both conditions.
 
@@
 }
-This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.
 On the other hand, there is a stark contrast between the results of item-level and condition-level data.
@@
 {to be continued}
+
+\begin{itemize}
+  \item This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.
+\end{itemize}
 

### 2023-11-23 03-paper/paper-LLM-probmod.tex
fixed typos §6

@@
 
-The \emph{average-scores predictor} first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities using the usual parameterize softmax function:\footnote{
+The \emph{average-scores predictor} first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities using the usual parameterized softmax function:\footnote{
   The reported results average over the multi-set \(\set{I_1, \dots I_m}\) of items that occurred in the human experiment for condition $C$.
@@
 Finally, the \emph{average-WTA predictor} considers the prediction of the WTA strategy as the basic item-level unit to aggregate over.
-To add parameterized scaling to this method, a power-law transformation with variable $\alpha$ is a natural choice:
+To add parameterized scaling to this method, a power-law transformation is a natural choice:
 \begin{align*}
@@
 The \emph{average-scores} and \emph{average-probabilities} predictions are equivalent if there is only one item, in which case the prediction of the \emph{average-WTA} method is the special case of $\alpha \rightarrow \infty$.
-For cases with more than one item, the predictions of these different predictors are not guaranteed to be the same.
-Conceptually, the \emph{average-probabilities} and the \emph{average-WTA} predictors, but not the \emph{averages-scores} predictor, are compatible with a picture in which condition-level predictions result from the actual predictions at the item level.
-The difference between the \emph{average-probabilities} and the \emph{average-WTA} predictors is that the latter is an aggregate of an item-level predictor that is demonstrably bad, as shown in the previous Section~\ref{sec:item-level-pred}.
+For cases with more than one item, the predictions of the three predictors are not guaranteed to be the same.
+Conceptually, the \emph{average-probabilities} and the \emph{average-WTA} predictors, but not the \emph{average-scores} predictor, are compatible with a picture in which condition-level predictions result from the actual predictions at the item level.
+The difference between the \emph{average-probabilities} and the \emph{average-WTA} predictor is that the latter is an aggregate of an item-level predictor that is demonstrably bad, as shown in the previous Section~\ref{sec:item-level-pred}.
 
@@
 This means that just averaging WTA-responses at the item-level yields a reasonable predictor for the production data at the condition-level.
-However, for the interpretation condition, the value $\alpha_{c}=1$ is clearly outside the range of credible parameter values, so that a simple recipe like ``always average WTA-repsonses without transformation'' is not a viable strategy for good condition-level predictions in general.
+However, for the interpretation condition, the value $\alpha_{c}=1$ is clearly outside the range of credible parameter values, so that a simple recipe like ``always average WTA-responses without transformation'' is not a viable strategy for good condition-level predictions in general.
 It is also worth noting that for the \emph{average-probability} model values of $\alpha_{c}$ close to one are virtually indistinguishable from the item-level predictions of the WTA strategy (see Figure~\ref{fig:closeness-target-item-level}).
@@
 On the other hand, there is a stark contrast between the results of item-level and condition-level data.
-The model that was able to properly fit the condition-level data relies on a model of item-level choice that is clearly incompatible with the human item-level data.
+The model that was able to properly fit the condition-level data builds on predictions for item-level choice that are clearly incompatible with the human item-level data.
 The model seems to be ``right for the wrong reasons.''

### 2023-11-23 03-paper/paper-LLM-probmod.tex
draft conclusion

@@
 
-It is not entirely ludicrous to use numerical predictions from LLMs are part of predictive probabilistic models.
-But we have to be careful.
-Ideally, we average out over low-level variation, such as from order of presentation or similar ``nuisance,'' at least as long as we do not understand what causes this variation in the predictions of models and further research that investigates when exactly this variation accords with empirically observed patterns.
+While the common practice in evaluating the capabilities of LLMs is based on accuracy averaged over large collections of data, this work took the alternative route to explore what we learn if we subjected LLMs to the same routines and strong demands on distributional quality of fit to human data as we normally do for probabilistic cognitive models.
+Taking this perspective, a main contribution of this paper is methodological, showing how statistical model criticism can be used for LLMs in the first place, and, more specifically, how it can be insightful in the detailed assessment of how LLMs might or might not be able to predict human data at the degree of accuracy that we would normally aspire for when dealing with explanatory statistical models in cognitive science.
+Applying this comparative approach to a minimal, but non-trivial data set, we find that the LLM predictions on a per-item level predict variance that is not attested in the human data.
+From several candidate predictor measures for aggregate condition-level data, only one was not refuted by the human data, but this was one that relied on the empirically implausible WTA-strategy at the item-level, incidentally the same strategy that is commonly used in accuracy-based benchmark testing \citep{srivastava2023-BIGbench}.
 
-This also implies that we should likely \emph{not} (yet) aspire to use
-LLMs are models or individual- or item-level predictors.
+\paragraph{Explanatory power.}
+The comparison of LLMs and PCMs led to a simple but consequential observation, namely that the atomic predictions or LLMs are at the item level, while explanatory PCMs usually make predictions at a more abstract level of conceptualization, similar to the level of an experimental condition.
+This is one sense in which LLMs may be felt to be less, or not at all, explanatory.
+They do not offer, at least not directly, a human-comprehensible compression of reality into a \emph{kind} of response pattern, over and beyond making a predicting for each particular situation.
+As this kind of compression is arguably important for a sense of understanding \citep{Dellsen2020:Beyond-Explanat,Grimm2021:Understanding}, the direct comparison of LLMs with common practices in experimental psychology and with probabilistic cognitive models, provides an interesting perspective on why LLMs are often felt to be lacking in explanatory power.
 
-( {TODO: rethink conclusion})
+This perspective on the explanatory role of LLMs goes beyond the factors of performance, indirect support and parsimony identified by \citet{SortNoopDeemtervan-Deemter2023:Dimensions-of-E}.
+It is also subtly different from considerations of an LLMs ability to generalize \citep{HupkesDankers2020:Compositionalit}.
+It rather suggests that \emph{transferability} is a dimension to ``explanatory power'' that is important as well.
+Imagine that models $M$ and $M'$ have been designed for and trained on data from a situation $S$, but need to be applied to a different situation $T$.
+Assume that for model $M$ the only way to make prediction for $T$ is to collect data pertaining to $T$, and either retrain or fine-tune the model.
+In contrast, for model $M'$ can make predictions for $T$ without novel data collection by recognizing a meaningful difference between $S$ and $T$ and consequently manually changing parameter values or model-internal mechanics to accommodate for this change.
+In that case, model $M'$ would be more transferable than model $M$.
+For example, if we change the experimental setting for a reference game to consist of data from a special population, such as very young children or language-impaired adults, a potentially reasonable architectural change to a model like the RSA model is to consider differences in the sets of alternative utterances for the speaker \citep[e.g.][]{Noveck2001:When-Children-a}.
+Even though this is only a vague explication of a notion of transferability, it suffices to corroborate the intuition that models like PCMs, which are designed to operate at a higher level of conceptual abstraction, will often appear more transferable than models like LLMs, which make predictions not for \emph{kinds} of situations but for \emph{particular} situations.
+Whether any given model's transfer-ability is correct, is an orthogonal empirical question.
+It is also an empirical question, exactly to which extent and in which areas LLMs are (not) transferable, especially if we consider prompting a transfer strategy \citep{LiuLiu2022:Generated-Knowl,XieRaghunathan2022:An-Explanation-}.
+In any case, the comparison between LLMs and PCMs started here suggests systematic research into the transfer-ability and explanatory value of LLMs, e.g., by prompting strategies that are empirically insightful or by their use in composite neuro-symbolic models that implement theoretically-meaningful conceptual differences between types of situations.
 
-{to be continued}
+\paragraph{Variability in predictions and performance measures.}
+Despite being anchored in a small, but detailed case-study, the fact that different plausible methods of aggregating item-level information led to condition-level predictions of variable quality is worrisome.
+Similarly, we argued that the wide-spread reliance on a winner-takes-all strategy might be inconsistent with the actual use practices of LLMs, which may not always rely on a temperature-zero sampling strategy (see Section~\ref{motivation}).
+In conclusion, research on LLM should systematically study the conceptual and empirical consequences of seemingly minor decisions in evaluation or application settings.
+Variability of performance measures also comes with a well-known risk for robust and reproducible science.
+The more researcher degrees of freedom there are, the higher the risk of false results, even in the absence of intentions to mislead \citep[e.g.][]{Ioannidis2005:Why-Most-Publis,Chambers2017:The-Seven-Deadl}.
+The comparison of LLMs with PCMs here raises awareness for the issue of robust research methods also in NLP \citep{WielingRawee2018:Reproducibility}.
 
-\begin{itemize}
-  \item This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.
-\end{itemize}
+\paragraph{Human-like predictions from LLM.}
+The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
+Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
+Based on the data set and the detailed analyses conducted here, it seems not infeasible to use numerical predictions from LLMs are part of predictive probabilistic models.
+But, ideally, low-level variation, such as from order of presentation or similar ``nuisance variables,'' should be averaged out or taken into account in some way or other, at least as long as we do not understand what causes this variation in the predictions of models and further research that investigates when exactly this variation accords with empirically observed patterns.
+In sum, we conjecture that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is taken into account.
+
+\paragraph{No substitute for human subjects.}
+Looking at LLM-derived predictions for human data from the perspective of cognitive modeling highlights the fact that LLMs predict item-level variation, but not subject-level variation, which is common in human data.
+We may consider variation introduced via softmax / temperature as analogous to between-human variation, but this likely falls short of reality, where pronounced differences in answer behavior may surface.
+For the particular case of pragmatic language use, prior research has shown that individual participants have markedly different behavioral profiles, often consistently behaving like literal language users or more sophisticated language users \citep[e.g.,][]{NieuwlandDitman2010:On-the-incremen,FrankeDegen2015:Reasoning-in-Re,SpychalskaKontinen2016:Investigating-s}.
+It is an open question whether predictions from LLMs reflect the same kind of variation.
+The results presented here recommend skepticism.
+We therefore side with the cautious voices that do not recommend replacing human participants with LLMs in psychological research \citep{DillionTandon2023:Can-AI-language,HardingDAlessandro2023:AI-language-mod}.
+In contrast, there is a challenge for LLM research.
+The fact that aggregate predictions can track aggregate human behaviour means that the variance in on both sides is washed out to achieve a similar result.
+This raises the issue of finding the systematic differences between LLMs and human answerers.
+The task then would be to find cases where the differences do \emph{not} wash out and ask: What if anything do these cases share?
+
+\paragraph{Limitations and follow-up work.}
+The scope of our experimental investigation was deliberately small but perspicuous.
+Focusing on the methodological contributions and details in performance assessments, we focused on a minimal non-trivial case study where we could triangulate LLMs, PCMs and human data.
+This work may therefore serve as starting point for a wider investigation of more complex data sets and case studies in which LLMs are analyzed as and directly compared to explanatory PCMs.
 
@@
 
-\newpage
-
-\section*{Snippets}
-
-\begin{itemize}
-  \item for conclusion?
-  \begin{itemize}
-    \item The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
-    Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
-  \end{itemize}
-  \item PT: the high-level take-away that item-level LLM predictions don't fit human predictions well might, as always, be met with the comment that the task is rather artificial.
-maybe this should go into the discussion
-  \item FC: The fact that the aggregate predictions track aggregate human behaviour means that the discrepancies are washing out.
-But there's also the interesting question of finding the systematic differences between LLMs and people.
-The task then would be to find sets of stimuli where the differences do \emph{not} wash out and ask: What if anything do these stimuli share?
-This paper could also be a jumping board for this line of research.
-  \item PT: maybe a conclusion in the spirit of "cogmod in the age of LLMs"/ "computational psycholinguistics \& careful assessment of reasoning / linguistic abilities of LLMs against human performance" might be not entirely ludicrous too
-\end{itemize}
-
 \end{document}

### 2023-11-25 01-code/llama_logprobs.py
add script for eliciting llama log probs

@@ -1,0 +1,187 @@
+import time
+import numpy as np
+import pandas as pd
+import os
+import argparse
+from tqdm import tqdm
+from pprint import pprint
+# import make_material
+from make_material import sample_vignette
+import torch
+from transformers import (
+    AutoModelForCausalLM, 
+    AutoTokenizer
+)
+# get device
+device = torch.device("cuda" if torch.cuda.is_available() else "cpu")
+logsoftmax = torch.nn.LogSoftmax(dim=-1)
+
+def getLogProbContinuation(
+        initialSequence, 
+        continuation, 
+        model,
+        tokenizer,
+        preface = ''):
+    """
+    Helper for retrieving log probability of different response types from Llama-2 of various sizes.
+    """
+    initialSequence = preface + initialSequence
+    # tokenize separately, so as to know the shape of the continuation
+    input_ids_prompt = tokenizer(
+        initialSequence, 
+        return_tensors="pt",
+    ).input_ids
+    input_ids_continuation = tokenizer(
+        continuation,
+        return_tensors="pt",
+    ).input_ids
+    print("input_ids prompt ", input_ids_prompt)
+    print("input ids continuation ", input_ids_continuation)
+    # cut off the first token of the continuation, as it is SOS
+    input_ids = torch.cat(
+        (input_ids_prompt, input_ids_continuation[:, 1:]), 
+        -1
+    ).to(device)
+    # pass through model
+    outputs = model(
+        input_ids,
+    )
+    # transform logits to probabilities
+    llama_output_scores = logsoftmax(
+        outputs.logits[0]
+    )
+    # retreive log probs at token ids
+    # transform input_ids to a tensor of shape [n_tokens, 1] for this
+    input_ids_probs = input_ids.squeeze().unsqueeze(-1)
+    # retreive at correct token positions
+    conditionalLogProbs = torch.gather(
+        llama_output_scores, 
+        dim=-1, 
+        index=input_ids_probs
+    ).flatten().tolist()        
+    # slice output to only get scores of the continuation, not the context
+    continuationConditionalLogProbs = conditionalLogProbs[
+        input_ids_prompt.shape[-1]:
+    ]
+    # compute continunation log prob
+    sentLogProb = torch.sum(continuationConditionalLogProbs).item()
+
+    return sentLogProb
+            
+
+def soft_max(scores, alpha=1):
+    scores = np.array(scores)
+    output = np.exp(scores * alpha)
[…]

### 2023-11-25 01-code/run_llama.sh

@@ -1,0 +1,31 @@
+#!/bin/bash
+#SBATCH --partition=single
+#SBATCH --ntasks=1
+#SBATCH --time=20:00:00
+#SBATCH --mem=20gb
+#SBATCH --gres=gpu:A40:1
+
+echo 'Running simulation'
+
+
+# Load conda
+module load devel/miniconda/3
+source $MINICONDA_HOME/etc/profile.d/conda.sh
+
+conda deactivate
+# Activate the conda environment
+conda activate llmlink
+
+echo "Conda environment activated:"
+echo $(conda env list)
+echo " "
+echo "Python version:"
+echo $(which python)
+echo " "
+
+# activate CUDA
+module load devel/cuda/11.6
+
+python3 -u llama_logprobs.py \
+    --model_name="meta-llama/Llama-2-7b-hf"
+done

### 2023-12-01 01-code/llama_logprobs.py
fix indexing of token log probs

@@ -55,11 +55,13 @@ def getLogProbContinuation(
     )
     # transform logits to probabilities
     llama_output_scores = logsoftmax(
-        outputs.logits[0]
+        outputs.logits[0][:, :-1]
     )
     # retreive log probs at token ids
     # transform input_ids to a tensor of shape [n_tokens, 1] for this
-    input_ids_probs = input_ids.squeeze().unsqueeze(-1)
+    # cut off the sos token so as to get predictions for the actual token conditioned on 
+    # preceding context
+    input_ids_probs = input_ids[:, 1:].squeeze().unsqueeze(-1)
     # retreive at correct token positions
     conditionalLogProbs = torch.gather(
         llama_output_scores, 

### 2023-12-01 03-paper/paper-LLM-probmod.tex

@@
 %
-where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
-\mf{FC objected to ``condition-specificity'' which is fair, but I leave it as it was; my reasoning: we use model checking to rule out models; in this setting the conservative strategy is to err on the side of making the models too powerful (even at the cost of conceptual plausibility)}
+where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.\footnote{
+  Since the RSA model predicts probability 0 for the distractor option, this model is, in principle, able to predict any probability distribution over the three choice categories that is compatible with the order: $P_{r}(R_{t}) \ge P_{r}(R_{c}) \ge P_{r}(R_{d})$.
+  Intuitively, this is because with $\epsilon=0$, $P_{r}(R_{d}, C; \alpha_{c}) = 0$, so that there is an $\alpha$ for any ratio of predicted choice probabilities for target and competitor, as long as the target probability is no smaller than the competitor probability.
+  The $\epsilon$-transformation is essentially a linear shift in the probability simplex towards the maximum entropy prediction, so that every prediction which obeys the ordering restriction above can be made for some pair of $\alpha$ and $\epsilon$.
+  This prediction-triviality is met in two ways.
+  For one, the Bayesian priors on model parameters soft-constrain the model, so that the \emph{ex ante} credible predictions do rule out many logically possible observations.
+  For another, we break the triviality by assigning a non-zero probability to the prediction for the distractor option.
+  The same triviality problem lurks for the \emph{average-WTA} model introduced in Section~\ref{llm-predictions-for-reference-games}, and the same solution is applied to it.
+}
 
@@
 Parameterized predictions, like in Equation~(\ref{eq:RSA-likelihood}), can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
-Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
+Let $\alpha_{c}\sim \text{log-Normal}(1,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,15)$ have a Beta prior favoring small values.
 Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{tab:sumStats}).\footnote{
@@
 
-  item-level  & RSA                & production     & 2.62 & 3.15 & 3.69 & 0.08 & 0.13 & 0.17 & 0.34 &  \\
-  item-level  & RSA                & interpretation & 0.24 & 0.66 & 1.05 & 0.10 & 0.15 & 0.20 & 0.25 &  \\
-  item-level  & LLM                & production     & 0.15 & 0.21 & 0.29 & 0.00 & 0.10 & 0.21 & 0.00 & * \\
-  item-level  & LLM                & interpretation & 0.08 & 0.13 & 0.18 & 0.00 & 0.17 & 0.33 & 0.00 & * \\
-  cond.-level & RSA                & production     & 2.59 & 3.16 & 3.71 & 0.08 & 0.13 & 0.18 & 0.49 &  \\
-  cond.-level & RSA                & interpretation & 0.24 & 0.66 & 1.07 & 0.10 & 0.15 & 0.20 & 0.52 &  \\
-  cond.-level & avg. scores        & production     & 0.19 & 0.22 & 0.26 & 0.00 & 0.04 & 0.11 & 0.05 & * \\
-  cond.-level & avg. scores        & interpretation & 0.15 & 0.18 & 0.20 & 0.00 & 0.02 & 0.06 & 0.00 & * \\
-  cond.-level & avg. probabilities & production     & 0.86 & 1.04 & 1.22 & 0.08 & 0.13 & 0.18 & 0.51 &  \\
-  cond.-level & avg. probabilities & interpretation & 0.36 & 0.48 & 0.60 & 0.00 & 0.05 & 0.12 & 0.00 & * \\
-  cond.-level & avg. WTA           & production     & 0.87 & 1.04 & 1.23 & 0.08 & 0.13 & 0.17 & 0.49 &  \\
-  cond.-level & avg. WTA           & interpretation & 0.17 & 0.39 & 0.61 & 0.10 & 0.15 & 0.20 & 0.26 &  \\
+  item-level  & RSA                & production     & 2.62 & 3.13 & 3.69 & 0.08 & 0.12 & 0.16 & 0.29 &  \\
+  item-level  & RSA                & interpretation & 0.21 & 0.67 & 1.08 & 0.08 & 0.14 & 0.19 & 0.21 &  \\
+  item-level  & LLM                & production     & 0.15 & 0.19 & 0.24 & 0.00 & 0.07 & 0.16 & 0.00 & * \\
+  item-level  & LLM                & interpretation & 0.08 & 0.11 & 0.14 & 0.00 & 0.09 & 0.22 & 0.00 & * \\
+  cond.-level & RSA                & production     & 2.62 & 3.14 & 3.70 & 0.08 & 0.12 & 0.17 & 0.50 &  \\
+  cond.-level & RSA                & interpretation & 0.27 & 0.68 & 1.13 & 0.09 & 0.14 & 0.19 & 0.51 &  \\
+  cond.-level & avg. scores        & production     & 0.19 & 0.22 & 0.25 & 0.00 & 0.03 & 0.08 & 0.05 &  \\
+  cond.-level & avg. scores        & interpretation & 0.15 & 0.18 & 0.20 & 0.00 & 0.02 & 0.05 & 0.00 & * \\
+  cond.-level & avg. probabilities & production     & 0.86 & 1.03 & 1.22 & 0.08 & 0.12 & 0.17 & 0.47 &  \\
+  cond.-level & avg. probabilities & interpretation & 0.36 & 0.46 & 0.59 & 0.00 & 0.04 & 0.10 & 0.00 & * \\
+  cond.-level & avg. WTA           & production     & 0.86 & 1.04 & 1.22 & 0.08 & 0.12 & 0.17 & 0.48 &  \\
+  cond.-level & avg. WTA           & interpretation & 0.15 & 0.37 & 0.58 & 0.03 & 0.12 & 0.18 & 0.49 &  \\
 
@@
 On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the \emph{average-WTA} model, which is able to recover the patterns in the data.
-In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.\footnote{
-  A potential worry is that the \emph{average-WTA} model is trivial in the sense that it could have predicted \emph{any} data observation.
-  Appendix~\ref{sec:range-prior-pred} shows that this is not the case.
-  \mf{todo: redo these calculations with actual WTA!}
-}
+In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.
 On the other hand, there is a stark contrast between the results of item-level and condition-level data.
@@
 
+The work presented here also contains considerable researcher degrees of freedom.
+We only considered one out of several conceivable ways of carving a parameterized likelihood function from LLM-based predictors.
+If future work would contribute to systematically exploring these, the main goal of this paper would have been met, which is to raise awareness for the possibility, perhaps even necessity, to scrutinize LLMs at the same level of rigorous detail as other models in cognitive science.
+
 \paragraph{Human-like predictions from LLM.}
@@
 This work may therefore serve as starting point for a wider investigation of more complex data sets and case studies in which LLMs are analyzed as and directly compared to explanatory PCMs.
+
+\printbibliography[heading=bibintoc]
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Range of prior predictions of intermediate-scope model}
-\label{sec:range-prior-pred}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section{Range of prior predictions of intermediate-scope model}
+% \label{sec:range-prior-pred}
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-It may seem that, given the freedom to adjust optimization $\alpha$ and random guessing rate $\epsilon$, some models might be able to explain \emph{any} data observation.
-The intermediate-scope model, which was not discredited by model criticism in terms of recovery-based posterior predictive checks, is not trivial in this sense.
-Figure~\ref{fig:prediction-range} shows the range of predictions that the intermediate-scope model is, in principle, capable of making (for the whole possible range of \(\alpha\) and \(\epsilon\) values, disregarding Bayesian priors).
-There are distributions over response types which the model does not predict \emph{ex ante}.
-For example, the model does not predict cases where the number of competitor choices is almost as high as the number of target choice, as would be the case if participants would consistently \emph{not} engage in pragmatic reasoning.
+% It may seem that, given the freedom to adjust optimization $\alpha$ and random guessing rate $\epsilon$, some models might be able to explain \emph{any} data observation.
+% The intermediate-scope model, which was not discredited by model criticism in terms of recovery-based posterior predictive checks, is not trivial in this sense.
+% Figure~\ref{fig:prediction-range} shows the range of predictions that the intermediate-scope model is, in principle, capable of making (for the whole possible range of \(\alpha\) and \(\epsilon\) values, disregarding Bayesian priors).
+% There are distributions over response types which the model does not predict \emph{ex ante}.
+% For example, the model does not predict cases where the number of competitor choices is almost as high as the number of target choice, as would be the case if participants would consistently \emph{not} engage in pragmatic reasoning.
 
-\begin{figure}[t]
-  \centering
+% \begin{figure}[t]
+%   \centering
 
-  \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-prod.png}
-  %
-  \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-inter.png}
+%   \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-prod.png}
+%   %
+%   \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-inter.png}
 
-  \caption{
-    Range of predictions the intermediate-scope model makes for any pair of parameter values $\alpha_{c}$ and $\epsilon_{c}$.
-    The light gray triangle with the red boundary is the probability simplex, i.e., the space of all possible 3-place probability vectors.
-    The darker gray area in gray boundary lines is the subspace that is compatible with the ordering that the probability of choosing the target is bigger than that of the competitor, which in turn is bigger than that of the distractor.
-    The blue area is the subspace of probabilistic predictions the intermediate-scope model makes under any value of its parameters.
-    The orange dot in the middle is the ``Laplace point'' of equal probability for all three options, which the model predicts for $\alpha=0$ or $\epsilon=1$.
-    The yellow dot is the ``vanilla prediction'' for $\alpha_{c}=1$ and $\epsilon_{c}=0$.
-  }
-  \label{fig:prediction-range}
-\end{figure}
-
-\printbibliography[heading=bibintoc]
+%   \caption{
+%     Range of predictions the intermediate-scope model makes for any pair of parameter values $\alpha_{c}$ and $\epsilon_{c}$.
+%     The light gray triangle with the red boundary is the probability simplex, i.e., the space of all possible 3-place probability vectors.
+%     The darker gray area in gray boundary lines is the subspace that is compatible with the ordering that the probability of choosing the target is bigger than that of the competitor, which in turn is bigger than that of the distractor.
+%     The blue area is the subspace of probabilistic predictions the intermediate-scope model makes under any value of its parameters.
+%     The orange dot in the middle is the ``Laplace point'' of equal probability for all three options, which the model predicts for $\alpha=0$ or $\epsilon=1$.
+%     The yellow dot is the ``vanilla prediction'' for $\alpha_{c}=1$ and $\epsilon_{c}=0$.
+%   }
+%   \label{fig:prediction-range}
+% \end{figure}
 
@@
 %%% End:
+
+   \hline
+\end{tabular}

### 2023-12-01 01-code/llama_logprobs.py
add some sanity check prints

@@ -43,25 +43,30 @@ def getLogProbContinuation(
         return_tensors="pt",
     ).input_ids
     #print("input_ids prompt ", input_ids_prompt)
-    #print("input ids continuation ", input_ids_continuation)
+    print("input ids continuation ", input_ids_continuation.shape)
     # cut off the first token of the continuation, as it is SOS
     input_ids = torch.cat(
         (input_ids_prompt, input_ids_continuation[:, 1:]), 
         -1
     ).to("cuda:0") # put input on the first device
+    print("input ids shape ", input_ids.shape)
     # pass through model
     outputs = model(
         input_ids,
     )
     # transform logits to probabilities
+    print("shape of logits ", outputs.logits.shape)
+    # remove the EOS logit which we aren't interested in
     llama_output_scores = logsoftmax(
-        outputs.logits[0][:, :-1]
+        outputs.logits[0][:-1]
     )
+    print("output log probs shape ", llama_output_scores.shape)
     # retreive log probs at token ids
     # transform input_ids to a tensor of shape [n_tokens, 1] for this
     # cut off the sos token so as to get predictions for the actual token conditioned on 
     # preceding context
     input_ids_probs = input_ids[:, 1:].squeeze().unsqueeze(-1)
+    print("shape of input ids for porb retrieval ", input_ids_probs.shape)
     # retreive at correct token positions
     conditionalLogProbs = torch.gather(
         llama_output_scores, 

### 2023-12-01 01-code/llama_logprobs.py
add 13 and 7b updated results

@@ -75,7 +75,7 @@ def getLogProbContinuation(
     ).flatten()
     # slice output to only get scores of the continuation, not the context
     continuationConditionalLogProbs = conditionalLogProbs[
-        input_ids_prompt.shape[-1]:
+        (input_ids_prompt.shape[-1]-1):
     ]
     print("Shape of retrieved log probs", continuationConditionalLogProbs.shape)
     # compute continunation log prob

### 2023-12-16 04-paper/02-prior-versions/paper-LLM-probmod.tex
restructured dirs; refactored model analysis

\documentclass[fleqn]{article}

\usepackage{arxiv}

% \usepackage[style=authoryear-comp, % Citation marks as [Jef65]
%             natbib=true,      % Natbib-style cite macros \citeauthor &c.
%             hyperref=true,    % Cites in pdf are links to bib
%                               % (hyperref conf.)
%             maxnames=3,       % truncate name lists if more than 2
%                               % names appear
%             doi=false,        % no doi's
%             url=false,        % no url's
%             sortcites=false,  % do NOT sort cites in the style of the
%                               % bibliography
%             %backref=true     % insert backrefs in reference section
%             ]{biblatex}
\usepackage{mypackages}
\bibliography{~/Desktop/icloud/MyRefGlobal}
\usepackage{mycommands}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{cleveref}       % smart cross-referencing
\usepackage{lipsum}         % Can be removed after putting your text content
\usepackage{graphicx}
\usepackage{doi}

\usepackage{float}

% Statistical models based on LLM predictions
% LLMs fuel statistical models for full distributional predictions of human experimental data
% Large language models for full distributional predictions of human experimental data
% Statistical modeling with predictors from large language models
\title{Beyond the best guess: Distributional predictions from large language models for human forced-choice data}

% Here you can change the date presented in the paper title
%\date{September 9, 1985}
% Or remove it
\date{}

\newif\ifuniqueAffiliation
% Uncomment to use multiple affiliations variant of author block
\uniqueAffiliationtrue

\ifuniqueAffiliation % Standard variant of author block
  \author{ Michael Franke\thanks{Corresponding author.} \\
	Department of Linguistics\\
	University of Tübingen\\
	\texttt{mchfranke@uni-tuebingen.de} \\
	%% examples of more authors
	\And
	Polina Tsvilodub \\
	Department of Linguistics\\
	University of Tübingen\\
	\texttt{polina.tsvilodub@gmail.com} \\
	\And
	Fausto Carcassi \\
	ILLC\\
	University of Amsterdam\\
	\texttt{fausto.carcassi@gmail.com} \\
	%% \AND
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
	%% \And
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
	%% \And
	%% Coauthor \\
	%% Affiliation \\
	%% Address \\
	%% \texttt{email} \\
}
\else
% Multiple affiliations variant of author block
\usepackage{authblk}
\renewcommand\Authfont{\bfseries}
\setlength{\affilsep}{0em}
% box is needed for correct spacing with authblk
% \newbox{\orcid}\sbox{\orcid}{\includegraphics[scale=0.06]{orcid.pdf}}
\author{Michael Franke, Polina Tsvilodub, Fausto Carcassi}
\affil{Department of Linguistics\\University of Tübingen\\
\texttt{[michael.franke|polina.tsvilodub|fausto.carcassi]@uni-tuebingen.de}}
\fi

% running right header
\renewcommand{\headeright}{}
% small title under big title
\renewcommand{\undertitle}{}
\renewcommand{\shorttitle}{Statistical modeling with LLM predictors}

%%% Add PDF metadata to help others organize their library
%%% Once the PDF is generated, you can check the metadata with
%%% $ pdfinfo template.pdf
\hypersetup{
pdftitle={Statistical modeling with LLM predictors},
pdfsubject={},
pdfauthor={Michael Franke, Polina Tsvilodub, Fausto Carcassi},
pdfkeywords={statistical models, language use, large language models, Bayesian data analysis, reference games},
}

\begin{document}
\maketitle

\begin{abstract}
	{\textcolor{gray}{fill me}}
\end{abstract}

% keywords can be removed
% \keywords{First keyword \and Second keyword \and More}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\mf{would be good to add references here}

The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} enabled a new generation of powerful large language models (LLMs) \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and} which excel on standard benchmarks and promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
% especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
Recent works increasingly go beyond using LLMs based on single-run input-output behavior, and instead utilize LLMs as a part of a larger computational process.
Examples include sophisticated prompting strategies \citep[e.g.,][]{LiuLiu2022:Generated-Knowl}, or structured reasoning models \citep[e.g.,][]{CreswellShanahan2022:Selection-Infer,GaoMadaan2023:PAL:-Program-ai,ParanjapeLundberg2023:ART:-Automatic-}.
Information from LLMs is also used  to rank or numerically score options in open-ended applications, e.g., to mimic human judgements of relevance or interestingness \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
Other work, uses LLMs as part of bigger programs to build towards something more akin to explanatory cognitive models \citep[e.g.,][]{WongGrand2023:From-Word-Model}.

For all of these applications, it is crucial to understand what LLMs can or cannot reliable do.
The prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' exists, averaged over many instances of this task.
Benchmark-driven assessments are very useful to systematically study the effects of scale, e.g., in an engineering-oriented context \citep[e.g.,][]{srivastava2023-BIGbench}.
Other recent work is more psychologically oriented and asks to what extent LLM performance is ``human-like.''
LLM performance is therefore compared to human choice behavior in psychological experiments to investigate whether LLMs predict patterns of human answer behavior qualitatively \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
Going further, there is also work investigating whether LLMs can make adequate \emph{quantitative predictions}.
For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of common EEG signals like the N400 component \citep{LindborgRabovsky2021:Meaning-in-brai}.

This work seeks to extend the investigation of the human-likeness of quantitative information derived from LLMs even further, by exploring whether LLM-derived measures can feed into probabilistic models predicting the full distribution of human answers in a forced-choice tasks.
The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.

A major contribution of this work is methodological.
By embedding numerical scores from LLMs into probabilistic models for human multiple-choice data, we can use Bayesian data analysis to train, test and compare different LLM-derived predictor scores.
\mf{describe high-level results here}.

The paper is structured as follows.
Section~\ref{motivation} further motivates the desire to assess quantitative information from LLMs, and it introduces the distinction between item-level and condition-level predictions.
Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a salient probabilistic cognitive model \citep{FrankGoodman2012:Predicting-Prag}.
Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions are adequate to capture the human data at the item-level and finds that they are not, even though a condition-level predictor from the RSA model is.
Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
We find that not every way of constructing condition-level predictions from item-level scores is equally good.
\mf{grant conclusion?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Motivation: Why care for distributional predictions?}
\label{motivation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The prevalent approach to assessing LLM performance is based on benchmark-testing, using large data sets for tasks in which a designated goal answer exists.
Tasks often contain multiple-choice options, and accuracy for a task is determined with a ``winner-takes-all'' (WTA) strategy \citep[e.g.,][]{srivastava2023-BIGbench}.
Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
% For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
Among the choice options is a designated \emph{target option} $y_{ki^{*}}$ that is assumed to be the ``true'' or to-be-selected goal answer.
The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
}
%
\begin{align*}
  S\left( y_{ki}, x_{k} \right) = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right) =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%
The WTA approach considers item $I_{k}$ to be answered correctly if the target option $y_{ki^{*}}$ is the option that maximizes the item-level score.
The WTA-based accuracy of the LLM on the given task is the proportion of items which are answered correctly.

\mf{the following paragraphs should be shortened and sharpened}
While generally a pragmatic and useful approach, there are several reasons why going beyond the best guess can be beneficial.
Consider first a high-level conceptual argument.
If task performance is categorical (in the most extreme case: binary), somewhere along the path from a numerical item-level score to accuracy a discontinuity has to be introduced.
Every such discontinuity bottleneck entails loss of information.
As this information might be useful, discontinuity should ideally happen as late as possible; or even better: not at all, unless it is irrelevant.
Consequently, the question becomes: when is the additional information inherent in the scores (not) relevant for assessing LLM performance?

The WTA approach implicitly assumes that item-level choices are resolved by a greedy-like choice of the best alternative.
It is therefore a good measure of accuracy when the LLM's main use cases are of a greedy-like nature as well.
When applications rely on a less extreme sampling based strategy, the WTA approach can be generalized to softmax sampling.
The probability of choosing option $y_{ki}$ for item $I_{k}$ is:
%
\begin{align*}
P_{\text{choose}}(y_{ki}) \propto \expo \left [ \alpha \ S \left( y_{ki}, x_{k} \right) \right ]\,,
\end{align*}
%
and the accuracy is the average probability of choosing the designated target option $y_{ki^{*}}$.
For $\alpha \rightarrow \infty$, this softmax strategy converges to the WTA strategy.
But in general, these two approaches can yield even categorically different predictions.

Imagine that there are two options, and that the target option's score is a small $\epsilon$ better in 80\% of the task's items, and otherwise worse.
The WTA-based accuracy is 0.8.
This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
As only the best option will be chosen, that information is irrelevant.
But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases. This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
This is a categorical shift from predominantly goal answer to predominantly non-goal answer.
The argument holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.

Differences in accuracy scores between a WTA-based or a more general softmax-based strategy depend on the variation in item-level scores, in particular the relation between score-ordering and score-differences.
(The example in the previous paragraph was set up deliberately to have one score-ordering for very small score-differences, and another for huge score-differences.)
This raises an important general question, namely how to deal with item-level variation in a system's predictions in the first place.
Data from forced-choice tasks with human participants will likely show item-level variation as well, which is why hierarchical modeling accommodating item-level differences is recommended in statistical analyses \citep[e.g.,][]{Jaeger2008:Categorical-dat,barr2013}.
But whether item-level variation in LLM scores aligns with differences in human choice rates for each item is an empirical question that has to be addressed by explicit modeling and comparison; an exercise that this paper starts to take on.
On \emph{a priori} grounds, the WTA-based strategy may be deemed implausible as a predictor of human-level choice behavior for a simple item: it is unlikely that, for each item, there is a single choice option that (almost) \emph{all} participants will choose.

To the extent that machine predictions show extreme, unsystematic, erratic item-level variation which is quite unlike that of human participants in comparable experiments, an alternative strategy suggests itself, namely to \emph{not} judge a system by its average item-level predictions, but based on its task- or condition-level predictions obtained from different ways of aggregating over item-level variation.
Section~\ref{llm-predictions-for-reference-games} therefore introduces and compares different ways of deriving condition-level predictions from (variable) item-level scores.
Properly constructed condition-level predictions might be more aligned with average human choice rates, despite the fact that each complex system (machine and human) shows quite different underlying item-level variation.
In other words, we should ask whether modern LLMs can predict aggregate human data by \emph{some} process of aggregating its item-level scores, because it might turn out that predicting aggregates is the best the LLMs can do (in particular cases).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experiment: Reference games}
\label{experiment-reference-games}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}
  \centering

  \includegraphics[width = 0.8\textwidth]{00-pics/reference-game.png}

  \caption{Structure of a reference game with human participants. Each trial consists of a set of objects, the so-called context. In production trials, participants choose a single word to describe a target object from the context. In interpretation trials, an object is selected as the likely object a trigger word referring to.}
  \label{fig:ref-game}
\end{figure}

Reference games are an established, well-understood and austere experimental paradigm to test human decision making in abstract communicative tasks \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,SikosVenhuizen2021:Reevaluating-pr}.
A reference game consists of two players, a speaker and an interpreter, who jointly observe a set of objects, usually referred to as context (see Figure~\ref{fig:ref-game}).
In the \textbf{production condition}, the speaker is assigned a \emph{target object} from the context set which they have to describe to the interpreter.
In the \textbf{interpretation condition}, the interpreter observes a description, here called \emph{trigger word}, and chooses one of the objects from the context set.
The goal of the game is, for the speaker, to choose a description that enables the interpreter to choose the target object; and, for the interpreter, to guess correctly which object the speaker had in mind.

The example in Figure~\ref{fig:ref-game} is a standard case, which we will use throughout, where choices are informative about the pragmatic reasoning that decision makers engage in.
In this example, there are two features that differ across three objects (here shape and color).
One object shares both its color and shape with one other object, while the two other objects have one unique feature (e.g., being the only circle, or the only green object).
%
In a critical production trial, the target object to describe is one of the two objects with a unique feature.
The speaker has four words to choose from.
The \textbf{target utterance} is the word which uniquely describes the target object.
The \textbf{distractor utterance} is the word that is true of the target object, but also true of another object.
The other utterances, both of which are false of the target are \textbf{competitor utterance}.
%
In a critical interpretation trial, the trigger word is one that is true of two of the three objects.
If participants engage in pragmatic thought, they might reason that \emph{if} the speaker had wanted to refer to one of the two objects of which the trigger word is true (blue square and blue circle in Figure~\ref{fig:ref-game}), the speaker could have used a more informative word for exactly one of those two objects (``circle''), so they are more likely to refer to the \textbf{target object} (the blue square in Figure~\ref{fig:ref-game}).
The \textbf{competitor object} is the other object of which the trigger word is true.
The \textbf{distractor object} is the object of which the trigger word is false.

We replicated a simple reference game with human participants in which each trial instantiated the structure of the example shown in Figure~\ref{fig:ref-game}.
While previous reference games with human participants used pictorial representations of objects, and sometimes even pictorial representations of messages, we implemented a text-only version in order to be able to compare the predictions of LLMs for the exact same stimuli.
The experiment was realized as an online task using \texttt{magpie} \citep{FrankeJi:magpie:-Minimal}.\footnote{
  The code for the experiment can be found at \href{https://github.com/magpie-ea/magpie3-text-refgame}{https://github.com/magpie-ea/magpie3-text-refgame}, and a live version of the experiment can be tested at \href{https://magpie-ea.github.io/magpie3-text-refgame/}{https://magpie-ea.github.io/magpie3-text-refgame/}.
}

\subsection{Participants}
\label{participants}

A total of 302 participants were recruited via Prolific for monetary
compensation (\textsterling0.45, corresponding to roughly \textsterling 15.40 per hour).
All participants self-identified as native speakers of English.

\subsection{Materials \& design}
\label{materials-design}

We created 100 different items as stimulus material via a stochastic process.
Each item is a different textual description of a reference game with the same logical structure as the example from Figure~\ref{fig:ref-game}.
For each item, the context consisted of three objects.
Objects are defined by a triple of properties, namely a color, a shape and a texture.
For each property, there were four possible values, e.g., blue, green, red, and orange for color.
The sampled items differed in terms of the properties of the objects in the context set, and in terms of the order in which the objects and expression alternatives were presented in the text.
Figures~\ref{fig:refgame-screenshot-production} and \ref{fig:refgame-screenshot-interpretation} from Appendix~\ref{sec:scre-from-online} show example screenshots from the experiment.

\subsection{Procedure}
\label{procedure}

For each participant the experiment sampled four different items.
Participants first played two of these in the production condition, then the other two in the interpretation condition.

\subsection{Results}\label{results}

The overall distribution of choices that correspond to the target, competitor, and distractor states is shown in Figure~\ref{fig:refgame-counts}.\footnote{
  The production condition actually has two distractor choices.
  Here and in the following, these are lumped together as a single category, also when modeling random errors in later models.
  This is a simplification but does not change anything of substance.}
It is interesting that the distractor options were chosen rather often.
We also see that the number of target choices is higher in the production condition than in the interpretation condition.
This is in line with previous experimental results on human reference games.
% \citep{QingFranke2013:Variations-on-a}.
For example, in previous forced-choice reference games with human participants with pictorial presentations of objects, \citet{QingFranke2013:Variations-on-a} observed the following proportions: $\tuple{0.882, 0.118, 0}$ in the production and $\tuple{0.492, 0.506, 0.003}$ in the interpretation condition (for 288 observations in each condition).

\begin{figure}[t]
  \centering

    \includegraphics[width=0.9\linewidth]{00-pics/PPC-alpha-eps-model.pdf}

    \caption{Counts of choices from reference games with human participants (colored bars), with summary statistics from the posterior predictive distribution of four models (shapes and error bars).
      Shapes show the mean of the posterior predictive distributions of the RSA model and three further models derived from item-level LLM-supplied scores.
      Error-bars show corresponding 95\% credible intervals of the posterior predictive.
    }
  \label{fig:refgame-counts}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Model predictions from probabilistic pragmatics}
\label{sec:model-pred-from}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Data from reference games with human participants have been variously analyzed with probabilistic models using inspiration from behavioral game theory \citep[e.g.,][]{DegenFranke2013:Cost-Based-Prag}, probabilistic Bayesian modeling \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag} or other forms of probabilistic modeling \citep[e.g.,][]{GattGompel2013:Are-we-Bayesian}.
Common to these approaches is that they derive or define, based on some explicit conceptual motivation, a parameterized stochastic speaker policy, $P_{S}(u \mid s; \theta_{S})$, modulated by parameters $\theta_{S}$, for a speaker's choice of expression or utterance $u$ given a referent or state $s$, which the speaker wants to communicate;
and a stochastic listener policy, $P_{L}(s \mid u; \theta_{L})$, capturing the probability of choosing a referent $s$ for utterance $u$.

As a concrete example, we introduce the Rational Speech Act (RSA) model first described in this form by \citet{FrankGoodman2012:Predicting-Prag} \citep[for overview see][]{FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu,StevensBenz2018:Game-Theoretic-,ScontrasTessler2021:A-practical-int,Degen2023:The-Rational-Sp}.
The RSA model defines pragmatic reasoning as a sequence of iterated (soft-)optmization of policies, grounding out in literal interpretation.
If $\mathfrak{L}(s,u) \mapsto \set{0,1}$ is a semantic meaning function mapping each pair of state $s$ and utterance $u$ to a (binary) truth-value, and if $P_{s}$ is a prior over states, a literal listener policy is defined as:
%
\begin{align*}
 P_{L_{0}}(s \mid u) \propto \mathfrak{L}(s,u) P(s)\,.
\end{align*}
%
The pragmatic speaker soft-optimizes the choice of utterance to minimize the literal listener's surprisal for the to-be-communicated state:
%
\begin{align*}
  P_{S}(u_{i} \mid s, \alpha) & \propto \expo \left [ \log P_{L_{0}}(s \mid u_{i}) \right ] \,.
\end{align*}
%
Finally, the pragmatic listener is defined as the policy resulting from applying Bayes rule, solving the inverse-problem for the previously defined speaker policy:
%
\begin{align*}
  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P(s) \,.
\end{align*}

\begin{figure}[t]
  \centering
  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
  \caption{
    Example of predictions from the RSA model (with $\alpha=1$).
    The semantic meaning is shown as a matrix of binary truth-values.
    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents), assume $\alpha=1$ and are shown as row-stochastic matrices.
  }
  \label{fig:RSA-example}
\end{figure}

Figure~\ref{fig:RSA-example} gives example calculations (assuming flat priors and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
For $\alpha=1$, the model predicts that the probabilities of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{0.6, 0.4, 0}$ for the interpretation condition.
Increasing $\alpha$ will increase the odds of target over competitor choices.
Yet, the model also predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
It is therefore common to include a random error probability for each choice \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}, such as via Laplace smoothing: if $P_{r}(R_{l}, C, \alpha_{c})$ is the RSA model's probabilistic prediction for response category $R_{l}$ (target, competitor, or distractor) for condition $C$ (production or interpretation) and condition-specific optimality $\alpha_{c}$, the random-error smoothed prediction is:
\begin{align*}
  P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c}) \propto P_{r}(R_{l}, C; \alpha_{c}) +  \frac{\epsilon_{c}}{3}\,,
\end{align*}
where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
The result is a four-parameter model, one pair of parameters per condition, which provides a likelihood function for the categorical choice data and so can be fitted to the data and compared against other probabilistic models providing a likelihood function for the same data.

Parameterized predictions like $P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c})$ can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{fig:posterior-stats}).
\mf{include information on MCMC, samples, warm-up etc.; follow Kruschke recipe}

To assess goodness-of-fit, we can use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
\mf{maybe insert formal definition of post-pred distribution?}
As a minimal bar, we would require a model trained on observed data $D_{\text{obs}}$ to not be surprised by $D_{\text{obs}}$.
Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models to be introduced later).
We see that the RSA model passes this ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-} for both conditions.
To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
\mf{maybe insert formal definition of BPPPV?}
These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
The results from Table~\ref{tab:Bppp-values} show that the RSA model passes this numerical test of model adequacy.

The following sections apply these test of Bayesian model criticism also to models built around predictor values from LLMs.
Section~\ref{sec:item-level-pred} first looks at the data from each item individually, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Item-level predictions from LLMs}
\label{sec:item-level-pred}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Figure~\ref{fig:refgame-counts} shows the condition-level counts for choices for all of the 100 different items combined.
An LLM, on the other hand, first and foremost makes predictions about each item.
The items of the reference game experiment differ in the task-relevant features (color, shape, texture), as well as the order of presentation of states and words.
We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelvant material \citep{DegenHawkins2020:When-redundancy} \mf{more refs on (color) over-specification}; and machines may be susceptible to the presentation of the order of choice options. \mf{references from the NLP literature on effects of prompt variation?}
The main question to be addressed in this section is whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.

For a set-up in which the task descriptions and choice options are all text-based, predictions of LLMs at the item-level can be derived as a function of the (log) probabilities assigned to the continuations corresponding to each choice option after processing the task description.
Let $C$ be the experimental condition of interest (here: the production or interpretation condition).
Let \(\set{I_1, \dots I_m}\) be the multi-set of items that occurred in the human experiment for condition $C$.\footnote{
  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.
}
Each item $I_{k}$ consists of a task description $T_{k}$ and $l$ choice options $O_{k1}, \dots, O_{kl}$, which are treated as textual continuations of the task description (see Appendix~\ref{sec:examples-items-llm} for an example).
For each item $I_{k}$, let $F(R_{l}, I_{k})$ be the choice option $O_{ki}$ that corresponds to \emph{response type} $R_{l}$ (here: target, competitor, distractor).\footnote{In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
The \emph{item-level score} $S(O_{{ki}}, I_{k})$ for option $O_{ki} = w_{ki1}, \dots w_{kin}$ of item $I_{k}$ is defined as the log-probability of continuation $O_{ki}$ after prompt $T_{k}$:\footnote{Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
%
\begin{align*}
S(O_{ki}, I_{k}) =  \sum_{j=1}^n \log P_{\text{LLM}} \left(w_{kij} \mid T_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%
The item-level score of response type $R_{l}$ is $S(R_{l}, I_{k}) = S(F(R_{l}, I_{k}))$.

Using item-level scores from GPT-3.5, we fit a model to the human data at the item-level.
The data to be explained is the set of all triples of response type counts $\tuple{t_{k}, c_{k}, d_{k}}$ (target, competitor, distractor), separately for each item $I_{k}$.
The model's predictions for the data from item $I_{k}$ from condition $C$ (production or interpretation) are:
%
\begin{align*}
  P_{\text{item}}(R_{l}, C, I_{k}, \alpha_{c}, \epsilon_{c}) \propto \frac{\expo \left( \alpha_{c} \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha_{c} \ S(R_{l'}, I_{k}) \right )} + \epsilon_{c}\,.
\end{align*}
%
In words, we use a single pair of parameters $\alpha_{c}$ and $\epsilon_{c}$, one for each condition, take the LLM's item-level scores for each item, and derive the response probabilities in the exact same way as before.
% Notice that item-level scores are the same for all three types of models introduced and compared in the previous section, so that this by-item analysis applies to the raw LLM prediction irrespective of down-stream aggregation.
Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.

\begin{figure}[t]
  \centering

  \includegraphics[width = 0.9\linewidth]{00-pics/item-combined-obs-pred.pdf}

  \caption{
    Item-level prediction-observation plot.
    Each dot represents an item.
    The $x$-coordinate represents the mean of the posterior prediction for the target choice probability for the given item.
    The $y$-coordinate represents the observed proportion of target choices for that item.
    The gray line (identity function) is the ideal prediction.
  }
  \label{fig:item-level-obs-pred}
\end{figure}

Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
Concretely, the plot shows, for each item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
So, the model itself predicts systematic variability at the item level.
The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.

Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (0.0001 for production, 0 for interpretation), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
In contrast, using the global, item-independent RSA model predictions to fit the item-level data, we obtain estimates of posterior predictive $p$-values that do not discredit the model ($0.338$ for production, $0.252$ for interpretation).
These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
Put more strongly, a model that seeks to predict what human participants choose on a by-item level, based on the most obvious item-level score, is ruled out by experimental data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Condition-level predictions}
\label{llm-predictions-for-reference-games}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section explores different ways of deriving a probabilistic likelihood function for the human data from an LLM at the aggregate, condition-level.
Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
In the following, we first describe the different options of deriving LLM-based predictor terms.
We then compare all probabilistic models based on their adequacy of explaining the human choice data.

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{00-pics/measures-overview.png}
  \caption{
    Schematic overview over three different approaches of obtaining condition-level predictions by aggregating over item-level scores.
    The basic item-level score is average next-word log probability.
    This can be taken as-is into averaging (narrow-scope aggregation), or first transformed from log- into probability-space, either with optimization before (wide scope) or after averaging (intermediate scope).
    The notation for the condition-level options uses pseudo-code to more clearly reflect the relevant operator scope differences (where triangles represent the pipe operator).
  }
  \label{fig:measures-overview}
\end{figure}

Condition-level predictions are a probability distribution over response types, obtained from item-level scores $S(R_{l}, I_{k})$ by averaging over all items belonging to the relevant condition.
To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
%
\begin{align*}
 \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
\end{align*}
%
The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
%
\begin{align*}
  \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
  \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
\end{align*}
%
This means that there are three conceivable scope-sites for aggregating item-level information as shown in Figure~\ref{fig:measures-overview}.
Narrow-scope aggregation first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities:
%
\begin{align*}
  % & \textbf{narrow-scope aggregation} \\
  & P_{n}(R_{l}, C ; \alpha) \propto \expo \left [  \alpha \ \frac{1}{m} \ \sum_{i = k}^{m} S(R_{l}, I_{k})  \right ]
    \tag*{\textcolor{gray}{[narrow-scope aggregation]}}
\end{align*}
%
Wide-scope aggregation, first transposes scores into probabilities, scales them and only aggregates over items last.
\begin{align*}
  % & \textbf{wide-scope aggregation} \\
  & P_{w}(R_{l}, C ; \alpha) \propto \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( \alpha \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha \ S(R_{l'}, I_{k}) \right )}
    \tag*{\textcolor{gray}{[wide-scope aggregation]}}
\end{align*}
%
Intermediate-scope aggregation performs item-level averaging after mapping scores onto probabilities (using softmax with $\alpha=1$), but before scaling (with a power-law transformation with variable $\alpha$):
\begin{align*}
  % & \textbf{intermediate-scope aggregation} \\
  & P_{i}(R_{l}, C ; \alpha) \propto  \left [ \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( S(R_{l'}, I_{k}) \right )} \right ]^{\alpha}
    \tag*{\textcolor{gray}{[intermediate-scope aggregation]}}
\end{align*}

All three condition-level predictors coincide when there is only one item, of course.
For more items, wide-scope and intermediate-scope aggregation coincide when $\alpha = 1$.
In all other cases, the predictors are not guaranteed to be identical.
Conceptually, the main difference is what each measure considers to be the basic item-level unit to aggregate over.
Narrow-scope aggregation considers raw scores, which are \emph{not} probabilistic predictions at the item level.
Wide-scope aggregation considers $\alpha$-optimized probabilities at the item level, which is compatible with a procedure of sampling, via softmax decoding, item-level answer options.
Wide-scope aggregation contains the ``winner-takes-all'' procedure for accuracy scoring in common benchmark tests for $\alpha \rightarrow \infty$.
Finally, like wide-scope aggregation, intermediate-scope aggregation is also compatible with a sampling based picture, but would use pure decoding (without $\alpha$-optimization) at the item level and reserves the scaling of probability distributions to a stage \emph{after} aggregation.

\begin{table}[t]
  \centering

  % \includegraphics[width=0.9\linewidth]{00-pics/posterior-stats.pdf}

  \begin{tabular}{llllcr}
    \toprule \addlinespace[1ex]
    model        & condition      & Parameter  & $|$95\% & mean & 95\%$|$ \\
    \midrule  \addlinespace[1ex]
    narrow       & production     & $\alpha$   & 0.19    & 0.22 & 0.26 \\
                 &                & $\epsilon$ & 0.00    & 0.04 & 0.11 \\
                 & interpretation & $\alpha$   & 0.15    & 0.18 & 0.20 \\
                 &                & $\epsilon$ & 0.00    & 0.02 & 0.06 \\ \addlinespace[0.75ex]
    intermediate & production     & $\alpha$   & 0.86    & 1.04 & 1.22 \\
                 &                & $\epsilon$ & 0.08    & 0.13 & 0.18 \\
                 & interpretation & $\alpha$   & 0.36    & 0.48 & 0.60 \\
                 &                & $\epsilon$ & 0.00    & 0.05 & 0.12 \\ \addlinespace[0.75ex]
    wide         & production     & $\alpha$   & 0.22    & 2.69 & 8.79 \\
                 &                & $\epsilon$ & 0.08    & 0.13 & 0.19 \\
                 & interpretation & $\alpha$   & 0.19    & 1.08 & 6.03 \\
                 &                & $\epsilon$ & 0.00    & 0.06 & 0.18 \\ \addlinespace[0.75ex]
    RSA          & production     & $\alpha$   & 2.59    & 3.16 & 3.71 \\
                 &                & $\epsilon$ & 0.08    & 0.13 & 0.18 \\
                 & interpretation & $\alpha$   & 0.24    & 0.66 & 1.07 \\
                 &                & $\epsilon$ & 0.10    & 0.15 & 0.20 \\ \addlinespace[0.25ex]
    \bottomrule \\
  \end{tabular}

  \caption{
    Summary statistics of posterior samples.
    Number indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
  }
  \label{fig:posterior-stats}
\end{table}

The three condition-level predictions yield three different probabilistic models, each with two pairs of condition-dependent parameters $\alpha_{c}$, and $\epsilon_{c}$.
Using priors and Bayesian posterior estimation as previously described for the RSA (see Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
The usual Bayesian summary statistics for posteriors over model parameters are shown in Table~\ref{fig:posterior-stats}.
What is noteworthy about these estimates is that the $\alpha_{c}$ parameters for the wide-scope model have a very large credible range.
This is because, already for values of $\alpha_{c} = 1$, the item-level predictions are virtually ``winner-takes-all'' choices, selecting the option with the maximum score with probability almost 1 for almost all items.
This implies that, by studying the posterior predictive distribution of the wide-scope model, we are virtually testing condition-level predictions from the standard WTA-strategy used for benchmark testing (see Section~\ref{motivation}).

Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
We find that only the theoretical model (RSA) and the intermediate-scope model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
Consequently, the results from Table~\ref{tab:Bppp-values} suggest that the narrow-scope model fails on the interpretation data, and is at most borderline compatible with the production data; that the wide-scope model is able to reproduce the production data, but fails on the interpretation data; and that only the intermediate-scope model is able to fully recover the data from both conditions.

\begin{table}[ht]
\centering

% \begin{tabular}{lrr}
%   \toprule
%   model        & production & interpretation \\ \midrule
%   narrow       & 0.05       & 0.00 \\
%   wide         & 0.51       & 0.00 \\
%   intermediate & 0.49       & 0.26 \\
%   RSA          & 0.49       & 0.52 \\
%   \bottomrule \\
% \end{tabular}

\begin{tabular}{lrrrr}
  \toprule
  & narrow & wide & intermediate & RSA \\ \midrule
  production & 0.05 & 0.51 & 0.49 & 0.49 \\
  interpretation & 0.00 & 0.00 & 0.26 & 0.52 \\ \bottomrule \\
\end{tabular}

\caption{Sample-based estimates of Bayesian posterior predictive $p$-values for each model and each condition, based on likelihood of the observed data as test statistic.}
\label{tab:Bppp-values}
\end{table}

These results tell us that not all ways of deriving condition-level predictions by averaging over item-level variation are equally good.
Some approaches clearly fail basic checks for statistical goodness-of-fit.
On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the intermediate-scope model, which is able to recover the patterns in the data.
In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.\footnote{
  A potential worry is that the intermediate-scope model is trivial in the sense that it could have predicted \emph{any} data observation.
  Appendix~\ref{sec:range-prior-pred} shows that this is not the case.
}
This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

It is not entirely ludicrous to use numerical predictions from LLMs are
part of predictive probabilistic models. But we have to be careful.
Ideally, we average out over low-level variation, such as from order of
presentation or similar ``nuisance,'' at least as long as we do not
understand what causes this variation in the predictions of models and
further research that investigates when exactly this variation accords
with empirically observed patterns.

This also implies that we should likely \emph{not} (yet) aspire to use
LLMs are models or individual- or item-level predictors.

( {TODO: rethink conclusion})

{to be continued}

\newpage
\appendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Screenshots from the online experiment with human participants}
\label{sec:scre-from-online}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Figure~\ref{fig:refgame-screenshot-production} shows a trial from the production condition, Figure~\ref{fig:refgame-screenshot-interpretation} one for the interpretation condition of the online experiment.

\begin{figure}[H]
  \centering
  \includegraphics[width = 0.8\textwidth]{00-pics/refgame-production.png}

  \caption{Screen shot from a production trial of the online experiment.}
  \label{fig:refgame-screenshot-production}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width = 0.8\textwidth]{00-pics/refgame-interpretation.png}

  \caption{Screen shot from an interpretation trial of the online experiment.}
  \label{fig:refgame-screenshot-interpretation}
\end{figure}

\section{Example item for the LLM experiment}
\label{sec:examples-items-llm}

The text-based input for the LLM predictions mirrors the text in the human experiment, except that the LLM input also lists the set of all available choice options (which for the human experiment is unnecessary since this information is given by the buttons for the forced-choice selection).
For example, the task description $T_{k}$ for the item that corresponds to the production trial shown in Figure~\ref{fig:refgame-screenshot-production} is shown below (the actual input has no line breaks in the first paragraph):

\begin{verbatim}
Your task is to play a conversation game. There are three objects that
you and your friend can see. You have to choose a single word to identify
one of the three objects for your friend.

The three objects are:

a blue triangle with spades
a blue circle with stripes
a blue triangle with stripes

Your task is to make your friend pick out the following target object:

the blue circle with stripes

Which of the following words would you choose:

spades
stripes
circle
triangle

Your answer:

I would choose the word
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Range of prior predictions of intermediate-scope model}
\label{sec:range-prior-pred}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

It may seem that, given the freedom to adjust optimization $\alpha$ and random guessing rate $\epsilon$, some models might be able to explain \emph{any} data observation.
The intermediate-scope model, which was not discredited by model criticism in terms of recovery-based posterior predictive checks, is not trivial in this sense.
Figure~\ref{fig:prediction-range} shows the range of predictions that the intermediate-scope model is, in principle, capable of making (for the whole possible range of \(\alpha\) and \(\epsilon\) values, disregarding Bayesian priors).
There are distributions over response types which the model does not explain predict \emph{ex ante}.
For example, the model does not predict cases where the number of competitor choices is almost as high as the number of target choice, as would be the case if participants would consistently \emph{not} engage in pragmatic reasoning.

\begin{figure}[t]
  \centering

  \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-prod.png}
  %
  \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-inter.png}

  \caption{
    Range of predictions the intermediate-scope model makes for any pair of parameter values $\alpha_{c}$ and $\epsilon_{c}$.
    The light gray triangle with the red boundary is the probability simplex, i.e., the space of all possible 3-place probability vectors.
    The darker gray area in gray boundary lines is the subspace that is compatible with the ordering that the probability of choosing the target is bigger than that of the competitor, which in turn is bigger than that of the distractor.
    The blue area is the subspace of probabilistic predictions the intermediate-scope model makes under any value of its parameters.
    The orange dot in the middle is the ``Laplace point'' of equal probability for all three options, which the model predicts for $\alpha=0$ or $\epsilon=1$.
    The yellow dot is the ``vanilla prediction'' for $\alpha_{c}=1$ and $\epsilon_{c}=0$.
  }
  \label{fig:prediction-range}
\end{figure}

\printbibliography[heading=bibintoc]

\end{document}

### 2023-12-16 01-code/data-analysis.r

@@ -1,86 +1,0 @@
-library(tidyverse)
-library(tidyboot)
-library(brms)
-
-d <- read_csv('results.csv')
-
-d |> select(starts_with("prob")) |> 
-  pivot_longer(everything()) |> 
-  group_by(name) |> 
-  tidyboot::tidyboot_mean(value)
-
-
-# sanity check if randomization worked
-
-table(d$trigger_feature)
-table(d$nuisance_feature)
-table(d$trigger_word)
-
-# effects of nuisance feature or target feature
-
-# there do seem to be differences based on the nuisance feature
-d |> group_by(nuisance_feature) |> 
-  tidyboot::tidyboot_mean(prob_interpretation_target)
-
-fit_nuisance_feature <- brms::brm(
-  formula = prob_interpretation_target ~ nuisance_feature, 
-  family = brms::brmsfamily("beta"), 
-  data = d)
-
-d |> ggplot(aes(x = prob_interpretation_target)) +
-  facet_grid(nuisance_feature ~.) +
-  geom_density(alpha = 0.3) +
-  geom_rug()
-
-# there might be trigger-feature effects!
-# interpretation seems to be better for 'texture' trigger words, than for 'shape' than for 'color'
-d |> group_by(trigger_feature) |> 
-  tidyboot::tidyboot_mean(prob_interpretation_target)
-
-fit_trigger_feature <- brms::brm(
-  formula = prob_interpretation_target ~ trigger_feature, 
-  family = brms::brmsfamily("beta"), 
-  data = d)
-
-faintr::compare_groups(
-  fit_trigger_feature, 
-  higher = trigger_feature == "texture",
-  lower = trigger_feature == "shape",
-)
-
-d |> ggplot(aes(x = prob_interpretation_target)) +
-  facet_grid(trigger_feature ~.) +
-  geom_density(alpha = 0.3) +
-  geom_rug()
-
-# positional effects
-
-d <- d |> 
-  mutate(interpretation_indices = str_c(
-    "[",
-    interpretation_index_target,
-    ",",
-    interpretation_index_distractor,
-    "]"
-  ))
-
-table(d$interpretation_indices)
-
-d |> group_by(interpretation_index_target, interpretation_index_competitor) |> 
-  tidyboot::tidyboot_mean(prob_interpretation_target)
-  
-d |> ggplot(aes(x = interpretation_index_target, y = prob_interpretation_target)) +
-  geom_jitter()
-  
[…]

### 2023-12-16 02-data-analysis/02-data-preparation.r

@@ -1,142 +1,0 @@
-source('00-premable.r')
-
-d_raw <- read_csv('01-data-raw.csv') |> 
-  select(- c(
-    "experiment_duration", "experiment_end_time", "experiment_start_time",
-    "prolific_pid", "prolific_session_id", "prolific_study_id"  )) |> 
-  rename(item = trial) |> 
-  mutate(item = factor(item, levels = 0:99))
-
-d_sim <- read_csv('01-simulation-results.csv') |> 
-  rename(item = trial) |> 
-  mutate(item = factor(item, levels = 0:99))
-
-d_join <- left_join(d_raw, d_sim, by = "item") 
-  
-d_join$response_type = map_chr(1:nrow(d_join), function(i) {
-  condition <-  d_join[i,] |> pull(condition)
-  response  <-  d_join[i,] |> pull(response)
-  if (condition == "production") {
-    if (response == d_join[i,] |> pull(production_target)) {
-      return("target")
-    } else {
-      if (response == d_join[i,] |> pull(production_competitor)) {
-        return("competitor")
-      } 
-      else {
-        return("distractor")
-      }  
-    }
-  }
-  if (condition == "interpretation") {
-    if (response |> stringr::str_sub(3) == d_join[i,] |> pull(interpretation_target)) {
-      return("target")
-    } else {
-      if (response |> stringr::str_sub(3) == d_join[i,] |> pull(interpretation_competitor)) {
-        return("competitor")
-      } else {
-        return("distractor")
-      }
-    }
-  }
-})
-
-d_join$prediction_type = map_chr(1:nrow(d_join), function(i) {
-  condition <-  d_join[i,] |> pull(condition)
-  string_match = ifelse (condition == "production", 'scores_production', 'scores_interpretation')
-  scores <-  d_join[i,] |> select(starts_with(string_match)) |> as.double()
-  return(c("target", "competitor", "distractor", "distractor")[which(scores == max(scores))])
-})
-
-d <- d_join |> 
-  rename(
-    prediction = prediction_type,
-    response_raw = response,
-    response = response_type
-  ) 
-
-## add position information string
-
-make_position_string <- function(target_index, competitor_index, condition = "interpretation") {
-  position <- c('d', 'd', 'd', 'd')
-  position[target_index] <- 't'
-  position[competitor_index] <- 'c'
[…]

### 2023-12-16 02-data-analysis/03-data-analysis.R

@@ -1,751 +1,0 @@
-# dependencies etc.
-source("00-premable.r")
-rerun = T
-source('00-stan-fit-helpers.R')
-
-#######################################################
-## read and prep data, get counts
-#######################################################
-
-tcd <- c("target", "competitor", "distractor")
-
-d <- read_csv('02-data-prepped.csv') |> 
-  mutate(
-    response   = factor(response  , levels = tcd),
-    prediction = factor(prediction, levels = tcd),
-    condition  = factor(condition,  levels = c("production", "interpretation"))
-  ) 
-
-d_counts <- d |> 
-  count(condition, response) |> 
-  group_by(condition) |> 
-  mutate(total = sum(n)) |> 
-  ungroup() |> 
-  mutate(proportion = n / total)
-
-#######################################################
-## plot empirical counts
-#######################################################
-
-refgame_counts <- d_counts |> 
-  ggplot(aes(x = response, y = n, fill = response)) +
-  geom_col(position = "dodge") +
-  facet_grid(condition ~ .) +
-  ylab("") + xlab("") +
-  theme(legend.position="none") +
-  ylim(c(0,604)) +
-  ggtitle("human data") +
-  theme(axis.text.x = element_text(angle = 25, vjust = 1, hjust=1)) +
-  theme(strip.text = element_text(size = 12)) + 
-  theme(plot.title = element_text(size = 12))
-
-refgame_counts
-
-ggsave(plot = refgame_counts, 
-       filename = "../03-paper/00-pics/refgame-counts.pdf", 
-       height = 4, width = 3, scale = 1.1)
-
-#######################################################
-## prepare model predictions (-> move to prep script?)
-#######################################################
-
-softmax_row <- function(matrix_of_scores, alpha = 1) {
-  exp_matrix <- exp(alpha * matrix_of_scores)
-  row_sums <- rowSums(exp_matrix)
-  softmax_matrix <- exp_matrix / row_sums
-  return(softmax_matrix)
-}
-
-wta_row <- function(matrix_of_scores) {
-  result_matrix <- matrix(0, nrow = nrow(matrix_of_scores), ncol = ncol(matrix_of_scores))
-  for (i in 1:nrow(matrix_of_scores)) {
-    max_val <- max(matrix_of_scores[i, ])
[…]

### 2023-12-16 03-data-analysis/03-data-analysis.R

@@ -1,0 +1,806 @@
+# dependencies etc.
+source("00-premable.r")
+rerun = T
+source('00-stan-fit-helpers.R')
+
+#######################################################
+## read and prep data, get counts, massage human data
+#######################################################
+
+tcd <- c("target", "competitor", "distractor")
+
+d <- read_csv('02-prepped-data/data-prepped-GPT.csv') |> 
+    mutate(
+      response   = factor(response  , levels = tcd),
+      prediction = factor(prediction, levels = tcd),
+      condition  = factor(condition,  levels = c("production", "interpretation"))
+    )  
+
+# raw counts
+d_counts <- d |> 
+  count(condition, response) |> 
+  group_by(condition) |> 
+  mutate(total = sum(n)) |> 
+  ungroup() |> 
+  mutate(proportion = n / total)
+
+
+# production - global 
+d_production <- d |> 
+  filter(condition == "production") |> 
+  mutate(response = case_when(response == "distractor1" ~ "distractor",
+                              response == "distractor2" ~ "distractor",
+                              TRUE ~ response)) |> 
+  pull(response) |> 
+  match(c("target", "competitor", "distractor"))
+d_prod_global <- matrix(table(d_production) |> as.integer(), nrow = 1)
+
+# production - item-level
+d_prod_item <- d |> 
+  filter(condition == "production") |> 
+  mutate(response = case_when(response == "distractor1" ~ "distractor",
+                              response == "distractor2" ~ "distractor",
+                              TRUE ~ response)) |> 
+  group_by(item) |> 
+  count(response) |> 
+  pivot_wider(id_cols = item, names_from = response, values_from = n) |> 
+  ungroup() |> 
+  arrange(item) |> 
+  select(-item) |> 
+  as.matrix()
+d_prod_item[is.na(d_prod_item)] = 0
+d_prod_item <- d_prod_item[,c(3,1,2)]
+
+# production - global 
+d_interpretation <- d |> 
+  filter(condition == "interpretation") |> 
+  pull(response) |> 
+  match(c("target", "competitor", "distractor"))
[…]

### 2023-12-17 01-code/run_llama.sh
add tweaks on helix

@@ -1,7 +1,7 @@
 #!/bin/bash
 #SBATCH --partition=single
 #SBATCH --ntasks=1
-#SBATCH --time=10:00:00
+#SBATCH --time=1:00:00
 #SBATCH --mem=20gb
 #SBATCH --gres=gpu:A40:1
 
@@ -27,4 +27,4 @@ echo " "
 module load devel/cuda/11.6
 
 python3 -u llama_logprobs.py \
-    --model_name="meta-llama/Llama-2-7b-hf"
+    --model_name="meta-llama/Llama-2-13b-chat-hf"

### 2023-12-17 01-code/run_llama_70b.sh

@@ -1,9 +1,9 @@
 #!/bin/bash
 #SBATCH --partition=single
 #SBATCH --ntasks=1
-#SBATCH --time=10:00:00
+#SBATCH --time=2:00:00
 #SBATCH --mem=237gb
-#SBATCH --gres=gpu:A100:3
+#SBATCH --gres=gpu:A100:2
 
 echo 'Running simulation'
 

### 2023-12-17 01-code-LLM-experiments/llama_logprobs.py
add sampling

@@ -43,7 +43,7 @@ def getLogProbContinuation(
         return_tensors="pt",
     ).input_ids
     #print("input_ids prompt ", input_ids_prompt)
-    print("input ids continuation ", input_ids_continuation.shape)
+    print("input ids continuation ", input_ids_continuation.shape, input_ids_continuation)
     # cut off the first token of the continuation, as it is SOS
     input_ids = torch.cat(
         (input_ids_prompt, input_ids_continuation[:, 1:]), 
@@ -51,9 +51,10 @@ def getLogProbContinuation(
     ).to("cuda:0") # put input on the first device
     print("input ids shape ", input_ids.shape)
     # pass through model
-    outputs = model(
-        input_ids,
-    )
+    with torch.no_grad():
+        outputs = model(
+            input_ids,
+        )
     # transform logits to probabilities
     print("shape of logits ", outputs.logits.shape)
     # remove the EOS logit which we aren't interested in
@@ -66,7 +67,7 @@ def getLogProbContinuation(
     # cut off the sos token so as to get predictions for the actual token conditioned on 
     # preceding context
     input_ids_probs = input_ids[:, 1:].squeeze().unsqueeze(-1)
-    print("shape of input ids for porb retrieval ", input_ids_probs.shape)
+    print("shape of input ids for porb retrieval ", input_ids_probs.shape, input_ids_probs)
     # retreive at correct token positions
     conditionalLogProbs = torch.gather(
         llama_output_scores, 
@@ -77,7 +78,7 @@ def getLogProbContinuation(
     continuationConditionalLogProbs = conditionalLogProbs[
         (input_ids_prompt.shape[-1]-1):
     ]
-    print("Shape of retrieved log probs", continuationConditionalLogProbs.shape)
+    print("Shape of retrieved log probs", continuationConditionalLogProbs.shape, continuationConditionalLogProbs)
     # compute continunation log prob
     sentLogProb = torch.sum(continuationConditionalLogProbs).item()
     print("sent log prob ", sentLogProb)
@@ -124,7 +125,15 @@ def get_model_predictions(
     lprob_distractor2 = getLogProbContinuation(
         context_production, vignette["production_distractor2"],
         model, tokenizer)
-
+    # for testing, also just sample a few productions
+    predictions_prompt_ids = tokenizer(context_production, return_tensors="pt")
+    production_samples = model.generate(
+        **predictions_prompt_ids,
+        do_sample = True,
+        num_return_sequences=5,
+    )
[…]

### 2023-12-17 01-code-LLM-experiments/llama_logprobs.py
add alternative method of llh retrieval for production

@@ -83,7 +83,25 @@ def getLogProbContinuation(
     sentLogProb = torch.sum(continuationConditionalLogProbs).item()
     print("sent log prob ", sentLogProb)
 
-    return sentLogProb
+    ### alternative method of retrieving log probs of single words via generate ###
+    # only pass the prompt and then retreive score of the respective tokens among the first predicted token
+    outputs_generate = model.generate(
+        input_ids_prompt, 
+        max_new_tokens=1,
+        output_scores=True,
+        num_return_sequences=1,
+        return_dict_in_generate=True
+    )
+    if isinstance(outputs_generate.scores, tuple):
+        logits = outputs_generate.scores[0][0]
+    else:
+        logits = outputs_generate.scores
+        
+    answer_logits = logits[input_ids_continuation[0][-1]].item()
+    print("input_ids_continuation[0][-1] ", input_ids_continuation[0][-1])
+    print("Answer logit retrieved with Jenn's method ", answer_logits)
+
+    return sentLogProb, answer_logits
             
 
 def soft_max(scores, alpha=1):
@@ -113,39 +131,42 @@ def get_model_predictions(
         context_interpretation = vignette['context_interpretation']
 
     # production
-    lprob_target      = getLogProbContinuation(
+    lprob_target, lprob_target_gen      = getLogProbContinuation(
         context_production, vignette["production_target"],
         model, tokenizer)
-    lprob_competitor  = getLogProbContinuation(
+    lprob_competitor, lprob_competitor_gen  = getLogProbContinuation(
         context_production, vignette["production_competitor"],
         model, tokenizer)
-    lprob_distractor1 = getLogProbContinuation(
+    lprob_distractor1, lprob_distractor1_gen = getLogProbContinuation(
         context_production, vignette["production_distractor1"],
         model, tokenizer)
-    lprob_distractor2 = getLogProbContinuation(
+    lprob_distractor2, lprob_distractor2_gen = getLogProbContinuation(
         context_production, vignette["production_distractor2"],
         model, tokenizer)
     # for testing, also just sample a few productions
-    predictions_prompt_ids = tokenizer(context_production, return_tensors="pt")
-    production_samples = model.generate(
-        **predictions_prompt_ids,
-        do_sample = True,
-        num_return_sequences=5,
-    )
-    production_decoded = tokenizer.batch_decode(production_samples)
-    print("productions decoded", production_decoded)
+    # predictions_prompt_ids = tokenizer(context_production, return_tensors="pt")
[…]

### 2023-12-17 01-code-LLM-experiments/llama_logprobs.py
another sanitycheck

@@ -101,6 +101,15 @@ def getLogProbContinuation(
     print("input_ids_continuation[0][-1] ", input_ids_continuation[0][-1])
     print("Answer logit retrieved with Jenn's method ", answer_logits)
 
+    ### sanity checking the llh results via nll loss comp
+    manual_llh = np.mean(np.array(conditionalLogProbs))
+    auto_llh = model(
+       input_ids,
+       labels=input_ids
+    ).loss
+    print("manually computed LL ", manual_llh)
+    print("loss computation based nll ", auto_llh)
+
     return sentLogProb, answer_logits
             
 

### 2023-12-18 01-code-LLM-experiments/llama_logprobs.py
add fixes of LLH retrieval and test run

@@ -33,27 +33,44 @@ def getLogProbContinuation(
     """
 
     initialSequence = preface + initialSequence
+    prompt = preface + initialSequence + continuation
+    print(initialSequence)
+    print(prompt)
     # tokenize separately, so as to know the shape of the continuation
     input_ids_prompt = tokenizer(
-        initialSequence, 
+        initialSequence.strip(), 
         return_tensors="pt",
+        # add_special_tokens=False
     ).input_ids
-    input_ids_continuation = tokenizer(
-        continuation,
+    input_ids = tokenizer(
+        prompt.strip(),
         return_tensors="pt",
-    ).input_ids
-    #print("input_ids prompt ", input_ids_prompt)
-    print("input ids continuation ", input_ids_continuation.shape, input_ids_continuation)
+    ).input_ids.to("cuda:0")
+    
+    print("input_ids prompt ",input_ids_prompt.shape, input_ids_prompt)
+    #prompt_to_words =  input_ids_prompt.word_ids()
+    #desired_output = []
+    #for w_idx in prompt_to_words:
+    #     start, end = input_ids_prompt.word_to_tokens(w_idx)
+    #     start+=1
+    #     end+=1
+    #     desired_output.append(list(range(start, end)))
+    #print("desired output ", desired_output)
+    #input_ids_to_words = input_ids.word_ids()
+    #print("input ids to words ", input_ids_to_words)
+    #print("input ids continuation ", input_ids_continuation.shape, input_ids_continuation)
     # cut off the first token of the continuation, as it is SOS
-    input_ids = torch.cat(
-        (input_ids_prompt, input_ids_continuation[:, 1:]), 
-        -1
-    ).to("cuda:0") # put input on the first device
-    print("input ids shape ", input_ids.shape)
+    #input_ids = torch.cat(
+    #    (input_ids_prompt, input_ids_continuation[:, 1:]), 
+    #    -1
+    #).to("cuda:0") # put input on the first device
+    print("input ids shape ", input_ids.shape, input_ids)
+    # print("----- decoded input ids -=-=== ", tokenizer.batch_decode(input_ids))
     # pass through model
     with torch.no_grad():
         outputs = model(
             input_ids,
+            labels=input_ids
         )
     # transform logits to probabilities
     print("shape of logits ", outputs.logits.shape)
@@ -67,7 +84,7 @@ def getLogProbContinuation(
     # cut off the sos token so as to get predictions for the actual token conditioned on 
[…]

### 2023-12-19 01-code-LLM-experiments/llama_logprobs.py
add npnlg approach saving for interpretation; finalized code

@@ -115,7 +115,8 @@ def getLogProbContinuation(
     else:
         logits = outputs_generate.scores
 
-    input_ids_continuation = input_ids[0][input_ids_prompt.shape[-1]-1:]        
+    input_ids_continuation = input_ids[0][input_ids_prompt.shape[-1]:]
+    print("indices of nonzero generation scores ", (logits > -torch.inf).nonzero())        
     answer_logits = logits[input_ids_continuation[0]].item()
     print("input_ids_continuation[0][-1] ", input_ids_continuation[0])
     print("outputs generate scores shape ", len(outputs_generate.scores), outputs_generate.scores[0][0].shape, outputs_generate.scores[0].shape)
@@ -132,13 +133,17 @@ def getLogProbContinuation(
 
     # another sanity check with MF's NPNLG code
     relevant_labels = torch.clone(input_ids)
-    for i in range(input_ids_prompt.shape[-1]-1):
+    for i in range(input_ids_prompt.shape[-1]):
         relevant_labels[0, i] = -100
     print("Relevant labels ", relevant_labels)
     output_masked = model(input_ids, labels=relevant_labels)
-    print("Output loss (i.e., mean) computed with NPNLG approach ", output_masked.loss.item(), output_masked.loss.item() * (input_ids_continuation.shape[-1]-1))
+    print("Output loss (i.e., mean) computed with NPNLG approach ", output_masked.loss.item(), output_masked.loss.item() * (input_ids_continuation.shape[-1]))
+    # for doubke checking, compute the same with only the last tokens
+    print("input ids continuation for checking NPNLG code" , input_ids_continuation)
+    output_last_tokens_loss = model(input_ids_continuation.unsqueeze(0), labels=input_ids_continuation.unsqueeze(0))
+    print("npnlg double checking loss ", output_last_tokens_loss.loss.item())
 
-    return sentLogProb, answer_logits
+    return sentLogProb, output_masked.loss.item()
             
 
 def soft_max(scores, alpha=1):
@@ -199,25 +204,29 @@ def get_model_predictions(
     # production_decoded = tokenizer.batch_decode(production_samples)
     # print("productions decoded", production_decoded)
     scores_production = np.array([lprob_target, lprob_competitor, lprob_distractor1, lprob_distractor2])
-    probs_production = soft_max(scores_production, alpha_production)
+    probs_production = soft_max(scores_production, alpha=1)
     # softmax the scores generated with alternative method
[…]

### 2023-12-19 01-code-LLM-experiments/llama_logprobs.py
tweak for temporary checking

@@ -166,10 +166,10 @@ def get_model_predictions(
     # no specific system prompt is passed
     # if one wanted to, the expected formatting would be: [INST]<<SYS>>{system prompt}<</SYS>>\n\n{user message}[/INST]
     if "chat" in model_name:
-        context_production = f"[INST]{vignette['context_production']}[/INST]"
+        # context_production = f"[INST]{vignette['context_production']}[/INST]"
         context_interpretation = f"[INST]{vignette['context_interpretation']}[/INST]"
     else:
-        context_production = vignette['context_production']
+        # context_production = vignette['context_production']
         context_interpretation = vignette['context_interpretation']
     # general sanity check
     #testing_prompt = getLogProbContinuation(
@@ -182,18 +182,18 @@ def get_model_predictions(
     #print("Christmas continuation ", testing_prompt2)
 
     # production
-    lprob_target, lprob_target_gen      = getLogProbContinuation(
-        context_production, vignette["production_target"],
-        model, tokenizer)
-    lprob_competitor, lprob_competitor_gen  = getLogProbContinuation(
-        context_production, vignette["production_competitor"],
-        model, tokenizer)
-    lprob_distractor1, lprob_distractor1_gen = getLogProbContinuation(
-        context_production, vignette["production_distractor1"],
-        model, tokenizer)
-    lprob_distractor2, lprob_distractor2_gen = getLogProbContinuation(
-        context_production, vignette["production_distractor2"],
-        model, tokenizer)
+    # lprob_target, lprob_target_gen      = getLogProbContinuation(
+    #     context_production, vignette["production_target"],
+    #     model, tokenizer)
+    # lprob_competitor, lprob_competitor_gen  = getLogProbContinuation(
+    #     context_production, vignette["production_competitor"],
+    #     model, tokenizer)
+    # lprob_distractor1, lprob_distractor1_gen = getLogProbContinuation(
+    #     context_production, vignette["production_distractor1"],
+    #     model, tokenizer)
+    # lprob_distractor2, lprob_distractor2_gen = getLogProbContinuation(
+    #     context_production, vignette["production_distractor2"],
+    #     model, tokenizer)
     # for testing, also just sample a few productions
     # predictions_prompt_ids = tokenizer(context_production, return_tensors="pt")
     # production_samples = model.generate(
@@ -203,12 +203,12 @@ def get_model_predictions(
     # )
[…]

### 2023-12-19 01-code-LLM-experiments/llama_logprobs.py
add color and shape checks

@@ -326,7 +326,7 @@ def main(model_name):
     pprint(results_df)
     # TODO format results_df name to include model name
     name_for_saving = model_name.split('/')[-1]
-    results_name = f'results_wholePrompt_{name_for_saving}.csv'
+    results_name = f'results_sanity_check_data_{name_for_saving}.csv'
     results_df.to_csv(results_name, index = False)
 
 if __name__ == '__main__':

### 2024-03-25 04-paper/paper-LLM-probmod.tex
rerun LLaMA analyses

--- 2023-12-01 03-paper/paper-LLM-probmod.tex
@@
 \centering
-\begin{tabular}{llllcrlcrcl}
+\begin{tabular}{cllllcrlcrlc}
   \toprule
   &&&& $\alpha$ &&& $\epsilon$ & \\ \cmidrule(r){4-6} \cmidrule(l){7-9}
-  data & model & condition & $|$95\%\ & mean\ & 95\%$|$\ & $|$95\%\ & mean\ & 95\%$|$\ & Bpppv & \\
+  & model & data & method & condition & |95\% & mean\ & 95\%| & |95\% & mean\ & 95\%| & Bpppv \\
   \midrule
-
-  item-level  & RSA                & production     & 2.62 & 3.13 & 3.69 & 0.08 & 0.12 & 0.16 & 0.29 &  \\
-  item-level  & RSA                & interpretation & 0.21 & 0.67 & 1.08 & 0.08 & 0.14 & 0.19 & 0.21 &  \\
-  item-level  & LLM                & production     & 0.15 & 0.19 & 0.24 & 0.00 & 0.07 & 0.16 & 0.00 & * \\
-  item-level  & LLM                & interpretation & 0.08 & 0.11 & 0.14 & 0.00 & 0.09 & 0.22 & 0.00 & * \\
-  cond.-level & RSA                & production     & 2.62 & 3.14 & 3.70 & 0.08 & 0.12 & 0.17 & 0.50 &  \\
-  cond.-level & RSA                & interpretation & 0.27 & 0.68 & 1.13 & 0.09 & 0.14 & 0.19 & 0.51 &  \\
-  cond.-level & avg. scores        & production     & 0.19 & 0.22 & 0.25 & 0.00 & 0.03 & 0.08 & 0.05 &  \\
-  cond.-level & avg. scores        & interpretation & 0.15 & 0.18 & 0.20 & 0.00 & 0.02 & 0.05 & 0.00 & * \\
-  cond.-level & avg. probabilities & production     & 0.86 & 1.03 & 1.22 & 0.08 & 0.12 & 0.17 & 0.47 &  \\
-  cond.-level & avg. probabilities & interpretation & 0.36 & 0.46 & 0.59 & 0.00 & 0.04 & 0.10 & 0.00 & * \\
-  cond.-level & avg. WTA           & production     & 0.86 & 1.04 & 1.22 & 0.08 & 0.12 & 0.17 & 0.48 &  \\
-  cond.-level & avg. WTA           & interpretation & 0.15 & 0.37 & 0.58 & 0.03 & 0.12 & 0.18 & 0.49 &  \\
-
+  ! & RSA & item  & ---         & prd. & 2.62 & 3.13 & 3.69  & 0.08 & 0.12 & 0.16 & 0.29 \\
+  ! & RSA & item  & ---         & int. & 0.21 & 0.67 & 1.08  & 0.08 & 0.14 & 0.19 & 0.21 \\
+  ! & RSA & cond. & ---         & prd. & 2.62 & 3.14 & 3.70  & 0.08 & 0.12 & 0.17 & 0.50 \\
+  ! & RSA & cond. & ---         & int. & 0.27 & 0.68 & 1.13  & 0.09 & 0.14 & 0.19 & 0.51 \\ \addlinespace[0.5em]
+    & GPT & item  & ---         & prd. & 0.40 & 0.51 & 0.62  & 0.07 & 0.12 & 0.17 & 0.00 \\
+    & GPT & item  & ---         & int. & 0.52 & 0.66 & 0.82  & 0.00 & 0.05 & 0.15 & 0.00 \\
+  ! & GPT & cond. & avg. scores & prd. & 0.61 & 0.74 & 0.87  & 0.08 & 0.12 & 0.17 & 0.49 \\
+    & GPT & cond. & avg. scores & int. & 0.99 & 1.17 & 1.35  & 0.00 & 0.02 & 0.06 & 0.00 \\
+  ! & GPT & cond. & avg. prob.  & prd. & 0.81 & 5.00 & 14.58 & 0.08 & 0.12 & 0.16 & 0.61 \\
+    & GPT & cond. & avg. prob.  & int. & 1.36 & 2.04 & 2.67  & 0.00 & 0.03 & 0.08 & 0.00 \\
+  ! & GPT & cond. & avg. WTA    & prd. & 0.86 & 1.04 & 1.22  & 0.08 & 0.12 & 0.17 & 0.48 \\
+  ! & GPT & cond. & avg. WTA    & int. & 0.16 & 0.43 & 0.69  & 0.06 & 0.13 & 0.19 & 0.50 \\
    \bottomrule \\
@@
   Summary statistics for model fits.
-  Columns show the means and 95\% credible intervals for posterior samples for each parameter, as well as the Bayesian posterior predictive $p$-values for each model, condition and data level.
-  Asterisks mark Bppp-values no higher than $0.05$, flagging relative inability to capture the structure of the data.
+  Columns show the means and 95\% credible intervals for posterior samples for each parameter, as well as the Bayesian posterior predictive $p$-values for each model, data type (item- or condition-level data), aggregation method (for condition-level data), and condition.
+  The final column shows the Bayesian posterior predictive $p$-values.
+  The first columns marks whether the row's Bayesian posterior predictive $p$-values are higher / lower than $0.05$, flagging relative inability to capture the structure of the data.
 }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Results from LLaMA2}
+\label{sec:LLaMA-results}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=0.9\textwidth]{00-pics/vPPC_all_LLaMA_models.pdf}
+  \caption{Visual posterior predictive checks on condition-level data for different models of the LLaMA family, paired with different aggregation strategies.}
+  \label{fig:LLaMA-results-vPPC}
+\end{figure}
+
+\begin{table}[t]
+\centering
+\begin{tabular}{cllllcrlcrlc}
+  \toprule
+  &&&& $\alpha$ &&& $\epsilon$ & \\ \cmidrule(r){4-6} \cmidrule(l){7-9}
+  & model & data & method & condition & |95\% & mean\ & 95\%| & |95\% & mean\ & 95\%| & Bpppv \\
+  \midrule
+
+    & L2-hf-7b    & item  & ---         & prd. & 1.09  & 1.30   & 1.51   & 0.00 & 0.02 & 0.06 & 0.00 \\
+    & L2-hf-7b    & item  & ---         & int. & 1.90  & 2.52   & 3.13   & 0.00 & 0.04 & 0.12 & 0.00 \\
+  ! & L2-hf-7b    & cond. & avg. scores & prd. & 9.65  & 11.50  & 13.62  & 0.08 & 0.12 & 0.16 & 0.48 \\
+    & L2-hf-7b    & cond. & avg. scores & int. & 6.80  & 7.96   & 9.14   & 0.00 & 0.02 & 0.05 & 0.00 \\
+    & L2-hf-7b    & cond. & avg. prob.  & prd. & 4.98  & 16.44  & 33.06  & 0.06 & 0.10 & 0.13 & 0.00 \\
+    & L2-hf-7b    & cond. & avg. prob.  & int. & 20.06 & 79.30  & 163.94 & 0.00 & 0.01 & 0.04 & 0.00 \\
+  ! & L2-hf-7b    & cond. & avg. WTA    & prd. & 4.17  & 4.96   & 5.85   & 0.07 & 0.12 & 0.16 & 0.48 \\
+    & L2-hf-7b    & cond. & avg. WTA    & int. & 0.87  & 1.02   & 1.18   & 0.00 & 0.02 & 0.05 & 0.00 \\ \addlinespace[0.5em]
+    & L2-hf-13b   & item  & ---         & prd. & 1.37  & 1.91   & 2.47   & 0.00 & 0.07 & 0.12 & 0.00 \\
+    & L2-hf-13b   & item  & ---         & int. & 2.13  & 2.90   & 3.69   & 0.00 & 0.05 & 0.14 & 0.00 \\
+  ! & L2-hf-13b   & cond. & avg. scores & prd. & 13.73 & 16.32  & 19.10  & 0.08 & 0.12 & 0.16 & 0.50 \\
+    & L2-hf-13b   & cond. & avg. scores & int. & 7.73  & 9.11   & 10.42  & 0.00 & 0.02 & 0.05 & 0.00 \\
+    & L2-hf-13b   & cond. & avg. prob.  & prd. & 69.84 & 185.80 & 360.44 & 0.06 & 0.09 & 0.13 & 0.00 \\
+    & L2-hf-13b   & cond. & avg. prob.  & int. & 13.63 & 39.72  & 88.28  & 0.00 & 0.02 & 0.05 & 0.00 \\
+  ! & L2-hf-13b   & cond. & avg. WTA    & prd. & 2.28  & 2.70   & 3.18   & 0.08 & 0.12 & 0.16 & 0.51 \\
+    & L2-hf-13b   & cond. & avg. WTA    & int. & 0.78  & 0.94   & 1.12   & 0.00 & 0.03 & 0.07 & 0.03 \\ \addlinespace[0.5em]
+    & L2-hf-70b   & item  & ---         & prd. & 1.53  & 1.86   & 2.22   & 0.00 & 0.03 & 0.07 & 0.00 \\
+    & L2-hf-70b   & item  & ---         & int. & 3.20  & 4.01   & 4.74   & 0.00 & 0.03 & 0.08 & 0.00 \\
+  ! & L2-hf-70b   & cond. & avg. scores & prd. & 44.57 & 53.17  & 62.44  & 0.08 & 0.12 & 0.17 & 0.48 \\
+    & L2-hf-70b   & cond. & avg. scores & int. & 6.86  & 8.14   & 9.32   & 0.00 & 0.02 & 0.05 & 0.00 \\
+    & L2-hf-70b   & cond. & avg. prob.  & prd. & 10.51 & 26.59  & 45.73  & 0.07 & 0.10 & 0.14 & 0.00 \\
+    & L2-hf-70b   & cond. & avg. prob.  & int. & 10.86 & 23.63  & 42.16  & 0.00 & 0.03 & 0.09 & 0.00 \\
+  ! & L2-hf-70b   & cond. & avg. WTA    & prd. & 5.88  & 7.03   & 8.19   & 0.08 & 0.12 & 0.17 & 0.50 \\
+    & L2-hf-70b   & cond. & avg. WTA    & int. & 0.55  & 0.68   & 0.81   & 0.00 & 0.03 & 0.07 & 0.04 \\ \addlinespace[0.5em]
+    & L2-chat-7b  & item  & ---         & prd. & 0.34  & 0.41   & 0.49   & 0.00 & 0.03 & 0.07 & 0.00 \\
+    & L2-chat-7b  & item  & ---         & int. & 0.79  & 1.05   & 1.34   & 0.00 & 0.05 & 0.15 & 0.00 \\
+  ! & L2-chat-7b  & cond. & avg. scores & prd. & 58.47 & 69.64  & 81.79  & 0.08 & 0.12 & 0.16 & 0.48 \\
+    & L2-chat-7b  & cond. & avg. scores & int. & 2.03  & 2.37   & 2.73   & 0.00 & 0.02 & 0.05 & 0.00 \\
+    & L2-chat-7b  & cond. & avg. prob.  & prd. & 8.05  & 27.26  & 55.30  & 0.07 & 0.11 & 0.15 & 0.00 \\
+    & L2-chat-7b  & cond. & avg. prob.  & int. & 2.86  & 19.24  & 51.41  & 0.00 & 0.01 & 0.04 & 0.00 \\
+  ! & L2-chat-7b  & cond. & avg. WTA    & prd. & 12.18 & 14.70  & 17.21  & 0.08 & 0.12 & 0.16 & 0.50 \\
+    & L2-chat-7b  & cond. & avg. WTA    & int. & 0.86  & 1.01   & 1.14   & 0.00 & 0.02 & 0.05 & 0.00 \\ \addlinespace[0.5em]
+    & L2-chat-13b & item  & ---         & prd. & 0.36  & 0.44   & 0.53   & 0.00 & 0.03 & 0.08 & 0.00 \\
+    & L2-chat-13b & item  & ---         & int. & 0.68  & 0.90   & 1.11   & 0.00 & 0.05 & 0.13 & 0.00 \\
+  ! & L2-chat-13b & cond. & avg. scores & prd. & 5.56  & 6.63   & 7.76   & 0.08 & 0.12 & 0.16 & 0.49 \\
+    & L2-chat-13b & cond. & avg. scores & int. & 1.70  & 2.05   & 2.40   & 0.00 & 0.02 & 0.06 & 0.01 \\
+    & L2-chat-13b & cond. & avg. prob.  & prd. & 3.90  & 22.25  & 74.12  & 0.06 & 0.10 & 0.14 & 0.00 \\
+  ! & L2-chat-13b & cond. & avg. prob.  & int. & 2.85  & 4.70   & 7.01   & 0.00 & 0.03 & 0.07 & 0.08 \\
+  ! & L2-chat-13b & cond. & avg. WTA    & prd. & 4.28  & 5.11   & 5.99   & 0.07 & 0.12 & 0.16 & 0.49 \\
+  ! & L2-chat-13b & cond. & avg. WTA    & int. & 0.15  & 0.32   & 0.51   & 0.02 & 0.11 & 0.18 & 0.50 \\ \addlinespace[0.5em]
+    & L2-chat-70b & item  & ---         & prd. & 0.28  & 0.35   & 0.42   & 0.00 & 0.03 & 0.08 & 0.00 \\
+    & L2-chat-70b & item  & ---         & int. & 1.03  & 1.24   & 1.47   & 0.00 & 0.03 & 0.08 & 0.00 \\
+    & L2-chat-70b & cond. & avg. scores & prd. & 0.40  & 0.55   & 0.74   & 0.00 & 0.04 & 0.10 & 0.00 \\
+  ! & L2-chat-70b & cond. & avg. scores & int. & 1.91  & 2.37   & 2.89   & 0.00 & 0.03 & 0.08 & 0.16 \\
+    & L2-chat-70b & cond. & avg. prob.  & prd. & 0.43  & 0.66   & 0.96   & 0.00 & 0.04 & 0.10 & 0.00 \\
+    & L2-chat-70b & cond. & avg. prob.  & int. & 2.82  & 5.21   & 8.43   & 0.00 & 0.01 & 0.04 & 0.02 \\
+    & L2-chat-70b & cond. & avg. WTA    & prd. & 0.14  & 0.21   & 0.35   & 0.00 & 0.06 & 0.13 & 0.00 \\
+    & L2-chat-70b & cond. & avg. WTA    & int. & 0.85  & 1.00   & 1.14   & 0.00 & 0.02 & 0.05 & 0.00 \\
+
+   \bottomrule \\
+\end{tabular}
+\caption{
+  Summary statistics for model fits based on LLaMA2 backends.
+  The information shown is the same as in Table~\ref{tab:sumStats}.
+}
+\label{tab:sumStats-LLaMA2}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Screenshots from the online experiment with human participants}
@@
 %%% End:
-
-   \hline
-\end{tabular}

### 2024-03-25 03-data-analysis/02-data-preparation.r

@@ -1,4 +1,5 @@
 source('00-premable.r')
+data_folder <- '../02-data/refgame/with_quotes/mean_scores_forward/'
 
 ##################################################
 ## helper functions 
@@ -27,7 +28,7 @@ wta_row <- function(matrix_of_scores) {
 
 get_prepped_data_for_model <- function(model_name, results_file) {
   
-  d_raw <- read_csv('../02-data/data-raw-human.csv') |> 
+  d_raw <- read_csv(str_c(data_folder, 'data-raw-human.csv')) |> 
     select(- c(
       "experiment_duration", "experiment_end_time", "experiment_start_time",
       "prolific_pid", "prolific_session_id", "prolific_study_id"  )) |> 
@@ -181,7 +182,8 @@ get_prepped_data_for_model <- function(model_name, results_file) {
     filter(condition == "production") |> 
     select(submission_id, trial_nr, item, condition, starts_with("scores_produ")) |> 
     mutate(scores_production_distractor = 
-             log(exp(scores_production_distractor1) + exp(scores_production_distractor2))) |> 
+             scores_production_distractor1 + scores_production_distractor2) |> 
+             # log(exp(scores_production_distractor1) + exp(scores_production_distractor2))) |> 
     select(-scores_production_distractor1, -scores_production_distractor2)
   
   # narrow- and intermediate-scope predictions for production
@@ -239,7 +241,8 @@ get_prepped_data_for_model <- function(model_name, results_file) {
     filter(condition == "production") |> 
     select(item, starts_with("scores_produ")) |> 
     mutate(scores_production_distractor = 
-             log(exp(scores_production_distractor1) + exp(scores_production_distractor2))) |> 
+             scores_production_distractor1 + scores_production_distractor2) |> 
+             # log(exp(scores_production_distractor1) + exp(scores_production_distractor2))) |> 
     select(-scores_production_distractor1, -scores_production_distractor2) |> 
     group_by(item) |> 
     mutate(n = n()) |>  
[…]

### 2024-03-25 03-data-analysis/03-model-fitting.R

@@ -152,7 +152,7 @@ get_model_fits_RSA <- function(rerun = FALSE) {
   
   # save everything
   save_fit_object(fit_prod_RSA, model_path, "fit_prod_RSA.RDS")
-  save_fit_object(fit_inter_RSA, model_path, "fit_iter_RSA.RDS")
+  save_fit_object(fit_inter_RSA, model_path, "fit_inter_RSA.RDS")
   save_fit_object(fit_items_prod_RSA, model_path, "fit_items_prod_RSA.RDS")
   save_fit_object(fit_items_inter_RSA, model_path, "fit_items_inter_RSA.RDS")
   saveRDS(pp_prod_RSA, file = str_c(model_path, "pp_prod_RSA.RDS"))
@@ -365,10 +365,3 @@ for (model in model_names) {
 # model_fits_13b      <- get_model_fits("LLaMA2-hf-13b", rerun = T)
 # model_fits_07b      <- get_model_fits("LLaMA2-hf-7b", rerun = T)
 # model_fits_GPT      <- get_model_fits("GPT", rerun = T)
-
-
-
-
-
-
-

### 2024-03-25 03-data-analysis/04-data-analysis.R

@@ -290,9 +290,6 @@ retrieve_results_model_fits <- function(model_name) {
   write_csv(sumStats_combined, file = str_c("04-analysis-results/", model_name, "sumStats.csv"))
   
   
- 
-  
-  
   #######################################################
   ## interpreting alpha-fits for item-level analysis
   #######################################################
@@ -394,6 +391,8 @@ sumStats_LLaMA2_hf_13b      <- read_csv("04-analysis-results/LLaMA2-hf-13bsumSta
 sumStats_LLaMA2_hf_7b       <- read_csv("04-analysis-results/LLaMA2-hf-7bsumStats.csv")
 sumStats_GPT                <- read_csv("04-analysis-results/GPTsumStats.csv")
 
+# table of all model results (LaTeX)
+
 # bind all sumStats together
 sumStats_all <- 
   rbind(
@@ -409,21 +408,183 @@ sumStats_all <-
   mutate(
     model = factor(model, levels = c("RSA", "GPT", 
                                      "LLaMA2-hf-7b", "LLaMA2-hf-13b", "LLaMA2-hf-70b", 
-                                     "LLaMA2-chat-hf-7b", "LLaMA2-chat-hf-13b", "LLaMA2-chat-hf-70b")),
+                                     "LLaMA2-chat-hf-7b", "LLaMA2-chat-hf-13b", "LLaMA2-chat-hf-70b"))
+  ) |> 
+    
+  # specify levels for 'backend models'
+  mutate(backend = case_when(
+    model == "LLaMA2-hf-7b" ~ "L2-hf-7b",
+    model == "LLaMA2-hf-13b" ~ "L2-hf-13b",
+    model == "LLaMA2-hf-70b" ~ "L2-hf-70b",
+    model == "LLaMA2-chat-hf-7b" ~ "L2-chat-7b",
+    model == "LLaMA2-chat-hf-13b" ~ "L2-chat-13b",
+    model == "LLaMA2-chat-hf-70b" ~ "L2-chat-70b",
+    T ~ model
+  )) |> 
+  mutate(model = factor(backend, 
+                          levels = c(
+                            "RSA", "GPT", 
+                            "L2-hf-7b", "L2-hf-13b", "L2-hf-70b", 
+                            "L2-chat-7b", "L2-chat-13b", "L2-chat-70b"
+                            ))) |> 
+  mutate(
     condition = factor(condition, levels = c("prd.", "int.")),
     method = factor(method, levels = c("avg. scores", "avg. prob.", "avg. WTA", "---")),
[…]

### 2024-03-27 04-paper/paper-LLM-probmod.tex
revised introduction

@@
 \renewcommand{\undertitle}{}
-\renewcommand{\shorttitle}{Statistical modeling with LLM predictors}
+\renewcommand{\shorttitle}{LLMs as statistical models for categorical experimental data}
 
@@
 
-The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} has enabled a new generation of powerful large language models  \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and}.
-State-of-the-art LLMs excel on many benchmark data sets \citep[e.g.,][]{srivastava2023-BIGbench,PerezRinger2023:Discovering-Lan}, and so promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
+Enabled by the invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al}, recent years have brought a new generation of powerful large language models  \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and}.
+State-of-the-art LLMs excel on many benchmark data sets \citep[e.g.,][]{srivastava2023-BIGbench,PerezRinger2023:Discovering-Lan}, and so promise to serve as foundation models for a vast and diverse set of applications, both in industry and academia \citep{BommasaniHudson2021:On-the-opportun}.
+Yet, for any downstream application of LLMs, it is crucial to understand what LLMs can or cannot reliably do.
+The way in which LLM capabilities are assessed depends on what the intended application is.
+For many industrial applications, the prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' or ``gold standard'' exists, averaged over many instances of this task.
+Benchmark-driven assessments are very useful for engineering purposes, when the main issue is to whether a given system can perform a particular task correctly.
+Benchmark testing is also useful to systematically study of the effects of scale \citep[e.g.,][]{srivastava2023-BIGbench}, to uncover areas in which LLMs might under-perform on some standard that is practically relevant, and to thereby direct the community's attention to what needs improvement.
+More encompassing approaches also highlight the importance of including multiple factors in holistic assessments of LLM, such as robustness, fairness and efficiency \citep{LiangBommasani2023:Holistic-Evalua}.
+
 % especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
+
+There are also applications of LLMs where benchmark testing on a ``gold standard'' is arguably not optimal.
 Recent works increasingly go beyond using LLMs based on single-run input-output behavior, and instead utilize LLMs as a part of a larger computational process.
-Examples include sophisticated prompting strategies \citep[e.g.,][]{LiuLiu2022:Generated-Knowl}, or structured reasoning models \citep[e.g.,][]{CreswellShanahan2022:Selection-Infer,GaoMadaan2023:PAL:-Program-ai,ParanjapeLundberg2023:ART:-Automatic-}.
-Information from LLMs is also used  to rank or numerically score options in open-ended applications, e.g., to mimic human judgements of relevance or interestingness \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
-Other work, uses LLMs as part of bigger programs to build towards something more akin to explanatory cognitive models \citep[e.g.,][]{WongGrand2023:From-Word-Model}.
+Simple examples include sophisticated prompting strategies \citep[e.g.,][]{LiuLiu2022:Generated-Knowl}, or structured reasoning models \citep[e.g.,][]{CreswellShanahan2022:Selection-Infer,GaoMadaan2023:PAL:-Program-ai,ParanjapeLundberg2023:ART:-Automatic-}.
+More sophisticated examples include neuro-symbolic models in which LLMs supply specific parts of the relevant information for some practical application \mf{include citations}, or where LLMs are a part of bigger programs to build towards something more akin to explanatory cognitive models \citep[e.g.,][]{WongGrand2023:From-Word-Model}.
+For example, information from LLMs can be used to generate alternatives for deliberation \citep[e.g.][]{TsvilodubCarcassi2024:Towards-Neuro-S}, arguably similar to human resource-rational reasoning in open-ended domains.
+\mf{include citations}
+LLMs may also be used to rank or numerically score options in open-ended applications, e.g., to mimic human judgements of relevance or interestingness \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
+In many of these applications, LLMs essentially serve as a cheap, compressed stand-in for (average) human judgements, associations or choice behavior.
+To assess the quality of LLMs in such contexts, it is less important to compare against a gold standard, and more important to compare against the full distribution of the human behavior that is to be captured by the LLM.
 
-For all of these applications, it is crucial to understand what LLMs can or cannot reliably do.
-The prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' exists, averaged over many instances of this task.
-Benchmark-driven assessments are very useful for engineering purposes, e.g., to systematically study of the effects of scale \citep[e.g.,][]{srivastava2023-BIGbench}, but is also used to ask to what extent LLM performance is ``human-like.''
-In more psychologically oriented work, LLM performance is therefore compared to human choice behavior in psychological experiments and investigates whether LLMs predict patterns of human answer behavior  \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
+Taking inspiration from experimental psychology, an increasing number of studies compares LLM predictions to human choice behavior in psychological experiments and investigates whether LLMs predict patterns of human answer behavior  \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
 The main focus is often to compare qualitative patterns in LLM predictions and human data, but there is also work investigating whether LLMs can make adequate \emph{quantitative predictions}.
-For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of the N400 component of event-related-potentials in EEG measurements \citep{LindborgRabovsky2021:Meaning-in-brai}.
-The work presented in this paper seeks to extend the investigation of the human-likeness of LLMs by looking at empirical data from human multiple-choice behavior and comparing an established, theoretically motivated probabilistic cognitive model, which is commonly use to predict or explain this kind of data, to structurally similar quantitative predictions derived from LLMs.
-In short, while a lot of previous work has used targeted assessment of LLM capabilities from the point of view of the experimental psychologist, we here adopt the more specific perspective of the cognitive modeller.
+Most notably, there is a strong tradition of relatively early work in computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As}, which investigates whether quantitative predictions derived from language models match quantitative aspects in human experimental data, such as reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of the N400 component of event-related-potentials in EEG measurements \citep{LindborgRabovsky2021:Meaning-in-brai}.
 
-For the purposes of this investigation, a probabilistic cognitive model (PCM) is characterized by providing a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each possible data observation $D_{C}$ for any experimental condition $C$ and parameter values $\theta$ \citep[e.g.,][]{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
-% Usually, conditions $C$ provide a categorization of different tasks which is meaningful against the theoretical background of experimental investigation, and parameters $\theta$ modulate the predicted likelihood of the data in human interpretable ways (see Section~\ref{motivation} for more detail).
-There are non-trivial, insightful differences between LLMs and PCMs, both at the conceptual level and in terms of common practices in evaluation, which we will explore in this paper in order to contribute to a better delineation of the potential and limitations of LLMs.
-For one, the ``primitive predictions'' that LLMs provide are not at the level of a condition $C$, but at a lower item-based level.
-% In this sense, LLMs could be said to be less explanatory than PCMs by their very nature (see Section~\ref{motivation}).
-For another, LLMs are often just assessed based on their ability to identify a target answer, not to predict the full distribution observed in the data.
-% Nevertheless, distributional predictions for condition-level data can in principle be derived from LLMs, but there are many ways of doing so, yet with little consensus of how to do so.
+The work presented in this paper seeks to extend the investigation of the human-likeness of predictions derived from LLMs.
+Our foremost concern is methodological: How can we derive full distributional predictions from information supplied by LLMs, and how can we stringently test whether these distributional predictions are adequate given suitable empirical data?
+In short, while a lot of previous like-minded work has used targeted assessment of LLM capabilities from the point of view of the experimental psychologist, we here adopt the more specific perspective of the probabilistic / statistical modeller.
+We introduce different ways of building a Bayesian probabilistic model around core predictor values derived from various LLMs.
+We then use the standard tools of Bayesian data analysis, to fit and check the resulting statistical models based on data from human multiple-choice experiments.
+Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single, small-scale case of pragmatic language production and interpretation, for which a widely used probabilistic cognitive model exists, the Rational Speech Act model \citep{FrankGoodman2012:Predicting-Prag}, so that we can compare it against the LLM-fueled statistical models.
 
-This paper introduces and systematically compares different ways of carving numerical information obtained from an LLM into a structure similar to a parameterized PCM and addresses the empirical question of whether such an ``LLM-based PCM'' adequately captures the distribution of human forced-choice data at different levels of aggregation.
-To obtain data for this investigation, we subjected human participants and an LLM (a version of OpenAI's GPT-3.5; see Section~\ref{sec:item-level-pred}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
-We introduce different ways of defining condition-level, distributional predictions based on LLM-provided numerical measures.
-We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare these ``LLM-based PCMs'' against each other, and against an established, theoretically informed PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
-This is quite different, and arguably more stringent and insightful, than the standard practice of evaluating LLMs based on accuracy scores.
-We find that variability predicted by the LLM at the item-level is not borne out by the human data, and that some but not all ways of constructing condition-level predictions are equally good.
-These results suggest that it is not impossible to build predictive models for condition-level data, similar to PCMs, on top of LLM-supplied information, but that seemingly subtle details in the derivation of likelihoods matter.
-The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not.
+The main results of our investigation are, first of all, that it is possible to build Bayesian statistical models around LLM-based predictors, and that some but not all ways of doing so yield models that adequately predict the human data.
+Importantly, this investigations highlights that LLMs are by design fundamentally different from common statistical or probabilistic cognitive models, in that LLMs make predictions for each individual item, rather than specifying a predictor of central tendency at a more aggregate level (referred to here as the condition-level; see Section~\ref{motivation}).
+This observation raises two important conceptual and empirical questions.
+First, we need to ask whether the item-level predictions made by LLMs are empirically correct.
+Second, the implied methodological challenge lies in specifying how item-level information can be used, e.g., by different aggregation methods, to make robust predictions at a more abstract level (condition, task, $\dots$).
+We find that variability predicted by the LLM at the item-level is not borne out by the human data, that some but not all ways of constructing condition-level predictions are equally good, and that different LLMs as backends may prefer to use different aggregation strategies.
+Our conceptual comparison of LLMs and PCMs also identifies a sense in which the latter are explanatory, while the former are not. \mf{double-check main take away}.
+
+% For the purposes of this investigation, a probabilistic cognitive model (PCM) is characterized by providing a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each possible data observation $D_{C}$ for any experimental condition $C$ and parameter values $\theta$ \citep[e.g.,][]{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
+% % Usually, conditions $C$ provide a categorization of different tasks which is meaningful against the theoretical background of experimental investigation, and parameters $\theta$ modulate the predicted likelihood of the data in human interpretable ways (see Section~\ref{motivation} for more detail).
+% There are non-trivial, insightful differences between LLMs and PCMs, both at the conceptual level and in terms of common practices in evaluation, which we will explore in this paper in order to contribute to a better delineation of the potential and limitations of LLMs.
+% For one, the ``primitive predictions'' that LLMs provide are not at the level of a condition $C$, but at a lower item-based level.
+% % In this sense, LLMs could be said to be less explanatory than PCMs by their very nature (see Section~\ref{motivation}).
+% For another, LLMs are often just assessed based on their ability to identify a target answer, not to predict the full distribution observed in the data.
+% % Nevertheless, distributional predictions for condition-level data can in principle be derived from LLMs, but there are many ways of doing so, yet with little consensus of how to do so.
+
+% This paper introduces and systematically compares different ways of carving numerical information obtained from an LLM into a structure similar to a parameterized PCM and addresses the empirical question of whether such an ``LLM-based PCM'' adequately captures the distribution of human forced-choice data at different levels of aggregation.
+% To obtain data for this investigation, we subjected human participants and an LLM (a version of OpenAI's GPT-3.5; see Section~\ref{sec:item-level-pred}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
+% We introduce different ways of defining condition-level, distributional predictions based on LLM-provided numerical measures.
+% We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare these ``LLM-based PCMs'' against each other, and against an established, theoretically informed PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
+% This is quite different, and arguably more stringent and insightful, than the standard practice of evaluating LLMs based on accuracy scores.
+% We find that variability predicted by the LLM at the item-level is not borne out by the human data, and that some but not all ways of constructing condition-level predictions are equally good.
+% These results suggest that it is not impossible to build predictive models for condition-level data, similar to PCMs, on top of LLM-supplied information, but that seemingly subtle details in the derivation of likelihoods matter.
+% The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not.
 
 The paper is structured as follows.
-Section~\ref{motivation} motivates the investigation in this paper by highlighting relevant differences in the way  LLMs and PCMs make predictions about data from forced-choice tasks.
-Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a commonly used, theoretically-motivate PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
+Section~\ref{motivation} motivates the investigation in this paper by highlighting relevant differences in the way LLMs and common probabilistic statistical models make predictions about data from forced-choice tasks.
+Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a commonly used, theoretically-motivate probabilistic cognitive model, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
 Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions adequately capture the human data at the item-level.
 Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
+\mf{include section to extend to LLaMA}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width = 0.8\textwidth]{00-pics/condition-item-level-prediction.png}
+  \caption{Schematic representation of key conceptual differences between different types of statistical models.}
+  \label{fig:stats-model-types}
+\end{figure}
 

### 2024-03-27 03-data-analysis/04-data-analysis.R
added diagnostic checks for posteriors

@@ -540,49 +540,57 @@ ggsave("../04-paper/00-pics/vPPC_all_LLaMA_models.pdf", width = 12, height = 6, 
 ##  inspecting model fit diagnostics
 ##################################################
 model = GPT
+model = LLaMA2_hf_13b
 
 extract_main_diagnostics <- function(model) {
   message("Extracting diagnostics for model: ", model$model_name)
+  
+  get_rhat_ess <- function(fit) {
+    summary <- fit$summary()
+    tibble(max_rhat = summary |> filter(variable != "at_least_as_extreme") |>  pull(rhat) |> max(),
+           min_ess  = summary |> filter(variable != "at_least_as_extreme") |> pull(ess_bulk) |> min())
+  }
+  
+  fits_ordered <- c(
+    model$fit_items_prod,
+    model$fit_items_inter,
+    model$fit_prod_avg_scores,
+    model$fit_inter_avg_scores,
+    model$fit_prod_avg_probs,
+    model$fit_inter_avg_probs,
+    model$fit_prod_WTA,
+    model$fit_inter_WTA  
+  )
+  
+  results <- map_df(fits_ordered, function(fit) get_rhat_ess(fit))
+  
   diagnostics <- tibble(
-    model = model$name_name,
+    model = model$model_name,
     data  = c("item", "item", "condition", "condition", "condition", "condition", "condition", "condition"),
     method = c("---", "---", "avg. scores", "avg. scores", "avg. prob.", "avg. prob.", "avg. WTA", "avg. WTA"),
     condition = c("production", "interpretation", "production", "interpretation", 
-                  "production", "interpretation", "production", "interpretation"),
-    max_rhat = c(model$fit_items_prod$summary() |> filter(variable != "at_least_as_extreme") |>  pull(rhat) |> max(),
-                 model$fit_items_inter$summary() |> filter(variable != "at_least_as_extreme") |> pull(rhat) |> max(),
-                 model$fit_prod_avg_scores$summary() |> filter(variable != "at_least_as_extreme") |> pull(rhat) |> max(),
-                 model$fit_inter_avg_scores$summary() |> filter(variable != "at_least_as_extreme") |> pull(rhat) |> max(),
-                 model$fit_prod_avg_probs$summary() |> filter(variable != "at_least_as_extreme") |> pull(rhat) |> max(),
[…]

### 2024-03-29 04-paper/paper-LLM-probmod.tex
minor tweaks introduction

@@
 Yet, for any downstream application of LLMs, it is crucial to understand what LLMs can or cannot reliably do.
-The way in which LLM capabilities are assessed depends on what the intended application is.
+The way in which LLM capabilities should be assessed depends on what their intended application is.
 For many industrial applications, the prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' or ``gold standard'' exists, averaged over many instances of this task.
-Benchmark-driven assessments are very useful for engineering purposes, when the main issue is to whether a given system can perform a particular task correctly.
+Benchmark-driven assessments are very useful for engineering purposes, when the main issue is whether a given system can perform a particular task correctly.
 Benchmark testing is also useful to systematically study of the effects of scale \citep[e.g.,][]{srivastava2023-BIGbench}, to uncover areas in which LLMs might under-perform on some standard that is practically relevant, and to thereby direct the community's attention to what needs improvement.
@@
 We then use the standard tools of Bayesian data analysis, to fit and check the resulting statistical models based on data from human multiple-choice experiments.
-Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single, small-scale case of pragmatic language production and interpretation, for which a widely used probabilistic cognitive model exists, the Rational Speech Act model \citep{FrankGoodman2012:Predicting-Prag}, so that we can compare it against the LLM-fueled statistical models.
+Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single case study of pragmatic language production and interpretation, so-called reference games \mf{references}, for which a widely used probabilistic cognitive model exists, the Rational Speech Act model \citep{FrankGoodman2012:Predicting-Prag}, so that we can compare a theoretically motivated probabilistic cognitive model (PCM) against a probabilistic model built on top of LLM predictions.
 
-The main results of our investigation are, first of all, that it is possible to build Bayesian statistical models around LLM-based predictors, and that some but not all ways of doing so yield models that adequately predict the human data.
-Importantly, this investigations highlights that LLMs are by design fundamentally different from common statistical or probabilistic cognitive models, in that LLMs make predictions for each individual item, rather than specifying a predictor of central tendency at a more aggregate level (referred to here as the condition-level; see Section~\ref{motivation}).
+The main results of our investigation are that it is possible to build Bayesian statistical models around LLM-based predictors, and that some but not all ways of doing so yield models that adequately predict the human data.
+More concretely, our investigation highlights that LLMs are by design fundamentally different from common statistical or probabilistic cognitive models, in that LLMs make predictions for each individual item, rather than specifying a predictor of central tendency at a more aggregate level (referred to here as the condition-level; see Section~\ref{motivation}).
 This observation raises two important conceptual and empirical questions.

### 2024-03-29 03-data-analysis/04-data-analysis.R
updated plotting (alpha-target)

@@ -14,7 +14,7 @@ model_names = c(
   "GPT"
 )
 
-model_name = model_names[4]
+model_name = model_names[7]
 
 retrieve_results_model_fits <- function(model_name) {
 
@@ -310,8 +310,10 @@ retrieve_results_model_fits <- function(model_name) {
       select(`|95%`,  mean, `95%|`) |> 
       as.numeric()
     
+    xMax <- sumStats[3]+ 0.1*sumStats[2]
+    
     plot_data <- tibble(
-      alpha = c(sumStats,seq(0,0.5, length.out = 1000)),
+      alpha = c(sumStats,seq(0, xMax, length.out = 1000)),
       mean_target_prob = map_dbl(alpha, function(x) mean_target_prob(x, current_condition))
     ) 
     
@@ -330,10 +332,10 @@ retrieve_results_model_fits <- function(model_name) {
         data = filter(plot_data, alpha >= sumStats[1] & alpha <= sumStats[3]), 
         fill = "gray", alpha = 0.5) +
       geom_polygon(data = polygon_df, fill = "gray", alpha = 0.5) +
-      geom_segment(aes(y = 1/3, yend = 1/3, x=0, xend = 0.5), 
+      geom_segment(aes(y = 1/3, yend = 1/3, x=0, xend = xMax), 
                    color = project_colors[1], 
                    linetype = "dotted", size = 1.1) +
-      geom_segment(aes(y = 1, yend = 1, x=0, xend = 0.5), 
+      geom_segment(aes(y = 1, yend = 1, x=0, xend = xMax), 
                    color = project_colors[4], 
                    linetype = "dotted", size = 1.1) +
       geom_line(size = 1.5) +

### 2024-03-29 04-paper/paper-LLM-probmod.tex
refactored §2 into §1 and 'item-level'-Sec

@@
 
-\title{Language models as probabilistic cognitive models}
+\title{Bayesian Statistical Modeling with Predictors from LLMs}
 
@@
 \renewcommand{\undertitle}{}
-\renewcommand{\shorttitle}{LLMs as statistical models for categorical experimental data}
+\renewcommand{\shorttitle}{Bayesian Statistical Modeling with Predictors from LLMs}
 
@@
 In short, while a lot of previous like-minded work has used targeted assessment of LLM capabilities from the point of view of the experimental psychologist, we here adopt the more specific perspective of the probabilistic / statistical modeller.
-We introduce different ways of building a Bayesian probabilistic model around core predictor values derived from various LLMs.
+Taking numerical predictor values generated from LLMs as input, we explore strategies of building a Bayesian statistical model around them, and to scrutinize these LLM-grounded Bayesian statistical models with the usual methods of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A}.
+A main conceptual take-away of this investigation, is that statistical models built around LLMs are by design fundamentally different from common statistical or probabilistic cognitive models; an observation which also reflects back on the possibility of seeing LLMs as potentially explanatory for human behavior or cognition.
+Concretely, LLMs make predictions for each individual item, rather than specifying a predictor of central tendency at a more aggregate level, such as at the level of an experimental condition, as standard statistical models or probabilistic cognitive models normally do (see Figure~\ref{fig:stats-model-types}).
+As this point is crucial for our investigation, the following elaborates briefly.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width = 0.8\textwidth]{00-pics/condition-item-level-prediction.png}
+  \caption{Schematic representation of key conceptual differences between different types of statistical models.}
+  \label{fig:stats-model-types}
+\end{figure}
+
+Psychological research into the workings of the human mind aims to find generalizable patterns in the way information is processed within or across different domains of cognition.
+Experimental work therefore often compares human performance in different \emph{experimental conditions}, which reflect the general factors that are hypothesized to influence behavior.
+Yet a single experimental condition is often instantiated with different \emph{experimental items}, which are not under scrutiny for any systematic, predictable effect on the observed measurements.
+For example, classical research on human memory \citep{AtkinsonShiffrin1968:Human-memory:-A} investigated the effect of rehearsal on memory consolidation.
+Relevant experiments compare recall with and without rehearsal (experimental conditions), while using different items (words, numbers, etc., to be memorized) in each instantiation of the same memory task.
+Likewise, when studying how hearing a color word can facilitate a same-or-different judgement of color swatches \citep{Rosch1975:The-Nature-of-M}, the main experimental manipulation concerns the typicality of shown color swatches, while the variability between different color words like ``blue'' or ``green'', is less important to this research question and so treated as \emph{item-level variation}.
+Consequently, a typical psychological experiment is mainly interested in assessing behavior at the level of the experimental condition, because that is where the distinctions relevant to the research question reside.
+Nevertheless, each experimental condition can be, and often is, instantiated with different items, variation among which is deemed less relevant to the research question at hand.
+
+Data from human participants for experiments of this kind usually show some variability between items, and also variability between participants.
+This variability is commonly incorporated in standard statistical models as random stochastic variation, e.g., by using hierarchical regression models \citep{Jaeger2008:Categorical-dat,barr2013,SorensenHohensteinb2016:Bayesian-linear}, as shown on the left side of Figure~\ref{fig:stats-model-types}.
+Still, the focus of interest usually remains at the condition-level effects, because it is this more abstract level of behavioral aggregation that is relevant for generalizable theory building.
+Similarly, when analyzing or explaining data from psychological experiments with a probabilistic cognitive model (PCM), the model's predictions will naturally be set-up to predict data by taking condition-level properties, possibly in conjunction with item-level properties, into account \citep[e.g.,][]{NilsonRieskamp2011:Hierarchical-Ba,Lee2011:How-Cognitive-M,ScheibehenneRieskamp2013:Testing-the-Ada}, as shown schematically in the middle of Figure~\ref{fig:stats-model-types}.
+On the other hand, LLMs first and foremost provide predictions about each item.
+While it may be the case that, in producing an item-level prediction, the internal computation of a powerful LLM is informed by computations that incorporate abstract knowledge roughly corresponding to the condition-level, the atomic predictions accessible to the common user are specific to each individual string tested, thus raising concerns about robustness of predictions under perturbations of input prompts \mf{cite}.
+
+These considerations raise two important conceptual and empirical questions.
+First, we need to ask whether the item-level predictions made by LLMs are empirically correct.
+Second, the implied methodological challenge lies in specifying how item-level information can be used, e.g., by different aggregation methods, to make robust predictions at a more abstract level (condition, task, $\dots$).
+To investigate these issues, this paper introduces different ways of building a Bayesian probabilistic model around core predictor values derived from various LLMs.
 We then use the standard tools of Bayesian data analysis, to fit and check the resulting statistical models based on data from human multiple-choice experiments.
 Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single case study of pragmatic language production and interpretation, so-called reference games \mf{references}, for which a widely used probabilistic cognitive model exists, the Rational Speech Act model \citep{FrankGoodman2012:Predicting-Prag}, so that we can compare a theoretically motivated probabilistic cognitive model (PCM) against a probabilistic model built on top of LLM predictions.
-
-The main results of our investigation are that it is possible to build Bayesian statistical models around LLM-based predictors, and that some but not all ways of doing so yield models that adequately predict the human data.
-More concretely, our investigation highlights that LLMs are by design fundamentally different from common statistical or probabilistic cognitive models, in that LLMs make predictions for each individual item, rather than specifying a predictor of central tendency at a more aggregate level (referred to here as the condition-level; see Section~\ref{motivation}).
-This observation raises two important conceptual and empirical questions.
-First, we need to ask whether the item-level predictions made by LLMs are empirically correct.
-Second, the implied methodological challenge lies in specifying how item-level information can be used, e.g., by different aggregation methods, to make robust predictions at a more abstract level (condition, task, $\dots$).
-We find that variability predicted by the LLM at the item-level is not borne out by the human data, that some but not all ways of constructing condition-level predictions are equally good, and that different LLMs as backends may prefer to use different aggregation strategies.
-Our conceptual comparison of LLMs and PCMs also identifies a sense in which the latter are explanatory, while the former are not. \mf{double-check main take away}.
+We find that variability predicted by the LLM at the item-level is generally not borne out by the human data, that some but not all ways of constructing condition-level predictions are equally good, and that different LLMs as backends may prefer to use different aggregation strategies.
 
@@
 Section~\ref{motivation} motivates the investigation in this paper by highlighting relevant differences in the way LLMs and common probabilistic statistical models make predictions about data from forced-choice tasks.
-Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a commonly used, theoretically-motivate probabilistic cognitive model, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
-Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions adequately capture the human data at the item-level.
-Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
-\mf{include section to extend to LLaMA}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Predictions for multiple-choice tasks from PCMs and LLMs}
-\label{motivation}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width = 0.8\textwidth]{00-pics/condition-item-level-prediction.png}
-  \caption{Schematic representation of key conceptual differences between different types of statistical models.}
-  \label{fig:stats-model-types}
-\end{figure}
-
-This section argues that a major conceptual difference between LLMs and PCMs is that they predict data at different levels of abstraction: the item-level and the condition-level, as introduced below.
-Another important difference is that PCMs are usually evaluated on a non-degenerate likelihood function, whereas common practice in LLM-evaluation is to use a winner-takes-all strategy.
-Problems with this latter strategy are highlighted and alternative ways of defining item- and condition-level predictions from LLMs are introduced for subsequent empirical testing.
-
-\paragraph{Item- and condition-level data.}
-Psychological research into the workings of the human mind aims to find generalizable patterns in the way information is processed within or across different domains of cognition.
-Experimental work therefore often compares human performance in different \emph{experimental conditions}, which reflect the general factors that are hypothesized to influence behavior.
-Yet a single experimental condition is often instantiated with different \emph{experimental items}, which are not under scrutiny for any systematic, predictable effect on the observed measurements.
-For example, classical research on human memory \citep{AtkinsonShiffrin1968:Human-memory:-A} investigated the effect of rehearsal on memory consolidation.
-Relevant experiments compare recall with and without rehearsal (experimental conditions), while using different items (words, numbers, etc., to be memorized) in each instantiation of the same memory task.
-Likewise, when studying how hearing a color word can facilitate a same-or-different judgement of color swatches \citep{Rosch1975:The-Nature-of-M}, the main experimental manipulation concerns the typicality of shown color swatches, while the variability between different color words like ``blue'' or ``green'', is less important to this research question and so treated as \emph{item-level variation}.
-Consequently, a typical psychological experiment is mainly interested in assessing behavior at the level of the experimental condition, because that is where the distinctions relevant to the research question reside.
-Nevertheless, each experimental condition can be, and often is, instantiated with different items, variation among which is deemed less relevant to the research question at hand.
-
-\paragraph{Probabilistic cognitive models.}
-Data from human participants for experiments of this kind usually shows some variability between items, and also variability between participants.
-This variability is commonly incorporated in statistical analyses as random stochastic variation, e.g., by using hierarchical (regression) models \citep{Jaeger2008:Categorical-dat,barr2013,SorensenHohensteinb2016:Bayesian-linear}.
-Still, the focus of interest usually remains at the condition-level effects, because it is this more abstract level of behavioral aggregation that is relevant for generalizable theory building.
-Similarly, when analyzing or explaining data from psychological experiments with a probabilistic cognitive model (PCM), the PCM will naturally be set-up to predict data at the level of the relevant experimental manipulation, i.e., at the condition-level.
-Even though item-level (and individual-level) variation can, and sometimes should, be incorporated into probabilistic cognitive modeling \citep[e.g.,][]{NilsonRieskamp2011:Hierarchical-Ba,Lee2011:How-Cognitive-M,ScheibehenneRieskamp2013:Testing-the-Ada}, the focus of cognitive modeling is to explain systematic variation between experimental conditions.
-Therefore, a basic PCM is essentially characterized by a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each logically possible data observation $D_{C}$ for any experimental condition $C$ and numerical parameters $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
-The difference between a theory-driven PCM and a standard, theory-neutral statistical model (like a regression model), which may also define a parameterized likelihood function of condition-level data, is that the PCM's likelihood function is defined in terms that are considered to be conceptually meaningful for the research domain at hand (see Section~\ref{sec:model-pred-from} for a concrete example).
-
-\paragraph{PCMs vs.~LLMs.}
-PCMs, thus construed, differ markedly from LLMs.
-LLMs make predictions about the likelihood of words given a preceding or surrounding text.
-They do not, without further derivation, provide predictions at a conceptually meaningful level of abstraction, similar to experimental conditions.
-In this way, LLMs are not explanatory models in the same sense as PCMs might be: they simply do not operate at a level of generalizable abstraction which is arguably necessary for increasing human understanding of the phenomenon in question \citep{Dellsen2020:Beyond-Explanat,Grimm2021:Understanding}.
-Nevertheless, we can derive predictions from LLM's word-level predictions, first at the item-level and also at the condition-level.
-Whether any of these predictions are adequate is ultimately an empirical question, which can be addressed by subjecting ``LLM-based PCMs'' to the same statistical model checking as the more usual theoretically motivated PCMs.
-
-\paragraph{Predictions from LLMs.}
-Standard benchmark-testing on multiple-choice tasks usually uses a winner-take-all (WTA) strategy \citep[e.g.,][]{srivastava2023-BIGbench} to define accuracy scores.
-The WTA-strategy can also be used to derive probabilistic item-level and condition-level predictions from LLMs.
-Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
-% For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
-Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ words, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$, and each of which .
-For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same response category $r_{i}$ and that $y_{k1}$ always corresponds to the designated \emph{target option}, $r_{1}$, that is assumed to be the ``true'' or to-be-selected goal answer.
-
-The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
-  \mf{TODO: Add quick word on how this can in principle be extended to masked language models.}
-  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
-  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
-  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
-}
-%
-\begin{align*}
-  % S\left( y_{ki}, x_{k} \right) =
-  \text{S}_{ki}
-  % = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right)
-  =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
-\end{align*}
-%
-The WTA-based approach selects answers that maximize the item-level score (with random tie-breaking).
-Concretely, if $B_{k} = \arg \max_{i} \text{S}_{ki}$ is the set of all options that maximize the item-level score for item $k$, then the \emph{item-level prediction} for item $I_{k}$ of the WTA is:
-%
-\begin{align*}
-  P_{\text{item}}^{\text{WTA}} =
-  \begin{cases}
-    \frac{1}{\card{B_{k}}} & \text{if } y_{ki} \in B_{k} \text{, and} \\
-    0                     & \text{otherwise.}
-  \end{cases}
-\end{align*}
-%
-Notice that the item-level WTA-prediction is usually a degenerate probability distribution placing all probability mass on a single choice option.
-
-The \emph{condition-level prediction} of the WTA approach is obtained by averaging over the item-level predictions for each response category:
-%
-\begin{align*}
-  P_{\text{cond}}^{\text{WTA}}\left(r_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{WTA}} \left(y_{ki} \right)\,.
-\end{align*}
-%
-The \emph{WTA-based accuracy} is the probability $P_{\text{cond}}^{\text{WTA}}\left(r_{1} \right)$ of selecting the target option $r_{1}$ over all instances of the task.
-This is the most prevalent measure of the quality of LLM predictions.
-
-% While generally a pragmatic and useful approach, assessing LLM performance with a WTA-based accuracy measure can misleading, because it ignores potentially relevant distributional information.
-% At an abstract level, the problem is that if task performance is categorical (in the most extreme case: binary), somewhere along the path from a numerical item-level score to accuracy a discontinuity has to be introduced.
-% Every such discontinuity bottleneck entails loss of information.
-% As this information might be useful, discontinuity should ideally happen as late as possible; or even better: not at all, unless we can be sure that it is always irrelevant.
-% But since we hardly can be sure that it is irrelevant for all circumstances, we should better assess performance based on a models' full distributional predictions.
-
-The WTA approach implicitly assumes that item-level choices are resolved by a greedy-like choice of the best alternatives.
-This procedure ignores information about relative differences between scores among choice options; information which may be conceptually and practically relevant.
-To generalize to less extreme sampling scenarios, the WTA approach can be generalized to softmax sampling, where the \emph{item-level prediction} of choosing option $y_{ki}$ is:
-%
-\begin{align*}
-P_{\text{item}}^{\text{SM}}\left ( y_{ki} \right ) \propto \expo \left (\alpha \ \text{S}_{ki} \right)\,.
-\end{align*}
-%
-The \emph{condition-level prediction} can be defined analogously to the above by averaging over items:
-\begin{align*}
-  P_{\text{cond}}^{\text{SM}}\left(r_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{ki} \right)\,.
-\end{align*}
-As before, the \emph{softmax-based accuracy}, $P_{\text{cond}}^{\text{SM}}\left(r_{1} \right)$, is the average probability of choosing the designated target response option $r_{1}$.
-
-For $\alpha \rightarrow \infty$, the softmax strategy converges to the WTA strategy, but for finite $\alpha$ the approaches differ.
-Indeed, WTA-based accuracy can even differ qualitatively from softmax-based accuracy, as shown by the following example.
-The example shows that differences between accuracy measures depend on the variation in item-level scores, in particular on the relation between score-ordering and score-differences.
-Note that the example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
-%
-\begin{quote}
-  % \ex \label{bsp:exmpl-diff-WTA-SM}
-  \textbf{Example:}
-  Imagine that there are two options, and that the target option's score is a small $\epsilon$ higher in 80\% of the task's items, and otherwise lower.
-  The WTA-based accuracy is 0.8.
-  This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
-  For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
-  As only the best option will be chosen, that information is irrelevant.
-  But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
-  If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases.
-This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
-\end{quote}
-%
-The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.
-
-When aspiring to predict experimental data from human participants with LLMs, the WTA-based strategy makes very strong predictions at the item-level, where it usually predicts a degenerate probability distribution according to which every data observation for a particular item should fall into the exact same response category.
-This may seem like an \emph{a priori} unlikely prediction, but needs to be tested empirically, just as the item-level predictions by the softmax-based approach should (see Section~\ref{sec:item-level-pred}).
-Whether item-level predictions by an LLM are empirically vindicated or not, is logically independent of whether these item-level predictions serve well as the backbone for deriving condition-level predictions.
-We therefore also need to test, against empirical data, whether different ways of deriving condition-level predictions are adequate (see Section~\ref{llm-predictions-for-reference-games}).
+Section~\ref{experiment-reference-games} describes a simple but non-trivial forced-choice experiment, namely a text-based reference game.
+Section~\ref{sec:model-pred-from} introduces the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag} for the human data from the reference game experiment.
+Section~\ref{sec:item-level-pred} exposes a statistical model for item-level predictions derived from LLM scores and investigates whether this adequately captures the human data at the item-level based on scores from GPT-3.5 \citep{BrownMann2020:Language-Models}.
+Section~\ref{llm-predictions-for-reference-games} discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
+Finally, Section~\ref{sec:gener-other-llm} explores whether previous results generalize to other LLM backends by investigating different versions of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}
 
@@
 
+\mf{metion (somewhere) where item-level effects could be added!}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The RSA model's predictions are defined at the condition-level.
-An LLM, on the other hand, first and foremost makes predictions about each item.
-As mentioned previously, the items of the reference game experiment differ in which levels of features (color, shape, texture) instantiate the structure of the task shown in Figure \ref{fig:ref-game}, as well as the order of presentation of objects and words.
-We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, such as over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options.
-\mf{@PT references from the NLP literature on effects of prompt variation?}
-It therefore becomes an empirical question of whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.
+An LLM is design to predict the next token given an input string.
+This is essentially an item-level prediction: next-token probabilities are specified for concrete strings after a concrete instance of a task, i.e., an item of the task, not for the task as such.
+From these next-token probabilities, we can derive probabilities for multiple-choice answers for each item (this section) and for a condition (next section).
 
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\textwidth]{00-pics/closeness-target-by-alpha-item-level.pdf}
-  \caption{
-    Mean predicted target probability for different values of softmax parameter $\alpha$.
-    The solid black lines show the probability of the target options, predicted for a fixed value of $\alpha$ ($x$-axis), averaged over all items for each condition.
-    The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$.
-    For reference, the target probability under a random choice strategy and under the WTA strategy are shown with dashed lines.
-    For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
+Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
+% For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
+Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ tokens, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
+For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same response category $r_{i}$.
+For the case at hand these categories are: target, competitor, distractor, so that $y_{k1}$ always corresponds to the designated \emph{target option}, $r_{1}$, that is the ``gold standard'' answer.
+
+The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
+  \mf{TODO: Add quick word on how this can in principle be extended to masked language models.}
+  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
+  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
+  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
+  \mf{check whether this is correct also after new LLaMA results.}
 }
-  \label{fig:closeness-target-item-level}
-\end{figure}
+%
+\begin{align*}
+  % S\left( y_{ki}, x_{k} \right) =
+  \text{S}_{ki}
+  % = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right)
+  =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
+\end{align*}
+%
 
-As described in Section~\ref{motivation}, the standard procedure for assessing LLM predictions in multi-choice tasks is to use a winner-takes-all (WTA) strategy that just picks the best option for each item.
-But since this strategy has conceptual and empirical weaknesses, we define the item-level LLM predictions using the more general softmax-approach, which contains the WTA-based approach as a limiting case, repeated here from above with slightly updated notation:
+Based on each option's score, we can define the \emph{item-level prediction} of choosing option $y_{ki}$ in term of soft-maximization as:
 %
@@
 %
-Item-level scores $\text{S}_{ki}$, as defined in Section~\ref{motivation}, were obtained from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} .
+Notice that the usual item-level prediction used in benchmark testing is actually a ``winner-takes-all'' strategy, but this is just a special case of the above for $\alpha_{C} \rightarrow \infty$.
+Here, the $\alpha_{C}$ parameter corresponds to inverse temperature, so that $\alpha_{C} \rightarrow \infty$ corresponds to greedy sampling with temperature zero.
+Like in the RSA model, we use an independent soft-max parameter for each condition.
+
+As mentioned previously, the items of the reference game experiment from Section~\ref{experiment-reference-games} differ in which levels of features (color, shape, texture) instantiate the structure of the task shown in Figure \ref{fig:ref-game}, as well as the order of presentation of objects and words.
+We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, such as over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options.
+\mf{references prompt variation}
+It therefore becomes an empirical question of whether item-level predictions from LLMs provide a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.
+
+To address this question, we assessed item-level scores $\text{S}_{ki}$ from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
 % \footnote{
@@
 Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
-Allowing for random errors with probability $\epsilon_{C}$, as before,:
+Similar to the RSA model fit, we allow for random errors with probability $\epsilon_{C}$:
 %
@@
 %
-The resulting likelihood function for item-level data with item-level predictions is:
+The resulting likelihood function for item-level data with item-level predictions is, where priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.:
 %
@@
 %
-Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.
 
-Summary statistics for samples from the posterior distribution over parameters are shown in Table~\ref{tab:sumStats}.
+Summary statistics for samples from the posterior distribution over parameters are shown in Table~\ref{tab:sumStats} (rows 4 and 5).
 Credible values of $\alpha$ parameters are comparatively low for the LLM-based model, suggesting that the predicted scores have rather large differences, which have to be compensated for by the softmax link-function to adequately fit the data.
-Figure~\ref{fig:closeness-target-item-level} shows the mean predicted target probabilities for different parameter values of $\alpha$, highlighting that for posterior credible values, the predictions are clearly distinct from the predictions of a WTA strategy.
+Figure~\ref{fig:closeness-target-item-level} shows the mean of the probabilities for the highest scoring answer category, averaged over all items, for different parameter values of $\alpha$, highlighting that for \textit{a posteriori} credible values of $\alpha$ (gray shaded areas), the predictions are clearly distinct from the predictions of a WTA strategy (which assigns probability 1 to the highest scoring option for each item).
 This suggests that the WTA strategy, which is the special limiting case for $\alpha \rightarrow \infty$, provides a substantially worse fit for item-level choices than those provided by less extreme values of $\alpha$.
@@
   \centering
+  \includegraphics[width=\textwidth]{00-pics/GPT-closeness-target-by-alpha-item-level.pdf}
+  \caption{
+    Predicted probability of highest-scoring answer category, averaged over items, for different values of softmax parameter $\alpha$ (black lines).
+    The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$, and the implied probabilistic prediction.
+    For reference, the target probability under a random choice strategy and under the WTA strategy are shown with dashed lines.
+    For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
+  }
+  \label{fig:closeness-target-item-level}
+\end{figure}
 
-  \includegraphics[width = 0.9\linewidth]{00-pics/item-combined-obs-pred.pdf}
+\begin{figure}[t]
+  \centering
+
+  \includegraphics[width = 0.9\linewidth]{00-pics/GPT-item-combined-obs-pred.pdf}
 
@@
 
-Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
-Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
+Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated LLM scores are inadequate predictors of the human data.
+To corroborate this conclusion, Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
 Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
@@
 These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
-Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by experimental data.
+Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by the current experimental data.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-\label{conclusion}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-While the common practice in evaluating the capabilities of LLMs is based on accuracy averaged over large collections of data, this work took the alternative route to explore what we learn if we subjected LLMs to the same routines and strong demands on distributional quality of fit to human data as we normally do for probabilistic cognitive models.
-Taking this perspective, a main contribution of this paper is methodological, showing how statistical model criticism can be used for LLMs in the first place, and, more specifically, how it can be insightful in the detailed assessment of how LLMs might or might not be able to predict human data at the degree of accuracy that we would normally aspire for when dealing with explanatory statistical models in cognitive science.
-Applying this comparative approach to a minimal, but non-trivial data set, we find that the LLM predictions on a per-item level predict variance that is not attested in the human data.
-From several candidate predictor measures for aggregate condition-level data, only one was not refuted by the human data, but this was one that relied on the empirically implausible WTA-strategy at the item-level, incidentally the same strategy that is commonly used in accuracy-based benchmark testing \citep{srivastava2023-BIGbench}.
-
-\paragraph{Explanatory power.}
-The comparison of LLMs and PCMs led to a simple but consequential observation, namely that the atomic predictions or LLMs are at the item level, while explanatory PCMs usually make predictions at a more abstract level of conceptualization, similar to the level of an experimental condition.
-This is one sense in which LLMs may be felt to be less, or not at all, explanatory.
-They do not offer, at least not directly, a human-comprehensible compression of reality into a \emph{kind} of response pattern, over and beyond making a predicting for each particular situation.
-As this kind of compression is arguably important for a sense of understanding \citep{Dellsen2020:Beyond-Explanat,Grimm2021:Understanding}, the direct comparison of LLMs with common practices in experimental psychology and with probabilistic cognitive models, provides an interesting perspective on why LLMs are often felt to be lacking in explanatory power.
-
-This perspective on the explanatory role of LLMs goes beyond the factors of performance, indirect support and parsimony identified by \citet{SortNoopDeemtervan-Deemter2023:Dimensions-of-E}.
-It is also subtly different from considerations of an LLMs ability to generalize \citep{HupkesDankers2020:Compositionalit}.
-It rather suggests that \emph{transferability} is a dimension to ``explanatory power'' that is important as well.
-Imagine that models $M$ and $M'$ have been designed for and trained on data from a situation $S$, but need to be applied to a different situation $T$.
-Assume that for model $M$ the only way to make prediction for $T$ is to collect data pertaining to $T$, and either retrain or fine-tune the model.
-In contrast, for model $M'$ can make predictions for $T$ without novel data collection by recognizing a meaningful difference between $S$ and $T$ and consequently manually changing parameter values or model-internal mechanics to accommodate for this change.
-In that case, model $M'$ would be more transferable than model $M$.
-For example, if we change the experimental setting for a reference game to consist of data from a special population, such as very young children or language-impaired adults, a potentially reasonable architectural change to a model like the RSA model is to consider differences in the sets of alternative utterances for the speaker \citep[e.g.][]{Noveck2001:When-Children-a}.
-Even though this is only a vague explication of a notion of transferability, it suffices to corroborate the intuition that models like PCMs, which are designed to operate at a higher level of conceptual abstraction, will often appear more transferable than models like LLMs, which make predictions not for \emph{kinds} of situations but for \emph{particular} situations.
-Whether any given model's transfer-ability is correct, is an orthogonal empirical question.
-It is also an empirical question, exactly to which extent and in which areas LLMs are (not) transferable, especially if we consider prompting a transfer strategy \citep{LiuLiu2022:Generated-Knowl,XieRaghunathan2022:An-Explanation-}.
-In any case, the comparison between LLMs and PCMs started here suggests systematic research into the transfer-ability and explanatory value of LLMs, e.g., by prompting strategies that are empirically insightful or by their use in composite neuro-symbolic models that implement theoretically-meaningful conceptual differences between types of situations.
-
-\paragraph{Variability in predictions and performance measures.}
-Despite being anchored in a small, but detailed case-study, the fact that different plausible methods of aggregating item-level information led to condition-level predictions of variable quality is worrisome.
-Similarly, we argued that the wide-spread reliance on a winner-takes-all strategy might be inconsistent with the actual use practices of LLMs, which may not always rely on a temperature-zero sampling strategy (see Section~\ref{motivation}).
-In conclusion, research on LLM should systematically study the conceptual and empirical consequences of seemingly minor decisions in evaluation or application settings.
-Variability of performance measures also comes with a well-known risk for robust and reproducible science.
-The more researcher degrees of freedom there are, the higher the risk of false results, even in the absence of intentions to mislead \citep[e.g.][]{Ioannidis2005:Why-Most-Publis,Chambers2017:The-Seven-Deadl}.
-The comparison of LLMs with PCMs here raises awareness for the issue of robust research methods also in NLP \citep{WielingRawee2018:Reproducibility}.
-
-The work presented here also contains considerable researcher degrees of freedom.
-We only considered one out of several conceivable ways of carving a parameterized likelihood function from LLM-based predictors.
-If future work would contribute to systematically exploring these, the main goal of this paper would have been met, which is to raise awareness for the possibility, perhaps even necessity, to scrutinize LLMs at the same level of rigorous detail as other models in cognitive science.
-
-\paragraph{Human-like predictions from LLM.}
-The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
-Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
-Based on the data set and the detailed analyses conducted here, it seems not infeasible to use numerical predictions from LLMs are part of predictive probabilistic models.
-But, ideally, low-level variation, such as from order of presentation or similar ``nuisance variables,'' should be averaged out or taken into account in some way or other, at least as long as we do not understand what causes this variation in the predictions of models and further research that investigates when exactly this variation accords with empirically observed patterns.
-In sum, we conjecture that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is taken into account.
-
-\paragraph{No substitute for human subjects.}
-Looking at LLM-derived predictions for human data from the perspective of cognitive modeling highlights the fact that LLMs predict item-level variation, but not subject-level variation, which is common in human data.
-We may consider variation introduced via softmax / temperature as analogous to between-human variation, but this likely falls short of reality, where pronounced differences in answer behavior may surface.
-For the particular case of pragmatic language use, prior research has shown that individual participants have markedly different behavioral profiles, often consistently behaving like literal language users or more sophisticated language users \citep[e.g.,][]{NieuwlandDitman2010:On-the-incremen,FrankeDegen2015:Reasoning-in-Re,SpychalskaKontinen2016:Investigating-s}.
-It is an open question whether predictions from LLMs reflect the same kind of variation.
-The results presented here recommend skepticism.
-We therefore side with the cautious voices that do not recommend replacing human participants with LLMs in psychological research \citep{DillionTandon2023:Can-AI-language,HardingDAlessandro2023:AI-language-mod}.
-In contrast, there is a challenge for LLM research.
-The fact that aggregate predictions can track aggregate human behaviour means that the variance in on both sides is washed out to achieve a similar result.
-This raises the issue of finding the systematic differences between LLMs and human answerers.
-The task then would be to find cases where the differences do \emph{not} wash out and ask: What if anything do these cases share?
-
-\paragraph{Limitations and follow-up work.}
-The scope of our experimental investigation was deliberately small but perspicuous.
-Focusing on the methodological contributions and details in performance assessments, we focused on a minimal non-trivial case study where we could triangulate LLMs, PCMs and human data.
-This work may therefore serve as starting point for a wider investigation of more complex data sets and case studies in which LLMs are analyzed as and directly compared to explanatory PCMs.
-
-\printbibliography[heading=bibintoc]
-
-\newpage
-\appendix
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results from LLaMA2}
-\label{sec:LLaMA-results}
+\section{Generalizing to other LLM backends}
+\label{sec:gener-other-llm}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+\label{sec:conclusion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+While the common practice in evaluating the capabilities of LLMs is based on accuracy averaged over large collections of data, this work took the alternative route to explore what we learn if we subjected LLMs to the same routines and strong demands on distributional quality of fit to human data as we normally do for probabilistic cognitive models.
+Taking this perspective, a main contribution of this paper is methodological, showing how statistical model criticism can be used for LLMs in the first place, and, more specifically, how it can be insightful in the detailed assessment of how LLMs might or might not be able to predict human data at the degree of accuracy that we would normally aspire for when dealing with explanatory statistical models in cognitive science.
+Applying this comparative approach to a minimal, but non-trivial data set, we find that the LLM predictions on a per-item level predict variance that is not attested in the human data.
+From several candidate predictor measures for aggregate condition-level data, only one was not refuted by the human data, but this was one that relied on the empirically implausible WTA-strategy at the item-level, incidentally the same strategy that is commonly used in accuracy-based benchmark testing \citep{srivastava2023-BIGbench}.
+
+\paragraph{Explanatory power.}
+The comparison of LLMs and PCMs led to a simple but consequential observation, namely that the atomic predictions or LLMs are at the item level, while explanatory PCMs usually make predictions at a more abstract level of conceptualization, similar to the level of an experimental condition.
+This is one sense in which LLMs may be felt to be less, or not at all, explanatory.
+They do not offer, at least not directly, a human-comprehensible compression of reality into a \emph{kind} of response pattern, over and beyond making a predicting for each particular situation.
+As this kind of compression is arguably important for a sense of understanding \citep{Dellsen2020:Beyond-Explanat,Grimm2021:Understanding}, the direct comparison of LLMs with common practices in experimental psychology and with probabilistic cognitive models, provides an interesting perspective on why LLMs are often felt to be lacking in explanatory power.
+
+This perspective on the explanatory role of LLMs goes beyond the factors of performance, indirect support and parsimony identified by \citet{SortNoopDeemtervan-Deemter2023:Dimensions-of-E}.
+It is also subtly different from considerations of an LLMs ability to generalize \citep{HupkesDankers2020:Compositionalit}.
+It rather suggests that \emph{transferability} is a dimension to ``explanatory power'' that is important as well.
+Imagine that models $M$ and $M'$ have been designed for and trained on data from a situation $S$, but need to be applied to a different situation $T$.
+Assume that for model $M$ the only way to make prediction for $T$ is to collect data pertaining to $T$, and either retrain or fine-tune the model.
+In contrast, for model $M'$ can make predictions for $T$ without novel data collection by recognizing a meaningful difference between $S$ and $T$ and consequently manually changing parameter values or model-internal mechanics to accommodate for this change.
+In that case, model $M'$ would be more transferable than model $M$.
+For example, if we change the experimental setting for a reference game to consist of data from a special population, such as very young children or language-impaired adults, a potentially reasonable architectural change to a model like the RSA model is to consider differences in the sets of alternative utterances for the speaker \citep[e.g.][]{Noveck2001:When-Children-a}.
+Even though this is only a vague explication of a notion of transferability, it suffices to corroborate the intuition that models like PCMs, which are designed to operate at a higher level of conceptual abstraction, will often appear more transferable than models like LLMs, which make predictions not for \emph{kinds} of situations but for \emph{particular} situations.
+Whether any given model's transfer-ability is correct, is an orthogonal empirical question.
+It is also an empirical question, exactly to which extent and in which areas LLMs are (not) transferable, especially if we consider prompting a transfer strategy \citep{LiuLiu2022:Generated-Knowl,XieRaghunathan2022:An-Explanation-}.
+In any case, the comparison between LLMs and PCMs started here suggests systematic research into the transfer-ability and explanatory value of LLMs, e.g., by prompting strategies that are empirically insightful or by their use in composite neuro-symbolic models that implement theoretically-meaningful conceptual differences between types of situations.
+
+% SNIPPET from previous Section 2
+% \paragraph{PCMs vs.~LLMs.}
+% PCMs, thus construed, differ markedly from LLMs.
+% LLMs make predictions about the likelihood of words given a preceding or surrounding text.
+% They do not, without further derivation, provide predictions at a conceptually meaningful level of abstraction, similar to experimental conditions.
+% In this way, LLMs are not explanatory models in the same sense as PCMs might be: they simply do not operate at a level of generalizable abstraction which is arguably necessary for increasing human understanding of the phenomenon in question \citep{Dellsen2020:Beyond-Explanat,Grimm2021:Understanding}.
+% Nevertheless, we can derive predictions from LLM's word-level predictions, first at the item-level and also at the condition-level.
+% Whether any of these predictions are adequate is ultimately an empirical question, which can be addressed by subjecting ``LLM-based PCMs'' to the same statistical model checking as the more usual theoretically motivated PCMs.
+
+\paragraph{Variability in predictions and performance measures.}
+Despite being anchored in a small, but detailed case-study, the fact that different plausible methods of aggregating item-level information led to condition-level predictions of variable quality is worrisome.
+Similarly, we argued that the wide-spread reliance on a winner-takes-all strategy might be inconsistent with the actual use practices of LLMs, which may not always rely on a temperature-zero sampling strategy (see Section~\ref{motivation}).
+In conclusion, research on LLM should systematically study the conceptual and empirical consequences of seemingly minor decisions in evaluation or application settings.
+Variability of performance measures also comes with a well-known risk for robust and reproducible science.
+The more researcher degrees of freedom there are, the higher the risk of false results, even in the absence of intentions to mislead \citep[e.g.][]{Ioannidis2005:Why-Most-Publis,Chambers2017:The-Seven-Deadl}.
+The comparison of LLMs with PCMs here raises awareness for the issue of robust research methods also in NLP \citep{WielingRawee2018:Reproducibility}.
+
+The work presented here also contains considerable researcher degrees of freedom.
+We only considered one out of several conceivable ways of carving a parameterized likelihood function from LLM-based predictors.
+If future work would contribute to systematically exploring these, the main goal of this paper would have been met, which is to raise awareness for the possibility, perhaps even necessity, to scrutinize LLMs at the same level of rigorous detail as other models in cognitive science.
+
+\paragraph{Human-like predictions from LLM.}
+The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
+Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
+Based on the data set and the detailed analyses conducted here, it seems not infeasible to use numerical predictions from LLMs are part of predictive probabilistic models.
+But, ideally, low-level variation, such as from order of presentation or similar ``nuisance variables,'' should be averaged out or taken into account in some way or other, at least as long as we do not understand what causes this variation in the predictions of models and further research that investigates when exactly this variation accords with empirically observed patterns.
+In sum, we conjecture that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is taken into account.
+
+\paragraph{No substitute for human subjects.}
+Looking at LLM-derived predictions for human data from the perspective of cognitive modeling highlights the fact that LLMs predict item-level variation, but not subject-level variation, which is common in human data.
+We may consider variation introduced via softmax / temperature as analogous to between-human variation, but this likely falls short of reality, where pronounced differences in answer behavior may surface.
+For the particular case of pragmatic language use, prior research has shown that individual participants have markedly different behavioral profiles, often consistently behaving like literal language users or more sophisticated language users \citep[e.g.,][]{NieuwlandDitman2010:On-the-incremen,FrankeDegen2015:Reasoning-in-Re,SpychalskaKontinen2016:Investigating-s}.
+It is an open question whether predictions from LLMs reflect the same kind of variation.
+The results presented here recommend skepticism.
+We therefore side with the cautious voices that do not recommend replacing human participants with LLMs in psychological research \citep{DillionTandon2023:Can-AI-language,HardingDAlessandro2023:AI-language-mod}.
+In contrast, there is a challenge for LLM research.
+The fact that aggregate predictions can track aggregate human behaviour means that the variance in on both sides is washed out to achieve a similar result.
+This raises the issue of finding the systematic differences between LLMs and human answerers.
+The task then would be to find cases where the differences do \emph{not} wash out and ask: What if anything do these cases share?
+
+\paragraph{Limitations and follow-up work.}
+The scope of our experimental investigation was deliberately small but perspicuous.
+Focusing on the methodological contributions and details in performance assessments, we focused on a minimal non-trivial case study where we could triangulate LLMs, PCMs and human data.
+This work may therefore serve as starting point for a wider investigation of more complex data sets and case studies in which LLMs are analyzed as and directly compared to explanatory PCMs.
+
+\printbibliography[heading=bibintoc]
+
+\newpage
+\appendix
 

### 2024-03-29 04-paper/paper-LLM-probmod.tex
updated §5, added LLaMA results intepretation

@@
   In the current set-up the response type ``distractor'' has two instantiations in the production condition.
-Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
+Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option.}
 Similar to the RSA model fit, we allow for random errors with probability $\epsilon_{C}$:
@@
 There are many ways of averaging item-level information.
-Figure~\ref{fig:measures-overview} shows three salient approaches, which differ in what the underlying item-level measure is: the raw scores $\text{S}_{ki}$, the item-level probability derived from it (as introduced in Section~\ref{motivation} and used in Section~\ref{sec:item-level-pred}), or the predictions from the winner-takes-all strategy commonly used in benchmark testing (see Section~\ref{motivation}).
+Figure~\ref{fig:measures-overview} shows three salient approaches, which differ in what the underlying item-level measure is: the raw scores $\text{S}_{ki}$, the item-level probability derived from it (as used in Section~\ref{sec:item-level-pred}), or the predictions from the winner-takes-all (WTA) strategy commonly used in benchmark testing.
 
@@
 %
-Finally, the \emph{average-WTA predictor} considers the prediction of the WTA strategy as the basic item-level unit to aggregate over.
+Finally, the \emph{average-WTA predictor} considers the prediction of the WTA strategy (a softmax with $\alpha \rightarrow \infty$) as the basic item-level unit to aggregate over.
 To add parameterized scaling to this method, a power-law transformation is a natural choice:
@@
 \end{align*}
-where $P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) = \lim_{\alpha} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha)$.
+where $P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) = \lim_{\alpha \rightarrow \infty} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha)$.
 \bigskip
 
-The \emph{average-scores} and \emph{average-probabilities} predictions are equivalent if there is only one item, in which case the prediction of the \emph{average-WTA} method is the special case of $\alpha \rightarrow \infty$.
+The \emph{average-scores} and \emph{average-probabilities} predictors are equivalent if there is only one item, in which case the prediction of the \emph{average-WTA} method is the special case of $\alpha \rightarrow \infty$.
 For cases with more than one item, the predictions of the three predictors are not guaranteed to be the same.
 Conceptually, the \emph{average-probabilities} and the \emph{average-WTA} predictors, but not the \emph{average-scores} predictor, are compatible with a picture in which condition-level predictions result from the actual predictions at the item level.
-The difference between the \emph{average-probabilities} and the \emph{average-WTA} predictor is that the latter is an aggregate of an item-level predictor that is demonstrably bad, as shown in the previous Section~\ref{sec:item-level-pred}.
+However, based on the results from the previous section, the item-level predictions for the WTA strategy are demonstrably incongruent with the human data.
+In general, the average-probability and average-WTA predictors can lead to qualitatively different results about task-level accuracy (see Appendix~\ref{sec:aver-prob-vs}).
+
+\mf{UP TO HERE}
 
@@
 However, for the interpretation condition, the value $\alpha_{c}=1$ is clearly outside the range of credible parameter values, so that a simple recipe like ``always average WTA-responses without transformation'' is not a viable strategy for good condition-level predictions in general.
-It is also worth noting that for the \emph{average-probability} model values of $\alpha_{c}$ close to one are virtually indistinguishable from the item-level predictions of the WTA strategy (see Figure~\ref{fig:closeness-target-item-level}).
+It is also worth noting that for the \emph{average-probability} model values of $\alpha_{c}$ of around five and above are virtually indistinguishable from the item-level predictions of the WTA strategy (see Figure~\ref{fig:closeness-target-item-level}).
 This suggests that, for the production data, aggregation of item-level predictions from the WTA-strategy gives good condition-level predictions.
-Since the production data is also very biased towards almost exclusively target choices by human participants, this is not too surprising.
 
@@
 To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
-Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} model fails on the interpretation data, and is at most borderline compatible with the production data; that the \emph{average-probabilities} model is able to reproduce the production data, but fails on the interpretation data; and that only the \emph{average-WTA} model does not fail to capture the data from both conditions.
+Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} and the \emph{average-probabilities} model are able to reproduce the production data, but fail on the interpretation data; and that only the \emph{average-WTA} model does not fail to capture the data from both conditions.
 
@@
 On the other hand, there is a stark contrast between the results of item-level and condition-level data.
-The model that was able to properly fit the condition-level data builds on predictions for item-level choice that are clearly incompatible with the human item-level data.
+The model that was able to properly fit the condition-level data builds on predictions for item-level choices that are clearly incompatible with the human item-level data.
 The model seems to be ``right for the wrong reasons.''
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+All previous results where obtained for LMM predictions derived from GPT-3.5 turbo.
+To investigate whether key results replicate for other LLMs and scales, we ran the exact same analyses for item- and condition-level data also for scores derived from variants of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}, in particular the 7, 13 and 70 billion parameter versions of the versions fine-tuned with human feedback only (marked as `hf' in the following) and based on human feedback and chat data (marked as `chat').
+Table~\ref{tab:sumStats-LLaMA2} shows the relevant summary statistics and Bayesian posterior predictive $p$-values.
+Figure~\ref{fig:LLaMA-results-vPPC} shows visual posterior predictive checks for the condition-level data.
+The plots show the differences between human counts for each condition and option and the models' posterior predictions.
+We find that the average-WTA predictor is consistently able to capture the data from the production condition except for one of the six models (LLaMA2-chat 70B).
+However, the average-WTA predictor only recovers the interpretation data for only one of the six models (LLaMA2-chat 13B).
+Interestingly, for this LLM as backend for scoring, also the average-probability predictor is able to recover the interpretation data, unlike for the GPT-3.5 backend.
+
+\mf{TODO: item-level; check estimated alphas for condition-level!?}
+
+In sum, we find general support for the previous conclusion that the average-WTA predictor is most successful in capturing the condition-level data, but we also find variability in which method of aggregation works well with which LLM as scoring model.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Average-probability vs.~average-WTA for accuracy scores}
+\label{sec:aver-prob-vs}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Standard benchmark testing looks at \emph{task accuracy}, defined as the probability of selecting the ``gold standard'' response, usually based on a WTA selection of the highest scoring option.
+We can generalize this and define the \emph{softmax-based accuracy}, $P_{\text{cond}}^{\text{SM}}\left(r_{1} \right)$, as the average probability of choosing the designated target response option $r_{1}$ for the \emph{condition-level soft-max prediction}:
+\begin{align*}
+  P_{\text{cond}}^{\text{SM}}\left(r_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{ki} \right)\,.
+\end{align*}
+As before, the average-WTA task accuracy is the special case of $\alpha \rightarrow\infty$.
+
+Indeed, WTA-based accuracy can differ qualitatively from softmax-based accuracy, as shown by the following example.
+The example shows that differences between accuracy measures depend on the variation in item-level scores, in particular on the relation between score-ordering and score-differences.
+Note that the example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
+%
+\begin{quote}
+  % \ex \label{bsp:exmpl-diff-WTA-SM}
+  \textbf{Example:}
+  Imagine that there are two options, and that the target option's score is a small $\epsilon$ higher in 80\% of the task's items, and otherwise lower.
+  The WTA-based accuracy is 0.8.
+  This number is useful as a performance measure for applications in which the LLM is used in exactly the way the WTA strategy describes, e.g., any implementation which is outcome equivalent to greedy decoding with rejection sampling on a domain that contains only the available options.
+  For such a case, it never matters how much worse the goal answer is scored in the 20\% of the cases where it is not the maximum.
+  As only the best option will be chosen, that information is irrelevant.
+  But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
+  If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases.
+This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
+\end{quote}
+%
+The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.
+
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-03-30 04-paper/paper-LLM-probmod.tex
completed §6 and revised appendix C

@@
 
-\mf{UP TO HERE}
-
 Using the same approach as for the RSA model in Section~\ref{sec:model-pred-from}, we can build likelihood functions for condition-level data around the three predictors introduced above.
@@
 
-All previous results where obtained for LMM predictions derived from GPT-3.5 turbo.
-To investigate whether key results replicate for other LLMs and scales, we ran the exact same analyses for item- and condition-level data also for scores derived from variants of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}, in particular the 7, 13 and 70 billion parameter versions of the versions fine-tuned with human feedback only (marked as `hf' in the following) and based on human feedback and chat data (marked as `chat').
-Table~\ref{tab:sumStats-LLaMA2} shows the relevant summary statistics and Bayesian posterior predictive $p$-values.
-Figure~\ref{fig:LLaMA-results-vPPC} shows visual posterior predictive checks for the condition-level data.
-The plots show the differences between human counts for each condition and option and the models' posterior predictions.
-We find that the average-WTA predictor is consistently able to capture the data from the production condition except for one of the six models (LLaMA2-chat 70B).
-However, the average-WTA predictor only recovers the interpretation data for only one of the six models (LLaMA2-chat 13B).
-Interestingly, for this LLM as backend for scoring, also the average-probability predictor is able to recover the interpretation data, unlike for the GPT-3.5 backend.
-
-\mf{TODO: item-level; check estimated alphas for condition-level!?}
-
-In sum, we find general support for the previous conclusion that the average-WTA predictor is most successful in capturing the condition-level data, but we also find variability in which method of aggregation works well with which LLM as scoring model.
-
 \begin{figure}[t]
@@
 \end{figure}
+
+All previous results where obtained for LMM predictions derived from GPT-3.5 turbo.
+To investigate whether key results replicate for other LLMs and scales, we ran the same analyses also for scores derived from variants of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}, in particular the 7, 13 and 70 billion parameter versions of the models fine-tuned with human feedback only (marked as `hf' in the following) and based on human feedback and chat data (marked as `chat').
+Table~\ref{tab:sumStats-LLaMA2} shows the relevant summary statistics and Bayesian posterior predictive $p$-values.
+Figure~\ref{fig:LLaMA-results-vPPC} shows visual posterior predictive checks for the condition-level data.
+The plots show the \emph{differences} between observed counts and the models' posterior predictions for expected counts for each condition and response option.
+We find that the average-WTA predictor is consistently able to capture the data from the production condition, except for one of the six models (LLaMA2-chat 70B).
+However, the average-WTA predictor recovers the interpretation data for only one of the six models (LLaMA2-chat 13B, see also Table~\ref{tab:sumStats-LLaMA2}).
+Interestingly, for LLaMA2-chat 13B also the average-probability predictor is able to recover the interpretation data, unlike for the GPT-3.5.
+As for GPT-3.5, no model captured the item-level variance observed in the human data.
+In sum, we find general support for the previous conclusions that the average-WTA predictor is most successful in capturing the condition-level data, and therefore that, where models manage to recover the condition-level data, they seem to be ``right for the wrong reasons.''
+Additionally, we also find variability in which method of aggregation works well with which LLM as scoring model.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Average-probability vs.~average-WTA for accuracy scores}
+\section{Excursion: Accuracy scores from average-probability vs.~average-WTA predictors}
 \label{sec:aver-prob-vs}
@@
 
-Standard benchmark testing looks at \emph{task accuracy}, defined as the probability of selecting the ``gold standard'' response, usually based on a WTA selection of the highest scoring option.
-We can generalize this and define the \emph{softmax-based accuracy}, $P_{\text{cond}}^{\text{SM}}\left(r_{1} \right)$, as the average probability of choosing the designated target response option $r_{1}$ for the \emph{condition-level soft-max prediction}:
+Standard benchmark testing looks at \emph{task accuracy}, defined as the probability of selecting the ``gold standard'' response, usually based on ``winner-takes-all'' (WTA) selection of the highest scoring option.
+We can generalize this and define the \emph{softmax-based accuracy} as the average probability of choosing the designated target response option $r_{1}$ for the \emph{condition-level soft-max prediction} (notation as defined in main text):
 \begin{align*}
-  P_{\text{cond}}^{\text{SM}}\left(r_{i} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{ki} \right)\,.
+  P_{\text{cond}}^{\text{SM}}\left(r_{1} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{k1} \right)\,.
 \end{align*}
-As before, the average-WTA task accuracy is the special case of $\alpha \rightarrow\infty$.
+The WTA-based accuracy (the standard measure in most benchmark testing) is the special case of $\alpha \rightarrow\infty$.
 
-Indeed, WTA-based accuracy can differ qualitatively from softmax-based accuracy, as shown by the following example.
+The WTA-based accuracy can differ qualitatively from the more general softmax-based accuracy, as shown by the following example.
 The example shows that differences between accuracy measures depend on the variation in item-level scores, in particular on the relation between score-ordering and score-differences.

### 2024-03-30 04-paper/paper-LLM-probmod.tex
revised conclusions

@@
   \includegraphics[width = 0.8\textwidth]{00-pics/condition-item-level-prediction.png}
-  \caption{Schematic representation of key conceptual differences between different types of statistical models.}
+  \caption{Schematic representation of key conceptual differences between different types of statistical models. \mf{picture is cropped}}
   \label{fig:stats-model-types}
@@
 
-While the common practice in evaluating the capabilities of LLMs is based on accuracy averaged over large collections of data, this work took the alternative route to explore what we learn if we subjected LLMs to the same routines and strong demands on distributional quality of fit to human data as we normally do for probabilistic cognitive models.
-Taking this perspective, a main contribution of this paper is methodological, showing how statistical model criticism can be used for LLMs in the first place, and, more specifically, how it can be insightful in the detailed assessment of how LLMs might or might not be able to predict human data at the degree of accuracy that we would normally aspire for when dealing with explanatory statistical models in cognitive science.
+While the common practice in evaluating the capabilities of LLMs is based on accuracy averaged over large collections of data, this work took the alternative route to explore what we learn if we subjected LLMs to the same routines and strong demands on distributional quality of fit to human data as we normally do for statistical or probabilistic cognitive models.
+Knowledge of the adequacy of LLMs' full distributional predictions for simple tasks that require human like judgement or decisions is important to gauge in how far LLMs can be trusted to provide such information in applications such as hybrid, neuro-symbolic models \mf{refs}.
+A main contribution of this paper is methodological, showing how statistical model criticism can be used for LLMs in the first place, and, more specifically, how it can be insightful in the detailed assessment of how LLMs might or might not be able to predict human data at the degree of accuracy that we would normally aspire for when dealing with explanatory statistical models in cognitive science.
 Applying this comparative approach to a minimal, but non-trivial data set, we find that the LLM predictions on a per-item level predict variance that is not attested in the human data.
@@
 \paragraph{Explanatory power.}
-The comparison of LLMs and PCMs led to a simple but consequential observation, namely that the atomic predictions or LLMs are at the item level, while explanatory PCMs usually make predictions at a more abstract level of conceptualization, similar to the level of an experimental condition.
-This is one sense in which LLMs may be felt to be less, or not at all, explanatory.
-They do not offer, at least not directly, a human-comprehensible compression of reality into a \emph{kind} of response pattern, over and beyond making a predicting for each particular situation.
+A basic observation brought to the foreground by our approach is that LLMs' atomic predictions are for individual items and that some aggregation method is needed to derive more abstract, condition- or task-level predictions.
+This is potential one sense in which LLMs may be felt to be less, or not at all, explanatory.
+They do not offer, at least not directly, a human-comprehensible compression of reality into a \emph{kind} of response pattern, over and beyond making a prediction for each particular situation.
 As this kind of compression is arguably important for a sense of understanding \citep{Dellsen2020:Beyond-Explanat,Grimm2021:Understanding}, the direct comparison of LLMs with common practices in experimental psychology and with probabilistic cognitive models, provides an interesting perspective on why LLMs are often felt to be lacking in explanatory power.
@@
 This perspective on the explanatory role of LLMs goes beyond the factors of performance, indirect support and parsimony identified by \citet{SortNoopDeemtervan-Deemter2023:Dimensions-of-E}.
-It is also subtly different from considerations of an LLMs ability to generalize \citep{HupkesDankers2020:Compositionalit}.
+It is also subtly different from considerations of an LLM's ability to generalize \citep{HupkesDankers2020:Compositionalit}.
 It rather suggests that \emph{transferability} is a dimension to ``explanatory power'' that is important as well.
 Imagine that models $M$ and $M'$ have been designed for and trained on data from a situation $S$, but need to be applied to a different situation $T$.
-Assume that for model $M$ the only way to make prediction for $T$ is to collect data pertaining to $T$, and either retrain or fine-tune the model.
-In contrast, for model $M'$ can make predictions for $T$ without novel data collection by recognizing a meaningful difference between $S$ and $T$ and consequently manually changing parameter values or model-internal mechanics to accommodate for this change.
+Assume that for model $M$ the only way to make predictions for $T$ is to collect data pertaining to $T$, and either retrain or fine-tune the model.
+In contrast, model $M'$ can make predictions for $T$ without novel data collection by recognizing a meaningful difference between $S$ and $T$ and consequently manually changing parameter values or model-internal mechanics to accommodate for this change.
 In that case, model $M'$ would be more transferable than model $M$.
 For example, if we change the experimental setting for a reference game to consist of data from a special population, such as very young children or language-impaired adults, a potentially reasonable architectural change to a model like the RSA model is to consider differences in the sets of alternative utterances for the speaker \citep[e.g.][]{Noveck2001:When-Children-a}.
-Even though this is only a vague explication of a notion of transferability, it suffices to corroborate the intuition that models like PCMs, which are designed to operate at a higher level of conceptual abstraction, will often appear more transferable than models like LLMs, which make predictions not for \emph{kinds} of situations but for \emph{particular} situations.
+Even though this is only a vague explication of a notion of transferability, it suffices to corroborate the intuition that probabilistic cognitive models like the Rational Speech Act model, which are designed to operate at a higher level of conceptual abstraction, will often appear more transferable than models like LLMs, which make predictions not for \emph{kinds} of situations but for \emph{particular} situations.
 Whether any given model's transfer-ability is correct, is an orthogonal empirical question.
 It is also an empirical question, exactly to which extent and in which areas LLMs are (not) transferable, especially if we consider prompting a transfer strategy \citep{LiuLiu2022:Generated-Knowl,XieRaghunathan2022:An-Explanation-}.
-In any case, the comparison between LLMs and PCMs started here suggests systematic research into the transfer-ability and explanatory value of LLMs, e.g., by prompting strategies that are empirically insightful or by their use in composite neuro-symbolic models that implement theoretically-meaningful conceptual differences between types of situations.
+In any case, the comparison between LLMs and other statistical or probabilistic cognitive models started here suggests systematic research into the transfer-ability and explanatory value of LLMs, e.g., by prompting strategies that are empirically insightful or by their use in composite neuro-symbolic models that implement theoretically-meaningful conceptual differences between types of situations.
 
@@
 Despite being anchored in a small, but detailed case-study, the fact that different plausible methods of aggregating item-level information led to condition-level predictions of variable quality is worrisome.
-Similarly, we argued that the wide-spread reliance on a winner-takes-all strategy might be inconsistent with the actual use practices of LLMs, which may not always rely on a temperature-zero sampling strategy (see Section~\ref{motivation}).
-In conclusion, research on LLM should systematically study the conceptual and empirical consequences of seemingly minor decisions in evaluation or application settings.
-Variability of performance measures also comes with a well-known risk for robust and reproducible science.
+The wide-spread reliance on a winner-takes-all strategy might be inconsistent with the actual use practices of LLMs, which may not always rely on a temperature-zero sampling strategy (see Appendix~\ref{sec:aver-prob-vs}).
+In conclusion, research on LLMs should systematically study the conceptual and empirical consequences of seemingly minor decisions in evaluation or application settings.
+Variability of performance measures also comes with a well-known risk for robust and reproducible science \citep[c.f.][]{HuLevy2023:Prompt-based-me,TsvilodubWang2024:Predictions-fro}.
 The more researcher degrees of freedom there are, the higher the risk of false results, even in the absence of intentions to mislead \citep[e.g.][]{Ioannidis2005:Why-Most-Publis,Chambers2017:The-Seven-Deadl}.
-The comparison of LLMs with PCMs here raises awareness for the issue of robust research methods also in NLP \citep{WielingRawee2018:Reproducibility}.
+Testing LLMs as predictors in statistical models should raise awareness for the issue of robust research methods also in NLP \citep{WielingRawee2018:Reproducibility}.
 
@@
 The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
-Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
-Based on the data set and the detailed analyses conducted here, it seems not infeasible to use numerical predictions from LLMs are part of predictive probabilistic models.
+Moreover, to the extent that LLMs are used as parts of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-,TsvilodubCarcassi2024:Towards-Neuro-S}.
+Based on the data set and the detailed analyses conducted here, it seems not infeasible to use numerical predictions from LLMs as part of predictive probabilistic models.
 But, ideally, low-level variation, such as from order of presentation or similar ``nuisance variables,'' should be averaged out or taken into account in some way or other, at least as long as we do not understand what causes this variation in the predictions of models and further research that investigates when exactly this variation accords with empirically observed patterns.
 In sum, we conjecture that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is taken into account.
+This, however, entails that each LLM component in a hybrid model is independently tested against at least a modestly sized empirical data set.
 
@@
 We therefore side with the cautious voices that do not recommend replacing human participants with LLMs in psychological research \citep{DillionTandon2023:Can-AI-language,HardingDAlessandro2023:AI-language-mod}.
-In contrast, there is a challenge for LLM research.
-The fact that aggregate predictions can track aggregate human behaviour means that the variance in on both sides is washed out to achieve a similar result.
+In contrast, this points to an important challenge for future LLM research.
+The fact that aggregate predictions can track aggregate human behaviour means that the variance on both sides is washed out to achieve a similar result.
 This raises the issue of finding the systematic differences between LLMs and human answerers.
@@
 The scope of our experimental investigation was deliberately small but perspicuous.
-Focusing on the methodological contributions and details in performance assessments, we focused on a minimal non-trivial case study where we could triangulate LLMs, PCMs and human data.
+Focusing on the methodological contributions and details in performance assessments, we investigated a minimal non-trivial case study where we could triangulate LLMs, probabilistic cognitive models and human data.
 This work may therefore serve as starting point for a wider investigation of more complex data sets and case studies in which LLMs are analyzed as and directly compared to explanatory PCMs.

### 2024-04-05 04-paper/paper-LLM-probmod.tex
revisions §1

@@
 Yet, for any downstream application of LLMs, it is crucial to understand what LLMs can or cannot reliably do.
+
 The way in which LLM capabilities should be assessed depends on what their intended application is.
-For many industrial applications, the prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' or ``gold standard'' exists, averaged over many instances of this task.
+For many industrial applications, the prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' or ``gold standard'' exists, averaged over many instances of this task \citep[e.g.,][]{srivastava2023-BIGbench}, but more encompassing approaches also highlight the importance of a more holistic assessments of LLM, including factors such as robustness, fairness and efficiency \citep{LiangBommasani2023:Holistic-Evalua}.
 Benchmark-driven assessments are very useful for engineering purposes, when the main issue is whether a given system can perform a particular task correctly.
-Benchmark testing is also useful to systematically study of the effects of scale \citep[e.g.,][]{srivastava2023-BIGbench}, to uncover areas in which LLMs might under-perform on some standard that is practically relevant, and to thereby direct the community's attention to what needs improvement.
-More encompassing approaches also highlight the importance of including multiple factors in holistic assessments of LLM, such as robustness, fairness and efficiency \citep{LiangBommasani2023:Holistic-Evalua}.
+
+% Benchmark testing is also useful to systematically study of the effects of scale \citep[e.g.,][]{srivastava2023-BIGbench}, to uncover areas in which LLMs might under-perform on some standard that is practically relevant, and to thereby direct the community's attention to what needs improvement.
 
@@
 To assess the quality of LLMs in such contexts, it is less important to compare against a gold standard, and more important to compare against the full distribution of the human behavior that is to be captured by the LLM.
+In sum, at least for some practical applications, the ``gold standard'' is not a single ``correct'' answer, but the full distribution of human responses.
 
@@
 Taking numerical predictor values generated from LLMs as input, we explore strategies of building a Bayesian statistical model around them, and to scrutinize these LLM-grounded Bayesian statistical models with the usual methods of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A}.
+
 A main conceptual take-away of this investigation, is that statistical models built around LLMs are by design fundamentally different from common statistical or probabilistic cognitive models; an observation which also reflects back on the possibility of seeing LLMs as potentially explanatory for human behavior or cognition.
@@
 We then use the standard tools of Bayesian data analysis, to fit and check the resulting statistical models based on data from human multiple-choice experiments.
-Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single case study of pragmatic language production and interpretation, so-called reference games \mf{references}, for which a widely used probabilistic cognitive model exists, the Rational Speech Act model \citep{FrankGoodman2012:Predicting-Prag}, so that we can compare a theoretically motivated probabilistic cognitive model (PCM) against a probabilistic model built on top of LLM predictions.
-We find that variability predicted by the LLM at the item-level is generally not borne out by the human data, that some but not all ways of constructing condition-level predictions are equally good, and that different LLMs as backends may prefer to use different aggregation strategies.
+Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single case study of pragmatic language production and interpretation, namely so-called reference games \citep[e.g.][]{DeemterSluis2006:Building-a-Sema,DegenFranke2013:Cost-Based-Prag,GrafDegen2016:Animal-dog-or-d,SikosVenhuizen2021:Reevaluating-pr}.,
+For reference games, a widely used probabilistic cognitive model exists, namely the Rational Speech Act model \citep{FrankGoodman2012:Predicting-Prag}, so that we can compare a theoretically motivated probabilistic cognitive model (PCM) against a probabilistic model built on top of LLM predictions.
+We find that variability predicted by the LLM at the item-level is generally not borne out by the human data, that not all ways of constructing condition-level predictions are equally good, and that different LLMs as backends may prefer to use different aggregation strategies.
 
@@
 The paper is structured as follows.
-Section~\ref{motivation} motivates the investigation in this paper by highlighting relevant differences in the way LLMs and common probabilistic statistical models make predictions about data from forced-choice tasks.
-Section~\ref{experiment-reference-games} describes a simple but non-trivial forced-choice experiment, namely a text-based reference game.
-Section~\ref{sec:model-pred-from} introduces the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag} for the human data from the reference game experiment.
+Section~\ref{experiment-reference-games} describes an experiment with human participants with a text-based reference game.
+Section~\ref{sec:model-pred-from} introduces the Rational Speech Act (RSA) model for the human data from the reference game experiment.
 Section~\ref{sec:item-level-pred} exposes a statistical model for item-level predictions derived from LLM scores and investigates whether this adequately captures the human data at the item-level based on scores from GPT-3.5 \citep{BrownMann2020:Language-Models}.
 Section~\ref{llm-predictions-for-reference-games} discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
-Finally, Section~\ref{sec:gener-other-llm} explores whether previous results generalize to other LLM backends by investigating different versions of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}
+Finally, Section~\ref{sec:gener-other-llm} explores whether previous results generalize to other LLM backends by investigating different versions of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}.
 

### 2024-04-05 04-paper/paper-LLM-probmod.tex
updated Fig 1 and abstract

@@
   This raises questions about the human-likeness of LLM-derived information and whether LLMs, if they deliver human-like behavior, could possibly be considered explanatory models of (aspects of) human cognition or language use.
-  To shed more light on these issues, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of probabilistic cognitive modeling.
-  We identify core differences, in terms of conceptual interpretation and usual practical evaluation, between LLMs and probabilistic cognitive models (PCMs).
-  Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level), and that this is a sense in which LLMs may be said to be not explanatory.
+  To shed more light on these issues, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of Bayesian statistical modeling.
+  % We identify core differences, in terms of conceptual interpretation and usual practical evaluation, between LLMs and probabilistic cognitive models (PCMs).
+  % Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level), and that this is a sense in which LLMs may be said to be not explanatory.
   Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs fail to capture the variance in the human data at the item-level.
-  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, thus building PCM-like predictions from LLM-derived information, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to our data.
+  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to our data.
   These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at an aggregate level, for which they are, however, not designed to make predictions in the first place.
@@
 
+\begin{figure}[t]
+  \centering
+  \includegraphics[width = 0.95\textwidth]{00-pics/condition-item-level-prediction.png}
+  \caption{
+    Schematic representation of key conceptual differences between different types of predictive models.
+    Standard statistical models, like hierarchical regression models, typically make predictions for aggregate data (e.g., at the condition level), and add random offsets for item-level variation.
+    More sophisticated models, like some probabilistic cognitive models, may holistically combine information from an aggregate level (task, condition) with specific information about items.
+    LLMs, in contrast, first and foremost give prediction about each individual item and must rely on proper aggregation to arrive at condition-level predictions.
+  }
+  \label{fig:stats-model-types}
+\end{figure}
+
 A main conceptual take-away of this investigation, is that statistical models built around LLMs are by design fundamentally different from common statistical or probabilistic cognitive models; an observation which also reflects back on the possibility of seeing LLMs as potentially explanatory for human behavior or cognition.
@@
 As this point is crucial for our investigation, the following elaborates briefly.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width = 0.8\textwidth]{00-pics/condition-item-level-prediction.png}
-  \caption{Schematic representation of key conceptual differences between different types of statistical models. \mf{picture is cropped}}
-  \label{fig:stats-model-types}
-\end{figure}
 

### 2024-04-05 04-paper/paper-LLM-probmod.tex
revised §2 and 3, updated tables and figs + caps

@@
 We then use the standard tools of Bayesian data analysis, to fit and check the resulting statistical models based on data from human multiple-choice experiments.
-Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single case study of pragmatic language production and interpretation, namely so-called reference games \citep[e.g.][]{DeemterSluis2006:Building-a-Sema,DegenFranke2013:Cost-Based-Prag,GrafDegen2016:Animal-dog-or-d,SikosVenhuizen2021:Reevaluating-pr}.,
+Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single case study of pragmatic language production and interpretation, namely so-called reference games \citep[e.g.,][]{DeemterSluis2006:Building-a-Sema,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,GrafDegen2016:Animal-dog-or-d,SikosVenhuizen2021:Reevaluating-pr}.
 For reference games, a widely used probabilistic cognitive model exists, namely the Rational Speech Act model \citep{FrankGoodman2012:Predicting-Prag}, so that we can compare a theoretically motivated probabilistic cognitive model (PCM) against a probabilistic model built on top of LLM predictions.
@@
 
-Reference games are an established, well-understood and austere experimental paradigm to test human decision making in abstract communicative tasks \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,SikosVenhuizen2021:Reevaluating-pr}.
+Reference games are an established, well-understood and austere experimental paradigm to test human decision making in abstract communicative tasks.
 A reference game consists of two players, a speaker and an interpreter, who jointly observe a set of objects, usually referred to as context (see Figure~\ref{fig:ref-game}).
@@
 
-We replicated a simple reference game with human participants in which each trial instantiated the structure of the example shown in Figure~\ref{fig:ref-game}.
+We implemented a simple reference game for human participants in which each trial instantiated the structure of the example shown in Figure~\ref{fig:ref-game}.
 While previous reference games with human participants used pictorial representations of objects, and sometimes even pictorial representations of messages, we implemented a text-only version in order to be able to compare the predictions of LLMs for human data, when both LLMs and humans processed the same textual representation of the stimuli.
@@
 
-\subsection{Participants}
-\label{participants}
-
+\paragraph{Participants.}
 A total of 302 participants were recruited via Prolific for monetary
@@
 
-\subsection{Materials \& design}
-\label{materials-design}
-
+\paragraph{Materials \& design.}
 We created 100 different items as stimulus material via a stochastic process.
 Each item is a different textual description of a reference game with the same logical structure as the example from Figure~\ref{fig:ref-game}.
-For each item, the context consisted of three objects.
+For each item, the context consists of three objects.
 As in the original paper by \citet{FrankGoodman2012:Predicting-Prag}, objects are defined by a triple of properties, namely a color, a shape and a texture.
@@
 
-\subsection{Procedure}
-\label{procedure}
-
-For each participant the experiment sampled four different items.
+\paragraph{Procedure.}
+Each participant saw four different items sampled randomly from the pre-generated item set.
 Participants first played two of these in the production condition, then the other two in the interpretation condition.
 
-\subsection{Results}\label{results}
-
-The overall distribution of choices that correspond to the target, competitor, and distractor states is shown in Figure~\ref{fig:refgame-counts}.\footnote{
+\paragraph{Results.}
+The overall distribution of choices that correspond to the target, competitor, and distractor states is shown in Figure~\ref{fig:refgame-counts} (together with model checking results to be introduced later).\footnote{
   The production condition actually has two distractor choices.
   Here and in the following, these are lumped together as a single category, also when modeling random errors in later models.
-  This is a simplification but does not change anything of substance.}
+  }
 It is interesting that the distractor options were chosen rather often.
@@
 
-\begin{figure}[t]
-  \centering
-
-    \includegraphics[width=0.9\linewidth]{00-pics/PPC-alpha-eps-model.pdf}
-
-    \caption{Counts of choices from reference games with human participants (colored bars), with summary statistics from the posterior predictive distribution of four models (shapes and error bars).
-      Shapes show the mean of the posterior predictive distributions of the RSA model and three further models derived from item-level LLM-supplied scores.
-      Error-bars show corresponding 95\% credible intervals of the posterior predictive.
-    }
-  \label{fig:refgame-counts}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\begin{figure}[t]
+  \centering
+  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
+  \caption{
+    Example of predictions from the RSA model.
+    The semantic meaning is shown as a matrix of binary truth-values.
+    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents) for $\alpha=1$, and are shown as row-stochastic matrices.
+  }
+  \label{fig:RSA-example}
+\end{figure}
+
 Data from reference games with human participants have been variously analyzed with probabilistic models using inspiration from behavioral game theory \citep[e.g.,][]{DegenFranke2013:Cost-Based-Prag}, probabilistic Bayesian modeling \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag} or other forms of probabilistic modeling \citep[e.g.,][]{GattGompel2013:Are-we-Bayesian}.
@@
 
+\paragraph{The Rational Speech Act model.}
 As a concrete example, we introduce the Rational Speech Act (RSA) model first described in this form by \citet{FrankGoodman2012:Predicting-Prag} \citep[for overview see][]{FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu,StevensBenz2018:Game-Theoretic-,ScontrasTessler2021:A-practical-int,Degen2023:The-Rational-Sp}.
@@
 
-\begin{figure}[t]
-  \centering
-  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
-  \caption{
-    Example of predictions from the RSA model (with $\alpha=1$).
-    The semantic meaning is shown as a matrix of binary truth-values.
-    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents) for $\alpha=1$ are shown as row-stochastic matrices.
-  }
-  \label{fig:RSA-example}
-\end{figure}
-
 Figure~\ref{fig:RSA-example} gives example calculations (assuming a flat prior and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
@@
 Increasing $\alpha$ will increase the odds of target over competitor choices.
+
+\paragraph{Condition-level predictions.}
 In sum, the condition-level predictions of the RSA model are a parameterized function $P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c})$, assigning a probability to each response category $R_{l}$ (target, competitor, or distractor) in each condition $C$ (production or interpretation).
@@
 \begin{align*}
-  P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon_{c}) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
+  P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon_{c}) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
 \end{align*}
@@
   \label{eq:RSA-likelihood}
- P^{\text{RSA}}_{\text{cond}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \text{Multinomial}(D_{C}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
+ P^{\text{RSA}}_{\text{cond}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \text{Multinomial}(D_{C}, \tuple{P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
 \end{align}
@@
 
+\paragraph{Bayesian posteriors \& model checking.}
 Parameterized predictions, like in Equation~(\ref{eq:RSA-likelihood}), can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
@@
   \toprule
-  &&&& $\alpha$ &&& $\epsilon$ & \\ \cmidrule(r){4-6} \cmidrule(l){7-9}
+  &&&&&& $\alpha$ &&& $\epsilon$ & \\ \cmidrule(r){6-8} \cmidrule(l){9-11}
   & model & data & method & condition & |95\% & mean\ & 95\%| & |95\% & mean\ & 95\%| & Bpppv \\
   \midrule
-  ! & RSA & item  & ---         & prd. & 2.62 & 3.13 & 3.69  & 0.08 & 0.12 & 0.16 & 0.29 \\
-  ! & RSA & item  & ---         & int. & 0.21 & 0.67 & 1.08  & 0.08 & 0.14 & 0.19 & 0.21 \\
-  ! & RSA & cond. & ---         & prd. & 2.62 & 3.14 & 3.70  & 0.08 & 0.12 & 0.17 & 0.50 \\
-  ! & RSA & cond. & ---         & int. & 0.27 & 0.68 & 1.13  & 0.09 & 0.14 & 0.19 & 0.51 \\ \addlinespace[0.5em]
-    & GPT & item  & ---         & prd. & 0.40 & 0.51 & 0.62  & 0.07 & 0.12 & 0.17 & 0.00 \\
-    & GPT & item  & ---         & int. & 0.52 & 0.66 & 0.82  & 0.00 & 0.05 & 0.15 & 0.00 \\
-  ! & GPT & cond. & avg. scores & prd. & 0.61 & 0.74 & 0.87  & 0.08 & 0.12 & 0.17 & 0.49 \\
-    & GPT & cond. & avg. scores & int. & 0.99 & 1.17 & 1.35  & 0.00 & 0.02 & 0.06 & 0.00 \\
-  ! & GPT & cond. & avg. prob.  & prd. & 0.81 & 5.00 & 14.58 & 0.08 & 0.12 & 0.16 & 0.61 \\
-    & GPT & cond. & avg. prob.  & int. & 1.36 & 2.04 & 2.67  & 0.00 & 0.03 & 0.08 & 0.00 \\
-  ! & GPT & cond. & avg. WTA    & prd. & 0.86 & 1.04 & 1.22  & 0.08 & 0.12 & 0.17 & 0.48 \\
-  ! & GPT & cond. & avg. WTA    & int. & 0.16 & 0.43 & 0.69  & 0.06 & 0.13 & 0.19 & 0.50 \\
+  \textcolor{fern-main}{$\checkmark$} & RSA & item  & ---         & prd. & 2.62 & 3.13 & 3.69  & 0.08 & 0.12 & 0.16 & 0.29 \\
+  \textcolor{fern-main}{$\checkmark$} & RSA & item  & ---         & int. & 0.21 & 0.67 & 1.08  & 0.08 & 0.14 & 0.19 & 0.21 \\
+  \textcolor{fern-main}{$\checkmark$} & RSA & cond. & ---         & prd. & 2.62 & 3.14 & 3.70  & 0.08 & 0.12 & 0.17 & 0.50 \\
+  \textcolor{fern-main}{$\checkmark$} & RSA & cond. & ---         & int. & 0.27 & 0.68 & 1.13  & 0.09 & 0.14 & 0.19 & 0.51 \\ \addlinespace[0.5em]
+  \textcolor{shimmer-main}{$\times$} & GPT & item  & ---         & prd. & 0.40 & 0.51 & 0.62  & 0.07 & 0.12 & 0.17 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & GPT & item  & ---         & int. & 0.52 & 0.66 & 0.82  & 0.00 & 0.05 & 0.15 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & GPT & cond. & avg. scores & prd. & 0.61 & 0.74 & 0.87  & 0.08 & 0.12 & 0.17 & 0.49 \\
+  \textcolor{shimmer-main}{$\times$} & GPT & cond. & avg. scores & int. & 0.99 & 1.17 & 1.35  & 0.00 & 0.02 & 0.06 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & GPT & cond. & avg. prob.  & prd. & 0.81 & 5.00 & 14.58 & 0.08 & 0.12 & 0.16 & 0.61 \\
+  \textcolor{shimmer-main}{$\times$} & GPT & cond. & avg. prob.  & int. & 1.36 & 2.04 & 2.67  & 0.00 & 0.03 & 0.08 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & GPT & cond. & avg. WTA    & prd. & 0.86 & 1.04 & 1.22  & 0.08 & 0.12 & 0.17 & 0.48 \\
+  \textcolor{fern-main}{$\checkmark$} & GPT & cond. & avg. WTA    & int. & 0.16 & 0.43 & 0.69  & 0.06 & 0.13 & 0.19 & 0.50 \\
    \bottomrule \\
@@
 Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models for condition-level data, which are to be introduced later).
-We see that for both conditions the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-}, which require that the distribution of posterior predictions includes the observed choice rates for each answer option.
+We see that for both conditions the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-}, which requires that the distribution of posterior predictions includes the observed choice rates for each answer option.
 To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values (Bppp values), using likelihood of the observed data as a test statistics.
@@
 
-We can also use the condition-level predictions for predicting the item-level data.
-If the whole data set for condition $C$ is chunked into item-level data, $D_{C} = \tuple{D^{1}_{C}, \dots, D^{m}_{C} }$, the likelihood function for analysis at the item-level uses the same (condition-level) predictor to separately predict data from each item, so that the likelihood function becomes:
+\begin{figure}[t]
+  \centering
+
+    \includegraphics[width=0.9\linewidth]{00-pics/GPT-PPC-alpha-eps-model.pdf}
+
+    \caption{
+      Counts of choices from reference games with human participants (colored bars), with summary statistics from the posterior predictive distribution of four models (shapes and error bars).
+      Shapes show the mean of the posterior predictive distributions of the RSA model and three aggregated condition-level predictors derived from item-level LLM scores (introduced in Section~\ref{llm-predictions-for-reference-games}).
+      Error-bars show corresponding 95\% credible intervals of the posterior predictive.
+    }
+  \label{fig:refgame-counts}
+\end{figure}
+
+\paragraph{Item-level predictions.}
+To test item-level predictions, the whole data set for condition $C$ is chunked into item-level data, $D_{C} = \tuple{D^{1}_{C}, \dots, D^{m}_{C} }$, where $D_{C}^{k}$ is the data collected for the $k$-th out of $m$ items for condition $C$.
+Probabilistic cognitive models, like the RSA model, may holistically combine condition- and item-level information in their predictions, as shown in the middle of Figure~\ref{fig:ref-game}, e.g., by incorporating further parameters to capture variance based on different item-classes, such as the empirically observed preference for selecting shape terms over color terms in reference games \citep[e.g.][]{QingFranke2013:Variations-on-a}.
+But for present purposes, we simply use the same (condition-level) predictor to separately predict data from each item, so that the likelihood function for item-level data becomes:
 %
 \begin{align*}
- P^{\text{RSA}}_{\text{item}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \prod_{k = 1}^{m} \text{Multinomial} \left (D_{C}^{k}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3} \right )\,.
+ P^{\text{RSA}}_{\text{item}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \prod_{k = 1}^{m} \text{Multinomial} \left (D_{C}^{k}, \tuple{P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3} \right )\,.
 \end{align*}
@@
 Fitting the RSA model to the partitioned data set, we find estimates of Bppp values that do not discredit the model (see Table~\ref{tab:sumStats}).
-This suggests that the item-level variation in the human data is not so pronounced as to provide strong evidence against the condition-level predictor from a theoretically motivated PCM.
+This suggests that the item-level variation in the human data is not so pronounced as to provide strong evidence against the condition-level predictor from the RSA model.
 In other words, the condition-level RSA predictions seem to adequately capture also the item-level data.
 
-The following sections apply these tests of Bayesian model criticism also to models built around predictor values from LLMs.
+The following sections apply the same methods of Bayesian model criticism also to models built around predictor values from LLMs.
 Section~\ref{sec:item-level-pred} first looks at the item-level data, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.
-
-\mf{metion (somewhere) where item-level effects could be added!}
 

### 2024-04-05 04-paper/paper-LLM-probmod.tex
revised section 4

@@
 
-An LLM is design to predict the next token given an input string.
+An (autoregressive) LLM is designed to predict the next token given an input string.\footnote{
+  Nothing of substance changes when a Bayesian statistical model is built around scores from a masked language model.
+  We focus on autoregressive, left-to-right language modeling and next-token prediction for ease of reference, since all models tested here are autoregressive.
+}
 This is essentially an item-level prediction: next-token probabilities are specified for concrete strings after a concrete instance of a task, i.e., an item of the task, not for the task as such.
@@
 
+\paragraph{Notation.}
 Let $\set{I_{1}, \dots, I_{m}}$ be $m$ be instances of the same task, or items belonging to the same (logical) condition in a behavioral experiment.
@@
 Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ tokens, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
-For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same response category $r_{i}$.
-For the case at hand these categories are: target, competitor, distractor, so that $y_{k1}$ always corresponds to the designated \emph{target option}, $r_{1}$, that is the ``gold standard'' answer.
+For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same response category $R_{i}$.
+For the case at hand these categories are: target, competitor, distractor, so that $y_{k1}$ always corresponds to the designated \emph{target option}, $R_{1}$.
 
+\paragraph{LLM scores \& predictions.}
 The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
-  \mf{TODO: Add quick word on how this can in principle be extended to masked language models.}
   More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
@@
   For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
-  \mf{check whether this is correct also after new LLaMA results.}
+  \mf{@PT: check whether this is correct also after new LLaMA results.}
 }
@@
 
+\paragraph{Bayesian statistical modeling with LLM predictors.}
 As mentioned previously, the items of the reference game experiment from Section~\ref{experiment-reference-games} differ in which levels of features (color, shape, texture) instantiate the structure of the task shown in Figure \ref{fig:ref-game}, as well as the order of presentation of objects and words.
@@
 %
-The resulting likelihood function for item-level data with item-level predictions is, where priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.:
+The resulting likelihood function for item-level data with item-level predictions is, where priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}:
 %
 \begin{align*}
- P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
+ P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le i \le 3} \right )\,.
 \end{align*}
@@
 Summary statistics for samples from the posterior distribution over parameters are shown in Table~\ref{tab:sumStats} (rows 4 and 5).
-Credible values of $\alpha$ parameters are comparatively low for the LLM-based model, suggesting that the predicted scores have rather large differences, which have to be compensated for by the softmax link-function to adequately fit the data.
+Credible values of $\alpha$ parameters are comparatively low for the LLM-based item-level model, suggesting that the predicted scores have rather large differences, which have to be compensated for by the softmax link-function to adequately fit the data.
 Figure~\ref{fig:closeness-target-item-level} shows the mean of the probabilities for the highest scoring answer category, averaged over all items, for different parameter values of $\alpha$, highlighting that for \textit{a posteriori} credible values of $\alpha$ (gray shaded areas), the predictions are clearly distinct from the predictions of a WTA strategy (which assigns probability 1 to the highest scoring option for each item).
@@
 
+Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated LLM scores are inadequate predictors of the human data.
+To corroborate this conclusion, Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM-based model predicts variance which is not borne out by the human data.
+Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
+There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
+So, the model itself predicts systematic variability at the item level.
+The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
+These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
+Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by the current experimental data.
+
 \begin{figure}[t]
@@
 \end{figure}
-
-Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated LLM scores are inadequate predictors of the human data.
-To corroborate this conclusion, Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
-Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
-There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
-So, the model itself predicts systematic variability at the item level.
-The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
-These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
-Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by the current experimental data.
 

### 2024-04-05 04-paper/paper-LLM-probmod.tex
revised secs 5-6, C, tables and figs

@@
 % For human observers, these items are interchangeable, all exemplifying the same conceptual problem.
-Each item $I_{k} = \tuple{x_{k}, \tuple{y_{k1}, \dots, y_{kl}}}$ consists of an input prompt $x_{k}$, which is a string of text, and $l$ choice options $\tuple{y_{k1}, \dots, y_{kl}}$, all of which are strings as well, possibly composed of $|y_{ki}|$ tokens, $y_{ki} = w_{ki1}, \dots, w_{ki|y_{ki}|}$.
-For simplicity of notation, we assume that the $i$-th choice option $y_{ki}$ for each item $k$ belongs to the same response category $R_{i}$.
+Each item $I_{k} = \tuple{x_{k}, \tuple{y_{kl}}_{1 \le l \le n}}$ consists of an input prompt $x_{k}$, which is a string of text, and $n$ choice options $\tuple{y_{kl}}$, all of which are strings as well, possibly composed of $|y_{kl}|$ tokens, $y_{kl} = w_{kl1}, \dots, w_{kl|y_{kl}|}$.
+For simplicity of notation, we assume that the $l$-th choice option $y_{kl}$ for each item $k$ belongs to the same response category $R_{l}$.
 For the case at hand these categories are: target, competitor, distractor, so that $y_{k1}$ always corresponds to the designated \emph{target option}, $R_{1}$.
@@
 \paragraph{LLM scores \& predictions.}
-The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{ki}$ is its log-probability:\footnote{
+The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{kl}$ is its log-probability:\footnote{
   More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
@@
   % S\left( y_{ki}, x_{k} \right) =
-  \text{S}_{ki}
+  \text{S}_{kl}
   % = \log P_{\text{LLM}} \left(y_{ki} \mid x_{k} \right)
-  =  \sum_{j=1}^{|y_{ki}|} \log P_{\text{LLM}} \left(w_{kij} \mid x_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
+  =  \sum_{i=1}^{|y_{kl}|} \log P_{\text{LLM}} \left(w_{kli} \mid x_{k}, w_{kl1}, \dots, w_{kl(i-1)} \right)  \,.
 \end{align*}
@@
 \begin{align*}
-P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
+P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{kl} \right)\,.
 \end{align*}
@@
 
-To address this question, we assessed item-level scores $\text{S}_{ki}$ from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
+To address this question, we assessed item-level scores $\text{S}_{kl}$ from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
 % \footnote{
@@
 \begin{align*}
-  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
-  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
+  P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c}, \epsilon_{c} \right )
+  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
 \end{align*}
@@
 \begin{align*}
- P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le i \le 3} \right )\,.
+ P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{kl} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
 \end{align*}
@@
   \caption{
-    Schematic overview over three different approaches of obtaining condition-level predictions by aggregating item-level information.
-    Approaches differ in the kind of item-level information they take into account, which entails differences in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
+    Schematic overview over three different methods of obtaining condition-level predictions by aggregating item-level information.
+    Methods differ in the kind of item-level information they take into account, which entails differences in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
   }
@@
 
-While LLMs do not make conditional-level predictions as such, they can be derived from item-level scores $\text{S}_{ki}$ by averaging over all items belonging to the relevant condition.
+While LLMs do not make conditional-level predictions as such, they can be derived from item-level scores $\text{S}_{kl}$ by averaging over all items belonging to the relevant condition.
 There are many ways of averaging item-level information.
-Figure~\ref{fig:measures-overview} shows three salient approaches, which differ in what the underlying item-level measure is: the raw scores $\text{S}_{ki}$, the item-level probability derived from it (as used in Section~\ref{sec:item-level-pred}), or the predictions from the winner-takes-all (WTA) strategy commonly used in benchmark testing.
+Figure~\ref{fig:measures-overview} shows three salient approaches, which differ in what the underlying item-level measure for aggregation is: the raw scores $\text{S}_{kl}$, the item-level probability derived from it (as used in Section~\ref{sec:item-level-pred}), or the predictions from the winner-takes-all (WTA) strategy commonly used in benchmark testing.
 
@@
 \begin{align*}
-  & P_{\text{cond}}^{\text{SCR}}(r_{i} \mid C, \alpha_{c})
-    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{ki}  \right ] \,.
+  & P_{\text{cond}}^{\text{SCR}}(R_{l} \mid C, \alpha_{c})
+    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{kl}  \right ] \,.
     \tag*{\textcolor{gray}{[average scores (narrow-scope aggregation)]}}
@@
   % & \textbf{wide-scope aggregation} \\
-  & P_{\text{cond}}^{\text{PRB}}(r_{i} \mid C, \alpha_{c})
-    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha_{c}) \,.
+  & P_{\text{cond}}^{\text{PRB}}(R_{l} \mid C, \alpha_{c})
+    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{kl} \mid C, \alpha_{c}) \,.
     \tag*{\textcolor{gray}{[average probabilities (wide-scope aggregation)]}}
@@
   % & \textbf{intermediate-scope aggregation} \\
-  & P_{\text{cond}}^{\text{WTA}}(r_{i} \mid C, \alpha_{c})
-    \propto  \left [ \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) \right ]^{\alpha_{c}}\,,
+  & P_{\text{cond}}^{\text{WTA}}(R_{l} \mid C, \alpha_{c})
+    \propto  \left [ \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{WTA}}(y_{kl} \mid C) \right ]^{\alpha_{c}}\,,
     \tag*{\textcolor{gray}{[average WTA (intermediate-scope aggregation)]}}
 \end{align*}
-where $P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) = \lim_{\alpha \rightarrow \infty} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha)$.
+where $P_{\text{item}}^{\text{WTA}}(y_{kl} \mid C) = \lim_{\alpha \rightarrow \infty} P_{\text{item}}^{\text{LLM}}(y_{kl} \mid C, \alpha)$.
 \bigskip
@@
 We find that the average-WTA predictor is consistently able to capture the data from the production condition, except for one of the six models (LLaMA2-chat 70B).
-However, the average-WTA predictor recovers the interpretation data for only one of the six models (LLaMA2-chat 13B, see also Table~\ref{tab:sumStats-LLaMA2}).
+When the average-WTA predictor passes the model checking criterion for the production data, the posteriors for the power-law parameter $\alpha$ are credibly different from 1, thus suggesting that, contrary to the results for GPT-3.5 a mere ``WTA average'' is not a good prediction strategy for the production data for the LLaMA2-based models.
+Looking at interpretation, the average-WTA predictor recovers the interpretation data for only one of the six models (LLaMA2-chat 13B, see also Table~\ref{tab:sumStats-LLaMA2}).
 Interestingly, for LLaMA2-chat 13B also the average-probability predictor is able to recover the interpretation data, unlike for the GPT-3.5.
-As for GPT-3.5, no model captured the item-level variance observed in the human data.
+Finally, for item-level data, as for GPT-3.5, no LLaMA2-based model captured the item-level variance observed in the human data.
 In sum, we find general support for the previous conclusions that the average-WTA predictor is most successful in capturing the condition-level data, and therefore that, where models manage to recover the condition-level data, they seem to be ``right for the wrong reasons.''
@@
 
-\begin{table}[t]
+\begin{table}[]
 \centering
@@
   \toprule
-  &&&& $\alpha$ &&& $\epsilon$ & \\ \cmidrule(r){4-6} \cmidrule(l){7-9}
+  &&&&&& $\alpha$ &&& $\epsilon$ & \\ \cmidrule(r){6-8} \cmidrule(l){9-11}
   & model & data & method & condition & |95\% & mean\ & 95\%| & |95\% & mean\ & 95\%| & Bpppv \\
@@
 
-    & L2-hf-7b    & item  & ---         & prd. & 1.09  & 1.30   & 1.51   & 0.00 & 0.02 & 0.06 & 0.00 \\
-    & L2-hf-7b    & item  & ---         & int. & 1.90  & 2.52   & 3.13   & 0.00 & 0.04 & 0.12 & 0.00 \\
-  ! & L2-hf-7b    & cond. & avg. scores & prd. & 9.65  & 11.50  & 13.62  & 0.08 & 0.12 & 0.16 & 0.48 \\
-    & L2-hf-7b    & cond. & avg. scores & int. & 6.80  & 7.96   & 9.14   & 0.00 & 0.02 & 0.05 & 0.00 \\
-    & L2-hf-7b    & cond. & avg. prob.  & prd. & 4.98  & 16.44  & 33.06  & 0.06 & 0.10 & 0.13 & 0.00 \\
-    & L2-hf-7b    & cond. & avg. prob.  & int. & 20.06 & 79.30  & 163.94 & 0.00 & 0.01 & 0.04 & 0.00 \\
-  ! & L2-hf-7b    & cond. & avg. WTA    & prd. & 4.17  & 4.96   & 5.85   & 0.07 & 0.12 & 0.16 & 0.48 \\
-    & L2-hf-7b    & cond. & avg. WTA    & int. & 0.87  & 1.02   & 1.18   & 0.00 & 0.02 & 0.05 & 0.00 \\ \addlinespace[0.5em]
-    & L2-hf-13b   & item  & ---         & prd. & 1.37  & 1.91   & 2.47   & 0.00 & 0.07 & 0.12 & 0.00 \\
-    & L2-hf-13b   & item  & ---         & int. & 2.13  & 2.90   & 3.69   & 0.00 & 0.05 & 0.14 & 0.00 \\
-  ! & L2-hf-13b   & cond. & avg. scores & prd. & 13.73 & 16.32  & 19.10  & 0.08 & 0.12 & 0.16 & 0.50 \\
-    & L2-hf-13b   & cond. & avg. scores & int. & 7.73  & 9.11   & 10.42  & 0.00 & 0.02 & 0.05 & 0.00 \\
-    & L2-hf-13b   & cond. & avg. prob.  & prd. & 69.84 & 185.80 & 360.44 & 0.06 & 0.09 & 0.13 & 0.00 \\
-    & L2-hf-13b   & cond. & avg. prob.  & int. & 13.63 & 39.72  & 88.28  & 0.00 & 0.02 & 0.05 & 0.00 \\
-  ! & L2-hf-13b   & cond. & avg. WTA    & prd. & 2.28  & 2.70   & 3.18   & 0.08 & 0.12 & 0.16 & 0.51 \\
-    & L2-hf-13b   & cond. & avg. WTA    & int. & 0.78  & 0.94   & 1.12   & 0.00 & 0.03 & 0.07 & 0.03 \\ \addlinespace[0.5em]
-    & L2-hf-70b   & item  & ---         & prd. & 1.53  & 1.86   & 2.22   & 0.00 & 0.03 & 0.07 & 0.00 \\
-    & L2-hf-70b   & item  & ---         & int. & 3.20  & 4.01   & 4.74   & 0.00 & 0.03 & 0.08 & 0.00 \\
-  ! & L2-hf-70b   & cond. & avg. scores & prd. & 44.57 & 53.17  & 62.44  & 0.08 & 0.12 & 0.17 & 0.48 \\
-    & L2-hf-70b   & cond. & avg. scores & int. & 6.86  & 8.14   & 9.32   & 0.00 & 0.02 & 0.05 & 0.00 \\
-    & L2-hf-70b   & cond. & avg. prob.  & prd. & 10.51 & 26.59  & 45.73  & 0.07 & 0.10 & 0.14 & 0.00 \\
-    & L2-hf-70b   & cond. & avg. prob.  & int. & 10.86 & 23.63  & 42.16  & 0.00 & 0.03 & 0.09 & 0.00 \\
-  ! & L2-hf-70b   & cond. & avg. WTA    & prd. & 5.88  & 7.03   & 8.19   & 0.08 & 0.12 & 0.17 & 0.50 \\
-    & L2-hf-70b   & cond. & avg. WTA    & int. & 0.55  & 0.68   & 0.81   & 0.00 & 0.03 & 0.07 & 0.04 \\ \addlinespace[0.5em]
-    & L2-chat-7b  & item  & ---         & prd. & 0.34  & 0.41   & 0.49   & 0.00 & 0.03 & 0.07 & 0.00 \\
-    & L2-chat-7b  & item  & ---         & int. & 0.79  & 1.05   & 1.34   & 0.00 & 0.05 & 0.15 & 0.00 \\
-  ! & L2-chat-7b  & cond. & avg. scores & prd. & 58.47 & 69.64  & 81.79  & 0.08 & 0.12 & 0.16 & 0.48 \\
-    & L2-chat-7b  & cond. & avg. scores & int. & 2.03  & 2.37   & 2.73   & 0.00 & 0.02 & 0.05 & 0.00 \\
-    & L2-chat-7b  & cond. & avg. prob.  & prd. & 8.05  & 27.26  & 55.30  & 0.07 & 0.11 & 0.15 & 0.00 \\
-    & L2-chat-7b  & cond. & avg. prob.  & int. & 2.86  & 19.24  & 51.41  & 0.00 & 0.01 & 0.04 & 0.00 \\
-  ! & L2-chat-7b  & cond. & avg. WTA    & prd. & 12.18 & 14.70  & 17.21  & 0.08 & 0.12 & 0.16 & 0.50 \\
-    & L2-chat-7b  & cond. & avg. WTA    & int. & 0.86  & 1.01   & 1.14   & 0.00 & 0.02 & 0.05 & 0.00 \\ \addlinespace[0.5em]
-    & L2-chat-13b & item  & ---         & prd. & 0.36  & 0.44   & 0.53   & 0.00 & 0.03 & 0.08 & 0.00 \\
-    & L2-chat-13b & item  & ---         & int. & 0.68  & 0.90   & 1.11   & 0.00 & 0.05 & 0.13 & 0.00 \\
-  ! & L2-chat-13b & cond. & avg. scores & prd. & 5.56  & 6.63   & 7.76   & 0.08 & 0.12 & 0.16 & 0.49 \\
-    & L2-chat-13b & cond. & avg. scores & int. & 1.70  & 2.05   & 2.40   & 0.00 & 0.02 & 0.06 & 0.01 \\
-    & L2-chat-13b & cond. & avg. prob.  & prd. & 3.90  & 22.25  & 74.12  & 0.06 & 0.10 & 0.14 & 0.00 \\
-  ! & L2-chat-13b & cond. & avg. prob.  & int. & 2.85  & 4.70   & 7.01   & 0.00 & 0.03 & 0.07 & 0.08 \\
-  ! & L2-chat-13b & cond. & avg. WTA    & prd. & 4.28  & 5.11   & 5.99   & 0.07 & 0.12 & 0.16 & 0.49 \\
-  ! & L2-chat-13b & cond. & avg. WTA    & int. & 0.15  & 0.32   & 0.51   & 0.02 & 0.11 & 0.18 & 0.50 \\ \addlinespace[0.5em]
-    & L2-chat-70b & item  & ---         & prd. & 0.28  & 0.35   & 0.42   & 0.00 & 0.03 & 0.08 & 0.00 \\
-    & L2-chat-70b & item  & ---         & int. & 1.03  & 1.24   & 1.47   & 0.00 & 0.03 & 0.08 & 0.00 \\
-    & L2-chat-70b & cond. & avg. scores & prd. & 0.40  & 0.55   & 0.74   & 0.00 & 0.04 & 0.10 & 0.00 \\
-  ! & L2-chat-70b & cond. & avg. scores & int. & 1.91  & 2.37   & 2.89   & 0.00 & 0.03 & 0.08 & 0.16 \\
-    & L2-chat-70b & cond. & avg. prob.  & prd. & 0.43  & 0.66   & 0.96   & 0.00 & 0.04 & 0.10 & 0.00 \\
-    & L2-chat-70b & cond. & avg. prob.  & int. & 2.82  & 5.21   & 8.43   & 0.00 & 0.01 & 0.04 & 0.02 \\
-    & L2-chat-70b & cond. & avg. WTA    & prd. & 0.14  & 0.21   & 0.35   & 0.00 & 0.06 & 0.13 & 0.00 \\
-    & L2-chat-70b & cond. & avg. WTA    & int. & 0.85  & 1.00   & 1.14   & 0.00 & 0.02 & 0.05 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & item  & ---         & prd. & 1.09  & 1.30   & 1.51   & 0.00 & 0.02 & 0.06 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & item  & ---         & int. & 1.90  & 2.52   & 3.13   & 0.00 & 0.04 & 0.12 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-hf-7b    & cond. & avg. scores & prd. & 9.65  & 11.50  & 13.62  & 0.08 & 0.12 & 0.16 & 0.48 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & cond. & avg. scores & int. & 6.80  & 7.96   & 9.14   & 0.00 & 0.02 & 0.05 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & cond. & avg. prob.  & prd. & 4.98  & 16.44  & 33.06  & 0.06 & 0.10 & 0.13 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & cond. & avg. prob.  & int. & 20.06 & 79.30  & 163.94 & 0.00 & 0.01 & 0.04 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-hf-7b    & cond. & avg. WTA    & prd. & 4.17  & 4.96   & 5.85   & 0.07 & 0.12 & 0.16 & 0.48 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & cond. & avg. WTA    & int. & 0.87  & 1.02   & 1.18   & 0.00 & 0.02 & 0.05 & 0.00 \\ \addlinespace[0.5em]
+  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & item  & ---         & prd. & 1.37  & 1.91   & 2.47   & 0.00 & 0.07 & 0.12 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & item  & ---         & int. & 2.13  & 2.90   & 3.69   & 0.00 & 0.05 & 0.14 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-hf-13b   & cond. & avg. scores & prd. & 13.73 & 16.32  & 19.10  & 0.08 & 0.12 & 0.16 & 0.50 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & cond. & avg. scores & int. & 7.73  & 9.11   & 10.42  & 0.00 & 0.02 & 0.05 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & cond. & avg. prob.  & prd. & 69.84 & 185.80 & 360.44 & 0.06 & 0.09 & 0.13 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & cond. & avg. prob.  & int. & 13.63 & 39.72  & 88.28  & 0.00 & 0.02 & 0.05 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-hf-13b   & cond. & avg. WTA    & prd. & 2.28  & 2.70   & 3.18   & 0.08 & 0.12 & 0.16 & 0.51 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & cond. & avg. WTA    & int. & 0.78  & 0.94   & 1.12   & 0.00 & 0.03 & 0.07 & 0.03 \\ \addlinespace[0.5em]
+  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & item  & ---         & prd. & 1.53  & 1.86   & 2.22   & 0.00 & 0.03 & 0.07 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & item  & ---         & int. & 3.20  & 4.01   & 4.74   & 0.00 & 0.03 & 0.08 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-hf-70b   & cond. & avg. scores & prd. & 44.57 & 53.17  & 62.44  & 0.08 & 0.12 & 0.17 & 0.48 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & cond. & avg. scores & int. & 6.86  & 8.14   & 9.32   & 0.00 & 0.02 & 0.05 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & cond. & avg. prob.  & prd. & 10.51 & 26.59  & 45.73  & 0.07 & 0.10 & 0.14 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & cond. & avg. prob.  & int. & 10.86 & 23.63  & 42.16  & 0.00 & 0.03 & 0.09 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-hf-70b   & cond. & avg. WTA    & prd. & 5.88  & 7.03   & 8.19   & 0.08 & 0.12 & 0.17 & 0.50 \\
+  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & cond. & avg. WTA    & int. & 0.55  & 0.68   & 0.81   & 0.00 & 0.03 & 0.07 & 0.04 \\ \addlinespace[0.5em]
+  \textcolor{shimmer-main}{$\times$} & L2-chat-7b  & item  & ---         & prd. & 0.34  & 0.41   & 0.49   & 0.00 & 0.03 & 0.07 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-7b  & item  & ---         & int. & 0.79  & 1.05   & 1.34   & 0.00 & 0.05 & 0.15 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-chat-7b  & cond. & avg. scores & prd. & 58.47 & 69.64  & 81.79  & 0.08 & 0.12 & 0.16 & 0.48 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-7b  & cond. & avg. scores & int. & 2.03  & 2.37   & 2.73   & 0.00 & 0.02 & 0.05 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-7b  & cond. & avg. prob.  & prd. & 8.05  & 27.26  & 55.30  & 0.07 & 0.11 & 0.15 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-7b  & cond. & avg. prob.  & int. & 2.86  & 19.24  & 51.41  & 0.00 & 0.01 & 0.04 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-chat-7b  & cond. & avg. WTA    & prd. & 12.18 & 14.70  & 17.21  & 0.08 & 0.12 & 0.16 & 0.50 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-7b  & cond. & avg. WTA    & int. & 0.86  & 1.01   & 1.14   & 0.00 & 0.02 & 0.05 & 0.00 \\ \addlinespace[0.5em]
+  \textcolor{shimmer-main}{$\times$} & L2-chat-13b & item  & ---         & prd. & 0.36  & 0.44   & 0.53   & 0.00 & 0.03 & 0.08 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-13b & item  & ---         & int. & 0.68  & 0.90   & 1.11   & 0.00 & 0.05 & 0.13 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-chat-13b & cond. & avg. scores & prd. & 5.56  & 6.63   & 7.76   & 0.08 & 0.12 & 0.16 & 0.49 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-13b & cond. & avg. scores & int. & 1.70  & 2.05   & 2.40   & 0.00 & 0.02 & 0.06 & 0.01 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-13b & cond. & avg. prob.  & prd. & 3.90  & 22.25  & 74.12  & 0.06 & 0.10 & 0.14 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-chat-13b & cond. & avg. prob.  & int. & 2.85  & 4.70   & 7.01   & 0.00 & 0.03 & 0.07 & 0.08 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-chat-13b & cond. & avg. WTA    & prd. & 4.28  & 5.11   & 5.99   & 0.07 & 0.12 & 0.16 & 0.49 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-chat-13b & cond. & avg. WTA    & int. & 0.15  & 0.32   & 0.51   & 0.02 & 0.11 & 0.18 & 0.50 \\ \addlinespace[0.5em]
+  \textcolor{shimmer-main}{$\times$} & L2-chat-70b & item  & ---         & prd. & 0.28  & 0.35   & 0.42   & 0.00 & 0.03 & 0.08 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-70b & item  & ---         & int. & 1.03  & 1.24   & 1.47   & 0.00 & 0.03 & 0.08 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-70b & cond. & avg. scores & prd. & 0.40  & 0.55   & 0.74   & 0.00 & 0.04 & 0.10 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-chat-70b & cond. & avg. scores & int. & 1.91  & 2.37   & 2.89   & 0.00 & 0.03 & 0.08 & 0.16 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-70b & cond. & avg. prob.  & prd. & 0.43  & 0.66   & 0.96   & 0.00 & 0.04 & 0.10 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-70b & cond. & avg. prob.  & int. & 2.82  & 5.21   & 8.43   & 0.00 & 0.01 & 0.04 & 0.02 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-70b & cond. & avg. WTA    & prd. & 0.14  & 0.21   & 0.35   & 0.00 & 0.06 & 0.13 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-chat-70b & cond. & avg. WTA    & int. & 0.85  & 1.00   & 1.14   & 0.00 & 0.02 & 0.05 & 0.00 \\
 
@@
 \caption{
-  Summary statistics for model fits based on LLaMA2 backends.
-  The information shown is the same as in Table~\ref{tab:sumStats}.
+  Summary statistics for models based on LLaMA2 variants.
+  Information shown is the same as in Table~\ref{tab:sumStats}.
 }
@@
 Standard benchmark testing looks at \emph{task accuracy}, defined as the probability of selecting the ``gold standard'' response, usually based on ``winner-takes-all'' (WTA) selection of the highest scoring option.
-We can generalize this and define the \emph{softmax-based accuracy} as the average probability of choosing the designated target response option $r_{1}$ for the \emph{condition-level soft-max prediction} (notation as defined in main text):
+We can generalize this and define the \emph{softmax-based accuracy} as the average probability of choosing the designated target response option $R_{1}$ for the \emph{condition-level soft-max prediction} (notation as defined in main text):
 \begin{align*}
-  P_{\text{cond}}^{\text{SM}}\left(r_{1} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{k1} \right)\,.
+  P_{\text{cond}}^{\text{SM}}\left(R_{1} \right) = \frac{1}{m} \sum_{k = 1}^{m} P_{\text{item}}^{\text{SM}} \left(y_{k1} \right)\,.
 \end{align*}
@@
 The WTA-based accuracy can differ qualitatively from the more general softmax-based accuracy, as shown by the following example.
-The example shows that differences between accuracy measures depend on the variation in item-level scores, in particular on the relation between score-ordering and score-differences.
-Note that the example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
 %
@@
   But if an application uses anything other than greedy-like responses, the accuracy score of 0.8 may be misleading.
-  If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases.
+  If the remaining 20\% of the items are such that the non-goal option is almost infinitely better, it would be chosen under a pure sampling strategy, where $\alpha = 1$, with virtual certainty, so the softmax-based accuracy would be around 0.4.\footnote{The probability of the target option in the 80\% of items where the goal answer is slightly better is 0.5 in the limit of $\epsilon \rightarrow 0$, and it is virtually 0 in the remaining 20\% of the cases.
 This gives an expected rate of: $\nicefrac{4}{5} \ \nicefrac{1}{2} + \nicefrac{1}{5} \ 0 = \nicefrac{2}{5}$.}
@@
 %
-The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions.
+The example shows that differences between accuracy measures depend on the variation in item-level scores, in particular on the relation between score-ordering and score-differences.
+Note that the example holds equally if numbers for the two options are reversed, so that there is no way of saying which of the two measures of accuracy would generally be more favorable for selecting the target option.
+
+The upshot of these considerations is that the standard practice of WTA-based performance assessment for LLMs gives false, or at least misleading or inaccurate results, whenever not all downstream applications use a greedy-like sampling strategy (which is almost certainly the case), and there is variability in item-level predictions (which may or may not be the case, depending on the domain of application).
 

### 2024-05-21 04-paper/paper-LLM-probmod.tex
integrated feedback / typos

@@
 	University of Tübingen\\
-	\texttt{mchfranke@uni-tuebingen.de} \\
+	\texttt{mchfranke@gmail.com} \\
 	%% examples of more authors
@@
   State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks and are increasingly used as components in larger applications, where LLM-based predictions serve as proxies for human judgements.
-  This raises questions about the human-likeness of LLM-derived information and whether LLMs, if they deliver human-like behavior, could possibly be considered explanatory models of (aspects of) human cognition or language use.
+  This raises questions about the human-likeness of LLM-derived information, alignment with human intuition in larger applications, and whether LLMs, if they deliver human-like behavior, could possibly be considered explanatory models of (aspects of) human cognition or language use.
   To shed more light on these issues, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of Bayesian statistical modeling.
@@
   % Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level), and that this is a sense in which LLMs may be said to be not explanatory.
-  Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs fail to capture the variance in the human data at the item-level.
-  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to our data.
-  These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at an aggregate level, for which they are, however, not designed to make predictions in the first place.
+  Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs do not capture the variance in the human data at the item-level.
+  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to human data.
+  These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at the aggregate condition-level, for which they are, however, not designed to make predictions in the first place.
 \end{abstract}
@@
 On the other hand, LLMs first and foremost provide predictions about each item.
-While it may be the case that, in producing an item-level prediction, the internal computation of a powerful LLM is informed by computations that incorporate abstract knowledge roughly corresponding to the condition-level, the atomic predictions accessible to the common user are specific to each individual string tested, thus raising concerns about robustness of predictions under perturbations of input prompts \mf{cite}.
+While it may be the case that, in producing an item-level prediction, the internal computation of a powerful LLM is informed by computations that incorporate abstract knowledge roughly corresponding to the condition-level, the atomic predictions accessible to the common user are specific to each individual string tested.
+Consequently, this raises concerns about robustness of predictions under perturbations of input prompts \mf{cite}.
 
 These considerations raise two important conceptual and empirical questions.
-First, we need to ask whether the item-level predictions made by LLMs are empirically correct.
+First, we need to ask whether the item-level predictions made by LLMs are empirically correct, i.e., match the human data at the item-level.
 Second, the implied methodological challenge lies in specifying how item-level information can be used, e.g., by different aggregation methods, to make robust predictions at a more abstract level (condition, task, $\dots$).
@@
 In production trials, participants choose a single word to describe a trigger object from the context.
-In interpretation trials, an object is selected as the likely object a trigger word referring to.}
+In interpretation trials, an object is selected as the likely object a trigger word is referring to.}
   \label{fig:ref-game}
@@
 The \textbf{target utterance} is the word which uniquely describes the trigger object.
-The \textbf{distractor utterance} is the word that is true of the trigger object, but also true of another object.
-The other utterances, both of which are false of the trigger object are \textbf{competitor utterance}.
+The \textbf{competitor utterance} is the word that is true of the trigger object, but also true of another object.
+The other utterances, both of which are false of the trigger object are \textbf{distractor utterance}.
 %
-In a critical interpretation trial, the trigger word is one that is true of two of the three objects.
+In a critical interpretation trial, the trigger word is the one that is true of two of the three objects.
 If participants engage in pragmatic thought, they might reason that \emph{if} the speaker had wanted to refer to one of the two objects of which the trigger word is true (blue square and blue circle in Figure~\ref{fig:ref-game}), the speaker could have used a more informative word for exactly one of those two objects (``circle''), so they are more likely to refer to the \textbf{target object} (the blue square in Figure~\ref{fig:ref-game}).
@@
     The semantic meaning is shown as a matrix of binary truth-values.
-    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents) for $\alpha=1$, and are shown as row-stochastic matrices.
+    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over states (referents) for $\alpha=1$, and are shown as row-stochastic matrices.
   }
@@
 Common to these approaches is that they derive or define, based on some explicit conceptual motivation, a parameterized stochastic speaker policy, $P_{S}(u \mid s; \theta_{S})$, modulated by parameters $\theta_{S}$, for a speaker's choice of expression or utterance $u$ given a referent or state $s$, which the speaker wants to communicate;
-and a stochastic listener policy, $P_{L}(s \mid u; \theta_{L})$, capturing the probability of choosing a referent $s$ for utterance $u$.
+and a parameterized stochastic listener policy, $P_{L}(s \mid u; \theta_{L})$, capturing the probability of choosing a referent $s$ for utterance $u$.
 
@@
 As a concrete example, we introduce the Rational Speech Act (RSA) model first described in this form by \citet{FrankGoodman2012:Predicting-Prag} \citep[for overview see][]{FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu,StevensBenz2018:Game-Theoretic-,ScontrasTessler2021:A-practical-int,Degen2023:The-Rational-Sp}.
-The RSA model defines pragmatic reasoning as a sequence of iterated (soft-)optmization of policies, grounding out in literal interpretation.
+The RSA model defines pragmatic reasoning as a sequence of iterated (soft-)optmization of policies and Bayesian inference, grounding out in literal interpretation.
 If $\mathfrak{L}(s,u) \mapsto \set{0,1}$ is a semantic meaning function mapping each pair of state $s$ and utterance $u$ to a (binary) truth-value, and if $P_{\text{prior}}(s)$ is a prior over states, a literal listener policy is defined as:
@@
 %
-The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated, i.e., to maximize the log-probability of the trigger object:
+The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated, i.e., to maximize the log-probability of the trigger object given the utterance:
 %
@@
 \paragraph{Condition-level predictions.}
-In sum, the condition-level predictions of the RSA model are a parameterized function $P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c})$, assigning a probability to each response category $R_{l}$ (target, competitor, or distractor) in each condition $C$ (production or interpretation).
+In sum, the condition-level predictions of the RSA model are a parameterized function $P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c})$, assigning a probability to each response category $R_{l}$ (target, competitor, or distractor) in each condition $C$ (production or interpretation) for a given $\alpha_{c}$.
 
@@
 
-Based on each option's score, we can define the \emph{item-level prediction} of choosing option $y_{ki}$ in term of soft-maximization as:
+Based on each option's score, we can define the \emph{item-level prediction} of choosing option $y_{kl}$ in terms of soft-maximization as:
 %
@@
 %
-Notice that the usual item-level prediction used in benchmark testing is actually a ``winner-takes-all'' strategy, but this is just a special case of the above for $\alpha_{C} \rightarrow \infty$.
+Notice that the usual item-level prediction used in benchmark testing is actually a ``winner-takes-all'' strategy, which would choose any option that maximizes the score, but this is just a special case of the above for $\alpha_{C} \rightarrow \infty$.
 Here, the $\alpha_{C}$ parameter corresponds to inverse temperature, so that $\alpha_{C} \rightarrow \infty$ corresponds to greedy sampling with temperature zero.
@@
 
-To address this question, we assessed item-level scores $\text{S}_{kl}$ from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
+To address this question, we assessed item-level scores $\text{S}_{kl}$ from the text-davinci-003 instance of GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} for a text-based version of the reference game from Section~\ref{experiment-reference-games}.
 % \footnote{
@@
   In the current set-up the response type ``distractor'' has two instantiations in the production condition.
-Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option.}
+Since choice options are a single word in the production condition, for simplicity, we treat the union of the distractor words as a single option.}
 Similar to the RSA model fit, we allow for random errors with probability $\epsilon_{C}$:
@@
   P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c}, \epsilon_{c} \right )
-  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
+  = (1- \epsilon_{c}) \  P_{\text{item}}^{\text{LLM}}\left ( y_{kl} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
 \end{align*}
 %
-The resulting likelihood function for item-level data with item-level predictions is, where priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}:
+With priors on parameters as for the RSA model described in Section~\ref{sec:model-pred-from}, the resulting likelihood function for item-level data with item-level predictions is:
 %
@@
     The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$, and the implied probabilistic prediction.
-    For reference, the target probability under a random choice strategy and under the WTA strategy are shown with dashed lines.
+    For reference, the target probability under a random choice strategy and under the ``winner-takes-all'' (WTA) strategy are shown with dashed lines.
     For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
@@
 These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
-Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by the current experimental data.
+Put more strongly, a model that uses the most obvious item-level scores, derived from the predicted log-probabilities, to predict what human participants choose on a by-item level is ruled out by the current experimental data.
 
@@
 However, based on the results from the previous section, the item-level predictions for the WTA strategy are demonstrably incongruent with the human data.
-In general, the average-probability and average-WTA predictors can lead to qualitatively different results about task-level accuracy (see Appendix~\ref{sec:aver-prob-vs}).
+In general, the average-probability and average-WTA predictors can lead to qualitatively different results for task-level accuracy (see Appendix~\ref{sec:aver-prob-vs}).
 
@@
 To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
-Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} and the \emph{average-probabilities} model are able to reproduce the production data, but fail on the interpretation data; and that only the \emph{average-WTA} model does not fail to capture the data from both conditions.
+Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} and the \emph{average-probabilities} models are able to reproduce the production data, but fail on the interpretation data; and that only the \emph{average-WTA} model does not fail to capture the data from both conditions.
 
@@
   \includegraphics[width=0.9\textwidth]{00-pics/vPPC_all_LLaMA_models.pdf}
-  \caption{Visual posterior predictive checks on condition-level data for different models of the LLaMA family, paired with different aggregation strategies.}
+  \caption{Visual posterior predictive checks on condition-level data for different models of the LLaMA family, paired with different aggregation strategies.
+\mf{update plot by replacing 'hf' with 'base'.}}
   \label{fig:LLaMA-results-vPPC}
@@
 
-All previous results where obtained for LMM predictions derived from GPT-3.5 turbo.
-To investigate whether key results replicate for other LLMs and scales, we ran the same analyses also for scores derived from variants of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}, in particular the 7, 13 and 70 billion parameter versions of the models fine-tuned with human feedback only (marked as `hf' in the following) and based on human feedback and chat data (marked as `chat').
+All previous results where obtained for LMM predictions derived from GPT-3.5 davinci.
+To investigate whether key results replicate for other LLMs and scales, we ran the same analyses also for scores derived from variants of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}, in particular the 7, 13 and 70 billion parameter base models (marked as `base'), and the models of similar size fine-tuned on chat data (marked as `chat').
 Table~\ref{tab:sumStats-LLaMA2} shows the relevant summary statistics and Bayesian posterior predictive $p$-values.
@@
 Looking at interpretation, the average-WTA predictor recovers the interpretation data for only one of the six models (LLaMA2-chat 13B, see also Table~\ref{tab:sumStats-LLaMA2}).
-Interestingly, for LLaMA2-chat 13B also the average-probability predictor is able to recover the interpretation data, unlike for the GPT-3.5.
+Interestingly, for LLaMA2-chat 13B also the average-probability predictor is able to recover the interpretation data, unlike for the GPT-3.5 model.
 Finally, for item-level data, as for GPT-3.5, no LLaMA2-based model captured the item-level variance observed in the human data.
@@
 
-  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & item  & ---         & prd. & 1.09  & 1.30   & 1.51   & 0.00 & 0.02 & 0.06 & 0.00 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & item  & ---         & int. & 1.90  & 2.52   & 3.13   & 0.00 & 0.04 & 0.12 & 0.00 \\
-  \textcolor{fern-main}{$\checkmark$} & L2-hf-7b    & cond. & avg. scores & prd. & 9.65  & 11.50  & 13.62  & 0.08 & 0.12 & 0.16 & 0.48 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & cond. & avg. scores & int. & 6.80  & 7.96   & 9.14   & 0.00 & 0.02 & 0.05 & 0.00 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & cond. & avg. prob.  & prd. & 4.98  & 16.44  & 33.06  & 0.06 & 0.10 & 0.13 & 0.00 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & cond. & avg. prob.  & int. & 20.06 & 79.30  & 163.94 & 0.00 & 0.01 & 0.04 & 0.00 \\
-  \textcolor{fern-main}{$\checkmark$} & L2-hf-7b    & cond. & avg. WTA    & prd. & 4.17  & 4.96   & 5.85   & 0.07 & 0.12 & 0.16 & 0.48 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-7b    & cond. & avg. WTA    & int. & 0.87  & 1.02   & 1.18   & 0.00 & 0.02 & 0.05 & 0.00 \\ \addlinespace[0.5em]
-  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & item  & ---         & prd. & 1.37  & 1.91   & 2.47   & 0.00 & 0.07 & 0.12 & 0.00 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & item  & ---         & int. & 2.13  & 2.90   & 3.69   & 0.00 & 0.05 & 0.14 & 0.00 \\
-  \textcolor{fern-main}{$\checkmark$} & L2-hf-13b   & cond. & avg. scores & prd. & 13.73 & 16.32  & 19.10  & 0.08 & 0.12 & 0.16 & 0.50 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & cond. & avg. scores & int. & 7.73  & 9.11   & 10.42  & 0.00 & 0.02 & 0.05 & 0.00 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & cond. & avg. prob.  & prd. & 69.84 & 185.80 & 360.44 & 0.06 & 0.09 & 0.13 & 0.00 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & cond. & avg. prob.  & int. & 13.63 & 39.72  & 88.28  & 0.00 & 0.02 & 0.05 & 0.00 \\
-  \textcolor{fern-main}{$\checkmark$} & L2-hf-13b   & cond. & avg. WTA    & prd. & 2.28  & 2.70   & 3.18   & 0.08 & 0.12 & 0.16 & 0.51 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-13b   & cond. & avg. WTA    & int. & 0.78  & 0.94   & 1.12   & 0.00 & 0.03 & 0.07 & 0.03 \\ \addlinespace[0.5em]
-  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & item  & ---         & prd. & 1.53  & 1.86   & 2.22   & 0.00 & 0.03 & 0.07 & 0.00 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & item  & ---         & int. & 3.20  & 4.01   & 4.74   & 0.00 & 0.03 & 0.08 & 0.00 \\
-  \textcolor{fern-main}{$\checkmark$} & L2-hf-70b   & cond. & avg. scores & prd. & 44.57 & 53.17  & 62.44  & 0.08 & 0.12 & 0.17 & 0.48 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & cond. & avg. scores & int. & 6.86  & 8.14   & 9.32   & 0.00 & 0.02 & 0.05 & 0.00 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & cond. & avg. prob.  & prd. & 10.51 & 26.59  & 45.73  & 0.07 & 0.10 & 0.14 & 0.00 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & cond. & avg. prob.  & int. & 10.86 & 23.63  & 42.16  & 0.00 & 0.03 & 0.09 & 0.00 \\
-  \textcolor{fern-main}{$\checkmark$} & L2-hf-70b   & cond. & avg. WTA    & prd. & 5.88  & 7.03   & 8.19   & 0.08 & 0.12 & 0.17 & 0.50 \\
-  \textcolor{shimmer-main}{$\times$} & L2-hf-70b   & cond. & avg. WTA    & int. & 0.55  & 0.68   & 0.81   & 0.00 & 0.03 & 0.07 & 0.04 \\ \addlinespace[0.5em]
+  \textcolor{shimmer-main}{$\times$} & L2-base-7b    & item  & ---         & prd. & 1.09  & 1.30   & 1.51   & 0.00 & 0.02 & 0.06 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-7b    & item  & ---         & int. & 1.90  & 2.52   & 3.13   & 0.00 & 0.04 & 0.12 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-base-7b    & cond. & avg. scores & prd. & 9.65  & 11.50  & 13.62  & 0.08 & 0.12 & 0.16 & 0.48 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-7b    & cond. & avg. scores & int. & 6.80  & 7.96   & 9.14   & 0.00 & 0.02 & 0.05 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-7b    & cond. & avg. prob.  & prd. & 4.98  & 16.44  & 33.06  & 0.06 & 0.10 & 0.13 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-7b    & cond. & avg. prob.  & int. & 20.06 & 79.30  & 163.94 & 0.00 & 0.01 & 0.04 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-base-7b    & cond. & avg. WTA    & prd. & 4.17  & 4.96   & 5.85   & 0.07 & 0.12 & 0.16 & 0.48 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-7b    & cond. & avg. WTA    & int. & 0.87  & 1.02   & 1.18   & 0.00 & 0.02 & 0.05 & 0.00 \\ \addlinespace[0.5em]
+  \textcolor{shimmer-main}{$\times$} & L2-base-13b   & item  & ---         & prd. & 1.37  & 1.91   & 2.47   & 0.00 & 0.07 & 0.12 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-13b   & item  & ---         & int. & 2.13  & 2.90   & 3.69   & 0.00 & 0.05 & 0.14 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-base-13b   & cond. & avg. scores & prd. & 13.73 & 16.32  & 19.10  & 0.08 & 0.12 & 0.16 & 0.50 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-13b   & cond. & avg. scores & int. & 7.73  & 9.11   & 10.42  & 0.00 & 0.02 & 0.05 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-13b   & cond. & avg. prob.  & prd. & 69.84 & 185.80 & 360.44 & 0.06 & 0.09 & 0.13 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-13b   & cond. & avg. prob.  & int. & 13.63 & 39.72  & 88.28  & 0.00 & 0.02 & 0.05 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-base-13b   & cond. & avg. WTA    & prd. & 2.28  & 2.70   & 3.18   & 0.08 & 0.12 & 0.16 & 0.51 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-13b   & cond. & avg. WTA    & int. & 0.78  & 0.94   & 1.12   & 0.00 & 0.03 & 0.07 & 0.03 \\ \addlinespace[0.5em]
+  \textcolor{shimmer-main}{$\times$} & L2-base-70b   & item  & ---         & prd. & 1.53  & 1.86   & 2.22   & 0.00 & 0.03 & 0.07 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-70b   & item  & ---         & int. & 3.20  & 4.01   & 4.74   & 0.00 & 0.03 & 0.08 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-base-70b   & cond. & avg. scores & prd. & 44.57 & 53.17  & 62.44  & 0.08 & 0.12 & 0.17 & 0.48 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-70b   & cond. & avg. scores & int. & 6.86  & 8.14   & 9.32   & 0.00 & 0.02 & 0.05 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-70b   & cond. & avg. prob.  & prd. & 10.51 & 26.59  & 45.73  & 0.07 & 0.10 & 0.14 & 0.00 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-70b   & cond. & avg. prob.  & int. & 10.86 & 23.63  & 42.16  & 0.00 & 0.03 & 0.09 & 0.00 \\
+  \textcolor{fern-main}{$\checkmark$} & L2-base-70b   & cond. & avg. WTA    & prd. & 5.88  & 7.03   & 8.19   & 0.08 & 0.12 & 0.17 & 0.50 \\
+  \textcolor{shimmer-main}{$\times$} & L2-base-70b   & cond. & avg. WTA    & int. & 0.55  & 0.68   & 0.81   & 0.00 & 0.03 & 0.07 & 0.04 \\ \addlinespace[0.5em]
   \textcolor{shimmer-main}{$\times$} & L2-chat-7b  & item  & ---         & prd. & 0.34  & 0.41   & 0.49   & 0.00 & 0.03 & 0.07 & 0.00 \\
@@
 While the common practice in evaluating the capabilities of LLMs is based on accuracy averaged over large collections of data, this work took the alternative route to explore what we learn if we subjected LLMs to the same routines and strong demands on distributional quality of fit to human data as we normally do for statistical or probabilistic cognitive models.
-Knowledge of the adequacy of LLMs' full distributional predictions for simple tasks that require human like judgement or decisions is important to gauge in how far LLMs can be trusted to provide such information in applications such as hybrid, neuro-symbolic models \mf{refs}.
+Knowledge of the adequacy of LLMs' full distributional predictions for simple tasks that require human-like judgement or decisions is important to gauge in how far LLMs can be trusted to provide such information in applications such as hybrid, neuro-symbolic models \mf{refs}.
 A main contribution of this paper is methodological, showing how statistical model criticism can be used for LLMs in the first place, and, more specifically, how it can be insightful in the detailed assessment of how LLMs might or might not be able to predict human data at the degree of accuracy that we would normally aspire for when dealing with explanatory statistical models in cognitive science.
 Applying this comparative approach to a minimal, but non-trivial data set, we find that the LLM predictions on a per-item level predict variance that is not attested in the human data.
-From several candidate predictor measures for aggregate condition-level data, only one was not refuted by the human data, but this was one that relied on the empirically implausible WTA-strategy at the item-level, incidentally the same strategy that is commonly used in accuracy-based benchmark testing \citep{srivastava2023-BIGbench}.
+From several candidate predictor measures for aggregate condition-level data, only one was not refuted by the human data (at least for the GPT-3.5 model), but this was one that relied on the empirically implausible WTA-strategy at the item-level, incidentally the same strategy that is commonly used in accuracy-based benchmark testing \citep{srivastava2023-BIGbench}.
 
@@
 A basic observation brought to the foreground by our approach is that LLMs' atomic predictions are for individual items and that some aggregation method is needed to derive more abstract, condition- or task-level predictions.
-This is potential one sense in which LLMs may be felt to be less, or not at all, explanatory.
+This is one sense in which LLMs may be felt to be less, or not at all, explanatory.
 They do not offer, at least not directly, a human-comprehensible compression of reality into a \emph{kind} of response pattern, over and beyond making a prediction for each particular situation.
@@
 The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
-Moreover, to the extent that LLMs are used as parts of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-,TsvilodubCarcassi2024:Towards-Neuro-S}.
+Moreover, to the extent that LLMs are used as parts of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-} \mf{insert references}.
 Based on the data set and the detailed analyses conducted here, it seems not infeasible to use numerical predictions from LLMs as part of predictive probabilistic models.
-But, ideally, low-level variation, such as from order of presentation or similar ``nuisance variables,'' should be averaged out or taken into account in some way or other, at least as long as we do not understand what causes this variation in the predictions of models and further research that investigates when exactly this variation accords with empirically observed patterns.
+But, ideally, low-level prompt variation, such as from order of presentation or similar ``nuisance variables,'' should be averaged out or taken into account in some way or other, as we do not understand what causes this variation in the predictions of models.
+Further research is necessary that investigates when exactly this variation accords with empirically observed patterns.
 In sum, we conjecture that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is taken into account.
-This, however, entails that each LLM component in a hybrid model is independently tested against at least a modestly sized empirical data set.
+This, however, entails that each LLM component in a hybrid model should be independently tested against at least a modestly sized empirical data set.
 

### 2024-05-22 04-paper/paper-LLM-probmod.tex
added refs

@@
 \begin{abstract}
-  State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks and are increasingly used as components in larger applications, where LLM-based predictions serve as proxies for human judgements.
-  This raises questions about the human-likeness of LLM-derived information, alignment with human intuition in larger applications, and whether LLMs, if they deliver human-like behavior, could possibly be considered explanatory models of (aspects of) human cognition or language use.
+  State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks and are increasingly used as components in larger applications, where LLM-based predictions serve as proxies for human judgements or decision.
+  This raises questions about the human-likeness of LLM-derived information, alignment with human intuition, and whether LLMs could possibly be considered (parts of) explanatory models of (aspects of) human cognition or language use.
   To shed more light on these issues, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of Bayesian statistical modeling.
@@
   % Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level), and that this is a sense in which LLMs may be said to be not explanatory.
-  Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs do not capture the variance in the human data at the item-level.
-  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to human data.
-  These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at the aggregate condition-level, for which they are, however, not designed to make predictions in the first place.
+  Using human data from a forced-choice experiment on pragmatic language use, we find that LLMs do not capture the variance in the human data at the item-level.
+  We suggest different ways of deriving full distributional predictions from LLMs for aggregate, condition-level data, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to human data.
+  These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at the aggregate, condition-level, for which they are, however, not designed to, or usually used to, make predictions in the first place.
 \end{abstract}
@@
 The way in which LLM capabilities should be assessed depends on what their intended application is.
-For many industrial applications, the prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' or ``gold standard'' exists, averaged over many instances of this task \citep[e.g.,][]{srivastava2023-BIGbench}, but more encompassing approaches also highlight the importance of a more holistic assessments of LLM, including factors such as robustness, fairness and efficiency \citep{LiangBommasani2023:Holistic-Evalua}.
+For many industrial applications, the prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' or ``gold standard'' exists, averaged over many instances of this task \citep[e.g.,][]{srivastava2023-BIGbench}, but more encompassing approaches also highlight the importance of more holistic assessments of LLM, including factors such as robustness, fairness and efficiency \citep{LiangBommasani2023:Holistic-Evalua}.
 Benchmark-driven assessments are very useful for engineering purposes, when the main issue is whether a given system can perform a particular task correctly.
@@
 Recent works increasingly go beyond using LLMs based on single-run input-output behavior, and instead utilize LLMs as a part of a larger computational process.
-Simple examples include sophisticated prompting strategies \citep[e.g.,][]{LiuLiu2022:Generated-Knowl}, or structured reasoning models \citep[e.g.,][]{CreswellShanahan2022:Selection-Infer,GaoMadaan2023:PAL:-Program-ai,ParanjapeLundberg2023:ART:-Automatic-}.
-More sophisticated examples include neuro-symbolic models in which LLMs supply specific parts of the relevant information for some practical application \mf{include citations}, or where LLMs are a part of bigger programs to build towards something more akin to explanatory cognitive models \citep[e.g.,][]{WongGrand2023:From-Word-Model}.
-For example, information from LLMs can be used to generate alternatives for deliberation \citep[e.g.][]{TsvilodubCarcassi2024:Towards-Neuro-S}, arguably similar to human resource-rational reasoning in open-ended domains.
-\mf{include citations}
-LLMs may also be used to rank or numerically score options in open-ended applications, e.g., to mimic human judgements of relevance or interestingness \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
+Simple examples include sophisticated prompting strategies \citep[e.g.,][]{LiuLiu2022:Generated-Knowl}, or structured reasoning models \citep[e.g.,][]{CreswellShanahan2022:Selection-Infer,GaoMadaan2023:PAL:-Program-ai,ParanjapeLundberg2023:ART:-Automatic-,YangKlein2023:DOC:-Improving-}.
+More sophisticated examples include neuro-symbolic models in which LLMs supply specific parts of the relevant information for some practical application \citep[e.g.,][]{NyeTessler2021:Improving-Coher}, or where LLMs are a part of bigger programs to build towards something more akin to explanatory cognitive models \citep[e.g.,][]{WongGrand2023:From-Word-Model}.
+For example, information from LLMs can be used to generate alternatives for deliberation \citep[e.g.][]{LewTessler2020:Leveraging-Unst,TsvilodubCarcassi2024:Towards-Neuro-S}, arguably similar to human resource-rational reasoning in open-ended domains \citep{VulGoodman2014:One-and-Done-Op}.
+LLMs may also be used to rank or numerically score options in large or open-ended applications, e.g., to mimic human judgements of desirability, relevance or interestingness \citep[e.g.,][]{BaiKadavath2022:Constitutional-,ParkOBrien2023:Generative-Agen,KwonXie2023:Reward-Design-w,ZhangLehman2023:OMNI:-Open-ende}.
 In many of these applications, LLMs essentially serve as a cheap, compressed stand-in for (average) human judgements, associations or choice behavior.

### 2024-05-23 04-paper/paper-LLM-probmod.tex
added references, updated plot

@@
 In short, while a lot of previous like-minded work has used targeted assessment of LLM capabilities from the point of view of the experimental psychologist, we here adopt the more specific perspective of the probabilistic / statistical modeller.
-Taking numerical predictor values generated from LLMs as input, we explore strategies of building a Bayesian statistical model around them, and to scrutinize these LLM-grounded Bayesian statistical models with the usual methods of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A}.
+Taking numerical predictor values generated from LLMs as input, we explore strategies of building a Bayesian statistical model around them, and to scrutinize these LLM-grounded Bayesian statistical models with the usual methods of Bayesian data analysis, in particular model criticism \citep{GelmanCarlin2014:Bayesian-Data-A}.
 
@@
 
-A main conceptual take-away of this investigation, is that statistical models built around LLMs are by design fundamentally different from common statistical or probabilistic cognitive models; an observation which also reflects back on the possibility of seeing LLMs as potentially explanatory for human behavior or cognition.
+A main conceptual take-away of this investigation, is that statistical models built around LLMs are, by design, fundamentally different from common statistical or probabilistic cognitive models; an observation which also reflects back on the possibility of seeing LLMs as potentially explanatory model for human behavior or cognition.
 Concretely, LLMs make predictions for each individual item, rather than specifying a predictor of central tendency at a more aggregate level, such as at the level of an experimental condition, as standard statistical models or probabilistic cognitive models normally do (see Figure~\ref{fig:stats-model-types}).
@@
 Experimental work therefore often compares human performance in different \emph{experimental conditions}, which reflect the general factors that are hypothesized to influence behavior.
-Yet a single experimental condition is often instantiated with different \emph{experimental items}, which are not under scrutiny for any systematic, predictable effect on the observed measurements.
+Yet a single experimental condition is often instantiated with different \emph{experimental items}, which are usually not under scrutiny for any systematic, predictable effect on the observed measurements.
 For example, classical research on human memory \citep{AtkinsonShiffrin1968:Human-memory:-A} investigated the effect of rehearsal on memory consolidation.
@@
 While it may be the case that, in producing an item-level prediction, the internal computation of a powerful LLM is informed by computations that incorporate abstract knowledge roughly corresponding to the condition-level, the atomic predictions accessible to the common user are specific to each individual string tested.
-Consequently, this raises concerns about robustness of predictions under perturbations of input prompts \mf{cite}.
+Consequently, this points to known concerns about robustness of predictions under perturbations of input prompts \citep[e.g.][]{ReynoldsMcDonell2021:Prompt-Programm,WebsonPavlick2022:Do-Prompt-Based,SalinasMorstatter2024:The-Butterfly-E,TsvilodubWang2024:Predictions-fro}.
 
@@
 To investigate these issues, this paper introduces different ways of building a Bayesian probabilistic model around core predictor values derived from various LLMs.
-We then use the standard tools of Bayesian data analysis, to fit and check the resulting statistical models based on data from human multiple-choice experiments.
+We then use the standard tools of Bayesian data analysis to fit and check the resulting statistical models based on data from human multiple-choice experiments.
 Rather than aiming for scale and large-coverage, we focus on transparency and zoom in on a single case study of pragmatic language production and interpretation, namely so-called reference games \citep[e.g.,][]{DeemterSluis2006:Building-a-Sema,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,GrafDegen2016:Animal-dog-or-d,SikosVenhuizen2021:Reevaluating-pr}.
-For reference games, a widely used probabilistic cognitive model exists, namely the Rational Speech Act model \citep{FrankGoodman2012:Predicting-Prag}, so that we can compare a theoretically motivated probabilistic cognitive model (PCM) against a probabilistic model built on top of LLM predictions.
+The Rational Speech Act framework \citep{FrankGoodman2012:Predicting-Prag} provides a widely used probabilistic model for human data from reference games, so that we can compare a theoretically motivated probabilistic cognitive model (PCM) against a probabilistic model built on top of LLM predictions.
 We find that variability predicted by the LLM at the item-level is generally not borne out by the human data, that not all ways of constructing condition-level predictions are equally good, and that different LLMs as backends may prefer to use different aggregation strategies.
@@
 We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, such as over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options.
-\mf{references prompt variation}
 It therefore becomes an empirical question of whether item-level predictions from LLMs provide a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.
@@
   \includegraphics[width=0.9\textwidth]{00-pics/vPPC_all_LLaMA_models.pdf}
-  \caption{Visual posterior predictive checks on condition-level data for different models of the LLaMA family, paired with different aggregation strategies.
-\mf{update plot by replacing 'hf' with 'base'.}}
+  \caption{Visual posterior predictive checks on condition-level data for different models of the LLaMA family, paired with different aggregation strategies.}
   \label{fig:LLaMA-results-vPPC}
@@
 While the common practice in evaluating the capabilities of LLMs is based on accuracy averaged over large collections of data, this work took the alternative route to explore what we learn if we subjected LLMs to the same routines and strong demands on distributional quality of fit to human data as we normally do for statistical or probabilistic cognitive models.
-Knowledge of the adequacy of LLMs' full distributional predictions for simple tasks that require human-like judgement or decisions is important to gauge in how far LLMs can be trusted to provide such information in applications such as hybrid, neuro-symbolic models \mf{refs}.
+Knowledge of the adequacy of LLMs' full distributional predictions for simple tasks that require human-like judgement or decisions is important to gauge in how far LLMs can be trusted to provide such information in applications such as hybrid, neuro-symbolic models \citep[e.g.,][]{GarneloShanahan2019:Reconciling-dee,LewTessler2020:Leveraging-Unst,CreswellShanahan2022:Selection-Infer,Frank2023:Large-language-}.
 A main contribution of this paper is methodological, showing how statistical model criticism can be used for LLMs in the first place, and, more specifically, how it can be insightful in the detailed assessment of how LLMs might or might not be able to predict human data at the degree of accuracy that we would normally aspire for when dealing with explanatory statistical models in cognitive science.
@@
 The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
-Moreover, to the extent that LLMs are used as parts of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-} \mf{insert references}.
+Moreover, to the extent that LLMs are used as parts of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{LewTessler2020:Leveraging-Unst,WongGrand2023:From-Word-Model,Frank2023:Large-language-}.
 Based on the data set and the detailed analyses conducted here, it seems not infeasible to use numerical predictions from LLMs as part of predictive probabilistic models.

### 2024-05-23 03-data-analysis/04-data-analysis.R

@@ -358,10 +358,12 @@ retrieve_results_model_fits <- function(model_name) {
     
   }
   
-  outplot <- make_alpha_plot("production") + make_alpha_plot("interpretation")
+  alpha_plot_production <- make_alpha_plot("production")
+  alpha_plot_interpretation <- make_alpha_plot("interpretation")
+  outplot <- gridExtra::grid.arrange(alpha_plot_production, alpha_plot_interpretation, ncol = 2)
   
   ggsave(
-    filename = str_c("../04-paper/00-pics/", model_name, "-closeness-target-by-alpha-item-level.pdf"),
+    filename = stringr::str_c("../04-paper/00-pics/", model_name, "-closeness-target-by-alpha-item-level.pdf"),
     plot = outplot, width = 9, height = 4, scale =1)
     
 }
@@ -487,9 +489,9 @@ vPPC_data_all_models <- rbind(
 ) |> 
   # specify levels for 'backend models'
   mutate(backend = case_when(
-    backend == "LLaMA2-hf-7b" ~ "L2-hf-7b",
-    backend == "LLaMA2-hf-13b" ~ "L2-hf-13b",
-    backend == "LLaMA2-hf-70b" ~ "L2-hf-70b",
+    backend == "LLaMA2-hf-7b" ~ "L2-7b",
+    backend == "LLaMA2-hf-13b" ~ "L2-13b",
+    backend == "LLaMA2-hf-70b" ~ "L2-70b",
     backend == "LLaMA2-chat-hf-7b" ~ "L2-chat-7b",
     backend == "LLaMA2-chat-hf-13b" ~ "L2-chat-13b",
     backend == "LLaMA2-chat-hf-70b" ~ "L2-chat-70b",
@@ -497,7 +499,7 @@ vPPC_data_all_models <- rbind(
   )) |> 
   mutate(backend = factor(backend, 
                          levels = c(
-                           "L2-hf-7b", "L2-hf-13b", "L2-hf-70b", 
+                           "L2-7b", "L2-13b", "L2-70b", 
                            "L2-chat-7b", "L2-chat-13b", "L2-chat-70b",
                            "GPT", "RSA"))) |> 
   # norm to human data

### 2024-05-27 04-paper/paper-LLM-probmod.tex
corrected passage on length-normalization

@@
 The most obvious \emph{item-level score} an (autoregressive) LLM provides for each choice option $y_{kl}$ is its log-probability:\footnote{
-  More elaborate item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
-  From the point of view of experimental psychology, these corrections are \emph{post hoc} fixes to improperly balanced experimental materials.
-  For the purposes of this paper, where answer options are all equally long and commensurable, these corrections may be temporarily ignored for simplicity.
-  \mf{@PT: check whether this is correct also after new LLaMA results.}
+  Commonly used item-level scores include corrections for variable length of answer options \citep[e.g.,][]{BrownMann2020:Language-Models} or variation in base rate among answer options \citep[e.g.,][]{HoltzmanWest2021:Surface-Form-Co}.
+  Following common practice in the current literature, we also use length-corrected, average log-probabilities as raw scores for all options.
+  As most options have identical number of tokens, this only affects the numerical values of softmax-parameter $\alpha$, as introduced below.}
 }

### 2024-05-31 04-paper/paper-LLM-probmod.tex
added refs subj-variation

@@
   As most options have identical number of tokens, this only affects the numerical values of softmax-parameter $\alpha$, as introduced below.}
-}
 %
@@
 For the particular case of pragmatic language use, prior research has shown that individual participants have markedly different behavioral profiles, often consistently behaving like literal language users or more sophisticated language users \citep[e.g.,][]{NieuwlandDitman2010:On-the-incremen,FrankeDegen2015:Reasoning-in-Re,SpychalskaKontinen2016:Investigating-s}.
-It is an open question whether predictions from LLMs reflect the same kind of variation.
+It is an open question whether predictions from LLMs reflect the same kind of variation \citep[][]{AherArriaga2023:Using-large-lan,SalewskiAlaniz2023:In-Context-Impe}.
 The results presented here recommend skepticism.

### 2024-06-13 04-paper/paper-LLM-probmod.tex
added pic and acknowledgements

@@
 Section~\ref{llm-predictions-for-reference-games} discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
-Finally, Section~\ref{sec:gener-other-llm} explores whether previous results generalize to other LLM backends by investigating different versions of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}.
+Finally, Section~\ref{sec:gener-other-llm} explores whether previous results generalize to other LLM backends by investigating different versions of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}.\footnote{
+  All data and code is available at \url{https://osf.io/f6j3a/?view_only=5e820cc8bbee4549aed58dc252ba61b9}.
+}
 
@@
 
+\section*{Acknowledgements}
+Thanks to Robert D. Hawkins and Thomas Brochhagen for insightful comments.
+We gratefully acknowledge support by the state of Baden-Württemberg, Germany, through the computing resources provided by bwHPC and the German Research Foundation (DFG) through grant INST 35/1597-1 FUGG.
+MF is a member of the Machine Learning Cluster of Excellence, EXC number 2064/1 – Project number 39072764.
+
 \printbibliography[heading=bibintoc]

### 2024-06-13 04-paper/snippets-sec2-tmp.tex
added pic and acknowledgements

--- 2023-11-23 03-paper/paper-LLM-probmod.tex
@@
-\documentclass[fleqn]{article}
+\documentclass[fleqn,reqno,10pt]{article}
 
-\usepackage{arxiv}
+\usepackage{myarticlestyledefault}
 
-\usepackage{mypackages}
-\bibliography{~/Desktop/icloud/MyRefGlobal}
-\usepackage{mycommands}
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{cleveref}       % smart cross-referencing
-\usepackage{lipsum}         % Can be removed after putting your text content
-\usepackage{graphicx}
-\usepackage{doi}
-
-\usepackage{float}
-
-\title{Language models as probabilistic cognitive models}
-
+\title{Snippets for reuse}
+\author{Michael Franke}
 \date{}
-
-\newif\ifuniqueAffiliation
-% Uncomment to use multiple affiliations variant of author block
-\uniqueAffiliationtrue
-
-\ifuniqueAffiliation % Standard variant of author block
-  \author{ Michael Franke\thanks{Corresponding author.} \\
-	Department of Linguistics\\
-	University of Tübingen\\
-	\texttt{mchfranke@uni-tuebingen.de} \\
-	%% examples of more authors
-	\And
-	Polina Tsvilodub \\
-	Department of Linguistics\\
-	University of Tübingen\\
-	\texttt{polina.tsvilodub@gmail.com} \\
-	\And
-	Fausto Carcassi \\
-	ILLC\\
-	University of Amsterdam\\
-	\texttt{fausto.carcassi@gmail.com} \\
-}
-\else
-% Multiple affiliations variant of author block
-\usepackage{authblk}
-\renewcommand\Authfont{\bfseries}
-\setlength{\affilsep}{0em}
-% box is needed for correct spacing with authblk
-% \newbox{\orcid}\sbox{\orcid}{\includegraphics[scale=0.06]{orcid.pdf}}
-\author{Michael Franke, Polina Tsvilodub, Fausto Carcassi}
-\affil{Department of Linguistics\\University of Tübingen\\
-\texttt{[michael.franke|polina.tsvilodub|fausto.carcassi]@uni-tuebingen.de}}
-\fi
-
-% running right header
-\renewcommand{\headeright}{}
-% small title under big title
-\renewcommand{\undertitle}{}
-\renewcommand{\shorttitle}{Statistical modeling with LLM predictors}
-
-%%% Add PDF metadata to help others organize their library
-%%% Once the PDF is generated, you can check the metadata with
-%%% $ pdfinfo template.pdf
-\hypersetup{
-pdftitle={Statistical modeling with LLM predictors},
-pdfsubject={},
-pdfauthor={Michael Franke, Polina Tsvilodub, Fausto Carcassi},
-pdfkeywords={statistical models, language use, large language models, Bayesian data analysis, reference games},
-}
 
@@
 \maketitle
-
-\begin{abstract}
-  State of the art large language models (LLMs) have shown impressive performance on a variety of benchmark tasks and are increasingly used as components in larger applications, where LLM-based predictions serve as proxies for human judgements.
-  This raises questions about the human-likeness of LLM-derived information and whether LLMs, if they deliver human-like behavior, could possibly be considered explanatory models of (aspects of) human cognition or language use.
-  To shed more light on these issues, we here investigate the human-likeness of LLMs' predictions for multiple-choice decision tasks from the perspective of probabilistic cognitive modeling.
-  We identify core differences, in terms of conceptual interpretation and usual practical evaluation, between LLMs and probabilistic cognitive models (PCMs).
-  Concretely, we argue that LLMs are, first and foremost, predictors of item-level data, whereas PCMs primarily make predictions at a more abstract level (the condition-level), and that this is a sense in which LLMs may be said to be not explanatory.
-  Using human data from a multiple-choice experiment on pragmatic language use, we find that LLMs fail to capture the variance in the human data at the item-level.
-  We suggest different ways of deriving full distributional predictions from LLMs for condition-level data, thus building PCM-like predictions from LLM-derived information, and find that some, but not all ways of obtaining condition-level predictions yield adequate fits to our data.
-  These results suggests that assessment of LLM performance depends strongly on seemingly subtle choices in methodology, and that LLMs are at best predictors of human behavior at an aggregate level, for which they are, however, not designed to make predictions in the first place.
-\end{abstract}
-
-% keywords can be removed
-% \keywords{First keyword \and Second keyword \and More}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-\label{sec:introduction}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The invention of deep neural transformer architectures \citep{VaswaniShazeer2017:Attention-is-Al} has enabled a new generation of powerful large language models  \citep{DevlinChang2019:BERT:-Pre-train,ChungHou2022:Scaling-Instruc,OpenAI2023:GPT-4-Technical,TouvronLavril2023:LLaMA:-Open-and}.
-State-of-the-art LLMs excel on many benchmark data sets \citep[e.g.,][]{srivastava2023-BIGbench,PerezRinger2023:Discovering-Lan}, and so promise to serve as foundation models for a vast and diverse set of applications \citep{BommasaniHudson2021:On-the-opportun}.
-% especially when augmented with supervised fine-tuning  \citep{ChungHou2022:Scaling-Instruc} and reinforcement learning from human \citep{StiennonOuyang2022:Learning-to-sum} or other \citep{BaiKadavath2022:Constitutional-} feedback.
-Recent works increasingly go beyond using LLMs based on single-run input-output behavior, and instead utilize LLMs as a part of a larger computational process.
-Examples include sophisticated prompting strategies \citep[e.g.,][]{LiuLiu2022:Generated-Knowl}, or structured reasoning models \citep[e.g.,][]{CreswellShanahan2022:Selection-Infer,GaoMadaan2023:PAL:-Program-ai,ParanjapeLundberg2023:ART:-Automatic-}.
-Information from LLMs is also used  to rank or numerically score options in open-ended applications, e.g., to mimic human judgements of relevance or interestingness \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
-Other work, uses LLMs as part of bigger programs to build towards something more akin to explanatory cognitive models \citep[e.g.,][]{WongGrand2023:From-Word-Model}.
-
-For all of these applications, it is crucial to understand what LLMs can or cannot reliably do.
-The prevalent approach towards characterizing the capabilities of LLMs relies on benchmark testing, which usually consists in assessing the accuracy of LLM predictions in tasks where a designated ``target answer'' exists, averaged over many instances of this task.
-Benchmark-driven assessments are very useful for engineering purposes, e.g., to systematically study of the effects of scale \citep[e.g.,][]{srivastava2023-BIGbench}, but is also used to ask to what extent LLM performance is ``human-like.''
-In more psychologically oriented work, LLM performance is therefore compared to human choice behavior in psychological experiments and investigates whether LLMs predict patterns of human answer behavior  \citep[e.g.,][]{BinzSchulz2023:Using-cognitive,Hagendorff2023:Machine-Psychol,ShiffrinMitchell2023:Probing-the-psy}.
-The main focus is often to compare qualitative patterns in LLM predictions and human data, but there is also work investigating whether LLMs can make adequate \emph{quantitative predictions}.
-For example, work at the interface between NLP and computational psycholinguistics \citep{MarvinLinzen2018:Targeted-Syntac,HuGauthier2020:A-Systematic-As} has evolved into investigations of whether quantitative predictions by LLMs match quantitative aspects in human experimental data, such as reading times \citep{WilcoxVani2021:A-Targeted-Asse} or the amplitude of the N400 component of event-related-potentials in EEG measurements \citep{LindborgRabovsky2021:Meaning-in-brai}.
-The work presented in this paper seeks to extend the investigation of the human-likeness of LLMs by looking at empirical data from human multiple-choice behavior and comparing an established, theoretically motivated probabilistic cognitive model, which is commonly use to predict or explain this kind of data, to structurally similar quantitative predictions derived from LLMs.
-In short, while a lot of previous work has used targeted assessment of LLM capabilities from the point of view of the experimental psychologist, we here adopt the more specific perspective of the cognitive modeller.
-
-For the purposes of this investigation, a probabilistic cognitive model (PCM) is characterized by providing a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each possible data observation $D_{C}$ for any experimental condition $C$ and parameter values $\theta$ \citep[e.g.,][]{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
-% Usually, conditions $C$ provide a categorization of different tasks which is meaningful against the theoretical background of experimental investigation, and parameters $\theta$ modulate the predicted likelihood of the data in human interpretable ways (see Section~\ref{motivation} for more detail).
-There are non-trivial, insightful differences between LLMs and PCMs, both at the conceptual level and in terms of common practices in evaluation, which we will explore in this paper in order to contribute to a better delineation of the potential and limitations of LLMs.
-For one, the ``primitive predictions'' that LLMs provide are not at the level of a condition $C$, but at a lower item-based level.
-% In this sense, LLMs could be said to be less explanatory than PCMs by their very nature (see Section~\ref{motivation}).
-For another, LLMs are often just assessed based on their ability to identify a target answer, not to predict the full distribution observed in the data.
-% Nevertheless, distributional predictions for condition-level data can in principle be derived from LLMs, but there are many ways of doing so, yet with little consensus of how to do so.
-
-This paper introduces and systematically compares different ways of carving numerical information obtained from an LLM into a structure similar to a parameterized PCM and addresses the empirical question of whether such an ``LLM-based PCM'' adequately captures the distribution of human forced-choice data at different levels of aggregation.
-To obtain data for this investigation, we subjected human participants and an LLM (a version of OpenAI's GPT-3.5; see Section~\ref{sec:item-level-pred}) to the same forced-choice task, a simple reference game \citep{FrankGoodman2012:Predicting-Prag}.
-We introduce different ways of defining condition-level, distributional predictions based on LLM-provided numerical measures.
-We use standard tools of Bayesian data analysis \citep{GelmanCarlin2014:Bayesian-Data-A} to train, test and compare these ``LLM-based PCMs'' against each other, and against an established, theoretically informed PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
-This is quite different, and arguably more stringent and insightful, than the standard practice of evaluating LLMs based on accuracy scores.
-We find that variability predicted by the LLM at the item-level is not borne out by the human data, and that some but not all ways of constructing condition-level predictions are equally good.
-These results suggest that it is not impossible to build predictive models for condition-level data, similar to PCMs, on top of LLM-supplied information, but that seemingly subtle details in the derivation of likelihoods matter.
-The conceptual comparison of LLMs and PCMs also clearly identifies a sense in which the latter are explanatory, while the former are not.
-
-The paper is structured as follows.
-Section~\ref{motivation} motivates the investigation in this paper by highlighting relevant differences in the way  LLMs and PCMs make predictions about data from forced-choice tasks.
-Section~\ref{experiment-reference-games} looks at human data from a simple but non-trivial forced-choice experiment, namely a text-based reference game, for which Section~\ref{sec:model-pred-from} introduces a commonly used, theoretically-motivate PCM, namely the Rational Speech Act (RSA) model \citep{FrankGoodman2012:Predicting-Prag}.
-Section~\ref{sec:item-level-pred} investigates whether LLM-derived item-level predictions adequately capture the human data at the item-level.
-Section~\ref{llm-predictions-for-reference-games} then discusses different ways of deriving probabilistic predictions from LLMs at the condition-level and compares them against the human data and each other.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Predictions for multiple-choice tasks from PCMs and LLMs}
-\label{motivation}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-This section argues that a major conceptual difference between LLMs and PCMs is that they predict data at different levels of abstraction: the item-level and the condition-level, as introduced below.
-Another important difference is that PCMs are usually evaluated on a non-degenerate likelihood function, whereas common practice in LLM-evaluation is to use a winner-takes-all strategy.
-Problems with this latter strategy are highlighted and alternative ways of defining item- and condition-level predictions from LLMs are introduced for subsequent empirical testing.
-
-\paragraph{Item- and condition-level data.}
-Psychological research into the workings of the human mind aims to find generalizable patterns in the way information is processed within or across different domains of cognition.
-Experimental work therefore often compares human performance in different \emph{experimental conditions}, which reflect the general factors that are hypothesized to influence behavior.
-Yet a single experimental condition is often instantiated with different \emph{experimental items}, which are not under scrutiny for any systematic, predictable effect on the observed measurements.
-For example, classical research on human memory \citep{AtkinsonShiffrin1968:Human-memory:-A} investigated the effect of rehearsal on memory consolidation.
-Relevant experiments compare recall with and without rehearsal (experimental conditions), while using different items (words, numbers, etc., to be memorized) in each instantiation of the same memory task.
-Likewise, when studying how hearing a color word can facilitate a same-or-different judgement of color swatches \citep{Rosch1975:The-Nature-of-M}, the main experimental manipulation concerns the typicality of shown color swatches, while the variability between different color words like ``blue'' or ``green'', is less important to this research question and so treated as \emph{item-level variation}.
-Consequently, a typical psychological experiment is mainly interested in assessing behavior at the level of the experimental condition, because that is where the distinctions relevant to the research question reside.
-Nevertheless, each experimental condition can be, and often is, instantiated with different items, variation among which is deemed less relevant to the research question at hand.
-
-\paragraph{Probabilistic cognitive models.}
-Data from human participants for experiments of this kind usually shows some variability between items, and also variability between participants.
-This variability is commonly incorporated in statistical analyses as random stochastic variation, e.g., by using hierarchical (regression) models \citep{Jaeger2008:Categorical-dat,barr2013,SorensenHohensteinb2016:Bayesian-linear}.
-Still, the focus of interest usually remains at the condition-level effects, because it is this more abstract level of behavioral aggregation that is relevant for generalizable theory building.
-Similarly, when analyzing or explaining data from psychological experiments with a probabilistic cognitive model (PCM), the PCM will naturally be set-up to predict data at the level of the relevant experimental manipulation, i.e., at the condition-level.
-Even though item-level (and individual-level) variation can, and sometimes should, be incorporated into probabilistic cognitive modeling \citep[e.g.,][]{NilsonRieskamp2011:Hierarchical-Ba,Lee2011:How-Cognitive-M,ScheibehenneRieskamp2013:Testing-the-Ada}, the focus of cognitive modeling is to explain systematic variation between experimental conditions.
-Therefore, a basic PCM is essentially characterized by a parameterized likelihood function $P_{M}(D_{C} \mid \theta, C)$, which assigns a probability to each logically possible data observation $D_{C}$ for any experimental condition $C$ and numerical parameters $\theta$ \citep{LewandowskyFarrell2011:Computational-M,LeeWagenmakers2013:Bayesian-Cognit}.
-The difference between a theory-driven PCM and a standard, theory-neutral statistical model (like a regression model), which may also define a parameterized likelihood function of condition-level data, is that the PCM's likelihood function is defined in terms that are considered to be conceptually meaningful for the research domain at hand (see Section~\ref{sec:model-pred-from} for a concrete example).
-
-\paragraph{PCMs vs.~LLMs.}
-PCMs, thus construed, differ markedly from LLMs.
-LLMs make predictions about the likelihood of words given a preceding or surrounding text.
-They do not, without further derivation, provide predictions at a conceptually meaningful level of abstraction, similar to experimental conditions.
-In this way, LLMs are not explanatory models in the same sense as PCMs might be: they simply do not operate at a level of generalizable abstraction which is arguably necessary for increasing human understanding of the phenomenon in question \citep{Dellsen2020:Beyond-Explanat,Grimm2021:Understanding}.
-Nevertheless, we can derive predictions from LLM's word-level predictions, first at the item-level and also at the condition-level.
-Whether any of these predictions are adequate is ultimately an empirical question, which can be addressed by subjecting ``LLM-based PCMs'' to the same statistical model checking as the more usual theoretically motivated PCMs.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Experiment: Reference games}
-\label{experiment-reference-games}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}
-  \centering
-
-  \includegraphics[width = 0.8\textwidth]{00-pics/reference-game.png}
-
-  \caption{Structure of a reference game with human participants.
-Each trial consists of a set of objects, the so-called context.
-In production trials, participants choose a single word to describe a trigger object from the context.
-In interpretation trials, an object is selected as the likely object a trigger word referring to.}
-  \label{fig:ref-game}
-\end{figure}
-
-Reference games are an established, well-understood and austere experimental paradigm to test human decision making in abstract communicative tasks \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,SikosVenhuizen2021:Reevaluating-pr}.
-A reference game consists of two players, a speaker and an interpreter, who jointly observe a set of objects, usually referred to as context (see Figure~\ref{fig:ref-game}).
-In the \textbf{production condition}, the speaker is assigned a \emph{trigger object} from the context set which they have to describe to the interpreter.
-In the \textbf{interpretation condition}, the interpreter observes a description, here called \emph{trigger word}, and chooses one of the objects from the context set.
-The goal of the game is, for the speaker, to choose a description that enables the interpreter to choose the trigger object; and, for the interpreter, to guess correctly which object the speaker had in mind when using the trigger word.
-
-The example in Figure~\ref{fig:ref-game} is a standard case, which we will use throughout, where human choices are informative about the pragmatic reasoning that human decision makers engage in.
-In this example, there are two features that differ across three objects (here shape and color).
-One object shares both its color and shape with one other object, while the two other objects have one unique feature (e.g., being the only circle, or the only green object).
-%
-In a critical production trial, the trigger object to describe is one of the two objects with a unique feature.
-The speaker has four words to choose from.
-The \textbf{target utterance} is the word which uniquely describes the trigger object.
-The \textbf{distractor utterance} is the word that is true of the trigger object, but also true of another object.
-The other utterances, both of which are false of the trigger object are \textbf{competitor utterance}.
-%
-In a critical interpretation trial, the trigger word is one that is true of two of the three objects.
-If participants engage in pragmatic thought, they might reason that \emph{if} the speaker had wanted to refer to one of the two objects of which the trigger word is true (blue square and blue circle in Figure~\ref{fig:ref-game}), the speaker could have used a more informative word for exactly one of those two objects (``circle''), so they are more likely to refer to the \textbf{target object} (the blue square in Figure~\ref{fig:ref-game}).
-The \textbf{competitor object} is the other object of which the trigger word is true.
-The \textbf{distractor object} is the object of which the trigger word is false.
-
-We replicated a simple reference game with human participants in which each trial instantiated the structure of the example shown in Figure~\ref{fig:ref-game}.
-While previous reference games with human participants used pictorial representations of objects, and sometimes even pictorial representations of messages, we implemented a text-only version in order to be able to compare the predictions of LLMs for human data, when both LLMs and humans processed the same textual representation of the stimuli.
-The experiment was realized as an online task using \texttt{magpie} \citep{FrankeJi:magpie:-Minimal}.\footnote{
-  The code for the experiment can be found at \href{https://github.com/magpie-ea/magpie3-text-refgame}{https://github.com/magpie-ea/magpie3-text-refgame}, and a live version of the experiment can be tested at \href{https://magpie-ea.github.io/magpie3-text-refgame/}{https://magpie-ea.github.io/magpie3-text-refgame/}.
-}
-
-\subsection{Participants}
-\label{participants}
-
-A total of 302 participants were recruited via Prolific for monetary
-compensation (\textsterling0.45, corresponding to roughly \textsterling 15.40 per hour).
-All participants self-identified as native speakers of English.
-
-\subsection{Materials \& design}
-\label{materials-design}
-
-We created 100 different items as stimulus material via a stochastic process.
-Each item is a different textual description of a reference game with the same logical structure as the example from Figure~\ref{fig:ref-game}.
-For each item, the context consisted of three objects.
-As in the original paper by \citet{FrankGoodman2012:Predicting-Prag}, objects are defined by a triple of properties, namely a color, a shape and a texture.
-For each property, there were four possible values, e.g., blue, green, red, and orange for color.
-The sampled items differed in terms of the properties of the objects in the context set, and in terms of the order in which the objects and expression alternatives were presented in the text.
-Figures~\ref{fig:refgame-screenshot-production} and \ref{fig:refgame-screenshot-interpretation} from Appendix~\ref{sec:scre-from-online} show example screenshots from the experiment.
-
-\subsection{Procedure}
-\label{procedure}
-
-For each participant the experiment sampled four different items.
-Participants first played two of these in the production condition, then the other two in the interpretation condition.
-
-\subsection{Results}\label{results}
-
-The overall distribution of choices that correspond to the target, competitor, and distractor states is shown in Figure~\ref{fig:refgame-counts}.\footnote{
-  The production condition actually has two distractor choices.
-  Here and in the following, these are lumped together as a single category, also when modeling random errors in later models.
-  This is a simplification but does not change anything of substance.}
-It is interesting that the distractor options were chosen rather often.
-We also see that the number of target choices is higher in the production condition than in the interpretation condition.
-This is in line with previous experimental results on human reference games.
-% \citep{QingFranke2013:Variations-on-a}.
-For example, in previous forced-choice reference games with human participants with pictorial presentations of objects, \citet{QingFranke2013:Variations-on-a} observed the following proportions of target, competitor and distractor options: $\tuple{0.882, 0.118, 0}$ in the production and $\tuple{0.492, 0.506, 0.003}$ in the interpretation condition (for 288 observations in each condition).
-
-\begin{figure}[t]
-  \centering
-
-    \includegraphics[width=0.9\linewidth]{00-pics/PPC-alpha-eps-model.pdf}
-
-    \caption{Counts of choices from reference games with human participants (colored bars), with summary statistics from the posterior predictive distribution of four models (shapes and error bars).
-      Shapes show the mean of the posterior predictive distributions of the RSA model and three further models derived from item-level LLM-supplied scores.
-      Error-bars show corresponding 95\% credible intervals of the posterior predictive.
-    }
-  \label{fig:refgame-counts}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Model predictions from probabilistic pragmatics}
-\label{sec:model-pred-from}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Data from reference games with human participants have been variously analyzed with probabilistic models using inspiration from behavioral game theory \citep[e.g.,][]{DegenFranke2013:Cost-Based-Prag}, probabilistic Bayesian modeling \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag} or other forms of probabilistic modeling \citep[e.g.,][]{GattGompel2013:Are-we-Bayesian}.
-Common to these approaches is that they derive or define, based on some explicit conceptual motivation, a parameterized stochastic speaker policy, $P_{S}(u \mid s; \theta_{S})$, modulated by parameters $\theta_{S}$, for a speaker's choice of expression or utterance $u$ given a referent or state $s$, which the speaker wants to communicate;
-and a stochastic listener policy, $P_{L}(s \mid u; \theta_{L})$, capturing the probability of choosing a referent $s$ for utterance $u$.
-
-As a concrete example, we introduce the Rational Speech Act (RSA) model first described in this form by \citet{FrankGoodman2012:Predicting-Prag} \citep[for overview see][]{FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu,StevensBenz2018:Game-Theoretic-,ScontrasTessler2021:A-practical-int,Degen2023:The-Rational-Sp}.
-The RSA model defines pragmatic reasoning as a sequence of iterated (soft-)optmization of policies, grounding out in literal interpretation.
-If $\mathfrak{L}(s,u) \mapsto \set{0,1}$ is a semantic meaning function mapping each pair of state $s$ and utterance $u$ to a (binary) truth-value, and if $P_{\text{prior}}(s)$ is a prior over states, a literal listener policy is defined as:
-%
-\begin{align*}
- P_{L_{0}}(s \mid u) \propto \mathfrak{L}(s,u)\  P_{\text{prior}}(s)\,.
-\end{align*}
-%
-The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated, i.e., to maximize the log-probability of the trigger object:
-%
-\begin{align}
-  \label{eq:RSA-speaker}
-  P_{S}(u \mid s, \alpha) & \propto \expo \left [ \log P_{L_{0}}(s \mid u) \right ] \,.
-\end{align}
-%
-Finally, the pragmatic listener is defined as the policy resulting from applying Bayes rule, solving the inverse-problem for the previously defined speaker policy:
-%
-\begin{align}
-  \label{eq:RSA-interpreter}
-  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P_{\text{prior}}(s) \,.
-\end{align}
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
-  \caption{
-    Example of predictions from the RSA model (with $\alpha=1$).
-    The semantic meaning is shown as a matrix of binary truth-values.
-    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents) for $\alpha=1$ are shown as row-stochastic matrices.
-  }
-  \label{fig:RSA-example}
-\end{figure}
-
-Figure~\ref{fig:RSA-example} gives example calculations (assuming a flat prior and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
-For $\alpha=1$, the model predicts that the probabilities of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{\nicefrac{3}{5}, \nicefrac{2}{5}, 0}$ for the interpretation condition.
-Increasing $\alpha$ will increase the odds of target over competitor choices.
-In sum, the condition-level predictions of the RSA model are a parameterized function $P_{\text{cond}}^{\text{RSA}}(R_{l}, C; \alpha_{c})$, assigning a probability to each response category $R_{l}$ (target, competitor, or distractor) in each condition $C$ (production or interpretation).
-
-The model constructed so far predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
-It is therefore common to include a small error probability $\epsilon$, with which a choice would be made at random \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}, so that we get:
-%
-\begin{align*}
-  P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon_{c}) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
-\end{align*}
-%
-where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
-\mf{FC objected to ``condition-specificity'' which is fair, but I leave it as it was; my reasoning: we use model checking to rule out models; in this setting the conservative strategy is to err on the side of making the models too powerful (even at the cost of conceptual plausibility)}
-
-The data $D_{C}$ from condition $C$, see Figure~\ref{fig:refgame-counts}, consists of counts for each response category.
-The parameterized likelihood function entailed by the RSA model for condition-level data $D_{C}$ is:
-%
-\begin{align}
-  \label{eq:RSA-likelihood}
- P^{\text{RSA}}_{\text{cond}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \text{Multinomial}(D_{C}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3})\,.
-\end{align}
-%
-The result is a four-parameter model, one pair of parameters per condition.
-
-Parameterized predictions, like in Equation~(\ref{eq:RSA-likelihood}), can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
-Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
-Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{tab:sumStats}).\footnote{
-  Posterior samples where generated for four chains with 2000 samples each, after a warm-up of 1000 samples.
-The ``adapt-delta'' value was set to 0.99.
-Converge was checked with $\hat{R}$-statistics \citep{GelmanRubin1992:Inference-from-}.
-}
-
-\begin{table}[t]
-\centering
-\begin{tabular}{llllcrlcrcl}
-  \toprule
-  &&&& $\alpha$ &&& $\epsilon$ & \\ \cmidrule(r){4-6} \cmidrule(l){7-9}
-  data & model & condition & $|$95\%\ & mean\ & 95\%$|$\ & $|$95\%\ & mean\ & 95\%$|$\ & Bpppv & \\
-  \midrule
-
-  item-level  & RSA                & production     & 2.62 & 3.15 & 3.69 & 0.08 & 0.13 & 0.17 & 0.34 &  \\
-  item-level  & RSA                & interpretation & 0.24 & 0.66 & 1.05 & 0.10 & 0.15 & 0.20 & 0.25 &  \\
-  item-level  & LLM                & production     & 0.15 & 0.21 & 0.29 & 0.00 & 0.10 & 0.21 & 0.00 & * \\
-  item-level  & LLM                & interpretation & 0.08 & 0.13 & 0.18 & 0.00 & 0.17 & 0.33 & 0.00 & * \\
-  cond.-level & RSA                & production     & 2.59 & 3.16 & 3.71 & 0.08 & 0.13 & 0.18 & 0.49 &  \\
-  cond.-level & RSA                & interpretation & 0.24 & 0.66 & 1.07 & 0.10 & 0.15 & 0.20 & 0.52 &  \\
-  cond.-level & avg. scores        & production     & 0.19 & 0.22 & 0.26 & 0.00 & 0.04 & 0.11 & 0.05 & * \\
-  cond.-level & avg. scores        & interpretation & 0.15 & 0.18 & 0.20 & 0.00 & 0.02 & 0.06 & 0.00 & * \\
-  cond.-level & avg. probabilities & production     & 0.86 & 1.04 & 1.22 & 0.08 & 0.13 & 0.18 & 0.51 &  \\
-  cond.-level & avg. probabilities & interpretation & 0.36 & 0.48 & 0.60 & 0.00 & 0.05 & 0.12 & 0.00 & * \\
-  cond.-level & avg. WTA           & production     & 0.87 & 1.04 & 1.23 & 0.08 & 0.13 & 0.17 & 0.49 &  \\
-  cond.-level & avg. WTA           & interpretation & 0.17 & 0.39 & 0.61 & 0.10 & 0.15 & 0.20 & 0.26 &  \\
-
-   \bottomrule \\
-\end{tabular}
-\caption{
-  Summary statistics for model fits.
-  Columns show the means and 95\% credible intervals for posterior samples for each parameter, as well as the Bayesian posterior predictive $p$-values for each model, condition and data level.
-  Asterisks mark Bppp-values no higher than $0.05$, flagging relative inability to capture the structure of the data.
-}
-\label{tab:sumStats}
-\end{table}
-
-To assess goodness-of-fit, we use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
-Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models for condition-level data, which are to be introduced later).
-We see that for both conditions the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-}, which require that the distribution of posterior predictions includes the observed choice rates for each answer option.
-To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values (Bppp values), using likelihood of the observed data as a test statistics.
-These Bppp values approximate the probability that a model conditioned on observed data $D_{\text{obs}}$ predicts future data $D_{rep}$, of the same size and format as $D_{obs}$, that is at least as likely as the data $D_{\text{obs}}$ itself is (given the posterior predictive distribution).
-Very small Bppp values indicate that the model might be inadequate for reproducing the data it was trained on, so to speak.
-This is clearly not the case for the RSA model with Bppp values close to 0.5, as shown in Table~\ref{tab:sumStats}.
-In sum, the condition-level predictions by the RSA model, a theoretically motivated PCM, are not discredited by the condition-level data.
-Reversely, the RSA model seems to adequately capture the condition-level data.
-
-We can also use the condition-level predictions for predicting the item-level data.
-If the whole data set for condition $C$ is chunked into item-level data, $D_{C} = \tuple{D^{1}_{C}, \dots, D^{m}_{C} }$, the likelihood function for analysis at the item-level uses the same (condition-level) predictor to separately predict data from each item, so that the likelihood function becomes:
-%
-\begin{align*}
- P^{\text{RSA}}_{\text{item}}(D_{C} \mid C, \alpha_{C}, \epsilon_{C}) = \prod_{k = 1}^{m} \text{Multinomial} \left (D_{C}^{k}, \tuple{P_{r}(R_{l}, C; \alpha_{c}, \epsilon_{c}}_{1 \le l \le 3} \right )\,.
-\end{align*}
-%
-Fitting the RSA model to the partitioned data set, we find estimates of Bppp values that do not discredit the model (see Table~\ref{tab:sumStats}).
-This suggests that the item-level variation in the human data is not so pronounced as to provide strong evidence against the condition-level predictor from a theoretically motivated PCM.
-In other words, the condition-level RSA predictions seem to adequately capture also the item-level data.
-
-The following sections apply these tests of Bayesian model criticism also to models built around predictor values from LLMs.
-Section~\ref{sec:item-level-pred} first looks at the item-level data, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Item-level predictions from LLMs}
-\label{sec:item-level-pred}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The RSA model's predictions are defined at the condition-level.
-An LLM, on the other hand, first and foremost makes predictions about each item.
-As mentioned previously, the items of the reference game experiment differ in which levels of features (color, shape, texture) instantiate the structure of the task shown in Figure \ref{fig:ref-game}, as well as the order of presentation of objects and words.
-We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep[e.g.,][]{QingFranke2013:Variations-on-a}, such as over-production of informationally irrelevant material \citep[e.g.,][]{DaviesKatsos2010:Over-informativ,Rubio-Fernandez2019:Overinformative,DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options.
-\mf{@PT references from the NLP literature on effects of prompt variation?}
-It therefore becomes an empirical question of whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.
-
-\begin{figure}[t]
-  \centering
-  \includegraphics[width=\textwidth]{00-pics/closeness-target-by-alpha-item-level.pdf}
-  \caption{
-    Mean predicted target probability for different values of softmax parameter $\alpha$.
-    The solid black lines show the probability of the target options, predicted for a fixed value of $\alpha$ ($x$-axis), averaged over all items for each condition.
-    The gray-shaded area indicates the posterior 95\% credible interval for $\alpha$.
-    For reference, the target probability under a random choice strategy and under the WTA strategy are shown with dashed lines.
-    For credible values of $\alpha$, the means of predicted probabilities for the target option are clearly distinct from the WTA strategy.
-  }
-  \label{fig:closeness-target-item-level}
-\end{figure}
-
-As described in Section~\ref{motivation}, the standard procedure for assessing LLM predictions in multi-choice tasks is to use a winner-takes-all (WTA) strategy that just picks the best option for each item.
-But since this strategy has conceptual and empirical weaknesses, we define the item-level LLM predictions using the more general softmax-approach, which contains the WTA-based approach as a limiting case, repeated here from above with slightly updated notation:
-%
-\begin{align*}
-P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c} \right ) \propto \expo \left (\alpha_{c} \ \text{S}_{ki} \right)\,.
-\end{align*}
-%
-Item-level scores $\text{S}_{ki}$, as defined in Section~\ref{motivation}, were obtained from the text-davinci-003 instance GPT-3.5 August 2023 \citep{BrownMann2020:Language-Models} .
-% \footnote{
-  % Numerically, we worked with \emph{average} log-probabilities.
-As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition.
-Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words.
-In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
-The task description $x_{k}$ for item $I_{k}$ is a text supplied as prompt (see Appendix~\ref{sec:examples-items-llm} for an example).
-The choice options are categorized as in the human experiment.\footnote{
-  In the current set-up the response type ``distractor'' has two instantiations in the production condition.
-Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
-Allowing for random errors with probability $\epsilon_{C}$, as before,:
-%
-\begin{align*}
-  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right )
-  = \epsilon_{c} \  P_{\text{item}}^{\text{LLM}}\left ( y_{ki} \mid C; \alpha_{c}, \epsilon_{c} \right ) \ \ + \ \ \nicefrac{\epsilon_{c}}{3}   \,.
-\end{align*}
-%
-The resulting likelihood function for item-level data with item-level predictions is:
-%
-\begin{align*}
- P^{\text{LLM}}_{\text{item}}(D_{c} \mid C; \alpha_{c}, \epsilon_{c}) = \prod_{k = 1}^{m} \text{Multinomial} \left ( D_{c}^{k}, \tuple{P^{\text{LLM}}_{\text{item}}(y_{ki} \mid C;  \alpha_{c}, \epsilon_{c})}_{1 \le l \le 3} \right )\,.
-\end{align*}
-%
-Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.
-
-Summary statistics for samples from the posterior distribution over parameters are shown in Table~\ref{tab:sumStats}.
-Credible values of $\alpha$ parameters are comparatively low for the LLM-based model, suggesting that the predicted scores have rather large differences, which have to be compensated for by the softmax link-function to adequately fit the data.
-Figure~\ref{fig:closeness-target-item-level} shows the mean predicted target probabilities for different parameter values of $\alpha$, highlighting that for posterior credible values, the predictions are clearly distinct from the predictions of a WTA strategy.
-This suggests that the WTA strategy, which is the special limiting case for $\alpha \rightarrow \infty$, provides a substantially worse fit for item-level choices than those provided by less extreme values of $\alpha$.
-
-\begin{figure}[t]
-  \centering
-
-  \includegraphics[width = 0.9\linewidth]{00-pics/item-combined-obs-pred.pdf}
-
-  \caption{
-    Item-level prediction-observation plot.
-    Each dot represents an item.
-    The $x$-coordinate represents the mean of the posterior prediction for the target choice probability for the given item.
-    The $y$-coordinate represents the observed proportion of target choices for that item.
-    The gray line (identity function) is the ideal prediction.
-  }
-  \label{fig:item-level-obs-pred}
-\end{figure}
-
-Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (see Table~\ref{tab:sumStats}), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
-Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
-Concretely, the plots show, for each condition and item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
-There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
-So, the model itself predicts systematic variability at the item level.
-The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.
-These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
-Put more strongly, a model that uses the most obvious item-level score to predict what human participants choose on a by-item level is ruled out by experimental data.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Condition-level predictions}
-\label{llm-predictions-for-reference-games}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% This section explores different ways of deriving a probabilistic likelihood function for the human data from an LLM at the aggregate, condition-level.
-% Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
-% In the following, we first describe the different options of deriving LLM-based predictors.
-% We then compare all probabilistic models based on their adequacy of explaining the human choice data.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.9\textwidth]{00-pics/measures-overview.png}
-  \caption{
-    Schematic overview over three different approaches of obtaining condition-level predictions by aggregating item-level information.
-    Approaches differ in the kind of item-level information they take into account, which entails differences in the order in which aggregation, transformation to probabilities and parameterized scaling occur.
-  }
-  \label{fig:measures-overview}
-\end{figure}
-
-While LLMs do not make conditional-level predictions as such, they can be derived from item-level scores $\text{S}_{ki}$ by averaging over all items belonging to the relevant condition.
-There are many ways of averaging item-level information.
-Figure~\ref{fig:measures-overview} shows three salient approaches, which differ in what the underlying item-level measure is: the raw scores $\text{S}_{ki}$, the item-level probability derived from it (as introduced in Section~\ref{motivation} and used in Section~\ref{sec:item-level-pred}), or the predictions from the winner-takes-all strategy commonly used in benchmark testing (see Section~\ref{motivation}).
-
-% In general, to map a vector non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
-% %
-% \begin{align*}
-%  \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
-% \end{align*}
-% %
-% The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
-% %
-% \begin{align*}
-%   \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
-%   \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
-% \end{align*}
-
-The \emph{average-scores predictor} first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities using the usual parameterized softmax function:\footnote{
-  The reported results average over the multi-set \(\set{I_1, \dots I_m}\) of items that occurred in the human experiment for condition $C$.
-  By using a multi-set, which may contain a single item multiple times, we produce aggregate predictions for exactly the set of items that the participant group saw, which provides the most fitting counterpart to the human data.}
-%
-\begin{align*}
-  & P_{\text{cond}}^{\text{SCR}}(r_{i} \mid C, \alpha_{c})
-    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{ki}  \right ] \,.
-    \tag*{\textcolor{gray}{[average scores (narrow-scope aggregation)]}}
-\end{align*}
-%
-The \emph{average-probabilities predictor} first transposes scores into probabilities with a parameterized softmax function, and only aggregates over items last:
-\begin{align*}
-  % & \textbf{wide-scope aggregation} \\
-  & P_{\text{cond}}^{\text{PRB}}(r_{i} \mid C, \alpha_{c})
-    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha_{c}) \,.
-    \tag*{\textcolor{gray}{[average probabilities (wide-scope aggregation)]}}
-\end{align*}
-%
-Finally, the \emph{average-WTA predictor} considers the prediction of the WTA strategy as the basic item-level unit to aggregate over.
-To add parameterized scaling to this method, a power-law transformation is a natural choice:
-\begin{align*}
-  % & \textbf{intermediate-scope aggregation} \\
-  & P_{\text{cond}}^{\text{WTA}}(r_{i} \mid C, \alpha_{c})
-    \propto  \left [ \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) \right ]^{\alpha_{c}}\,,
-    \tag*{\textcolor{gray}{[average WTA (intermediate-scope aggregation)]}}
-\end{align*}
-where $P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) = \lim_{\alpha} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha)$.
-\bigskip
-
-The \emph{average-scores} and \emph{average-probabilities} predictions are equivalent if there is only one item, in which case the prediction of the \emph{average-WTA} method is the special case of $\alpha \rightarrow \infty$.
-For cases with more than one item, the predictions of the three predictors are not guaranteed to be the same.
-Conceptually, the \emph{average-probabilities} and the \emph{average-WTA} predictors, but not the \emph{average-scores} predictor, are compatible with a picture in which condition-level predictions result from the actual predictions at the item level.
-The difference between the \emph{average-probabilities} and the \emph{average-WTA} predictor is that the latter is an aggregate of an item-level predictor that is demonstrably bad, as shown in the previous Section~\ref{sec:item-level-pred}.
-
-Using the same approach as for the RSA model in Section~\ref{sec:model-pred-from}, we can build likelihood functions for condition-level data around the three predictors introduced above.
-With the same priors and methods used before, we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
-Summary statistics for posteriors over model parameters are shown in Table~\ref{tab:sumStats}.
-What is noteworthy is that for the \emph{average-WTA} model, the estimates of $\alpha_{c}$ in the production condition do not rule out, in fact lie close to, the special value $\alpha_{c}=1$, for which the power-law transformation is the identity function.
-This means that just averaging WTA-responses at the item-level yields a reasonable predictor for the production data at the condition-level.
-However, for the interpretation condition, the value $\alpha_{c}=1$ is clearly outside the range of credible parameter values, so that a simple recipe like ``always average WTA-responses without transformation'' is not a viable strategy for good condition-level predictions in general.
-It is also worth noting that for the \emph{average-probability} model values of $\alpha_{c}$ close to one are virtually indistinguishable from the item-level predictions of the WTA strategy (see Figure~\ref{fig:closeness-target-item-level}).
-This suggests that, for the production data, aggregation of item-level predictions from the WTA-strategy gives good condition-level predictions.
-Since the production data is also very biased towards almost exclusively target choices by human participants, this is not too surprising.
-
-Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
-We find that only the theoretical model (RSA) and the \emph{average-WTA} model pass this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
-To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
-Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} model fails on the interpretation data, and is at most borderline compatible with the production data; that the \emph{average-probabilities} model is able to reproduce the production data, but fails on the interpretation data; and that only the \emph{average-WTA} model does not fail to capture the data from both conditions.
-
-These results tell us that not all ways of deriving condition-level predictions by averaging over item-level variation are equally good.
-Some approaches clearly fail basic checks for statistical goodness-of-fit.
-On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the \emph{average-WTA} model, which is able to recover the patterns in the data.
-In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.\footnote{
-  A potential worry is that the \emph{average-WTA} model is trivial in the sense that it could have predicted \emph{any} data observation.
-  Appendix~\ref{sec:range-prior-pred} shows that this is not the case.
-  \mf{todo: redo these calculations with actual WTA!}
-}
-On the other hand, there is a stark contrast between the results of item-level and condition-level data.
-The model that was able to properly fit the condition-level data builds on predictions for item-level choice that are clearly incompatible with the human item-level data.
-The model seems to be ``right for the wrong reasons.''
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-\label{conclusion}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-It is not entirely ludicrous to use numerical predictions from LLMs are part of predictive probabilistic models.
-But we have to be careful.
-Ideally, we average out over low-level variation, such as from order of presentation or similar ``nuisance,'' at least as long as we do not understand what causes this variation in the predictions of models and further research that investigates when exactly this variation accords with empirically observed patterns.
-
-This also implies that we should likely \emph{not} (yet) aspire to use
-LLMs are models or individual- or item-level predictors.
-
-( {TODO: rethink conclusion})
-
-{to be continued}
-
-\begin{itemize}
-  \item This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.
-\end{itemize}
-
-\newpage
-\appendix
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Screenshots from the online experiment with human participants}
-\label{sec:scre-from-online}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Figure~\ref{fig:refgame-screenshot-production} shows a trial from the production condition, Figure~\ref{fig:refgame-screenshot-interpretation} one for the interpretation condition of the online experiment.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width = 0.8\textwidth]{00-pics/refgame-production.png}
-
-  \caption{Screen shot from a production trial of the online experiment.}
-  \label{fig:refgame-screenshot-production}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width = 0.8\textwidth]{00-pics/refgame-interpretation.png}
-
-  \caption{Screen shot from an interpretation trial of the online experiment.}
-  \label{fig:refgame-screenshot-interpretation}
-\end{figure}
-
-\section{Example item for the LLM experiment}
-\label{sec:examples-items-llm}
-
-The text-based input for the LLM predictions mirrors the text in the human experiment, except that the LLM input also lists the set of all available choice options (which for the human experiment is unnecessary since this information is given by the buttons for the forced-choice selection).
-For example, the task description $T_{k}$ for the item that corresponds to the production trial shown in Figure~\ref{fig:refgame-screenshot-production} is shown below (the actual input has no line breaks in the first paragraph):
-
-\begin{verbatim}
-Your task is to play a conversation game. There are three objects that
-you and your friend can see. You have to choose a single word to identify
-one of the three objects for your friend.
-
-The three objects are:
-
-a blue triangle with spades
-a blue circle with stripes
-a blue triangle with stripes
-
-Your task is to make your friend pick out the following target object:
-
-the blue circle with stripes
-
-Which of the following words would you choose:
-
-spades
-stripes
-circle
-triangle
-
-Your answer:
-
-I would choose the word
-\end{verbatim}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Range of prior predictions of intermediate-scope model}
-\label{sec:range-prior-pred}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-It may seem that, given the freedom to adjust optimization $\alpha$ and random guessing rate $\epsilon$, some models might be able to explain \emph{any} data observation.
-The intermediate-scope model, which was not discredited by model criticism in terms of recovery-based posterior predictive checks, is not trivial in this sense.
-Figure~\ref{fig:prediction-range} shows the range of predictions that the intermediate-scope model is, in principle, capable of making (for the whole possible range of \(\alpha\) and \(\epsilon\) values, disregarding Bayesian priors).
-There are distributions over response types which the model does not predict \emph{ex ante}.
-For example, the model does not predict cases where the number of competitor choices is almost as high as the number of target choice, as would be the case if participants would consistently \emph{not} engage in pragmatic reasoning.
-
-\begin{figure}[t]
-  \centering
-
-  \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-prod.png}
-  %
-  \includegraphics[width=0.45\textwidth]{00-pics/prediction-range-inter.png}
-
-  \caption{
-    Range of predictions the intermediate-scope model makes for any pair of parameter values $\alpha_{c}$ and $\epsilon_{c}$.
-    The light gray triangle with the red boundary is the probability simplex, i.e., the space of all possible 3-place probability vectors.
-    The darker gray area in gray boundary lines is the subspace that is compatible with the ordering that the probability of choosing the target is bigger than that of the competitor, which in turn is bigger than that of the distractor.
-    The blue area is the subspace of probabilistic predictions the intermediate-scope model makes under any value of its parameters.
-    The orange dot in the middle is the ``Laplace point'' of equal probability for all three options, which the model predicts for $\alpha=0$ or $\epsilon=1$.
-    The yellow dot is the ``vanilla prediction'' for $\alpha_{c}=1$ and $\epsilon_{c}=0$.
-  }
-  \label{fig:prediction-range}
-\end{figure}
-
 \printbibliography[heading=bibintoc]
 
-\newpage
-
-\section*{Snippets}
-
-\begin{itemize}
-  \item for conclusion?
-  \begin{itemize}
-    \item The question after the human-likeness of quantitative LLM-derived information matters for applications which use numerical scores to rank or weigh options \citep[e.g.,][]{ParkOBrien2023:Generative-Agen,ZhangLehman2023:OMNI:-Open-ende}.
-    Moreover, to the extent that LLMs are used as part of explanatory ``neuro-symbolic models'' of information processing \citep{GarcezLamb2020:Neurosymbolic-A}, understanding whether and how LLMs might yield full-fledged distributional predictions is important, e.g., to explore their integration into probabilistic (cognitive) models \citep[c.f.,][]{Frank2023:Large-language-}.
-  \end{itemize}
-  \item PT: the high-level take-away that item-level LLM predictions don't fit human predictions well might, as always, be met with the comment that the task is rather artificial.
-maybe this should go into the discussion
-  \item FC: The fact that the aggregate predictions track aggregate human behaviour means that the discrepancies are washing out.
-But there's also the interesting question of finding the systematic differences between LLMs and people.
-The task then would be to find sets of stimuli where the differences do \emph{not} wash out and ask: What if anything do these stimuli share?
-This paper could also be a jumping board for this line of research.
-  \item PT: maybe a conclusion in the spirit of "cogmod in the age of LLMs"/ "computational psycholinguistics \& careful assessment of reasoning / linguistic abilities of LLMs against human performance" might be not entirely ludicrous too
-\end{itemize}
-
 \end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:

### 2024-06-13 04-paper/paper-LLM-probmod.tex
fixed typo in section heading

@@
 \usepackage{mypackages}
-\bibliography{~/Desktop/icloud/MyRefGlobal}
+\bibliography{MyRefGlobal}
 \usepackage{mycommands}
@@
   \centering
-  \includegraphics[width = 0.95\textwidth]{00-pics/condition-item-level-prediction.png}
+  \includegraphics[width = 0.95\textwidth]{pics-final/condition-item-level-prediction.png}
   \caption{
@@
 Finally, Section~\ref{sec:gener-other-llm} explores whether previous results generalize to other LLM backends by investigating different versions of LLaMA2 \citep{TouvronLavril2023:LLaMA:-Open-and}.\footnote{
-  All data and code is available at \url{https://osf.io/f6j3a/?view_only=5e820cc8bbee4549aed58dc252ba61b9}.
+  All data and code is available at this OSF repository:  \href{https://osf.io/f6j3a/?view_only=5e820cc8bbee4549aed58dc252ba61b9}{https://osf.io/f6j3a/?view\_only=5e820cc8bbee4549aed58dc252ba61b9}.
 }
@@
 
-  \includegraphics[width = 0.8\textwidth]{00-pics/reference-game.png}
+  \includegraphics[width = 0.8\textwidth]{pics-final/reference-game.png}
 
@@
   \centering
-  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
+  \includegraphics[width = 0.9 \textwidth]{pics-final/RSA-example.png}
   \caption{
@@
 
-    \includegraphics[width=0.9\linewidth]{00-pics/GPT-PPC-alpha-eps-model.pdf}
+    \includegraphics[width=0.9\linewidth]{pics-final/GPT-PPC-alpha-eps-model.pdf}
 
@@
   \centering
-  \includegraphics[width=\textwidth]{00-pics/GPT-closeness-target-by-alpha-item-level.pdf}
+  \includegraphics[width=\textwidth]{pics-final/GPT-closeness-target-by-alpha-item-level.pdf}
   \caption{
@@
 
-  \includegraphics[width = 0.9\linewidth]{00-pics/GPT-item-combined-obs-pred.pdf}
+  \includegraphics[width = 0.9\linewidth]{pics-final/GPT-item-combined-obs-pred.pdf}
 
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{00-pics/measures-overview.png}
+  \includegraphics[width=0.9\textwidth]{pics-final/measures-overview.png}
   \caption{
@@
   \centering
-  \includegraphics[width=0.9\textwidth]{00-pics/vPPC_all_LLaMA_models.pdf}
+  \includegraphics[width=0.9\textwidth]{pics-final/vPPC_all_LLaMA_models.pdf}
   \caption{Visual posterior predictive checks on condition-level data for different models of the LLaMA family, paired with different aggregation strategies.}
@@
   \centering
-  \includegraphics[width = 0.8\textwidth]{00-pics/refgame-production.png}
+  \includegraphics[width = 0.8\textwidth]{pics-final/refgame-production.png}
 
@@
   \centering
-  \includegraphics[width = 0.8\textwidth]{00-pics/refgame-interpretation.png}
+  \includegraphics[width = 0.8\textwidth]{pics-final/refgame-interpretation.png}
 
